### 2011-12-07 paper.tex
put our outline in

% THIS IS SIGPROC-SP.TEX - VERSION 3.1
% WORKS WITH V3.2SP OF ACM_PROC_ARTICLE-SP.CLS
% APRIL 2009
%
% It is an example file showing how to use the 'acm_proc_article-sp.cls' V3.2SP
% LaTeX2e document class file for Conference Proceedings submissions.
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V3.2SP) *DOES NOT* produce:
%       1) The Permission Statement
%       2) The Conference (location) Info information
%       3) The Copyright Line with ACM data
%       4) Page numbering
% ---------------------------------------------------------------------------------------------------------------
% It is an example which *does* use the .bib file (from which the .bbl file
% is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission,
% you need to 'insert'  your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% Questions regarding SIGS should be sent to
% Adrienne Griscti ---> griscti@acm.org
%
% Questions/suggestions regarding the guidelines, .tex and .cls files, etc. to
% Gerald Murray ---> murray@hq.acm.org
%
% For tracking purposes - this is V3.1SP - APRIL 2009

\documentclass{acm_proc_article-sp}

\begin{document}

\title{Cache Oblivious B-trees}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Salman Ahmad\\
\affaddr{MIT CSAIL}
% 2nd. author
\alignauthor
Leilani Battle\\
\affaddr{MIT CSAIL}
% 3rd. author
\alignauthor
Stephen Tu\\
\affaddr{MIT CSAIL}
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.
\date{7 December 2011}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

\maketitle
\begin{abstract}
Fill me in
\end{abstract}

\section{Introduction}

\subsection{External Memory Algorithms}

\subsection{B-trees}

\subsection{CO vs CA}

\subsection{CO B-trees}

\section{Background}

\subsection{Original CO B-trees}

\subsection{Structure}

\subsection{Layout}

\subsection{Complexity}

\subsection{Operations}

\section{Variable Length Strings}

\subsection{Description}

\subsection{Operators}

\subsection{Complexity}

\section{Streaming B-trees}

\subsection{Description}

\subsection{Operators}

\subsection{Complexity}

\section{Concurrency}

\subsection{Description}

\subsection{Operators}

\subsection{Complexity}

%\end{document}  % This is where a 'short' article might terminate

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
Thanks Karger

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
%\balancecolumns
%\appendix
%%Appendix A
%\section{Headings in Appendices}
%The rules about hierarchical headings discussed above for
%the body of the article are different in the appendices.
%In the \textbf{appendix} environment, the command
%\textbf{section} is used to
%indicate the start of each Appendix, with alphabetic order
%designation (i.e. the first is A, the second B, etc.) and
%a title (if you include one).  So, if you need
%hierarchical structure
%\textit{within} an Appendix, start with \textbf{subsection} as the
%highest level. Here is an outline of the body of this
%document in Appendix-appropriate form:
%\subsection{Introduction}
%\subsection{The Body of the Paper}
%\subsubsection{Type Changes and  Special Characters}
%\subsubsection{Math Equations}
%\paragraph{Inline (In-text) Equations}
%\paragraph{Display Equations}
%\subsubsection{Citations}
%\subsubsection{Tables}
%\subsubsection{Figures}
%\subsubsection{Theorem-like Constructs}
%\subsubsection*{A Caveat for the \TeX\ Expert}
%\subsection{Conclusions}
%\subsection{Acknowledgments}
%\subsection{Additional Authors}
%This section is inserted by \LaTeX; you do not insert it.
%You just add the names and information in the
%\texttt{{\char'134}additionalauthors} command at the start
%of the document.
%\subsection{References}
%Generated by bibtex from your ~.bib file.  Run latex,
%then bibtex, then latex twice (to resolve references)
%to create the ~.bbl file.  Insert that ~.bbl file into
%the .tex source file and comment out
%the command \texttt{{\char'134}thebibliography}.
%% This next section command marks the start of
%% Appendix B, and does not continue the present hierarchy
%\section{More Help for the Hardy}
%The acm\_proc\_article-sp document class file itself is chock-full of succinct
%and helpful comments.  If you consider yourself a moderately
%experienced to expert user of \LaTeX, you may find reading
%it useful but please remember not to change it.
\balancecolumns
% That's all folks!
\end{document}

### 2011-12-07 paper.tex
Adding the citations
small fixes
initial work on concurrent paper

@@
 
-\documentclass{acm_proc_article-sp}
+\documentclass{style}
+
+\usepackage{paralist}
 
@@
 \affaddr{MIT CSAIL}
+\email{saahmad@mit.edu}
 % 2nd. author
@@
 \affaddr{MIT CSAIL}
+\email{leibatt@mit.edu}
 % 3rd. author
@@
 \affaddr{MIT CSAIL}
+\email{stephent@mit.edu}
 }
@@
 
+\newcommand{\lhyperceil}{\lceil\lceil}
+\newcommand{\rhyperceil}{\rceil\rceil}
+\newcommand{\lhyperfloor}{\lfloor\lfloor}
+\newcommand{\rhyperfloor}{\rfloor\rfloor}
+
+\newcommand{\Search}{\textsc{Search(k)}}
+\newcommand{\Insert}{\textsc{Insert(k, v)}}
+\newcommand{\Scan}{\textsc{Scan(a, b)}}
+
 \maketitle
@@
 \section{Introduction}
+
+\cite{BenderFaFi07}
 
@@
 
+\subsection{Motivation}
+B-trees are often used as an indexing data structure for various
+database and file systems. Thus, for a cache-oblivious B-tree data structure
+to be of practical value, it must be able to handle concurrent access and modification
+of its internal state. The straw-man solution is to simply ensure mutual exclusion
+on the entire data structure; this approach, however, yields poor performance.
+
+Various concurrency schemes for traditional cache-aware B-trees have been
+proposed \cite{BayerS77, LehmanY81}. We now survey a concurrency scheme for CO
+B-trees, first presented in \cite{BenderFiGi05}. Bender develops two different
+data structures to solve this problem: the first is an exponential CO B-tree,
+the second is a CO B-tree based on a packed memory array (PMA). Bender
+also presents a lock-free variant of the latter data structure. For
+the remainder of the discussion, assume that the keys are fixed-size integer
+keys.
+
 \subsection{Description}
+Before describing the data structures in more detail, we first consider
+the concurrency model which is used throughout \cite{BenderFiGi05}.
 
-\subsection{Operators}
+\subsubsection{Concurrency Model}
+The concurrency model starts with a system that has a total cache
+size of $M$ and block size of $B$. There are $P$ processors in the
+system, each with a cache of size $\frac{M}{P}$. A particular block
+can reside in multiple caches in \textit{shared} mode. However, if a
+processor wants to perform a write to a block, it must acquire it in 
+\textit{exclusive} mode, in which case that block can only reside in
+a single cache. In this model, a request for a block costs a single
+memory transfer.
 
-\subsection{Complexity}
+For concurrency control, two primitives are used. The first is
+locks. The second is \textit{load-linked/store-conditional} (LL/SC).
+The authors argue that LL/SC provides better performance than 
+using a \textit{compare-and-swap} (CAS) primitive, but their 
+techniques can be extended to use CAS.
+
+\subsection{Exponential CO B-tree}
+This data structure is based on a strongly weight-balanced 
+exponential tree found in \cite{Bender2002}. A tuning parameter
+$1 < \alpha < 2$ is chosen for the data structure. Each node
+in the tree maintains both keys which divide up its children,
+in addition to a \textit{right-link} pointer, which contains
+both a pointer to a node's right sibling, plus the key
+value of the minimum key in the node's right sibling. If no such
+sibling exists, a sentinel $\infty$ pointer is used.
+
+A node's layout in memory is as follows. Suppose a node has $k$
+elements in it. The first part of the node consists of a static
+CO search B-tree, containing the $k$ elements. The tree is sized
+to be a complete $\lhyperceil k \rhyperceil$-leaf tree. The second
+part of the node consists of a $\lhyperceil k \rhyperceil$-size array
+contained the $k$ elements in sorted order. The leaf nodes of the
+CO search tree point to the associated entries in the array.
+
+\subsubsection{Operations}
+The operations supported by the exponential CO B-tree are \Search{}
+and \Insert{}. The authors do not discuss range-queries, and this
+is presumably due to the concurrency issues in implementing a fully-consistent
+range scan (weakly-consistent range scans could be implemented in the data
+structure with very little modification), although interestingly enough
+the PMA variant of B-tree (discussed in Section \ref{sec:pma}) does
+cover range scans.
+
+\Search{} works as follows. For a non-leaf node, first check the
+\textit{right-link} key. If the \textit{right-link} key is less than
+$k$, then traverse the right link, and recurse. Otherwise, follow
+the appropriate child pointer. When a leaf node is reached, only
+follow \textit{right-link} pointers, until either $k$ has been located
+or some element greater than $k$ is reached. No locks are acquired
+during the entire search operation.
+
+\Insert{} works as follows. First, \Search{} is used to locate the leaf node
+where $k$ will be inserted into. A lock is acquired on that node, $k$ is
+inserted, and the lock is released. Unlike a standard B-tree, there is no
+maximum number of elements in a node; instead we split nodes probabilistically.
+With probability ${1}/{2^{\alpha^h}}$, a node at height $h$ is split (leaf
+nodes have $h = 0$). A split works as follows.  We reacquire a lock on the node
+to be split. We split the node into $u$ and $u'$, where $u$ contains all the
+keys $< k$, and $u'$ contains the keys $\geq k$. We release the lock, and
+recursively insert $k$ into the parent of the split node. It is important
+to note that, since $u$ contains a prefix of the original node to be split,
+that we reuse the original node in constructing $u$ (that way, concurrent
+searches can still locate the keys in $u$).
+
+%TODO: some section about linearizability?
+
+\subsubsection{Complexity}
+
+\subsection{Packed Memory Array CO B-tree}
+\label{sec:pma}
+The PMA CO B-tree is a two level data structure, containing a static
+CO B-tree which indexes into a packed-memory array. Bender introduces
+a new packed-memory array, called a \textit{one-way packed-memory array},
+which they argue is easier to manage concurrently.
+
+\subsubsection{One-Way PMA Primer}
+
+A one-way packed-memory array is a contiguous memory structure which has three
+distinct regions: a leftmost region of empty space, an \textit{active region}
+which contains the elements of the array, and a rightmost region of empty
+space. The array only grows into the rightmost region, and shrinks from the
+leftmost region (hence one-way).  The PMA is of size $m = \Theta(N)$, subject
+to $m$ being a power of $2$, where $N$ is the number of elements.
+
+When the \textit{active region} becomes too large/small due to inserts/deletes,
+the PMA is rebalanced. A new array is allocated, and the elements are
+spread out evenly over the new array (too minimize element density).
+
+Each leaf in the static CO B-tree used is a pointer into a $\Theta(\log
+N)$-size region in the packed-memory array.
+
+\subsubsection{Operations}
+
+\Search{} works as follows. Search the static tree for $k$ until
+a leaf node is reached. Follow the leaf node pointer into the PMA.
+Scan right in the PMA until either $k$ is located, or some element
+$> k$ is reached. Do not acquire any locks during the search.
+
+\Insert{} works as follows. Use \Search{} to locate the $\Theta(\log N)$-region
+of the PMA which (would) contain $k$. Lock the region, and scan right to find
+the appropriate slot $s$ to place $k$ into. $s$ is defined such that slot $s -
+1$ contains the largest key smaller than $k$. If scanning forward requires
+going into the next $\Theta(\log N)$-region, lock the next region and release
+the current region (\textit{hand-over-hand} locking technique). Once $s$
+is located, either $s$ is free, in which case just insert $k$, or $s$ is not free,
+in which case a \textit{rebalance} operation is required.
+
+To do a rebalance operation, we scan right (locking as necessary) until the
+region from $s$ to the current position $s'$ is not \textit{too dense}. The
+notion of density is defined in detail in \cite{BenderFiGi05}. The region
+$[s, s']$ is then rebalanced as follows. Start from the rightmost element,
+and move elements to the right (maintaining order) such that no suffix of
+$[s, s']$ is too dense. This rebalancing creates room for $k$; we insert $k$
+and then release the locks.
+
+\subsubsection{Complexity}
+
+\subsubsection{Lock-free Variant}
+The lock-free variant is based on the PMA CO tree, except the locks acquired
+during the PMA modification are replaced by augmenting each element in the PMA
+with a marker. A marker is used to denote the ongoing (concurrent) operation
+of a particular cell.
+
+In addition, four non-blocking primitives are introduced:
+\begin{inparaenum}[(a)]
+  \item \textit{move},
+  \item \textit{cell-insertion},
+  \item \textit{cell-deletion}, and
+  \item \textit{read}
+\end{inparaenum}. These primitives are self-explanatory, except they
+all work in a non-blocking manner (they can fail at any point).
+
+The concurrency protocol works as follows. Any of the non-blocking primitives
+are initiated by denoting the cell in question with the marker. If a process
+ever encounters a cell with a marker, it drops what it is doing and first helps
+to complete the operation. The implementation of the non-blocking primitives
+is spelled out in more detail in \cite{BenderFiGi05}, but the basic idea is
+using LL/SC to detect when concurrent operations are on-going.
+
+\subsubsection{Lock-free Variant Operations}
+
+\Search{} is un-modified from the lock-based version.
+
+\Insert{} is modified as follows. Cell $s$ is located as before. A
+\textit{cell-insertion} operation is performed on $s$. If it succeeds, we are
+done. If it fails, then a rebalance is necessary. Rebalance proceeds as before,
+scanning right to find the region $[s, s']$. However, whenever a non-empty cell
+is encountered, a load-link (LL) is performed on the cell.  Once the region is
+found, the rebalancing-to-the-right is performed as before.  However, a
+store-conditional (SC) is performed on the destination of each cell.  If any SC
+fails, then this means concurrent modification had occurred, in which case the
+algorithm restarts from the beginning.
 
@@
 \bibliographystyle{abbrv}
-\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
+\bibliography{citations}  % sigproc.bib is the name of the Bibliography in this case
 % You must have a proper ".bib" file

### 2011-12-07 paper.tex
more progress on concurrency paper

@@
 \subsubsection{Complexity}
+Sequential search takes $O(\log_B{N} + \log_{\alpha}{\log_2{B}})$ block
+transfers (with high probability). The probabilistic bound comes from the
+probabilistic key promotion property. The intuition behind the proof of this
+bound comes from the argument that when the tree has height $h =
+\Omega(\log_{\alpha}{\log_2{\log_2{N}}})$, then the total number of keys in the
+tree is $O(2^{\alpha^h})$. The proof is then completed by summing over the
+range of $h$ which is probabilistic $O(\log_{\alpha}{\log_2{N}})$, the access
+time from some level $h$ to find the correct child.
+
+Sequential insert takes $O(\log_B{N} + \log{\alpha}{\log_2{B}})$ block
+transfers (in expectation). This proof is constructed in a similar 
+fashion as the search proof: consider inserting a tree at some height $h$.
+By noting that the probability a particular node reaches height $h$
+is $2^{-(\alpha^h-1)/(\alpha-1)}$, a summation over all levels
+of the cost of rebuilding a node yields the time-bound.
 
@@
 
+% TODO: can somebody check this time-bound? seems kind of bad
+Sequential search time-bounds are not presented in the paper,
+but presumably since search is a static CO tree lookup followed
+by a scan, the time bound is $O(\log_{B}{N} + {\log_2{N}}/{B} + 1)$.
+
+Sequential insert is $O(\log_{B}{N} + \log_2^2{N}/{B} + 1)$. The proof
+is based off an accounting argument which places $\Theta(\log_2^2{m}/B)$
+dollars in each array slot. The proof is not presented in any detail in the
+paper, but rather it is a slightly-modified version of the proof presented
+in \cite{Katriel02}.
+
 \subsubsection{Lock-free Variant}
@@
 ever encounters a cell with a marker, it drops what it is doing and first helps
-to complete the operation. The implementation of the non-blocking primitives
-is spelled out in more detail in \cite{BenderFiGi05}, but the basic idea is
-using LL/SC to detect when concurrent operations are on-going.
+to complete the operation. When the originating process gets context-switched
+back in, any SCs will fail, and then the process can check to see if the
+operation was already completed. The implementation of the non-blocking
+primitives is spelled out in more detail in \cite{BenderFiGi05}, but the basic
+idea is using LL/SC to detect when concurrent operations are on-going.
 
@@
 algorithm restarts from the beginning.
+
+\section{Conclusion}
 

### 2011-12-08 paper.tex
Streaming section.

@@
 \newcommand{\Scan}{\textsc{Scan(a, b)}}
+\newcommand{\Delete}{\textsc{Delete(k)}}
 
@@
 \section{Introduction}
-
-\cite{BenderFaFi07}
 
@@
 \subsection{Original CO B-trees}
+\label{sec:original}
 
 \subsection{Structure}
+\label{sec:structure}
 
 \subsection{Layout}
+\label{sec:layout}
 
@@
 
-\subsection{Operators}
+While B-trees achieve good performance for searches, they do not perform as
+well for updates and insertions. The buffered-repository tree (BRT) is a
+common alternative to the B-tree achieves an improved amortized $O(log(N)/B)$
+memory transfers for inserts compared to the $O(log_{B+1}N)$ memory access of
+B-trees.
 
-\subsection{Complexity}
+The search vs insertion trade off also exists in the cache-oblivious model.
+The original cache-oblivious B-trees described in Section~\ref{sec:original}
+achieves the search end of this trade off. However, Bender et al. propose two
+data structures that and achieve the opposite end: the shuttle tree and the
+cache-oblivious lookahead array (COLA) \cite{BenderFaFi07}. These two data
+structures are refereed to as ``streaming B-trees'', indicating how they are
+optimized for insertions at the cost of search.
+
+\subsection{Shuttle Trees}
+
+The shuttle tree is similar in layout and structure as the original CO B-Tree.
+It uses a strongly weight-balanced search tree described previously (see
+Section~\ref{sec:structure}) that is laid out in a vEB layout into a PMA. A
+notable difference is that the CO B-tree splits the tree at height $h/2$ when
+creating the vEB layout while the shuttle tree splits at height $F_k$ where
+$F_k$ is the $kth$ Fibonacci number that is stickily less than the height of
+the tree. Thus if a tree's height happens to be a Fibonacci number, $F_k$,
+then the shuttle tree will choose to split the tree at height $F_{k-1}$. The
+top and bottom ``halves'' produced by this split are organized just like
+before and shown in Figure ???.
+
+The shuttle tree improves the efficiency of insertion by adding buffers to
+each node in the tree. In fact, shuttle tree nodes have a linked list of
+buffers rather than a single buffer like BRT and each of these buffers is
+another shuttle tree. The length of a node's linked list is related to the
+node's height in the tree. A node has a buffer for every time it the leaf of a
+sub trees that is split when creating the vEB layout. This idea is illustrated
+in Figure ???.
+
+The shuttle tree's node's buffers have a maximum capacity that is enforced by
+bounding it height. The size of the buffers increases going across the linked
+list. Thus, a node's first buffer is smaller than its last buffer. The maximum
+height of a buffer is given by $F_{H(k)} = F_{k-2 \lceil log_{\phi} k \rceil}
+$ where $\phi$ is the golden ratio and $k$ is chosen such that $F_k$ =
+$\xi(h)$ and $\xi(x)$ is defined recursively. If $h$ is a Fibonacci number
+then $\xi(h) = h$ and otherwise $\xi(h) = \xi(h-f)$ where f is the largest
+Fibonacci number less than $h$. The size of the buffer is related to the size
+of the recursive subtree for which the node is a leaf. Since the recursive
+subtrees are split based on the largest Fibonacci number less than $h$ it
+follows that the buffer sizes are also based on Fibonacci numbers.
+
+\subsubsection{Operators}
+
+The operations that are described by Bender et al. for the shuttle tree are
+\Search, \Insert, and \Scan. An explicit discussion on \Delete is missing;
+however, it is assumed that it is symmetric to \Insert.
+
+\Search and \Scan work virtually the same way as the CO B-tree. The only
+difference is that you have to search a child node's buffer before following
+the pointer down the tree.
+
+\Insert is also similar to the CO B-tree except you buffer elements as you add
+new elments. When you insert an element you move down the tree from the root
+to leaves and stop when you find a node that has a buffer (it is possible that
+a node does not have any buffers because of the height of the tree). You then
+attempt to insert the element into the smallest buffer in the node's linked
+list. If this insertion causes the buffer to overflow (e.g. the height of the
+buffer exceeds its maximum), you move all of the elements from this buffer to
+the next largest buffer. When the largest buffer overflows you take all of its
+elements and ``shuttle'' them down the tree. Eventually you will hit a leaf
+node and inserting here may cause the balance constraints on the SWBST to
+become invalidated. The shuttle tree rebalances itself just like any other
+SWBST: the leaf node is split into two and this cascades up the tree until
+balance is restored.
+
+When a node $u$ is split into two nodes $u1$ and $u2$ we immediately allocate
+enough space for all of the $u2$'s buffers. We then scan over $u$'s buffered
+sub trees and find the ones rooted in $u1$ and $u2$ and move them
+respectively. This can be achieved by a single scan. Once we have found the
+set of subtrees that correspond to $u1$ and $u2$ we replace the original sub
+trees for $u$ in the vEB layout.
+
+The core problem of maintain room for inserts in the shuttle tree does not
+change from the CO B-tree and is handled by the PMA just like in described in
+Section~\ref{sec:layout}.
+
+\subsubsection{Complexity}
+
+The complexity of \Search is $O(log_B N)$ where $N = \Theta(c^{F_k})$ and $c$
+is your fanout. Note that $c$ is not treated as a constant since this is a
+cache-oblivious data structure. This bound is achieved by searching over the
+top and bottom recursive subtrees at each level. A key trick is that when you
+visit a node you only search its largest buffer (that last buffer in the
+linked list). The smaller buffers will be visited eventually as you recurse
+down the tree. Eventually you will get down to a single subtree that can fit
+into $O(1)$ blocks and thus can be search with $O(1)$ memory accesses. Since
+the fanout is proportional to our block size it will take us $O(log_B N)$
+block accesses until we get down to a subtree and thus, by induction, the
+worst case runtime is $O(log_B N)$.
+
+\Insert uses $O(log_{B+1} N / B^{\Theta(1/(loglogB)^2)} + (log^2 N) / B)$
+memory transfers. This improvement of the the CO B-tree is achieved because
+the shuttle tree's buffers are used to maintain locality of the elements long
+enough to amortize the cost of moving them down the tree and possible into
+another block. The proof is based on the fact that when we move an item from
+one block to another, caused by splitting or buffers being overfilled, at
+least $B^{\Theta(1/(loglogB)^2)}$ will move with that item. This is because,
+with insert, when a buffer overflows we move all of the elements from that
+buffer down the tree, not just the single one that caused the overflow. Thus,
+we amortize the cost of a single insert by the number of elements in a node's
+largest buffer.
+
+The complexity to \Scan $L$ elements is amortized to $O(L/B)$ memory
+transfers. First you perform a \Search to get to the starting element of your
+range. Once you have a block in memory you can start scanning in elements and
+since you only have to read a new block every $B$ accesses each element access
+is amortized to $1/B$ and thus overall the number of memory access is
+$O(L/B)$. This, of course, assumes that you just performed a search for the
+starting point of your scan.
+
+The last thing that should be considered is the space requirement of the
+shuttle tree. However, the overall space requirements of the shuttle tree is
+actually just $O(n)$. The high level intuition for this proof is that the
+majority of the buffers are really small and thus do not impact the overall
+size of the tree.
+
+\subsection{COLA}
+
+The cache-oblivious lookahead array (COLA) is a cache-oblivious dictionary
+that is actually quite different from the other cache-oblivious data
+structures seen so far. Instead of an explicit tree structure with child and
+parent pointers, the COLA is organized as a list of arrays or levels. Each
+level has arrays of size $2^k$. Each level is either completely full or
+completely empty. As such, the COLA is very similar to the binomial list
+structure described in \cite{BentleySaxe}.
+
+\subsubsection{Operators}
+
+The operations supported by COLA are \Search, \Insert, and \Scan.
+
+\Insert works by initially inserting the element into the first-level array.
+This array has size $2^0 = 1$, thus the element is in an array on it own.
+Whenever you insert an element you check to see if there is another list of
+the same size and merge them together in sorted order. This merging will
+produce a new array of twice the size which goes into the next level. We
+continue to merge until there are no two arrays of the same length.
+
+\Search can naively be performed by simply scanning each of the levels in our
+data structure. Since everything is sorted we can use a binary search to
+quickly find the element we are looking for. Obviously this is not very
+efficient, thus COLA, as the name suggest, uses look ahead pointers to reduce
+the number of memory accesses required. Every 8th element from the $k+1$ level
+exists in the $k$ level array in its proper sorted location. This element also
+contains a pointer to its real location in the $k+1$ level, thus it is
+referred to as a ``real lookahead pointer''. Additionally, each every 4th cell
+in $k$ level array contains two pointers pointing to the next and previous
+``real lookahead pointers'' on this level. Therefore, to search a level we
+perform a binary search and automatically cascade up to the next level.
+
+\subsubsection{Complexity}
+
+The complexity of \Insert is amortized to $O(log N / B)$ block transfers.
+First, it is important to know that an element will never be merged more than
+$O(log(N))$ times since at that point, it will be in the last array. The first
+$log(B)$ merges that a element participates is constant time since we can
+assume that the relevant blocks are already in memory. Thus, we only need to
+worry about memory accesses when we merge levels that have size greater than
+$B$. If the size of the array is $k$ then the cost of merging is $O(k/B)$ and
+thus the amortized cost of a merging a single element is simply $1/B$. Since
+an element will only be merged $log(N)$ time, the overall cost is amortized to
+$O(logN/B)$.
+
+The complexity of \Search is $O(log N)$ memory transfers. The proof is based
+on the fact that we only ever need to search at most eight items on each
+level. Since there are $O(log(N))$ levels and since eight is a constant it
+follows that overall, search accesses $O(log N)$ blocks. A proof by induction
+shows that only eight items need to be searched on each level. The base case
+is trivial. The first three levels of the COLA each have at most eight
+elements that either contain the actual elements or a look ahead pointer to
+the next level. Obviously we can search these levels with no more than eight
+member transfers. As we follow the lookahead pointers to the high levels we
+will eventual get to a level where $k \geq 3$. Note that we do not enter this
+level at the start; rather, we arrived here by following the real lookahead
+pointers from the levels below and are at some index $i$ that contains an
+element $e_i$ such that $e_i < e$ where $e$ is the element we are looking for.
+Our goal is to find a new $i$ such that $e_i \leq e < e_{i+1}$. Once we find
+this $i$ then we can use the duplicate look ahead pointer to find the ``next''
+and previous look ahead pointers in the next level and restrict our search to
+that range. Since we have a real look ahead pointer for every eighth element
+in the next level, it follow that when we get to the next level, we only need
+to search eight elements. Furthermore, since we arrived at our current level
+from a lower level, we only need to search $8$ elements to find this $i$.
+Thus, overall, we only need to perform $O(8log(N)) = O(log(N))$ memory
+transfers.
+
+\subsubsection{Performance}
+
+In experimental results, COLA achieved significant improvements for random
+inserts, performing 790 times faster than a normal B-tree. Sorted inserts are
+3.1 times slower and searches are 3.5 times slower. These experimental results
+indicate that the the COLA is a suitable data structure for streaming data
+structure where random insertions are common.
 

### 2011-12-08 paper.tex
Adding the figures

@@
 \usepackage{paralist}
+\usepackage{graphicx}
 
@@
 \begin{abstract}
-Fill me in
+
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
 \end{abstract}
@@
 
-\subsection{Structure}
+Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in \cite{cob tree}. The
+CO B-tree was able to match the optimum search bound of $\Theta(log_B N)$ memory transfers 
+without including any memory-specific parameterization describe the speed or sizes
+of the machines memory hierarchy.
+
+The CO B-tree is able to achieve this bound by structuring its elements as a strongly weight-balanced
+search tree (SWBST). The SWBST is then laid out into a packed memory array (PMA) using a van Emde Boas (vEB) layout.
+
+\subsection{Structure - SWBST}
 \label{sec:structure}
 
-\subsection{Layout}
+\subsection{Layout - PMA}
 \label{sec:layout}
+
+\begin{figure}
+
+\begin{center}
+	\includegraphics[width=0.8\columnwidth]{figures/veb.png}
+\end{center}
+
+\caption{An illustration of the van Emde Boas (vEB) layout}
+\label{fig:veb}
+\end{figure}
 
@@
 top and bottom ``halves'' produced by this split are organized just like
-before and shown in Figure ???.
+before and shown in Figure \ref{fig:veb}.
+
+\begin{figure}
+
+\begin{center}
+	\includegraphics[width=0.8\columnwidth]{figures/veb.png}
+\end{center}
+
+\caption{An illustration of how the shuttle tree determines the size of a node's linked list}
+\label{fig:buffers}
+\end{figure}
 
@@
 sub trees that is split when creating the vEB layout. This idea is illustrated
-in Figure ???.
+in Figure \ref{fig:buffers}.
 

### 2011-12-08 paper.tex
Adding my stuff.

@@
 
+This paper presents a survey of cache-oblivious B-tree data structures. Memory hierarchies . 
+Since memory transfers often dominate the runtimes of an algorithm it makes sense to
+develop data structure that are optimized not in terms of the number of instructions it executes; but rather, 
+by the number of memory accesses and transfers it has to perform. Among these so called ``external-memory'' data structures,
+the B-tree is widely used as it is able to perform \Search in an optimal number of memory transfers. 
+However, a short coming of B-trees is that they require knowledge of the size of the memory blocks, thus 
+it is a cache aware data structure. This gave rise to the notion of cache-oblivious algorithms, e.g. data structures
+that perform well regardless of the size of the memory blocks or the number hierarchies. Bender et al. 
+introduced the cache-oblivious B-tree which ...
+
 Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
@@
 
-Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in \cite{cob tree}. The
+Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in \cite{cobtree}. The
 CO B-tree was able to match the optimum search bound of $\Theta(log_B N)$ memory transfers 
-without including any memory-specific parameterization describe the speed or sizes
-of the machines memory hierarchy.
+without including any specific parameterization describing the speed or sizes
+of the machine's memory hierarchy.
 
-The CO B-tree is able to achieve this bound by structuring its elements as a strongly weight-balanced
-search tree (SWBST). The SWBST is then laid out into a packed memory array (PMA) using a van Emde Boas (vEB) layout.
+The CO B-tree paper presents three different data structures that make 
+different trade off in the number of memory transfer they perform. However, all of them 
+achieve their performance by structuring their elements as a strongly weight-balanced
+search tree (SWBST). The SWBST is then laid out into a packed memory array (PMA) 
+using a van Emde Boas (vEB) layout. The SWBST and PMA play crucial roles in the
+other cache-oblivious data structures that were developed after the CO B-tree and thus are
+described at length in the next two sections.
 
@@
 
+Strongly weight-balanced search trees play a key role in CO B-tree. The weight of a node is $u$ is equal to
+the total number of $u$'s descendants plus 1. It can als be thought of as the sum of the weights of $u$'s children plus 1.
+Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.
+
+Most weight-balanced search trees which requires that the heights of a node's left and right subtrees are
+simply off by a constant factor. SWBST enforce the stronger property that a node at height $h$ in the tree (height is the node's
+distance from the leaves; leaves have a height of 1) has $\Theta c^h$ where $c$ is some constant.
+
+A known data structure that achieves this property is the ``weight-balanced B-tree'' \cite{paper}. The CO B-tree leverages this
+data structure to maintain its elements. The next section describes how SWBST perform various operations
+while maintain their strongly weight-balanced property.
+
+\subsection{Operations}
+
+The CO B-tree supports \Search, \Insert, \Delete, and \Scan operations.
+
+\subsection{Complexity}
+
+It should be noted, that keeping the level of indirection sorted is only needed in case we want
+to perform \Scan operations. If we do not need \Scan, then we can keep the indirection array in any arbitrary order
+reducing the cost of \Insert and \Delete by the $(log^2 N) /B$ factor making them incur $O(log_{B} N)$ memory transfers
+overall.
+
 \subsection{Layout - PMA}
 \label{sec:layout}
+
+SWBST is only one half of the CO B-tree story...
 
@@
 \end{figure}
-
-\subsection{Complexity}
-
-\subsection{Operations}
 

### 2011-12-08 paper.tex
adding my changes

@@
 
+\subsection{Motivation}
+In order for B-trees to remain versatile, they must be able to accommodate various forms of data.
+However, traditional B-trees and string B-trees have suboptimal performance when variable length strings are used as keys.
+In \cite{BenderFaKu06}, Bender et al.
+describe 4 major issues:
+\begin{itemize}
+\item B-trees handle variable length keys poorly
+\item string B-trees do not support compression, and the front compression used in many other B-trees is handled sub-optimally
+\item range queries are handled sub-optimally by regular B-trees, since they result in random disk seeks
+\item traditional and string B-trees assume fixed block-transfer size B at all levels of the memory hierarchy, resulting in sub-optimal performance over hierarchies with varied block-transfer sizes.
+\end{itemize}
 \subsection{Description}
+They address these issues by designing a Cache Oblivious String B-tree.
 
+\subsubsection{General COSB-tree Structure}
+The static COSB-tree consists of 2 layers: a centroid tree at the top, and an array of the keys stored in lexicographic order on the bottom.
+The leaves of the centroid tree contain pointers to keys in the array.
+The centroid tree is structured such that a search through the centroid tree identifies the local area of the desired key in the array.
+Then a sequential scan is performed across the array to find the key.
+
+The Dynamic COSB-tree consists of 3 major components: a centroid tree at the top; a packed memory array (PMA) structure of hash data for the keys known as the \emph{hashdata} layer; and on the bottom a PMA structure for the original key values, compressed using an augmented version of forward compression known as the \emph{keydata} layer.
+
+The centroid tree on a high level behaves similarly for static and dynamic COSB-trees.
+The leaves of the centroid tree point into the hashdata layer, where a sequential scan is performed to find the correct key by hash value.
+The hashdata layer points directly into the keydata layer, where the appropriate key is decoded, compared and returned.
+
+In the following subsections, centroid trees are explained, and a summary of how to modify them to preserve weight balance in the dynamic case is provided.
+Two iterations on forward compression are also summarized to show how to incrementally add support for variable length strings as keys for dynamic COSB-trees.
+Lastly, the design and use of the hashdata and keydata are explained.
+
+\subsubsection{The Dictionary Matching Problem and Centroid Trees} %todo:cite dictionary matching problem--sources in cosb-tree paper
+In this approach to solving the dictionary matching problem, a dictionary $D$ of $N$ keys is preprocessed using hashing and divide-and-conquer via the construction of a centroid tree over the keys.
+This technique forms the basis for the Cache Oblivious String B-tree (COSB-tree) design.
+Consider the compacted trie $T$ of $D$.
+Bender et al.
+observe that there exists a centroid vertex $\rho$ in $T$ such that $\rho$ has between N/3 and 2N/3 descendants, inclusive.
+When searching for a key $k$ in the compacted trie, you traverse the tree as follows: for each node $y$ that you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on into the trie rooted at $y$ or \emph{down trie} of $y$.
+Otherwise traverse the trie resulting from excluding $y$ and its subtree.
+The \emph{centroid tree} is obtained by rooting the tree at \emph{centroid vertex} $\rho$, and making its children recursively defined \emph{centroid trees} of its up and down tries.
+
+At each step through the centroid tree, a consistent fraction of the nodes are eliminated from the search, resulting in an $O(logN)$ time bound when ignoring the time required to operate on keys.
+To make key comparisons faster, the keys are hashed and hash values are compared in place of the original key values.
+To avoid mismatches due to overlap in hash values, the original key values are compared after finding a match in hash values.
+The resulting time bound for \Search{} (interchangeable with \Member{}) is $O(||k||+logN)$.
+If each node in the centroid tree knows its successor and predecessor descendants, \Pred{} and \Succ{} can be achieved in $O(||k||+||k'||+logN)$, where $k'$ is the successor or predecessor of $k$.
+Note that the additive $logN$ incurred by the use of centroid trees can be avoided when using main memory.
+Centroid trees are used to capitalize on locality for the cache oblivious string B-tree design, described later in this section.
+
+For COSB-trees, the centroid tree is built using $O(N/logN)$ keys, and the centroid tree is layed out in memory using a modified van Emde Boas (vEB) layout for weight-balanced binary trees.
+The \emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into equal-size subtrees for the vEB layout.
+The \emph{hyperhyperfloor} of x is denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to the nearest power of a power of 2.
+This ensures trees of equal size on all \emph{levels of detail}, as the centroid tree is traversed.
+
+\subsubsection{Modifying Front Compression to Preserve Locality}
+This section provides a brief summary of how front compression is modified to optimize compression and decoding for static and dynamic cache-oblivious string B-trees.
+Consider a sequence of $j$ keys to store $k_{1}$,  $k_{2}$, ...,  $k_{j}$.
+Without compression, this requires $\sum_{j}||k_{j}||$ space in memory.
+To compress these keys using front compression, let $\pi_{j}$ be the longest common prefix between $k_{j}$ and $k_{j+1}$.
+Now instead we store the keys as follows:
+\begin{center}
+$k_{1}$,$||\pi_{2}||$,$\sigma_{2}$,$||\pi_{3}||$$\sigma_{3}$,...,$||\pi_{j}||$,$\sigma_{j}$
+\end{center}
+where $\sigma_{j}$ is the suffix of $k_{j}$ after removing the first $\pi_{j}$ bits.
+This saves almost $\sum_{j}||\pi_{j}||$ memory.
+However, since decoding $k_{j}$ involves concatenating the first $\pi_{j}$ bits of $k_{j-1}$ to $\sigma_{j}$, $k_{j-1}$ must be decoded.
+$k_{j-1}$ may depend on previous keys as well, making $k_{j}$ expensive to decode.
+
+The goal of modifying front compression is to maintain some of the space benefits to compression while reducing the high decoding cost and keeping the size of D small.
+Bender et al.
+provide a modification that reduces the high decoding cost of front compression of D to $O(1+||k||B)$, with size $O(\ll{D}\gg)$ after compression.
+The modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be the first key added to the compression stream.
+For each subsequent key $k_{i}$, if $k_{i}$ can be decoded with just the last $c||k_{i}||$ characters, add $\pi_{i}$,$\sigma_{i}$ like before.
+Otherwise, add 0,$k_{i}$, which means add $k_{i}$ to the stream in its uncompressed form.
+By construction, this scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.%todo: add something about why this isO(<<D>>)?
+%todo: add another paragraph or two about the second iteration
 \subsection{Operators}
+The operations supported are
 

### 2011-12-09 paper.tex
added my macros
Adding the first pass at the introduction

@@
 \newcommand{\Delete}{\textsc{Delete(k)}}
+\newcommand{\Succ}{\textsc{Succ(k)}}
+\newcommand{\Pred}{\textsc{Pred(k)}}
+\newcommand{\Member}{\textsc{Member(k)}}
+%\newcommand{\Range}{\textsc{Range-Query(k$_{i}$,k$_{j}$)}}
 
@@
 
-This paper presents a survey of cache-oblivious B-tree data structures. Memory hierarchies . 
-Since memory transfers often dominate the runtimes of an algorithm it makes sense to
-develop data structure that are optimized not in terms of the number of instructions it executes; but rather, 
-by the number of memory accesses and transfers it has to perform. Among these so called ``external-memory'' data structures,
-the B-tree is widely used as it is able to perform \Search in an optimal number of memory transfers. 
-However, a short coming of B-trees is that they require knowledge of the size of the memory blocks, thus 
-it is a cache aware data structure. This gave rise to the notion of cache-oblivious algorithms, e.g. data structures
-that perform well regardless of the size of the memory blocks or the number hierarchies. Bender et al. 
-introduced the cache-oblivious B-tree which ...
-
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+This paper presents a survey of cache-oblivious B-tree data structures. Memory
+hierarchies . Since memory transfers often dominate the runtimes of an
+algorithm it makes sense to develop data structure that are optimized not in
+terms of the number of instructions it executes; but rather, by the number of
+memory accesses and transfers it has to perform. Among these so called
+``external-memory'' data structures, the B-tree is widely used as it is able
+to perform \Search in an optimal number of memory transfers. However, a short
+coming of B-trees is that they require knowledge of the size of the memory
+blocks, thus it is a cache aware data structure. This gave rise to the notion
+of cache-oblivious algorithms, e.g. data structures that perform well
+regardless of the size of the memory blocks or the number hierarchies. Bender
+et al. introduced the cache-oblivious B-tree which ...
 
@@
 
+Most algorithms assume an idealized model of computation that consists of a
+single CPU, which performs instructions, and memory, which is a flat array
+that can be read and written. However, in reality memory is not a flat array
+but rather a hierarchy from super fast on chip caches, to slower main memory,
+and even all the way down to the disk. When a program requests data, the data
+is found somewhere in the memory hierarchy and typically carried up the
+hierarchy to the fastest memory possible layer when the CPU is able to access
+it. Unfortunately, this fast memory often has small capacity and once it is
+filled and new data is requests some old entries must be evicted. Bring data
+from lower to higher levels in the memory hierarchy often takes a long time.
+In fact, access times between the different layers of the hierarchy can differ
+by many orders of magnitude and often dominate the overall runtime performance
+of an algorithm. External memory algorithms are wary of this fact and aim to
+solve problem by optimizing the number of memory transfers it performs rather
+than the number of instructions it executes. They often try to exploit the
+fact the memory hierarchies transfer data in blocks rather than individual
+atomic units. Thus, external memory algorithms are designed to not only limit
+the number of memory transfers but also preserve the locality of data so that
+when a memory access is performed, its cost is amortized.
+
 \subsection{B-trees}
 
-\subsection{CO vs CA}
+Since External memory algorithms are primarily concerned with efficient access
+to data, it is not surprise that most problems can be reduced to finding a
+suitable data structure that efficiently implement certain operations. A well
+known external memory data structure is the B-tree. At a high level, B-tree
+work by making their nodes the same size as a memory block in the hierarchy.
+Furthermore, the fanout in the tree (the number of children) is also
+proportional to the block size as well. The B-tree incurs $O(log(N))$ memory
+transfers for \Search{}, \Insert{}, and \Delete{} and is widely used in
+databases and filesystems.
+
+\subsection{DAM vs CO}
+
+The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in
+a two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}. However,
+in reality computing architectures typically have memory hierarchies that are
+much larger than two. The argument for the DAM model is that in practice only
+one level in the memory hierarchy tends to dominate the performance and thus a
+two-level memory model is sufficient for all intents and purposes. However,
+this does mean that, theoretically, B-trees can begin to incur an increasing
+number of memory transfers if the dominating level in the hierarchy changes.
+
+Additionally, the B-tree expects to know the the block size in the memory
+hierarchy in advanced. This means that the parameterized in code need to be
+changed for every level in the memory hierarchy which can be very difficult
+and error prone. Additionally, tuning this parameter may even be impossible in
+a heterogeneous computing environment.
+
+The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
+of computation. In the CO model the algorithm / data structure has no
+information about the memory hierarchy. It does not know anything about the
+speed of any level or the block size. At a conceptual level, the CO model
+allow algorithm designers to reason about the simple two-level model like DAM
+but prove results for any arbitrary memory hierarchy. In short, algorithms
+that perform well in a CO model should, in theory, perform well regardless of
+the block size in a memory hierarchy and thus should gracefully scale to
+multi-level hierarchies as well.
 
@@
 
+In \cite{BenderDemainColton} Bender et al. introduce the cache-oblivious B-tree (CO B-tree). The
+CO B-tree is similar to normal the B-tree except that it does not have the
+block size of levels in the memory hierarchy as explicitly parameters.
+
+The CO B-tree is able to achieve the optimal search bound of $\Theta(log(N))$
+memory transfers, the optimal scan bound of $\Theta(N/B)$ memory transfers and
+efficient insertion and deletion operations. The CO B-tree was a seminal data
+structure that gave rise to many others in the future.
+
+This paper provides a survey of these different data structures. We start by
+analyzing the original CO B-tree and explain how it works. This includes a
+detailed description of the packed memory array (PMA) and strongly
+weight-balance binary search trees (SWSBT) that are integral to the CO
+B-tree's performance. Following this discussion we provide an analysis of the
+CO String B-tree,a CO B-tree with variable length keys, the CO Streaming
+B-tree, a CO B-tree that improves the cost of insertion at the expense of
+search, and the CO Concurrent B-tree (which allows multiple processes to
+access the data at the same time).
+
 \section{Background}
+
+High level idea:
+
+We want to have a B-tree that is cache oblivious. Okay, well, first we still
+need a way to actually store our data. We should use a tree. But what type of
+tree? Obviously it needs to be balanced but anything else? Yes, it needs to
+maintain the property that the number of descendants are $O(d^h)$. Why? It has
+to do with how we will ultimately layout the tree in memory, but we will get
+to that later.
+
+Okay, so, we find the tree that we want to use. Now what? Well, first of all,
+we need to make sure that it supports all of the operations that we want while
+maintain this key property. Lets start with searching. We will layout the tree
+in a vEB layout. This is helpful for two reasons. First, it is a nice way to
+store it in memory and second because it gives us a search bound for free. Now
+we need to make sure that it supports all of the other operations while
+maintain the key property. I can show you how it handles insertions and
+delete. Bam!
+
+Okay, so now, we have a data structure that holds the amount of memory that we
+want and it has the property that we will use later. Great! Now what? Well, we
+need to actually store this in memory some how. We said that we will use a vEB
+layout but that is only half of the story. We want a data structure that
+maintain the locality of as many elements as once while also keeping extra
+space to allow for future inserts. This is seemingly contradictory but we can
+do it!
+
+How? we use a PMA. The main problem is what inserts and delete may cause
+splits and merges which may cause huge parts and nodes of the tree to move
+around. So we need to first show how it is able to handle splits and merges
+efficiently in a cache oblivious manner. bam! done! Great!
+
+Now what? We also have the problem of updating pointers. If the points are
+close to one anther that is not a big deal. However, it is possible that the
+nodes are really far apart. Close means they are in the same block and thus
+only incur a single memory transfer. Far apart means they are more. To solve
+this problem we use buffer nodes and that works in a weird way. However, this
+only gives us a bad bound, can we do better? ****
+
+Yes! We use indirection. Indirection allows us to use another array and this
+makes out pointer updates much more efficient.
+
+We can now perform inserts and deletions faster. Additionally, our scans are
+super fast as well. Search have not changed at all because we still have our
+vEB structure and thus we are okay. But, we can also do a little better. If we
+are okay without having scans, we can improve inserts and deletion since we no
+longer have to maintain our indirection array.
+
+The important part is that our analysis above we never treat the block size as
+a constant. It is always a variable in our analysis but we are able to achieve
+the desired bounds. Thus, we have a cache-oblivious B-tree that does
+everything that we want.
 
@@
 
-Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in \cite{cobtree}. The
-CO B-tree was able to match the optimum search bound of $\Theta(log_B N)$ memory transfers 
-without including any specific parameterization describing the speed or sizes
-of the machine's memory hierarchy.
+Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
+\cite{cobtree}. The CO B-tree was able to match the optimum search bound of
+$\Theta(log_B N)$ memory transfers without including any specific
+parameterization describing the speed or sizes of the machine's memory
+hierarchy.
 
 The CO B-tree paper presents three different data structures that make
-different trade off in the number of memory transfer they perform. However, all of them 
-achieve their performance by structuring their elements as a strongly weight-balanced
-search tree (SWBST). The SWBST is then laid out into a packed memory array (PMA) 
-using a van Emde Boas (vEB) layout. The SWBST and PMA play crucial roles in the
-other cache-oblivious data structures that were developed after the CO B-tree and thus are
-described at length in the next two sections.
+different trade off in the number of memory transfer they perform. However,
+all of them achieve their performance by structuring their elements as a
+strongly weight-balanced search tree (SWBST). The SWBST is then laid out into
+a packed memory array (PMA) using a van Emde Boas (vEB) layout. The SWBST and
+PMA play crucial roles in the other cache-oblivious data structures that were
+developed after the CO B-tree and thus are described at length in the next two
+sections.
 
@@
 
-Strongly weight-balanced search trees play a key role in CO B-tree. The weight of a node is $u$ is equal to
-the total number of $u$'s descendants plus 1. It can als be thought of as the sum of the weights of $u$'s children plus 1.
+Strongly weight-balanced search trees play a key role in CO B-tree. The weight
+of a node is $u$ is equal to the total number of $u$'s descendants plus 1. It
+can als be thought of as the sum of the weights of $u$'s children plus 1.
 Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.
 
-Most weight-balanced search trees which requires that the heights of a node's left and right subtrees are
-simply off by a constant factor. SWBST enforce the stronger property that a node at height $h$ in the tree (height is the node's
-distance from the leaves; leaves have a height of 1) has $\Theta c^h$ where $c$ is some constant.
+Most weight-balanced search trees which requires that the heights of a node's
+left and right subtrees are simply off by a constant factor. SWBST enforce the
+stronger property that a node at height $h$ in the tree (height is the node's
+distance from the leaves; leaves have a height of 1) has $\Theta c^h$ where
+$c$ is some constant.
 
-A known data structure that achieves this property is the ``weight-balanced B-tree'' \cite{paper}. The CO B-tree leverages this
-data structure to maintain its elements. The next section describes how SWBST perform various operations
+A known data structure that achieves this property is the ``weight-balanced
+B-tree'' \cite{paper}. The CO B-tree leverages this data structure to maintain
+its elements. The next section describes how SWBST perform various operations
 while maintain their strongly weight-balanced property.
@@
 
-It should be noted, that keeping the level of indirection sorted is only needed in case we want
-to perform \Scan operations. If we do not need \Scan, then we can keep the indirection array in any arbitrary order
-reducing the cost of \Insert and \Delete by the $(log^2 N) /B$ factor making them incur $O(log_{B} N)$ memory transfers
-overall.
+It should be noted, that keeping the level of indirection sorted is only
+needed in case we want to perform \Scan operations. If we do not need \Scan,
+then we can keep the indirection array in any arbitrary order reducing the
+cost of \Insert and \Delete by the $(log^2 N) /B$ factor making them incur
+$O(log_{B} N)$ memory transfers overall.
 

### 2011-12-09 paper.tex
Adding the abstract

@@
 
-This paper presents a survey of cache-oblivious B-tree data structures. Memory
-hierarchies . Since memory transfers often dominate the runtimes of an
-algorithm it makes sense to develop data structure that are optimized not in
-terms of the number of instructions it executes; but rather, by the number of
-memory accesses and transfers it has to perform. Among these so called
-``external-memory'' data structures, the B-tree is widely used as it is able
-to perform \Search in an optimal number of memory transfers. However, a short
-coming of B-trees is that they require knowledge of the size of the memory
-blocks, thus it is a cache aware data structure. This gave rise to the notion
-of cache-oblivious algorithms, e.g. data structures that perform well
-regardless of the size of the memory blocks or the number hierarchies. Bender
-et al. introduced the cache-oblivious B-tree which ...
+This paper present a survey on cache-oblivious B-trees (CO B-trees). The CO B-tree
+was first introduce by Bender et al in 2000. It is an external memory data structure meaning
+that it optimizes the number of memory transfers if performs rather than the number of
+CPU instructions that it executions. Unlike other external memory data structure, like
+the widely use B-tree, the CO B-tree does not require any information about the
+number of level in the machine's memory hierarchy, the block size and capacity at each level, or the 
+relative speeds for memory transfer. Hence, the CO B-tree is ``cache-oblivious''.
+Despite this, the CO B-tree is able to achieve the same
+optimal search bound as the B-tree and near optimal insertion and deletion bounds as well.
+In this paper we first describe the original CO B-tree and explain how it is able to achieve 
+its memory performance. We then present and analyze three followup CO B-tree data structures
+ that improved upon shortcoming of the original: the cache-oblivious string B-tree, which introduces support 
+for variable length keys, the cache-oblivious streaming B-tree, which incorporates improvements to the
+insertion bound at the expense of the search bound, and lastly, the cache-oblivious concurrent B-tree, which 
+introduces support for concurrency.
 

### 2011-12-09 paper.tex
Updating the abstract and introduction.

@@
 
-This paper present a survey on cache-oblivious B-trees (CO B-trees). The CO B-tree
-was first introduce by Bender et al in 2000. It is an external memory data structure meaning
-that it optimizes the number of memory transfers if performs rather than the number of
-CPU instructions that it executions. Unlike other external memory data structure, like
-the widely use B-tree, the CO B-tree does not require any information about the
-number of level in the machine's memory hierarchy, the block size and capacity at each level, or the 
-relative speeds for memory transfer. Hence, the CO B-tree is ``cache-oblivious''.
-Despite this, the CO B-tree is able to achieve the same
-optimal search bound as the B-tree and near optimal insertion and deletion bounds as well.
-In this paper we first describe the original CO B-tree and explain how it is able to achieve 
-its memory performance. We then present and analyze three followup CO B-tree data structures
- that improved upon shortcoming of the original: the cache-oblivious string B-tree, which introduces support 
-for variable length keys, the cache-oblivious streaming B-tree, which incorporates improvements to the
-insertion bound at the expense of the search bound, and lastly, the cache-oblivious concurrent B-tree, which 
-introduces support for concurrency.
+This paper present a survey on cache-oblivious B-trees (CO B-trees). The CO
+B-tree was first introduce by Bender et al in 2000. It is an external memory
+data structure - meaning that it optimizes the number of memory transfers it
+performs rather than the number of CPU instructions that it executions. Unlike
+other external memory data structures, like the widely use B-tree, the CO
+B-tree does not require any information about the number of level in the
+machine's memory hierarchy, the block size and capacity at each level, or the
+relative speeds at each level. Hence, the CO B-tree is ``cache-oblivious''.
+Despite this, the CO B-tree is able to achieve the same optimal search bound
+as the B-tree and near optimal insertion and deletion bounds as well. In this
+paper we first describe the original CO B-tree and explain how it is able to
+achieve its memory performance. We then present and analyze three followup CO
+B-tree data structures that improved upon shortcoming of the original: the
+cache-oblivious string B-tree, which introduces support for variable length
+keys, the cache-oblivious streaming B-tree, which incorporates improvements to
+the insertion bound at the expense of the search bound, and lastly, the
+cache-oblivious concurrent B-tree, which introduces support for concurrency.
 
@@
 but rather a hierarchy from super fast on chip caches, to slower main memory,
-and even all the way down to the disk. When a program requests data, the data
-is found somewhere in the memory hierarchy and typically carried up the
-hierarchy to the fastest memory possible layer when the CPU is able to access
-it. Unfortunately, this fast memory often has small capacity and once it is
-filled and new data is requests some old entries must be evicted. Bring data
-from lower to higher levels in the memory hierarchy often takes a long time.
-In fact, access times between the different layers of the hierarchy can differ
-by many orders of magnitude and often dominate the overall runtime performance
-of an algorithm. External memory algorithms are wary of this fact and aim to
-solve problem by optimizing the number of memory transfers it performs rather
-than the number of instructions it executes. They often try to exploit the
-fact the memory hierarchies transfer data in blocks rather than individual
-atomic units. Thus, external memory algorithms are designed to not only limit
-the number of memory transfers but also preserve the locality of data so that
+to even slower disks. When a program requests data, the data is found in the
+memory hierarchy and typically carried up the hierarchy to the fastest memory
+layer possible, usually a cache. The CPU is then able to access it to perform
+any necessary operations. Unfortunately, this cache often has a small capacity
+and once it is filled any new data requests cause some existing cached data to
+be evicted. Transfering data from lower to higher levels in the memory
+hierarchy often takes a long time. In fact, access times between the different
+layers of the hierarchy can differ by many orders of magnitude and thus the
+time to transfer data often dominates the overall runtime performance of an
+algorithm. External memory algorithms are wary of this fact and aim to
+optimizing the number of memory transfers it performs rather than the number
+of instructions it executes. They often try to exploit the fact the memory
+hierarchies transfer data in blocks rather than individual atomic units. Thus,
+external memory algorithms are designed to not only limit the number of memory
+transfers it performs but also preserve the locality of related data so that
 when a memory access is performed, its cost is amortized.
@@
 
-Since External memory algorithms are primarily concerned with efficient access
-to data, it is not surprise that most problems can be reduced to finding a
-suitable data structure that efficiently implement certain operations. A well
-known external memory data structure is the B-tree. At a high level, B-tree
-work by making their nodes the same size as a memory block in the hierarchy.
-Furthermore, the fanout in the tree (the number of children) is also
-proportional to the block size as well. The B-tree incurs $O(log(N))$ memory
-transfers for \Search{}, \Insert{}, and \Delete{} and is widely used in
-databases and filesystems.
+Since external memory algorithms are primarily concerned with accessing data,
+it is no surprise that many external memory problems can be reduced to finding
+a suitable data structure that efficiently implement certain operations. A
+well known external memory data structure is the B-tree. At a high level, the
+B-tree works by making each one of its nodes the same size as a memory block
+in a particular level of the hierarchy. Furthermore, the fanout in the tree
+(the number of children of a node) is proportional to the block size as well.
+The B-tree incurs $O(log(N))$ memory transfers for \Search{}, \Insert{}, and
+\Delete{} and is widely used in databases and filesystems.
 
@@
 The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in
-a two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}. However,
-in reality computing architectures typically have memory hierarchies that are
-much larger than two. The argument for the DAM model is that in practice only
-one level in the memory hierarchy tends to dominate the performance and thus a
-two-level memory model is sufficient for all intents and purposes. However,
-this does mean that, theoretically, B-trees can begin to incur an increasing
-number of memory transfers if the dominating level in the hierarchy changes.
+a two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}.
+However, in reality, computing architectures typically have memory hierarchies
+that are much larger than two. The argument for the DAM model is that in
+practice only one level in the memory hierarchy tends to dominate the
+performance and thus a two-level memory model is sufficient for all intents
+and purposes. However, this does mean that, theoretically, B-trees can begin
+to incur an increasing number of memory transfers if the dominating level in
+the hierarchy changes.
 
-Additionally, the B-tree expects to know the the block size in the memory
-hierarchy in advanced. This means that the parameterized in code need to be
-changed for every level in the memory hierarchy which can be very difficult
-and error prone. Additionally, tuning this parameter may even be impossible in
-a heterogeneous computing environment.
+Additionally, the B-tree expects to know the block size of its memory layer in
+advanced. B-Trees are implemented with parameterized code that needs to be
+changed for every level in the memory hierarchy. Needless to say, this tuning
+can be very difficult and error prone. Additionally, tuning this parameter may
+even be impossible in certain heterogeneous computing environments.
 
 The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
-of computation. In the CO model the algorithm / data structure has no
-information about the memory hierarchy. It does not know anything about the
-speed of any level or the block size. At a conceptual level, the CO model
-allow algorithm designers to reason about the simple two-level model like DAM
-but prove results for any arbitrary memory hierarchy. In short, algorithms
-that perform well in a CO model should, in theory, perform well regardless of
-the block size in a memory hierarchy and thus should gracefully scale to
-multi-level hierarchies as well.
+of computation. In the CO model an algorithm has no information about the
+memory hierarchy. It does not know anything about the speed or the block size
+of any level. Conceptually, the CO model allow algorithm designers to reason
+about the simple two-level model like DAM but prove results for any arbitrary
+multi-level memory hierarchy. This is because algorithms that perform well in
+a CO model should, in theory, perform well regardless of changes in block
+size, cache capacity, and cache speed. Consequently, CO algorithm should
+gracefully scale to multi-level hierarchies as well.
 
@@
 
-In \cite{BenderDemainColton} Bender et al. introduce the cache-oblivious B-tree (CO B-tree). The
-CO B-tree is similar to normal the B-tree except that it does not have the
-block size of levels in the memory hierarchy as explicitly parameters.
+In \cite{BenderDemainColton} Bender et al. introduce the cache-oblivious
+B-tree (CO B-tree). The CO B-tree is similar to the normal B-tree except that
+it does not require explicit parameters about the memory hierarch.
 
@@
 memory transfers, the optimal scan bound of $\Theta(N/B)$ memory transfers and
-efficient insertion and deletion operations. The CO B-tree was a seminal data
-structure that gave rise to many others in the future.
+near optimal insertion and deletion bound of $\Theta(log_B N +
+\frac{log^2{N}}{B})$. The CO B-tree was a seminal data structure that gave
+rise to many others.
 
-This paper provides a survey of these different data structures. We start by
-analyzing the original CO B-tree and explain how it works. This includes a
-detailed description of the packed memory array (PMA) and strongly
-weight-balance binary search trees (SWSBT) that are integral to the CO
-B-tree's performance. Following this discussion we provide an analysis of the
-CO String B-tree,a CO B-tree with variable length keys, the CO Streaming
-B-tree, a CO B-tree that improves the cost of insertion at the expense of
-search, and the CO Concurrent B-tree (which allows multiple processes to
-access the data at the same time).
+This paper provides a survey of the different types of CO B-trees and
+discusses the various trade offs that they make. We start by analyzing the
+original CO B-tree and explain how it works. This includes a detailed
+description of the packed memory array (PMA) and strongly weight-balance
+binary search trees (SWSBT) that are integral to the CO B-tree's memory
+performance. Following this discussion we provide an analysis of the CO String
+B-tree,a CO B-tree with variable length keys, the CO Streaming B-tree, a CO
+B-tree that improves the cost of insertion at the expense of search, and the
+CO Concurrent B-tree, which allows multiple processes to access the data at
+the same time.
 

### 2011-12-09 paper.tex
Fixing the streaming section exception for the insert section

@@
 
-While B-trees achieve good performance for searches, they do not perform as
-well for updates and insertions. The buffered-repository tree (BRT) is a
-common alternative to the B-tree achieves an improved amortized $O(log(N)/B)$
-memory transfers for inserts compared to the $O(log_{B+1}N)$ memory access of
-B-trees.
+While B-trees achieve an optimal bound for the number of memory transfer for
+search, they do not perform as well for updates and insertions. The
+buffered-repository tree (BRT) is a common alternative to the B-tree that
+achieves an improved amortized $O(log(N)/B)$ memory transfers for inserts
+compared to the $O(log_{B+1}N)$ memory access of B-trees. Of course, the
+B-tree and BRT are analyzed with the DMA model and are not cache-oblivious.
 
-The search vs insertion trade off also exists in the cache-oblivious model.
-The original cache-oblivious B-trees described in Section~\ref{sec:original}
-achieves the search end of this trade off. However, Bender et al. propose two
+This search vs insertion trade off also exists in the cache-oblivious model.
+The original BO B-tree described in Section~\ref{sec:original} is designed to
+optimize for the search in this trade off. However, Bender et al. propose two
 data structures that and achieve the opposite end: the shuttle tree and the
@@
 It uses a strongly weight-balanced search tree described previously (see
-Section~\ref{sec:structure}) that is laid out in a vEB layout into a PMA. A
-notable difference is that the CO B-tree splits the tree at height $h/2$ when
-creating the vEB layout while the shuttle tree splits at height $F_k$ where
-$F_k$ is the $kth$ Fibonacci number that is stickily less than the height of
-the tree. Thus if a tree's height happens to be a Fibonacci number, $F_k$,
-then the shuttle tree will choose to split the tree at height $F_{k-1}$. The
-top and bottom ``halves'' produced by this split are organized just like
-before and shown in Figure \ref{fig:veb}.
+Section~\ref{sec:structure}) that is laid out in a vEB (see
+Section~\ref{sec:veb}) layout into a PMA (see Section~\ref{sec:layout}). A
+notable difference between the shuttle tree and the CO B-tree is that the CO
+B-tree splits the tree at height $h/2$ when creating the vEB layout while the
+shuttle tree splits at height $F_k$, where $F_k$ is the $kth$ Fibonacci number
+that is stickily less than the height of the tree. Thus if a tree's height
+happens to be a Fibonacci number, $F_k$, then the shuttle tree will choose to
+split the tree at height $F_{k-1}$. The top and bottom ``halves'' produced by
+this split are organized just like before and shown in Figure \ref{fig:veb}.
 
@@
 
-The shuttle tree improves the efficiency of insertion by adding buffers to
-each node in the tree. In fact, shuttle tree nodes have a linked list of
-buffers rather than a single buffer like BRT and each of these buffers is
-another shuttle tree. The length of a node's linked list is related to the
-node's height in the tree. A node has a buffer for every time it the leaf of a
-sub trees that is split when creating the vEB layout. This idea is illustrated
-in Figure \ref{fig:buffers}.
+At a high level, the shuttle tree improves the efficiency of insertion by
+adding buffers to each node in the tree. In fact, shuttle tree nodes have a
+linked list of buffers rather than just a single buffer like the BRT. Each of
+these buffers is another shuttle tree. The a shuttle tree node enforces two
+constraints: the length of its linked list as well as the overall height (and
+thus capacity) of each of the buffered shuttle trees contained within its
+linked list.
+
+The length of a node's linked list is related to the node's height in the
+tree. A node has a buffer for every time it the leaf of a sub trees that is
+split when creating the vEB layout. This idea is illustrated in Figure
+\ref{fig:buffers}.
 
 The shuttle tree's node's buffers have a maximum capacity that is enforced by
-bounding it height. The size of the buffers increases going across the linked
+bounding its height. The size of the buffers increases going across the linked
 list. Thus, a node's first buffer is smaller than its last buffer. The maximum
@@
 then $\xi(h) = h$ and otherwise $\xi(h) = \xi(h-f)$ where f is the largest
-Fibonacci number less than $h$. The size of the buffer is related to the size
-of the recursive subtree for which the node is a leaf. Since the recursive
-subtrees are split based on the largest Fibonacci number less than $h$ it
-follows that the buffer sizes are also based on Fibonacci numbers.
+Fibonacci number less than $h$. The maximum height of an individual buffer is
+related to the size of the recursive subtree for which the node is a leaf.
+Since the recursive subtrees are split based on the largest Fibonacci number
+less than $h$ (as described previously) it follows that the buffer sizes are
+also based on Fibonacci numbers.
 
@@
 The operations that are described by Bender et al. for the shuttle tree are
-\Search, \Insert, and \Scan. An explicit discussion on \Delete is missing;
-however, it is assumed that it is symmetric to \Insert.
+\Search{}, \Insert{}, and \Scan{}. An explicit discussion on \Delete{} is
+missing; however, it is assumed that it is symmetric to \Insert{}.
 
-\Search and \Scan work virtually the same way as the CO B-tree. The only
+\Search{} and \Scan{} work virtually the same way as the CO B-tree. The only
 difference is that you have to search a child node's buffer before following
@@
 
-\Insert is also similar to the CO B-tree except you buffer elements as you add
-new elments. When you insert an element you move down the tree from the root
-to leaves and stop when you find a node that has a buffer (it is possible that
-a node does not have any buffers because of the height of the tree). You then
-attempt to insert the element into the smallest buffer in the node's linked
-list. If this insertion causes the buffer to overflow (e.g. the height of the
-buffer exceeds its maximum), you move all of the elements from this buffer to
-the next largest buffer. When the largest buffer overflows you take all of its
-elements and ``shuttle'' them down the tree. Eventually you will hit a leaf
-node and inserting here may cause the balance constraints on the SWBST to
-become invalidated. The shuttle tree rebalances itself just like any other
-SWBST: the leaf node is split into two and this cascades up the tree until
-balance is restored.
+\Insert{} is also similar to the CO B-tree except you buffer elements as you
+add them. When you insert an element you move down the tree from the root to
+leaves and stop when you find a node that has a buffer (note that because of
+the nature of Fibonacci numbers that it is possible that a node does not have
+any buffers). You then attempt to insert the element into the smallest buffer
+in the node's linked list. If this insertion causes the buffer to overflow
+(e.g. the height of the buffer exceeds its maximum), you move all of the
+elements from this buffer to the next largest buffer. When the largest buffer
+overflows you take all of its elements and ``shuttle'' them down the tree.
+Eventually you will hit a leaf node and inserting here may cause the balance
+constraints on the SWBST to become invalidated. The shuttle tree rebalances
+itself just like any other SWBST: the leaf node is split into two and this
+split cascades up the tree until balance is restored.
 
-When a node $u$ is split into two nodes $u1$ and $u2$ we immediately allocate
-enough space for all of the $u2$'s buffers. We then scan over $u$'s buffered
-sub trees and find the ones rooted in $u1$ and $u2$ and move them
-respectively. This can be achieved by a single scan. Once we have found the
-set of subtrees that correspond to $u1$ and $u2$ we replace the original sub
-trees for $u$ in the vEB layout.
+The shuttle tree also needs to take care of the buffers belonging to a split
+node. When a node $u$ is split into two nodes $u_1$ and $u_2$ the shuttle tree
+immediately allocates enough space for all of the $u_2$'s buffers ($u_1$
+simply reuses the space allocated for $u$'s buffers). We then scan over $u$'s
+buffers and find the ones rooted in $u_1$ and $u_2$ and move them
+respectively. This can be achieved by a single scan.
 
-The core problem of maintain room for inserts in the shuttle tree does not
-change from the CO B-tree and is handled by the PMA just like in described in
-Section~\ref{sec:layout}.
+The core problem of maintaining room for inserts, bounding the costs of
+merging and splitting, and bounding the costs of updating parent and child
+pointers in the shuttle tree is achieved using the same PMA used with the CO
+B-tree. This is decried in detail in Section~\ref{sec:layout}.
 
@@
 
-The complexity of \Search is $O(log_B N)$ where $N = \Theta(c^{F_k})$ and $c$
+\Search uses $O(log_B N)$ memory transfers where $N = \Theta(c^{F_k})$ and $c$
 is your fanout. Note that $c$ is not treated as a constant since this is a
 cache-oblivious data structure. This bound is achieved by searching over the
-top and bottom recursive subtrees at each level. A key trick is that when you
-visit a node you only search its largest buffer (that last buffer in the
-linked list). The smaller buffers will be visited eventually as you recurse
-down the tree. Eventually you will get down to a single subtree that can fit
-into $O(1)$ blocks and thus can be search with $O(1)$ memory accesses. Since
-the fanout is proportional to our block size it will take us $O(log_B N)$
-block accesses until we get down to a subtree and thus, by induction, the
-worst case runtime is $O(log_B N)$.
+top and bottom recursive subtrees at each level when traversing the tree. A
+key trick is that when you visit a node you only search its largest buffer
+(that last buffer in the linked list for that level). The smaller buffers will
+be visited eventually as you recurse down the tree. Eventually you will get
+down to a single subtree that can fit into $O(1)$ blocks and thus can be
+searched with $O(1)$ memory accesses. Since the fanout is proportional to our
+block size it will take us $O(log_B N)$ block accesses until we get down to a
+subtree and thus, by induction, the worst case runtime is $O(log_B N)$.
+
+*****
 
@@
 
-The complexity to \Scan $L$ elements is amortized to $O(L/B)$ memory
-transfers. First you perform a \Search to get to the starting element of your
-range. Once you have a block in memory you can start scanning in elements and
-since you only have to read a new block every $B$ accesses each element access
-is amortized to $1/B$ and thus overall the number of memory access is
-$O(L/B)$. This, of course, assumes that you just performed a search for the
-starting point of your scan.
+******
 
-The last thing that should be considered is the space requirement of the
-shuttle tree. However, the overall space requirements of the shuttle tree is
-actually just $O(n)$. The high level intuition for this proof is that the
-majority of the buffers are really small and thus do not impact the overall
-size of the tree.
+\Scan{} uses an amortized $O(L/B)$ memory transfers to scan in $L$ consecutive
+elements. This cost obviously does not include the single \Search{} operation
+that is required to find the first starting element of the range. Once that
+first element's block is transferred, the shuttle tree proceeds by traversing
+the tree and scanning in elements. Since the vEB layout enforces the spatial
+locality of nearby nodes in the tree, the shuttle tree only has to read in a
+new block once every $B$ memory accesses. Thus, each individual memory access
+is is amortized to $1/B$ and the overall number of memory transfers is
+$O(L/B)$.
+
+The last thing that is worth mentioning is that the overall space requirements
+of the shuttle tree is $O(n)$. The high level intuition for this proof is that
+the majority of the buffers are really small and thus do not impact the
+overall size of the tree.
 
@@
 
-The cache-oblivious lookahead array (COLA) is a cache-oblivious dictionary
-that is actually quite different from the other cache-oblivious data
-structures seen so far. Instead of an explicit tree structure with child and
-parent pointers, the COLA is organized as a list of arrays or levels. Each
-level has arrays of size $2^k$. Each level is either completely full or
-completely empty. As such, the COLA is very similar to the binomial list
-structure described in \cite{BentleySaxe}.
+The cache-oblivious lookahead array (COLA) is actually quite different from
+the other cache-oblivious data structures seen so far. Instead of an explicit
+tree structure with child and parent pointers, the COLA is organized as a list
+of arrays or levels. Each array has size $2^k$. Furthermore, each array is
+either completely full or completely empty. As such, the COLA is very similar
+to the binomial list structure described in \cite{BentleySaxe}.
 
@@
 
-The operations supported by COLA are \Search, \Insert, and \Scan.
+The operations supported by COLA are \Search{}, \Insert{}, and \Scan{}. Once
+again, \Delete{} can be inferred symmetrically from \Insert{}.
 
-\Insert works by initially inserting the element into the first-level array.
+\Insert{} works by initially inserting an element into the first-level array.
 This array has size $2^0 = 1$, thus the element is in an array on it own.
-Whenever you insert an element you check to see if there is another list of
-the same size and merge them together in sorted order. This merging will
-produce a new array of twice the size which goes into the next level. We
-continue to merge until there are no two arrays of the same length.
+Whenever you insert an element you check to see if there is another array of
+the same size (this can be though of as checking to see if there is another
+array in the same level). If there is another array with the same size, the
+COLA merges the two together in sorted order. This merging will produce a new
+array that is twice the size and thus goes into the next level. The COLA
+continues to merge until there are no two arrays of the same size.
 
-\Search can naively be performed by simply scanning each of the levels in our
-data structure. Since everything is sorted we can use a binary search to
-quickly find the element we are looking for. Obviously this is not very
-efficient, thus COLA, as the name suggest, uses look ahead pointers to reduce
-the number of memory accesses required. Every 8th element from the $k+1$ level
-exists in the $k$ level array in its proper sorted location. This element also
-contains a pointer to its real location in the $k+1$ level, thus it is
-referred to as a ``real lookahead pointer''. Additionally, each every 4th cell
-in $k$ level array contains two pointers pointing to the next and previous
-``real lookahead pointers'' on this level. Therefore, to search a level we
-perform a binary search and automatically cascade up to the next level.
+\Search{} can naively be performed by simply scanning each of the levels in
+the COLA. Since everything is sorted the COLA could can use a binary search to
+quickly find the element it is looking for. However, this is not very
+efficient. As an improvement COLA, as the name suggest, uses look ahead
+pointers to reduce the number of memory accesses required. Every 8th element
+from the $(k+1)$st level exists in the $k$ level array in its proper sorted
+location. This element also contains a pointer to its real location in the
+$(k+1)$st level. Thus this pointer is referred to as ``real lookahead
+pointer''. Additionally, each every 4th cell in $k$th level array contains two
+pointers pointing to the next and previous ``real lookahead pointers'' on this
+level. The next and previous pointers are referred to as ``duplicate lookahead
+pointers''. To search a level the COLA performs a binary search within an
+appropriate range, determine the range in the next level using the ``duplicate
+lookahead pointers'' and then recursively cascade up the tree.
 
@@
 
-The complexity of \Insert is amortized to $O(log N / B)$ block transfers.
-First, it is important to know that an element will never be merged more than
-$O(log(N))$ times since at that point, it will be in the last array. The first
-$log(B)$ merges that a element participates is constant time since we can
-assume that the relevant blocks are already in memory. Thus, we only need to
-worry about memory accesses when we merge levels that have size greater than
-$B$. If the size of the array is $k$ then the cost of merging is $O(k/B)$ and
-thus the amortized cost of a merging a single element is simply $1/B$. Since
-an element will only be merged $log(N)$ time, the overall cost is amortized to
-$O(logN/B)$.
+\Insert{} achieves an amortized $O(log N / B)$ memory transfers. The proof is
+based on the far that an element will never be merged more than $O(log(N))$
+times. This is because at that point it will be in the last array.
+Additionally, the first $log(B)$ merges that a element participates in uses a
+constant number of memory transfer since we can assume that the relevant
+blocks are already cached. Thus, we only need to worry about memory transfers
+when we merge levels that have size greater than $B$. If the size of the array
+is $k$ then the cost of merging is $O(k/B)$ and thus the amortized cost of a
+merging a single element is simply $1/B$. Since we just proved that an element
+will only be merged $O(log(N))$ time, it follows that the overall amortized
+memory transfers to insert an element is $O((log N)/B)$.
 
-The complexity of \Search is $O(log N)$ memory transfers. The proof is based
-on the fact that we only ever need to search at most eight items on each
-level. Since there are $O(log(N))$ levels and since eight is a constant it
-follows that overall, search accesses $O(log N)$ blocks. A proof by induction
-shows that only eight items need to be searched on each level. The base case
-is trivial. The first three levels of the COLA each have at most eight
-elements that either contain the actual elements or a look ahead pointer to
-the next level. Obviously we can search these levels with no more than eight
-member transfers. As we follow the lookahead pointers to the high levels we
-will eventual get to a level where $k \geq 3$. Note that we do not enter this
-level at the start; rather, we arrived here by following the real lookahead
-pointers from the levels below and are at some index $i$ that contains an
-element $e_i$ such that $e_i < e$ where $e$ is the element we are looking for.
-Our goal is to find a new $i$ such that $e_i \leq e < e_{i+1}$. Once we find
-this $i$ then we can use the duplicate look ahead pointer to find the ``next''
-and previous look ahead pointers in the next level and restrict our search to
-that range. Since we have a real look ahead pointer for every eighth element
-in the next level, it follow that when we get to the next level, we only need
-to search eight elements. Furthermore, since we arrived at our current level
-from a lower level, we only need to search $8$ elements to find this $i$.
-Thus, overall, we only need to perform $O(8log(N)) = O(log(N))$ memory
+\Search{} achieves $O(log N)$ memory transfers. The proof uses induction to
+show that the COLA never needs to search more than eight elements on each
+level. Since there are $O(log(N))$ levels and since eight is a constant, it
+follows that overall, search uses $O(log N)$ memory transfers. The fact that
+only eight elements need to be search per level is shown through induction.
+The base case is trivial. The first three levels of the COLA each have at most
+eight elements (note that these elements are either actual real elements or
+lookahead pointers). The COLA can obviously search these levels with no more
+than eight member transfers. As the COLA follows lookahead pointers to the
+higher levels, it will eventual get to a level where $k \geq 3$. Note that the
+COLA does not ``enter'' this level at the start of the array; rather, its
+arrived here by following the real lookahead pointers from the levels below
+and thus is at some index $i$ that contains an element $e_i$ such that $e_i <
+e$ and $e$ is the element being searched for. The goal at this point is to
+find a new $i$ such that $e_i \leq e < e_{i+1}$. Once this $i$ is found the
+duplicate lookahead pointers can be used to locate the ``next'' and
+``previous'' real lookahead pointers in the next level. The two real lookahead
+pointers are used to and restrict the search in the next level. Recall that at
+each level in the COLA every eighth element in a level is present in the
+previous level. Thus, it follows that in next level only 8 elements need to be
+searched. Furthermore, since the current level was reach from a lower level,
+it follows that only 8 elements need to be search on this level to find the
+appropriate $i$. Thus, overall, the COLA uses $O(8log(N)) = O(log(N))$ memory
 transfers.
@@
 
-In experimental results, COLA achieved significant improvements for random
-inserts, performing 790 times faster than a normal B-tree. Sorted inserts are
-3.1 times slower and searches are 3.5 times slower. These experimental results
-indicate that the the COLA is a suitable data structure for streaming data
-structure where random insertions are common.
+In experimental results, the COLA achieved significant improvements over
+normal (cache-aware) B-trees, performing 790 times faster. However this
+improvement comes at the cost of sorted inserts and normal searches. Sorted
+inserts are 3.1 times slower and searches are 3.5 times slower. Nevertheless,
+these experimental results indicate that the the COLA is a suitable data
+structure when random insertions are common.
 

### 2011-12-09 paper.tex
adding my more completed rough draft

--- 2011-12-09 paper.tex
@@
 \newcommand{\rhyperfloor}{\rfloor\rfloor}
+\newcommand{\llangle}{\langle\langle}
+\newcommand{\rrangle}{\rangle\rangle}
 
@@
 \newcommand{\Insert}{\textsc{Insert(k, v)}}
+\newcommand{\Insertkonly}{\textsc{Insert(k)}}
 \newcommand{\Scan}{\textsc{Scan(a, b)}}
@@
 \newcommand{\Member}{\textsc{Member(k)}}
-%\newcommand{\Range}{\textsc{Range-Query(k$_{i}$,k$_{j}$)}}
+\newcommand{\Range}{\textsc{Range-Query(k$_{i}$,k$_{j}$)}}
 
@@
 
-This paper present a survey on cache-oblivious B-trees (CO B-trees). The CO B-tree
-was first introduce by Bender et al in 2000. It is an external memory data structure meaning
-that it optimizes the number of memory transfers if performs rather than the number of
-CPU instructions that it executions. Unlike other external memory data structure, like
-the widely use B-tree, the CO B-tree does not require any information about the
-number of level in the machine's memory hierarchy, the block size and capacity at each level, or the 
-relative speeds for memory transfer. Hence, the CO B-tree is ``cache-oblivious''.
-Despite this, the CO B-tree is able to achieve the same
-optimal search bound as the B-tree and near optimal insertion and deletion bounds as well.
-In this paper we first describe the original CO B-tree and explain how it is able to achieve 
-its memory performance. We then present and analyze three followup CO B-tree data structures
- that improved upon shortcoming of the original: the cache-oblivious string B-tree, which introduces support 
-for variable length keys, the cache-oblivious streaming B-tree, which incorporates improvements to the
-insertion bound at the expense of the search bound, and lastly, the cache-oblivious concurrent B-tree, which 
-introduces support for concurrency.
+This paper presents a survey of cache-oblivious B-tree data structures. Memory hierarchies . 
+Since memory transfers often dominate the runtimes of an algorithm it makes sense to
+develop data structure that are optimized not in terms of the number of instructions it executes; but rather, 
+by the number of memory accesses and transfers it has to perform. Among these so called ``external-memory'' data structures,
+the B-tree is widely used as it is able to perform \Search in an optimal number of memory transfers. 
+However, a short coming of B-trees is that they require knowledge of the size of the memory blocks, thus 
+it is a cache aware data structure. This gave rise to the notion of cache-oblivious algorithms, e.g. data structures
+that perform well regardless of the size of the memory blocks or the number hierarchies. Bender et al. 
+introduced the cache-oblivious B-tree which ...
+
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
 
@@
 
-Most algorithms assume an idealized model of computation that consists of a
-single CPU, which performs instructions, and memory, which is a flat array
-that can be read and written. However, in reality memory is not a flat array
-but rather a hierarchy from super fast on chip caches, to slower main memory,
-and even all the way down to the disk. When a program requests data, the data
-is found somewhere in the memory hierarchy and typically carried up the
-hierarchy to the fastest memory possible layer when the CPU is able to access
-it. Unfortunately, this fast memory often has small capacity and once it is
-filled and new data is requests some old entries must be evicted. Bring data
-from lower to higher levels in the memory hierarchy often takes a long time.
-In fact, access times between the different layers of the hierarchy can differ
-by many orders of magnitude and often dominate the overall runtime performance
-of an algorithm. External memory algorithms are wary of this fact and aim to
-solve problem by optimizing the number of memory transfers it performs rather
-than the number of instructions it executes. They often try to exploit the
-fact the memory hierarchies transfer data in blocks rather than individual
-atomic units. Thus, external memory algorithms are designed to not only limit
-the number of memory transfers but also preserve the locality of data so that
-when a memory access is performed, its cost is amortized.
-
 \subsection{B-trees}
 
-Since External memory algorithms are primarily concerned with efficient access
-to data, it is not surprise that most problems can be reduced to finding a
-suitable data structure that efficiently implement certain operations. A well
-known external memory data structure is the B-tree. At a high level, B-tree
-work by making their nodes the same size as a memory block in the hierarchy.
-Furthermore, the fanout in the tree (the number of children) is also
-proportional to the block size as well. The B-tree incurs $O(log(N))$ memory
-transfers for \Search{}, \Insert{}, and \Delete{} and is widely used in
-databases and filesystems.
-
-\subsection{DAM vs CO}
-
-The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in
-a two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}. However,
-in reality computing architectures typically have memory hierarchies that are
-much larger than two. The argument for the DAM model is that in practice only
-one level in the memory hierarchy tends to dominate the performance and thus a
-two-level memory model is sufficient for all intents and purposes. However,
-this does mean that, theoretically, B-trees can begin to incur an increasing
-number of memory transfers if the dominating level in the hierarchy changes.
-
-Additionally, the B-tree expects to know the the block size in the memory
-hierarchy in advanced. This means that the parameterized in code need to be
-changed for every level in the memory hierarchy which can be very difficult
-and error prone. Additionally, tuning this parameter may even be impossible in
-a heterogeneous computing environment.
-
-The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
-of computation. In the CO model the algorithm / data structure has no
-information about the memory hierarchy. It does not know anything about the
-speed of any level or the block size. At a conceptual level, the CO model
-allow algorithm designers to reason about the simple two-level model like DAM
-but prove results for any arbitrary memory hierarchy. In short, algorithms
-that perform well in a CO model should, in theory, perform well regardless of
-the block size in a memory hierarchy and thus should gracefully scale to
-multi-level hierarchies as well.
+\subsection{CO vs CA}
 
@@
 
-In \cite{BenderDemainColton} Bender et al. introduce the cache-oblivious B-tree (CO B-tree). The
-CO B-tree is similar to normal the B-tree except that it does not have the
-block size of levels in the memory hierarchy as explicitly parameters.
-
-The CO B-tree is able to achieve the optimal search bound of $\Theta(log(N))$
-memory transfers, the optimal scan bound of $\Theta(N/B)$ memory transfers and
-efficient insertion and deletion operations. The CO B-tree was a seminal data
-structure that gave rise to many others in the future.
-
-This paper provides a survey of these different data structures. We start by
-analyzing the original CO B-tree and explain how it works. This includes a
-detailed description of the packed memory array (PMA) and strongly
-weight-balance binary search trees (SWSBT) that are integral to the CO
-B-tree's performance. Following this discussion we provide an analysis of the
-CO String B-tree,a CO B-tree with variable length keys, the CO Streaming
-B-tree, a CO B-tree that improves the cost of insertion at the expense of
-search, and the CO Concurrent B-tree (which allows multiple processes to
-access the data at the same time).
-
 \section{Background}
-
-High level idea:
-
-We want to have a B-tree that is cache oblivious. Okay, well, first we still
-need a way to actually store our data. We should use a tree. But what type of
-tree? Obviously it needs to be balanced but anything else? Yes, it needs to
-maintain the property that the number of descendants are $O(d^h)$. Why? It has
-to do with how we will ultimately layout the tree in memory, but we will get
-to that later.
-
-Okay, so, we find the tree that we want to use. Now what? Well, first of all,
-we need to make sure that it supports all of the operations that we want while
-maintain this key property. Lets start with searching. We will layout the tree
-in a vEB layout. This is helpful for two reasons. First, it is a nice way to
-store it in memory and second because it gives us a search bound for free. Now
-we need to make sure that it supports all of the other operations while
-maintain the key property. I can show you how it handles insertions and
-delete. Bam!
-
-Okay, so now, we have a data structure that holds the amount of memory that we
-want and it has the property that we will use later. Great! Now what? Well, we
-need to actually store this in memory some how. We said that we will use a vEB
-layout but that is only half of the story. We want a data structure that
-maintain the locality of as many elements as once while also keeping extra
-space to allow for future inserts. This is seemingly contradictory but we can
-do it!
-
-How? we use a PMA. The main problem is what inserts and delete may cause
-splits and merges which may cause huge parts and nodes of the tree to move
-around. So we need to first show how it is able to handle splits and merges
-efficiently in a cache oblivious manner. bam! done! Great!
-
-Now what? We also have the problem of updating pointers. If the points are
-close to one anther that is not a big deal. However, it is possible that the
-nodes are really far apart. Close means they are in the same block and thus
-only incur a single memory transfer. Far apart means they are more. To solve
-this problem we use buffer nodes and that works in a weird way. However, this
-only gives us a bad bound, can we do better? ****
-
-Yes! We use indirection. Indirection allows us to use another array and this
-makes out pointer updates much more efficient.
-
-We can now perform inserts and deletions faster. Additionally, our scans are
-super fast as well. Search have not changed at all because we still have our
-vEB structure and thus we are okay. But, we can also do a little better. If we
-are okay without having scans, we can improve inserts and deletion since we no
-longer have to maintain our indirection array.
-
-The important part is that our analysis above we never treat the block size as
-a constant. It is always a variable in our analysis but we are able to achieve
-the desired bounds. Thus, we have a cache-oblivious B-tree that does
-everything that we want.
 
@@
 
-Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
-\cite{cobtree}. The CO B-tree was able to match the optimum search bound of
-$\Theta(log_B N)$ memory transfers without including any specific
-parameterization describing the speed or sizes of the machine's memory
-hierarchy.
+Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in \cite{cobtree}. The
+CO B-tree was able to match the optimum search bound of $\Theta(log_B N)$ memory transfers 
+without including any specific parameterization describing the speed or sizes
+of the machine's memory hierarchy.
 
 The CO B-tree paper presents three different data structures that make 
-different trade off in the number of memory transfer they perform. However,
-all of them achieve their performance by structuring their elements as a
-strongly weight-balanced search tree (SWBST). The SWBST is then laid out into
-a packed memory array (PMA) using a van Emde Boas (vEB) layout. The SWBST and
-PMA play crucial roles in the other cache-oblivious data structures that were
-developed after the CO B-tree and thus are described at length in the next two
-sections.
+different trade off in the number of memory transfer they perform. However, all of them 
+achieve their performance by structuring their elements as a strongly weight-balanced
+search tree (SWBST). The SWBST is then laid out into a packed memory array (PMA) 
+using a van Emde Boas (vEB) layout. The SWBST and PMA play crucial roles in the
+other cache-oblivious data structures that were developed after the CO B-tree and thus are
+described at length in the next two sections.
 
@@
 
-Strongly weight-balanced search trees play a key role in CO B-tree. The weight
-of a node is $u$ is equal to the total number of $u$'s descendants plus 1. It
-can als be thought of as the sum of the weights of $u$'s children plus 1.
+Strongly weight-balanced search trees play a key role in CO B-tree. The weight of a node is $u$ is equal to
+the total number of $u$'s descendants plus 1. It can als be thought of as the sum of the weights of $u$'s children plus 1.
 Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.
 
-Most weight-balanced search trees which requires that the heights of a node's
-left and right subtrees are simply off by a constant factor. SWBST enforce the
-stronger property that a node at height $h$ in the tree (height is the node's
-distance from the leaves; leaves have a height of 1) has $\Theta c^h$ where
-$c$ is some constant.
+Most weight-balanced search trees which requires that the heights of a node's left and right subtrees are
+simply off by a constant factor. SWBST enforce the stronger property that a node at height $h$ in the tree (height is the node's
+distance from the leaves; leaves have a height of 1) has $\Theta c^h$ where $c$ is some constant.
 
-A known data structure that achieves this property is the ``weight-balanced
-B-tree'' \cite{paper}. The CO B-tree leverages this data structure to maintain
-its elements. The next section describes how SWBST perform various operations
+A known data structure that achieves this property is the ``weight-balanced B-tree'' \cite{paper}. The CO B-tree leverages this
+data structure to maintain its elements. The next section describes how SWBST perform various operations
 while maintain their strongly weight-balanced property.
@@
 
-It should be noted, that keeping the level of indirection sorted is only
-needed in case we want to perform \Scan operations. If we do not need \Scan,
-then we can keep the indirection array in any arbitrary order reducing the
-cost of \Insert and \Delete by the $(log^2 N) /B$ factor making them incur
-$O(log_{B} N)$ memory transfers overall.
+It should be noted, that keeping the level of indirection sorted is only needed in case we want
+to perform \Scan operations. If we do not need \Scan, then we can keep the indirection array in any arbitrary order
+reducing the cost of \Insert and \Delete by the $(log^2 N) /B$ factor making them incur $O(log_{B} N)$ memory transfers
+overall.
 
@@
 However, traditional B-trees and string B-trees have suboptimal performance when variable length strings are used as keys.
+In addition, string B-trees do not support compression, and commonly used front compression is handled sub-optimally.
+Regular B-trees and string B-trees are not cache oblivious, and thus incur poorer performance as described previously in *cite original b-tree paper*.
+Range queries have poor performance in B-trees.
 In \cite{BenderFaKu06}, Bender et al.
-describe 4 major issues:
-\begin{itemize}
-\item B-trees handle variable length keys poorly
-\item string B-trees do not support compression, and the front compression used in many other B-trees is handled sub-optimally
-\item range queries are handled sub-optimally by regular B-trees, since they result in random disk seeks
-\item traditional and string B-trees assume fixed block-transfer size B at all levels of the memory hierarchy, resulting in sub-optimal performance over hierarchies with varied block-transfer sizes.
-\end{itemize}
+address these issues by designing a Cache Oblivious String B-tree (COSB-tree), which they argue works efficiently in these problem areas for variable length string keys.
+They also provide a modified front-compression scheme, which gives their approach the advantages of using less space and having improved locality within the compressed data.
+
+Two experiments are performed, in which a regular B-tree implementation tuned to various block sizes is compared to a static COSB-tree implementation.
+The experiment involved measuring the amount of time required to perform: 440,000 and 450,000 inserts of random values, a range query of all values, and 1000 random searches.
+The results of the experiment show that COSB-trees outperformed B-trees most of the time for random searches, range queries and random inserts.
+However, some point between 440,000 and 450,000 inserts the COSB-tree had to rebalance the entire tree, causing it to fall behind in this measurement.
+However, Bender et al.
+argue that the experiment was biased towards the B-tree implementation, since the trees are "young", and thus still well organized.
+The COSB-tree implementation is also compared to Berkeley DB **cite berkeley db**, and achieves performance comparable to Berkeley DB with tuned parameters.
+The results seem to show that there is a window in which large and small-block size B-trees perform poorly, and that COSB-trees may be a reasonable alternative to improve performance.
 \subsection{Description}
-They address these issues by designing a Cache Oblivious String B-tree.
 
@@
 
-For COSB-trees, the centroid tree is built using $O(N/logN)$ keys, and the centroid tree is layed out in memory using a modified van Emde Boas (vEB) layout for weight-balanced binary trees.
+The centroid tree is built using $O(N/logN)$ keys, and the centroid tree is layed out in memory using a modified van Emde Boas (vEB) layout for weight-balanced binary trees.
 The \emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into equal-size subtrees for the vEB layout.
@@
 This ensures trees of equal size on all \emph{levels of detail}, as the centroid tree is traversed.
+
+The following approach is used to rebalance the centroid tree: Put all the elements in DFS or Euler tour order for the trie.
+Given this order, scan the trie to identify the centroid vertex, partition the trie into its respective up and down tries, and recurse on the up and down tries.
+The Euler tour order can be achieved via $O(\log{}N)$ scans, since the trie is only $O(\log{}N)$ deep.
+Thus updates have amortized cost of $O(1+\log^{2}N/B)$ memory transfers.
+Note that this doesn't include updating successor or predecessor pointers, which is described in more detail below.
+
+The following approach is used for maintaining the vEB layout for centroid trees in the dynamic COSB-tree: first, weight balance is relaxed from a maximum weight difference of 2 between left and right subtrees, to a larger constant $c$.
+Thus a node $v$ only becomes out of balance every $\Omega(\textsc{weight(v)})$ insertions or deletions of descendants of v, reducing the frequency of rebuilds.
+The nodes are stored in a PMA structure, preserving order of the nodes using $\Theta(N)$space, apart from inserts and deletes.
+On insert or delete, the structure is scanned in both directions starting from the update point for sections of the array to rebalance based on a "density" threshold of the section.
+Rebalance here refers to evenly distributing the nodes across the section.
+To ensure that updating pointers isn't too expensive, any node included in the section must have its subtrees included.
+Specifically, the section is built up from the the lowest subtree level that contains or contained the update point, and neighboring subtrees are continuously added until either the density threshold is met or the entire current subtree layout at this "level of detail" is included.
+For example, suppose a node is inserted at the leaf of some lower recursive subtree $L_{i}$.
+To continue building up this section, $L_{i-1}$ or $L_{i+1}$ could be added.
+If all the lower recursive subtrees $L_{1}$,$L_{2}$,...,$L_{x}$ have been added, the top level tree $U$ is added.
+This corresponds to a subtree $L'_{j}$ one level above.
+Thus repeat this process until the density threshold is met, and then rebalance the tree.
+This approach depends on the fact that all the recursive subtrees at the same level of detail asymptotically have the same size, which Bender et al.
+prove about the hyperhyperfloor-based vEB layout described previously.
+This approach results in an amortized update cost of $O(1+(\log{}N)/B)$ memory transfers per update.
+In conjunction with the cost of rebuilds described above, this is only an additive $O(\log{}N/B)$ factor, so the overall bound is amortized $O(1+\log^{2}N/B)$ memory transfers.
+
+Note that making the COSB-trees dynamic causes updates to successor and predecessor pointers in the centroid tree to be expensive.
+Thus the notion of ancestors is relaxed to the following (Lemma 10 in \cite{BenderFaKu06}): Let $\alpha$ be the parent node of $\gamma$ in the compressed trie.
+Then either $\alpha$ is a descendant of $\gamma$ or $\gamma$ is a descendant of $\alpha$.
+The proof is as follows: As you recursively traverse the trie, both $\alpha$ and $\gamma$ are in the up trie or down trie, until at some point one of them must be the centroid vertex.
+This is used to redefine what the predecessor/ancestor pointers refer to.
+The centroid may no loner point to the correct predecessor or successor with respect to the relevant subtrie, but one of the matched ancestors by construction must point to the correct predecessor/successor.
+Thus for the dynamic case, when dealing with successors and predecessors, $O(logN)$ ancestors' successor and predecessor pointers in addition to the successor and predecessor pointers of the node in question must be checked.
+However, this is an additive factor, so the same time bound for update described above is still achieved.
 
@@
 Bender et al.
-provide a modification that reduces the high decoding cost of front compression of D to $O(1+||k||B)$, with size $O(\ll{D}\gg)$ after compression.
+provide a modification for static COSB-trees that reduces the high decoding cost of front compression of D to $O(1+||k||B)$, with size $O(\llangle{D}\rrangle{})$ after compression.
 The modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be the first key added to the compression stream.
@@
 By construction, this scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.%todo: add something about why this isO(<<D>>)?
-%todo: add another paragraph or two about the second iteration
+
+However, the above modification falls short for anticipating insertions and deletions.
+The previous modification relies on each key being decoded within $c||k^{*}||$ elements of the compressed array, but this is no longer guaranteed when new keys can be inserted to the left of $k$.
+To make the locality-preserving front compression technique suitable for dynamic COSB-trees, Bender et al.
+modify the algorithm to include an extra copied prefix for each key.
+Consider the $3c||k||$ characters that would be to the left of a new key$k^{*}$ after insertion.
+Let $k'$ be the key at the end of this range.
+The idea is to prevent the insertion of $k^{*}$ from affecting any non-copied keys just to the right of $k'$, since this could potentially increase their decode cost.
+Compute the length $l$ of the largest common prefix (lcp) between $k^{*}$ and $k'$, and compare $l$ to all the other lcp's in this range.
+$l$ represents the minimum length for any lcp between $k^{*}$ and $k'$, since the keys are stored in lexicographic order.
+If there is a copied prefix of length at least $l$ or a copied key within this range, insert $k^{*}$ as usual.
+Inserting $k^{*}$ will have no negative effect in this respect, since the decode length is effectively reset by the copied key or prefix like before.
+However, if inserting $k^{*}$ causes the minimum length of the lcp's in this range to increase, we need to fix the compression.
+To do this, we update the lcp of the key directly to the right of $k^{*}$ and store an extra copy of this lcp with $k^{*}$ itself.
+In this way, we can charge the work of copying extra prefixes for a new key to the appropriate number of elements before it.
+This number is determined by the length of the extra copied prefix stored with the key.
+
 \subsection{Operators}
-The operations supported are
+%todo: figure out how to do range queries from this paper somehow...
+The operations supported are dictated by the Dictionary Matching problem.
+These operations are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{}, and \Range{}.
+The behavior of each operation for dynamic COSB-trees is described below.
+\Insertkonly{} and \Delete{} are also supported but not described here.
+
+\Search{}: works as follows for a given key $k$.
+Compute the hash value for $k$ and call it $\rho$, and traverse the centroid tree of the COSB-tree using $\rho$ to obtain the successor $\rho_{s}$ and predecessor $\rho_{p}$ of $k$.
+Index into the hashdata layer and perform a sequential scan across the PMA structure from $\rho_{p}$  to $\rho_{s}$ to find $\rho$.
+On a match with value $\rho'$, index into the keydata layer and decode $\rho'$ to obtain the resulting key $k'$.
+If $k' = k$, return true.
+If $k' \ne k$ or $k$ was not found during the sequential scan return false.
+
+\Pred{},\Succ{}: works in the same way as \Search{} until the sequential scan in the hashdata layer.
+When scanning the hashdata layer, look for $\rho_{p}$ or $\rho_{s}$, respectively, in place of $\rho$, and compare the appropriate keys in the keydata layer.
+
+\Range{}: The paper doesn't say, but presumably this is very similar to \Search{}.
+Let $k_{b}$ be the beginning of the range and let $k_{e}$ be the end of the range.
+Hash the range keys and identify the predecessor of $\rho_{b}$ and the successor of $\rho_{e}$ using the centroid tree.
+Scan the hashdata layer across from  $\rho_{b,p}$ to  $\rho_{e,s}$.
+Only start noting the beginning of range when the first key with hash value after $\rho_{b}$, and stop once we reach the first key with hash value after $\rho_{e}$.
+Decode and return the set of keys within the given range in the keydata layer.
 
 \subsection{Complexity}
+\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high probability (w.h.p.).
+Traversing the centroid tree requires  $O(1+\log_{B}N)$ memory transfers, since the tree is built using$O(N/logN)$ keys and is balanced.
+Once in the hashdata layer, we are already within $O(logN)$ keys of $k$.
+Thus the local scan requires $O(1+(||k||+\log{}N)/B)$ memory transfers.
+Obtaining $k$ from the keydata layer requires no more memory transfers than the scan in the hashdata layer.
+The \Pred{} and \Succ{} operations are almost identical in behavior to searches, and thus only have the added cost of having to operate on $k'$, where $k'$ is the predecessor or successor, respectively.
+These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers w.h.p..
+Similarly, range queries take $O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory transfers, where in this case $k'$ is the other end of the range and $Q$ is the set of keys within the given range.
+
+\Insertkonly{} and \Delete{} w.h.p.
+require $O(1+\log_{B}N+\log^{2}N||k||/B)$ memory transfers.
+As mentioned previously, due to rebuilds updates have amortized cost of $O(1+(\log^{2}N)/B)$ memory transfers.
+The rest of the cost for insertions and deletions is due to having to find the correct key to delete or location for insert, which can be seen from the cost of \Search{}.
 

### 2011-12-09 paper.tex
Fixing the insert section. I may add updates to it later.

--- 2011-12-09 paper.tex
@@
 
-\Search uses $O(log_B N)$ memory transfers where $N = \Theta(c^{F_k})$ and $c$
-is your fanout. Note that $c$ is not treated as a constant since this is a
+\Search{} uses $O(log_B N)$ memory transfers where $N = \Theta(c^{F_k})$ and
+$c$ is your fanout. Note that $c$ is not treated as a constant since this is a
 cache-oblivious data structure. This bound is achieved by searching over the
@@
 
-*****
+% TODO - redo the insert section. Namely, include a discussion on how the
+% shuttle tree maintains its structure dynamically. I mention the node
+% splitting from pg 85 in the paper but do not include that in the analysis. I
+% should basically mention that the cost of dynamically maintaining the
+% shuttle tree O(logN / B) is not greater than the PMA log^2 N / B and thus
+% the dynamic updates do not change the number of memory tranfers.
 
-\Insert uses $O(log_{B+1} N / B^{\Theta(1/(loglogB)^2)} + (log^2 N) / B)$
-memory transfers. This improvement of the the CO B-tree is achieved because
-the shuttle tree's buffers are used to maintain locality of the elements long
-enough to amortize the cost of moving them down the tree and possible into
-another block. The proof is based on the fact that when we move an item from
-one block to another, caused by splitting or buffers being overfilled, at
-least $B^{\Theta(1/(loglogB)^2)}$ will move with that item. This is because,
-with insert, when a buffer overflows we move all of the elements from that
-buffer down the tree, not just the single one that caused the overflow. Thus,
-we amortize the cost of a single insert by the number of elements in a node's
-largest buffer.
-
-******
+\Insert{} uses $O(log_{B+1} N / B^{\Theta(1/(loglogB)^2)} + (log^2 N) / B)$
+memory transfers. This improvement over the CO B-tree is achieved because the
+shuttle tree's buffers are used to maintain locality of the elements long
+enough to amortize the cost of moving them down the tree and possibly into
+another memory block. The proof is based on the fact that when an element is
+moved from one block to another it is due to either a leaf node splitting or a
+node's buffer overflowing. Recall that when inserting a new element causes a
+buffer to overflow, all of the the elements from that buffer move down the
+tree, not just the single one that caused the overflow. Thus, the cost of
+inserting an element is amortize by the number of elements in a node's largest
+buffer. The number of elements in the largest buffer is
+$B^{\Theta(1/(loglogB)^2)}$ leading to the $B^{\Theta(1/(loglogB)^2)}$ factor
+in the big-Oh notation.
 

### 2011-12-09 paper.tex
Adding the first pass at the background section. It is a piece of crap right now.

@@
 
-High level idea:
-
-We want to have a B-tree that is cache oblivious. Okay, well, first we still
-need a way to actually store our data. We should use a tree. But what type of
-tree? Obviously it needs to be balanced but anything else? Yes, it needs to
-maintain the property that the number of descendants are $O(d^h)$. Why? It has
-to do with how we will ultimately layout the tree in memory, but we will get
-to that later.
-
-Okay, so, we find the tree that we want to use. Now what? Well, first of all,
-we need to make sure that it supports all of the operations that we want while
-maintain this key property. Lets start with searching. We will layout the tree
-in a vEB layout. This is helpful for two reasons. First, it is a nice way to
-store it in memory and second because it gives us a search bound for free. Now
-we need to make sure that it supports all of the other operations while
-maintain the key property. I can show you how it handles insertions and
-delete. Bam!
-
-Okay, so now, we have a data structure that holds the amount of memory that we
-want and it has the property that we will use later. Great! Now what? Well, we
-need to actually store this in memory some how. We said that we will use a vEB
-layout but that is only half of the story. We want a data structure that
-maintain the locality of as many elements as once while also keeping extra
-space to allow for future inserts. This is seemingly contradictory but we can
-do it!
-
-How? we use a PMA. The main problem is what inserts and delete may cause
-splits and merges which may cause huge parts and nodes of the tree to move
-around. So we need to first show how it is able to handle splits and merges
-efficiently in a cache oblivious manner. bam! done! Great!
-
-Now what? We also have the problem of updating pointers. If the points are
-close to one anther that is not a big deal. However, it is possible that the
-nodes are really far apart. Close means they are in the same block and thus
-only incur a single memory transfer. Far apart means they are more. To solve
-this problem we use buffer nodes and that works in a weird way. However, this
-only gives us a bad bound, can we do better? ****
-
-Yes! We use indirection. Indirection allows us to use another array and this
-makes out pointer updates much more efficient.
-
-We can now perform inserts and deletions faster. Additionally, our scans are
-super fast as well. Search have not changed at all because we still have our
-vEB structure and thus we are okay. But, we can also do a little better. If we
-are okay without having scans, we can improve inserts and deletion since we no
-longer have to maintain our indirection array.
-
-The important part is that our analysis above we never treat the block size as
-a constant. It is always a variable in our analysis but we are able to achieve
-the desired bounds. Thus, we have a cache-oblivious B-tree that does
-everything that we want.
-
 \subsection{Original CO B-trees}
@@
 Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
-\cite{cobtree}. The CO B-tree was able to match the optimum search bound of
-$\Theta(log_B N)$ memory transfers without including any specific
-parameterization describing the speed or sizes of the machine's memory
-hierarchy.
-
-The CO B-tree paper presents three different data structures that make
-different trade off in the number of memory transfer they perform. However,
+\cite{cobtree}. The paper presents three different data structures that make
+different trade offs in the number of memory transfer they perform. However,
 all of them achieve their performance by structuring their elements as a
@@
 
-Strongly weight-balanced search trees play a key role in CO B-tree. The weight
-of a node is $u$ is equal to the total number of $u$'s descendants plus 1. It
-can als be thought of as the sum of the weights of $u$'s children plus 1.
-Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.
-
-Most weight-balanced search trees which requires that the heights of a node's
-left and right subtrees are simply off by a constant factor. SWBST enforce the
-stronger property that a node at height $h$ in the tree (height is the node's
-distance from the leaves; leaves have a height of 1) has $\Theta c^h$ where
-$c$ is some constant.
-
-A known data structure that achieves this property is the ``weight-balanced
-B-tree'' \cite{paper}. The CO B-tree leverages this data structure to maintain
-its elements. The next section describes how SWBST perform various operations
-while maintain their strongly weight-balanced property.
-
-\subsection{Operations}
-
-The CO B-tree supports \Search, \Insert, \Delete, and \Scan operations.
-
-\subsection{Complexity}
-
-It should be noted, that keeping the level of indirection sorted is only
-needed in case we want to perform \Scan operations. If we do not need \Scan,
-then we can keep the indirection array in any arbitrary order reducing the
-cost of \Insert and \Delete by the $(log^2 N) /B$ factor making them incur
-$O(log_{B} N)$ memory transfers overall.
-
-\subsection{Layout - PMA}
-\label{sec:layout}
-
-SWBST is only one half of the CO B-tree story...
-
 \begin{figure}
@@
 \end{figure}
+
+The ultimate goal of a CO B-tree is to reduce the number of memory transfers
+while supporting \Search{}, \Insert{}, and \Delete{} operations. Thus, at a
+fundamental level the CO B-tree still needs to maintain its data in some
+structure. That structure is a balanced search tree. The search tree is laid
+out according to a van Emde Boas (vEB) layout [???, ???]. The idea of a vEB
+layout is fairly straight forward. Let $h$ denote the height of a tree. the
+vEB layout splits the tree approximately at height $h/2$. This splits the
+original tree into a ``top'' subtree and many ``bottom'' subtrees. Let $A$
+denote the top subtree and $B_1... B_k$ denote the bottom subtrees. The vEB
+layout is defined recursively using the order: $A, B1, ... B_k$. This layout
+is also shown in Figure~\ref{fig:veb}.
+
+A nice consequence of using a vEB layout is that we know that it will use
+$O(log_B N)$ memory transfers for searches [???, ???]. Thus, off the bat we
+have achieve the desired number of memory transfers for search queries. It
+still remains to be shown that the CO B-tree achieves the desired number of
+memory transfers for dynamic inserts and deletes.
+
+We know that the CO B-tree is going to use a balance search tree, but what
+kind?. Strongly weight-balanced search trees play a key role in CO B-tree. The
+weight of a node is $u$ is equal to the total number of $u$'s descendants plus
+1. It can also be thought of as the sum of the weights of $u$'s children plus
+1. Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.
+
+Most weight-balanced search trees which requires that the heights of a node's
+left and right subtrees are simply off by a constant factor. SWBST enforce the
+stronger property that a node at height $h$ in the tree (height is the node's
+distance from the leaves; leaves have a height of 1) has $\Theta c^h$
+descendants where $c$ is some constant. This property becomes important later
+on when laying the tree out in memory (see Section~\ref{}).
+
+A known data structure that achieves this property is the ``weight-balanced
+B-tree'' \cite{????}. The CO B-tree leverages this data structure to maintain
+its elements. Before SWBST can be used, however, it remains to be proven that
+it maintains the the $c^h$ descendants property even after inserts and
+deletes.
+
+The definition of a SWBST given in [????] provides us with this bound for the
+weight of a node:
+
+\begin{equation}
+\frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}}
+\end{equation}
+
+Lets now consider insertions. To insert an element into a WSBST we search down
+the tree and find the natural place to insert the element - just like in any
+tree structure. After inserting the element into some node $w$, we have to
+check to see if any of $w$'s ancestors have become unbalanced by having a
+weight greater than $2^{d-1}$. If such an ancestor exists, it follows that
+every node along the path between that ancestor and $w$ will be un balanced as
+well. Thus, we start with $w$ and move up the tree.
+
+When we split a node, we naturally need to determine which children belong to
+which of the two nodes. We cannot just divide the children evenly because it
+may cause another violation. Thus, instead, we find the longest sequence of
+elements such that the weight of that subsequence is $\lceil w(u)/2 \rceil$.
+By doing this, we maintain the bound that $\frac{c^{h-1}{2} \leq w(u) \leq
+2c^{h-1}}$. We cascade these changes up the tree until all of the ancestors
+are balanced.
+
+Deletions are supported in the exact same way, except, we check is a node has
+a weight $w(u) < \frac{c^{h-1}}{2}$. In this case, we will merge two neighbors
+together and proceed up the tree. There is one edge case that needs to be
+handled. After merging, you may produce a new node that has a weight larger
+than the bounds. If this happens we immediately split the node again. Thus,
+this can be thought of as stealing nodes from a neighbor.
+
+WE have now shown that a SWBST achieves the desired property that the number
+of descendants of a node is $c^h$. We can now show how to lay this tree out in
+memory.
+
+\subsection{Layout - PMA}
+\label{sec:layout}
+
+Now that we have our elements structured in a SWBST, we are still left with
+mapping this tree to memory and proving that it achieves our desired number of
+memory transfers. The packed memory array (PMA) was developed to solve this
+problem [???]. There are three main costs that the PMA is concerned with.
+
+Fist, the PMA needs to maintain open space in its memory for newly inserted
+items. The PMA is able to accomplish this using $O(1 + (log^2 N) / B)$ memory
+transfers. Doing so may seem like attempting to achieve contradictory goals:
+maintaining the locality of elements while keeping space available for new
+inserts. The PMA uses a technique described in [???]. At a high level, the PMA
+maintains a windows into the array. When a window becomes too full or too
+empty, the elements are spread out within the context of a larger window. Two
+parameters are required, the window size and the threshold for trigger the
+spreading out. However, these parameters are not related to the block size of
+our memory level and thus do not violate the CO model of computation. Overall,
+the PMA is able to keep extra space in $O(1 + (log^2 N) / B)$ memory
+transfers.
+
+Second, the PMA needs to handle the splitting and merging that will be caused
+by the SWBST discussed in the previous section. THe PMA is able to accomplish
+this using $O(1 + (log N) /B)$ memory transfers. The proof is based on the
+fact that each node has $c^h$ descendants (recall, this is the key property
+that SWBST provided). Thus, merging or splitting a node will cause $c^h$ nodes
+to move, requiring $O(1 + c^h/B)$ memory transfers. It follows from the
+defitinot of SWBST in [???] that when a node is rebalanced, it will require
+$c^h$ insertions or deletions until it is rebalanced again. Thus, the
+amortized cost of each insertion and/or deletion is $O(1/B)$. When you split a
+node, it may go all the way up the tree and thus overall it will incur $O((log
+N) / B)$ memory transfers.
+
+Third, the PMA needs to handle updating the parent and child pointers and may
+become invalidated because of how the nodes moved around. The PMA is able to
+update the parent and child pointers. If the pointers are all close together
+in memory (e.g. within $B$ distance of each other) it is not big deal because
+only a constant number of memory transfers are required. However, if they are
+further apart the PMA introduces buffer nodes. The buffer nodes don't actually
+do anything. All they do is serve to ``stretch'' the vEB layout so that nodes
+do not move far away from one another. By using these buffers the PMA is able
+to achieve a the pointer update cost using $O(1+ logB / \sqrt{B}) log^2 N)$
+memory transfers.
+
+Of these three terms, the dominating factor is maintain room for newly
+inserted elements - $O((log^2N)/B)$. We know that SWBST achieve insertions and
+deletitions in $O(log_B N)$ if they are cache-aware. Thus, adding PMA makes
+the cache-oblivious and thereby incurring an additional $O((log^2N)/B)$ memory
+transfers.
+
+Therefore, overall, CO B-Tree achieve insertions and deletions in $O(log_B N +
+(log^2N)/B)$ memory transfers. However, they still achieve the optimal
+$O(log_B N)$ memory transfers for searches, matching the performance of
+ordinary, cache-aware, B-trees.
 

### 2011-12-09 paper.tex
Adding updates to the background section.

@@
 Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
-\cite{cobtree}. The paper presents three different data structures that make
-different trade offs in the number of memory transfer they perform. However,
-all of them achieve their performance by structuring their elements as a
+\cite{cobtree}. The CO B-tree achieves its number of memory transfers by structuring its elements as a
 strongly weight-balanced search tree (SWBST). The SWBST is then laid out into
@@
 
-A nice consequence of using a vEB layout is that we know that it will use
-$O(log_B N)$ memory transfers for searches [???, ???]. Thus, off the bat we
-have achieve the desired number of memory transfers for search queries. It
+A nice consequence of using a vEB layout is that it is known that it will use
+$O(log_B N)$ memory transfers for searches [???, ???]. Thus, the CO B-tree has already
+achieved the desired number of memory transfers for search queries. It
 still remains to be shown that the CO B-tree achieves the desired number of
@@
 
-We know that the CO B-tree is going to use a balance search tree, but what
-kind?. Strongly weight-balanced search trees play a key role in CO B-tree. The
+It is known that the CO B-tree is going to use a balance search tree, but what
+kind? This is where the strongly weight-balanced search tree becomes useful. The
 weight of a node is $u$ is equal to the total number of $u$'s descendants plus
@@
 Most weight-balanced search trees which requires that the heights of a node's
-left and right subtrees are simply off by a constant factor. SWBST enforce the
+left and right subtrees are simply off by a constant factor. The SWBST enforces the
 stronger property that a node at height $h$ in the tree (height is the node's
@@
 descendants where $c$ is some constant. This property becomes important later
-on when laying the tree out in memory (see Section~\ref{}).
+on when laying the tree out in memory (see Section~\ref{sec:layout}).
 
@@
 \begin{equation}
-\frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}}
+\frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}
 \end{equation}
@@
 elements such that the weight of that subsequence is $\lceil w(u)/2 \rceil$.
-By doing this, we maintain the bound that $\frac{c^{h-1}{2} \leq w(u) \leq
-2c^{h-1}}$. We cascade these changes up the tree until all of the ancestors
+By doing this, we maintain the bound that $\frac{c^{h-1}}{2} \leq w(u) \leq
+2c^{h-1}$. We cascade these changes up the tree until all of the ancestors
 are balanced.
 
-Deletions are supported in the exact same way, except, we check is a node has
+Deletions are supported in the exact same way, except, we check if a node has
 a weight $w(u) < \frac{c^{h-1}}{2}$. In this case, we will merge two neighbors
@@
 
-WE have now shown that a SWBST achieves the desired property that the number
+We have now shown that the SWBST achieves the desired property that the number
 of descendants of a node is $c^h$. We can now show how to lay this tree out in
@@
 spreading out. However, these parameters are not related to the block size of
-our memory level and thus do not violate the CO model of computation. Overall,
+the memory level and thus do not violate the CO model of computation. Overall,
 the PMA is able to keep extra space in $O(1 + (log^2 N) / B)$ memory
@@
 Second, the PMA needs to handle the splitting and merging that will be caused
-by the SWBST discussed in the previous section. THe PMA is able to accomplish
+by the SWBST discussed in the previous section. The PMA is able to accomplish
 this using $O(1 + (log N) /B)$ memory transfers. The proof is based on the
@@
 to move, requiring $O(1 + c^h/B)$ memory transfers. It follows from the
-defitinot of SWBST in [???] that when a node is rebalanced, it will require
+definition of SWBST in [???] that when a node is rebalanced, it will require
 $c^h$ insertions or deletions until it is rebalanced again. Thus, the
@@
 
-Of these three terms, the dominating factor is maintain room for newly
-inserted elements - $O((log^2N)/B)$. We know that SWBST achieve insertions and
-deletitions in $O(log_B N)$ if they are cache-aware. Thus, adding PMA makes
-the cache-oblivious and thereby incurring an additional $O((log^2N)/B)$ memory
+Of these three terms, the dominating factor is maintaining room for newly
+inserted elements - $O((log^2N)/B)$. We know that the SWBST achieves insertions and
+deletions in $O(log_B N)$ if it is cache-aware. Thus, adding PMA makes
+the SWBST cache-oblivious at the cost of incurring an additional $O((log^2N)/B)$ memory
 transfers.
 
-Therefore, overall, CO B-Tree achieve insertions and deletions in $O(log_B N +
-(log^2N)/B)$ memory transfers. However, they still achieve the optimal
+Therefore, overall, the CO B-Tree achieve insertions and deletions in $O(log_B N +
+(log^2N)/B)$ memory transfers. However, it still achieves the optimal
 $O(log_B N)$ memory transfers for searches, matching the performance of

### 2011-12-09 paper.tex
Adding the cituations

@@
 Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
-\cite{cobtree}. The CO B-tree achieves its number of memory transfers by structuring its elements as a
+\cite{BenderDemainColton}. The CO B-tree achieves its number of memory transfers by structuring its elements as a
 strongly weight-balanced search tree (SWBST). The SWBST is then laid out into
@@
 structure. That structure is a balanced search tree. The search tree is laid
-out according to a van Emde Boas (vEB) layout [???, ???]. The idea of a vEB
+out according to a van Emde Boas (vEB) layout \cite{veb1}, \cite{veb2}. The idea of a vEB
 layout is fairly straight forward. Let $h$ denote the height of a tree. the
@@
 A nice consequence of using a vEB layout is that it is known that it will use
-$O(log_B N)$ memory transfers for searches [???, ???]. Thus, the CO B-tree has already
+$O(log_B N)$ memory transfers for searches \cite{veb1}, \cite{veb2}. Thus, the CO B-tree has already
 achieved the desired number of memory transfers for search queries. It
@@
 A known data structure that achieves this property is the ``weight-balanced
-B-tree'' \cite{????}. The CO B-tree leverages this data structure to maintain
+B-tree'' \cite{swbst}. The CO B-tree leverages this data structure to maintain
 its elements. Before SWBST can be used, however, it remains to be proven that
@@
 
-The definition of a SWBST given in [????] provides us with this bound for the
+The definition of a SWBST given in \cite{swbst} provides us with this bound for the
 weight of a node:
@@
 memory transfers. The packed memory array (PMA) was developed to solve this
-problem [???]. There are three main costs that the PMA is concerned with.
+problem \cite{BenderDemainColton}. There are three main costs that the PMA is concerned with.
 
@@
 maintaining the locality of elements while keeping space available for new
-inserts. The PMA uses a technique described in [???]. At a high level, the PMA
+inserts. The PMA uses a technique described in \cite{pma}. At a high level, the PMA
 maintains a windows into the array. When a window becomes too full or too
@@
 to move, requiring $O(1 + c^h/B)$ memory transfers. It follows from the
-definition of SWBST in [???] that when a node is rebalanced, it will require
+definition of SWBST in \cite{swbst} that when a node is rebalanced, it will require
 $c^h$ insertions or deletions until it is rebalanced again. Thus, the
@@
 
-Therefore, overall, the CO B-Tree achieve insertions and deletions in $O(log_B N +
-(log^2N)/B)$ memory transfers. However, it still achieves the optimal
+Therefore, overall, the CO B-Tree achieves insertions and deletions in 
+$O(log_B N + (log^2N)/B)$ memory transfers. However, it still achieves the optimal
 $O(log_B N)$ memory transfers for searches, matching the performance of

### 2011-12-09 paper.tex
Got LaTeX to work
Finally getting the stupid citations to work.
Adding the new figures

@@
 \begin{center}
-	\includegraphics[width=0.8\columnwidth]{figures/veb.png}
+	\includegraphics[width=0.8\columnwidth]{figures/veb.pdf}
 \end{center}
@@
 maintaining the locality of elements while keeping space available for new
-inserts. The PMA uses a technique described in \cite{pma}. At a high level, the PMA
+inserts. The PMA uses a technique described in \cite{packedmemoryarray}. At a high level, the PMA
 maintains a windows into the array. When a window becomes too full or too
@@
 \begin{center}
-	\includegraphics[width=0.8\columnwidth]{figures/veb.png}
+	\includegraphics[width=0.8\columnwidth]{figures/buffers.pdf}
 \end{center}
 
-\caption{An illustration of how the shuttle tree determines the size of a node's linked list}
+\caption{An illustration of how the shuttle tree determines the size of a node's linked list. The 
+node $n2$ will have a linked list of size 2 because it is the leaf of two of the recursive subtrees. 
+On the other hand $n1$ will only have a linked list of size 1.}
 \label{fig:buffers}
@@
 \bibliographystyle{abbrv}
+
 \bibliography{citations}  % sigproc.bib is the name of the Bibliography in this case
+
 % You must have a proper ".bib" file

### 2011-12-09 paper.tex
made some grammar edits to the first section, still need to edit the rest of the paper for grammar.

--- 2011-12-09 paper.tex
@@
 
-This paper present a survey on cache-oblivious B-trees (CO B-trees). The CO
-B-tree was first introduce by Bender et al in 2000. It is an external memory
+This paper presents a survey of cache-oblivious B-trees (CO B-trees). The CO
+B-tree was first introduced by Bender et al. in 2000. It is an external memory
 data structure - meaning that it optimizes the number of memory transfers it
-performs rather than the number of CPU instructions that it executions. Unlike
-other external memory data structures, like the widely use B-tree, the CO
+performs rather than the number of CPU instructions that it executes. Unlike
+other external memory data structures, like the widely used B-tree, the CO
 B-tree does not require any information about the number of level in the
@@
 achieve its memory performance. We then present and analyze three followup CO
-B-tree data structures that improved upon shortcoming of the original: the
+B-tree data structures that improve upon weaknesses of the original: the
 cache-oblivious string B-tree, which introduces support for variable length
@@
 Most algorithms assume an idealized model of computation that consists of a
-single CPU, which performs instructions, and memory, which is a flat array
+single CPU and memory as a flat array
 that can be read and written. However, in reality memory is not a flat array
@@
 any necessary operations. Unfortunately, this cache often has a small capacity
-and once it is filled any new data requests cause some existing cached data to
-be evicted. Transfering data from lower to higher levels in the memory
+and once it is full any new data requests cause some existing cached data to
+be evicted. Transferring data from lower to higher levels in the memory
 hierarchy often takes a long time. In fact, access times between the different
@@
 Additionally, the B-tree expects to know the block size of its memory layer in
-advanced. B-Trees are implemented with parameterized code that needs to be
+advance. B-Trees are implemented with parameterized code that needs to be
 changed for every level in the memory hierarchy. Needless to say, this tuning
@@
 memory hierarchy. It does not know anything about the speed or the block size
-of any level. Conceptually, the CO model allow algorithm designers to reason
-about the simple two-level model like DAM but prove results for any arbitrary
+of any level. Conceptually, the CO model allows algorithm designers to reason
+about the simple two-level model like DAM but also prove results for any arbitrary
 multi-level memory hierarchy. This is because algorithms that perform well in
@@
 
-Most weight-balanced search trees which requires that the heights of a node's
+Most weight-balanced search trees require that the heights of a node's
 left and right subtrees are simply off by a constant factor. The SWBST enforces the
@@
 
-Lets now consider insertions. To insert an element into a WSBST we search down
+Let us now consider insertions. To insert an element into a WSBST we search down
 the tree and find the natural place to insert the element - just like in any
@@
 handled. After merging, you may produce a new node that has a weight larger
-than the bounds. If this happens we immediately split the node again. Thus,
-this can be thought of as stealing nodes from a neighbor.
+than the bounds. If this happens we immediately split the node again. This can 
+be thought of as stealing nodes from a neighbor.
 
@@
 
-Fist, the PMA needs to maintain open space in its memory for newly inserted
+First, the PMA needs to maintain open space in its memory for newly inserted
 items. The PMA is able to accomplish this using $O(1 + (log^2 N) / B)$ memory
@@
 maintaining the locality of elements while keeping space available for new
-inserts. The PMA uses a technique described in \cite{pma}. At a high level, the PMA
-maintains a windows into the array. When a window becomes too full or too
+inserts. The PMA uses a technique described in \cite{packedmemoryarray}. At a high level, the PMA
+maintains windows into the array. When a window becomes too full or too
 empty, the elements are spread out within the context of a larger window. Two
@@
 
-Second, the PMA needs to handle the splitting and merging that will be caused
+Second, the PMA needs to handle the splitting and merging caused
 by the SWBST discussed in the previous section. The PMA is able to accomplish
@@
 In addition, string B-trees do not support compression, and commonly used front compression is handled sub-optimally.
-Regular B-trees and string B-trees are not cache oblivious, and thus incur poorer performance as described previously in *cite original b-tree paper*.
+Regular B-trees and string B-trees are not cache oblivious, and thus incur poorer performance as described previously.
 Range queries have poor performance in B-trees.
@@
 
-In the following subsections, centroid trees are explained, and a summary of how to modify them to preserve weight balance in the dynamic case is provided.
+In the following subsections, centroid trees are explained, and a summary of how to modify them to preserve balance and structure in the dynamic case is provided.
 Two iterations on forward compression are also summarized to show how to incrementally add support for variable length strings as keys for dynamic COSB-trees.
-Lastly, the design and use of the hashdata and keydata are explained.
 
 \subsubsection{The Dictionary Matching Problem and Centroid Trees} %todo:cite dictionary matching problem--sources in cosb-tree paper
-In this approach to solving the dictionary matching problem, a dictionary $D$ of $N$ keys is preprocessed using hashing and divide-and-conquer via the construction of a centroid tree over the keys.
+In this approach to solving the dictionary matching problem *cite dictionary matching problem paper*, a dictionary $D$ of $N$ keys is preprocessed using hashing and divide-and-conquer via the construction of a centroid tree over the keys.
 This technique forms the basis for the Cache Oblivious String B-tree (COSB-tree) design.
@@
 Otherwise traverse the trie resulting from excluding $y$ and its subtree.
-The \emph{centroid tree} is obtained by rooting the tree at \emph{centroid vertex} $\rho$, and making its children recursively defined \emph{centroid trees} of its up and down tries.
+The \emph{centroid tree} is obtained by rooting the tree at \emph{centroid vertex} $\rho$, and making its children recursively defined centroid trees of its up and down tries.
 
@@
 To avoid mismatches due to overlap in hash values, the original key values are compared after finding a match in hash values.
-The resulting time bound for \Search{} (interchangeable with \Member{}) is $O(||k||+logN)$.
+The resulting time bound for \Search{} (interchangeable with \Member{}) is $O(||k||+\log{}N)$.
 If each node in the centroid tree knows its successor and predecessor descendants, \Pred{} and \Succ{} can be achieved in $O(||k||+||k'||+logN)$, where $k'$ is the successor or predecessor of $k$.
@@
 
-The centroid tree is built using $O(N/logN)$ keys, and the centroid tree is layed out in memory using a modified van Emde Boas (vEB) layout for weight-balanced binary trees.
+The centroid tree is built using $O(N/\log{}N)$ keys, and the centroid tree is layed out in memory using a modified van Emde Boas (vEB) layout for weight-balanced binary trees.
 The \emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into equal-size subtrees for the vEB layout.
@@
 The nodes are stored in a PMA structure, preserving order of the nodes using $\Theta(N)$space, apart from inserts and deletes.
-On insert or delete, the structure is scanned in both directions starting from the update point for sections of the array to rebalance based on a "density" threshold of the section.
-Rebalance here refers to evenly distributing the nodes across the section.
+On insert or delete, the structure is scanned in both directions starting from the update point for sections of the array to rearrange based on a "density" threshold of the section.
+Rearranging here refers to evenly distributing the nodes across the section to make room for more nodes.
 To ensure that updating pointers isn't too expensive, any node included in the section must have its subtrees included.
@@
 The proof is as follows: As you recursively traverse the trie, both $\alpha$ and $\gamma$ are in the up trie or down trie, until at some point one of them must be the centroid vertex.
-This is used to redefine what the predecessor/ancestor pointers refer to.
-The centroid may no loner point to the correct predecessor or successor with respect to the relevant subtrie, but one of the matched ancestors by construction must point to the correct predecessor/successor.
+This is used to redefine what the predecessor successor pointers refer to.
+The centroid may no loner point to the correct predecessor or successor with respect to the relevant subtrie, but one of the matched ancestors by construction must point to the correct predecessor or successor.
 Thus for the dynamic case, when dealing with successors and predecessors, $O(logN)$ ancestors' successor and predecessor pointers in addition to the successor and predecessor pointers of the node in question must be checked.
@@
 To compress these keys using front compression, let $\pi_{j}$ be the longest common prefix between $k_{j}$ and $k_{j+1}$.
-Now instead we store the keys as follows:
+Now instead we store the keys in the following format:
 \begin{center}
@@
 
-While B-trees achieve an optimal bound for the number of memory transfer for
+While B-trees achieve an optimal bound for the number of memory transfers for
 search, they do not perform as well for updates and insertions. The
@@
 
-This search vs insertion trade off also exists in the cache-oblivious model.
+This search vs insertion trade-off also exists in the cache-oblivious model.
 The original BO B-tree described in Section~\ref{sec:original} is designed to

### 2011-12-09 paper.tex
made more minor grammar edits

@@
 
-\title{Cache Oblivious B-trees}
+\title{Survey of Cache Oblivious B-trees}
 %
@@
 Whenever you insert an element you check to see if there is another array of
-the same size (this can be though of as checking to see if there is another
+the same size (this can be thought of as checking to see if there is another
 array in the same level). If there is another array with the same size, the
@@
 \Search{} can naively be performed by simply scanning each of the levels in
-the COLA. Since everything is sorted the COLA could can use a binary search to
+the COLA. Since everything is sorted the COLA can use a binary search to
 quickly find the element it is looking for. However, this is not very
-efficient. As an improvement COLA, as the name suggest, uses look ahead
+efficient. As an improvement COLA, as the name suggests, uses look ahead
 pointers to reduce the number of memory accesses required. Every 8th element
@@
 $(k+1)$st level. Thus this pointer is referred to as ``real lookahead
-pointer''. Additionally, each every 4th cell in $k$th level array contains two
+pointer''. Additionally, every 4th cell in $k$th level array contains two
 pointers pointing to the next and previous ``real lookahead pointers'' on this
@@
 \Insert{} achieves an amortized $O(log N / B)$ memory transfers. The proof is
-based on the far that an element will never be merged more than $O(log(N))$
+based on the fact that an element will never be merged more than $O(log(N))$
 times. This is because at that point it will be in the last array.
 Additionally, the first $log(B)$ merges that a element participates in uses a
-constant number of memory transfer since we can assume that the relevant
+constant number of memory transfers since we can assume that the relevant
 blocks are already cached. Thus, we only need to worry about memory transfers
@@
 follows that overall, search uses $O(log N)$ memory transfers. The fact that
-only eight elements need to be search per level is shown through induction.
+only eight elements need to be searched per level is shown through induction.
 The base case is trivial. The first three levels of the COLA each have at most
@@
 than eight member transfers. As the COLA follows lookahead pointers to the
-higher levels, it will eventual get to a level where $k \geq 3$. Note that the
+higher levels, it will eventually get to a level where $k \geq 3$. Note that the
 COLA does not ``enter'' this level at the start of the array; rather, its
@@
 each level in the COLA every eighth element in a level is present in the
-previous level. Thus, it follows that in next level only 8 elements need to be
+previous level. Thus, it follows that in the next level only 8 elements need to be
 searched. Furthermore, since the current level was reach from a lower level,
-it follows that only 8 elements need to be search on this level to find the
+it follows that only 8 elements need to be searched on this level to find the
 appropriate $i$. Thus, overall, the COLA uses $O(8log(N)) = O(log(N))$ memory
@@
 \section{Conclusion}
-
+Fill this in.
 %\end{document}  % This is where a 'short' article might terminate
@@
 \section{Acknowledgments}
-Thanks Karger
+Fill this in.
 

### 2011-12-11 paper.tex
added some figures for string b-tree section

@@
 \section{Variable Length Strings}
-
+\begin{figure}
+
+\begin{center}
+	\includegraphics[width=0.8\columnwidth]{figures/COSB-treegeneralstructurediagram.pdf}
+\end{center}
+
+\caption{An illustration of the dynamic COSB-tree structure}
+\label{fig:string_structure}
+\end{figure}
 \subsection{Motivation}
@@
 \subsubsection{The Dictionary Matching Problem and Centroid Trees} %todo:cite dictionary matching problem--sources in cosb-tree paper
+\begin{figure}
+
+\begin{center}
+	\includegraphics[width=0.8\columnwidth]{figures/centroidtreeexample.pdf}
+\end{center}
+
+\caption{A centroid tree example}
+\label{fig:centroid}
+\end{figure}
 In this approach to solving the dictionary matching problem *cite dictionary matching problem paper*, a dictionary $D$ of $N$ keys is preprocessed using hashing and divide-and-conquer via the construction of a centroid tree over the keys.

### 2011-12-11 paper.tex
Changing the column width

@@
 of any level. Conceptually, the CO model allows algorithm designers to reason
-about the simple two-level model like DAM but also prove results for any arbitrary
-multi-level memory hierarchy. This is because algorithms that perform well in
-a CO model should, in theory, perform well regardless of changes in block
-size, cache capacity, and cache speed. Consequently, CO algorithm should
-gracefully scale to multi-level hierarchies as well.
+about the simple two-level model like DAM but also prove results for any
+arbitrary multi-level memory hierarchy. This is because algorithms that
+perform well in a CO model should, in theory, perform well regardless of
+changes in block size, cache capacity, and cache speed. Consequently, CO
+algorithm should gracefully scale to multi-level hierarchies as well.
 
@@
 Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
-\cite{BenderDemainColton}. The CO B-tree achieves its number of memory transfers by structuring its elements as a
-strongly weight-balanced search tree (SWBST). The SWBST is then laid out into
-a packed memory array (PMA) using a van Emde Boas (vEB) layout. The SWBST and
-PMA play crucial roles in the other cache-oblivious data structures that were
-developed after the CO B-tree and thus are described at length in the next two
-sections.
+\cite{BenderDemainColton}. The CO B-tree achieves its number of memory
+transfers by structuring its elements as a strongly weight-balanced search
+tree (SWBST). The SWBST is then laid out into a packed memory array (PMA)
+using a van Emde Boas (vEB) layout. The SWBST and PMA play crucial roles in
+the other cache-oblivious data structures that were developed after the CO
+B-tree and thus are described at length in the next two sections.
 
@@
 structure. That structure is a balanced search tree. The search tree is laid
-out according to a van Emde Boas (vEB) layout \cite{veb1}, \cite{veb2}. The idea of a vEB
-layout is fairly straight forward. Let $h$ denote the height of a tree. the
-vEB layout splits the tree approximately at height $h/2$. This splits the
-original tree into a ``top'' subtree and many ``bottom'' subtrees. Let $A$
-denote the top subtree and $B_1... B_k$ denote the bottom subtrees. The vEB
-layout is defined recursively using the order: $A, B1, ... B_k$. This layout
-is also shown in Figure~\ref{fig:veb}.
+out according to a van Emde Boas (vEB) layout \cite{veb1}, \cite{veb2}. The
+idea of a vEB layout is fairly straight forward. Let $h$ denote the height of
+a tree. the vEB layout splits the tree approximately at height $h/2$. This
+splits the original tree into a ``top'' subtree and many ``bottom'' subtrees.
+Let $A$ denote the top subtree and $B_1... B_k$ denote the bottom subtrees.
+The vEB layout is defined recursively using the order: $A, B1, ... B_k$. This
+layout is also shown in Figure~\ref{fig:veb}.
 
 A nice consequence of using a vEB layout is that it is known that it will use
-$O(log_B N)$ memory transfers for searches \cite{veb1}, \cite{veb2}. Thus, the CO B-tree has already
-achieved the desired number of memory transfers for search queries. It
-still remains to be shown that the CO B-tree achieves the desired number of
-memory transfers for dynamic inserts and deletes.
+$O(log_B N)$ memory transfers for searches \cite{veb1}, \cite{veb2}. Thus, the
+CO B-tree has already achieved the desired number of memory transfers for
+search queries. It still remains to be shown that the CO B-tree achieves the
+desired number of memory transfers for dynamic inserts and deletes.
 
 It is known that the CO B-tree is going to use a balance search tree, but what
-kind? This is where the strongly weight-balanced search tree becomes useful. The
-weight of a node is $u$ is equal to the total number of $u$'s descendants plus
-1. It can also be thought of as the sum of the weights of $u$'s children plus
-1. Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.
-
-Most weight-balanced search trees require that the heights of a node's
-left and right subtrees are simply off by a constant factor. The SWBST enforces the
+kind? This is where the strongly weight-balanced search tree becomes useful.
+The weight of a node is $u$ is equal to the total number of $u$'s descendants
+plus 1. It can also be thought of as the sum of the weights of $u$'s children
+plus 1. Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.
+
+Most weight-balanced search trees require that the heights of a node's left
+and right subtrees are simply off by a constant factor. The SWBST enforces the
 stronger property that a node at height $h$ in the tree (height is the node's
@@
 
-The definition of a SWBST given in \cite{swbst} provides us with this bound for the
-weight of a node:
+The definition of a SWBST given in \cite{swbst} provides us with this bound
+for the weight of a node:
 
@@
 
-Let us now consider insertions. To insert an element into a WSBST we search down
-the tree and find the natural place to insert the element - just like in any
-tree structure. After inserting the element into some node $w$, we have to
+Let us now consider insertions. To insert an element into a WSBST we search
+down the tree and find the natural place to insert the element - just like in
+any tree structure. After inserting the element into some node $w$, we have to
 check to see if any of $w$'s ancestors have become unbalanced by having a
@@
 By doing this, we maintain the bound that $\frac{c^{h-1}}{2} \leq w(u) \leq
-2c^{h-1}$. We cascade these changes up the tree until all of the ancestors
-are balanced.
+2c^{h-1}$. We cascade these changes up the tree until all of the ancestors are
+balanced.
 
@@
 memory transfers. The packed memory array (PMA) was developed to solve this
-problem \cite{BenderDemainColton}. There are three main costs that the PMA is concerned with.
+problem \cite{BenderDemainColton}. There are three main costs that the PMA is
+concerned with.
 
@@
 maintaining the locality of elements while keeping space available for new
-inserts. The PMA uses a technique described in \cite{packedmemoryarray}. At a high level, the PMA
-maintains windows into the array. When a window becomes too full or too
-empty, the elements are spread out within the context of a larger window. Two
-parameters are required, the window size and the threshold for trigger the
-spreading out. However, these parameters are not related to the block size of
-the memory level and thus do not violate the CO model of computation. Overall,
-the PMA is able to keep extra space in $O(1 + (log^2 N) / B)$ memory
-transfers.
-
-Second, the PMA needs to handle the splitting and merging caused
-by the SWBST discussed in the previous section. The PMA is able to accomplish
-this using $O(1 + (log N) /B)$ memory transfers. The proof is based on the
-fact that each node has $c^h$ descendants (recall, this is the key property
-that SWBST provided). Thus, merging or splitting a node will cause $c^h$ nodes
-to move, requiring $O(1 + c^h/B)$ memory transfers. It follows from the
-definition of SWBST in \cite{swbst} that when a node is rebalanced, it will require
-$c^h$ insertions or deletions until it is rebalanced again. Thus, the
-amortized cost of each insertion and/or deletion is $O(1/B)$. When you split a
-node, it may go all the way up the tree and thus overall it will incur $O((log
-N) / B)$ memory transfers.
+inserts. The PMA uses a technique described in \cite{packedmemoryarray}. At a
+high level, the PMA maintains windows into the array. When a window becomes
+too full or too empty, the elements are spread out within the context of a
+larger window. Two parameters are required, the window size and the threshold
+for trigger the spreading out. However, these parameters are not related to
+the block size of the memory level and thus do not violate the CO model of
+computation. Overall, the PMA is able to keep extra space in $O(1 + (log^2 N)
+/ B)$ memory transfers.
+
+Second, the PMA needs to handle the splitting and merging caused by the SWBST
+discussed in the previous section. The PMA is able to accomplish this using
+$O(1 + (log N) /B)$ memory transfers. The proof is based on the fact that each
+node has $c^h$ descendants (recall, this is the key property that SWBST
+provided). Thus, merging or splitting a node will cause $c^h$ nodes to move,
+requiring $O(1 + c^h/B)$ memory transfers. It follows from the definition of
+SWBST in \cite{swbst} that when a node is rebalanced, it will require $c^h$
+insertions or deletions until it is rebalanced again. Thus, the amortized cost
+of each insertion and/or deletion is $O(1/B)$. When you split a node, it may
+go all the way up the tree and thus overall it will incur $O((log N) / B)$
+memory transfers.
 
@@
 Of these three terms, the dominating factor is maintaining room for newly
-inserted elements - $O((log^2N)/B)$. We know that the SWBST achieves insertions and
-deletions in $O(log_B N)$ if it is cache-aware. Thus, adding PMA makes
-the SWBST cache-oblivious at the cost of incurring an additional $O((log^2N)/B)$ memory
-transfers.
+inserted elements - $O((log^2N)/B)$. We know that the SWBST achieves
+insertions and deletions in $O(log_B N)$ if it is cache-aware. Thus, adding
+PMA makes the SWBST cache-oblivious at the cost of incurring an additional
+$O((log^2N)/B)$ memory transfers.
 
 Therefore, overall, the CO B-Tree achieves insertions and deletions in
-$O(log_B N + (log^2N)/B)$ memory transfers. However, it still achieves the optimal
-$O(log_B N)$ memory transfers for searches, matching the performance of
-ordinary, cache-aware, B-trees.
+$O(log_B N + (log^2N)/B)$ memory transfers. However, it still achieves the
+optimal $O(log_B N)$ memory transfers for searches, matching the performance
+of ordinary, cache-aware, B-trees.
 
@@
 \subsection{Motivation}
-In order for B-trees to remain versatile, they must be able to accommodate various forms of data.
-However, traditional B-trees and string B-trees have suboptimal performance when variable length strings are used as keys.
-In addition, string B-trees do not support compression, and commonly used front compression is handled sub-optimally.
-Regular B-trees and string B-trees are not cache oblivious, and thus incur poorer performance as described previously.
-Range queries have poor performance in B-trees.
-In \cite{BenderFaKu06}, Bender et al.
-address these issues by designing a Cache Oblivious String B-tree (COSB-tree), which they argue works efficiently in these problem areas for variable length string keys.
-They also provide a modified front-compression scheme, which gives their approach the advantages of using less space and having improved locality within the compressed data.
-
-Two experiments are performed, in which a regular B-tree implementation tuned to various block sizes is compared to a static COSB-tree implementation.
-The experiment involved measuring the amount of time required to perform: 440,000 and 450,000 inserts of random values, a range query of all values, and 1000 random searches.
-The results of the experiment show that COSB-trees outperformed B-trees most of the time for random searches, range queries and random inserts.
-However, some point between 440,000 and 450,000 inserts the COSB-tree had to rebalance the entire tree, causing it to fall behind in this measurement.
-However, Bender et al.
-argue that the experiment was biased towards the B-tree implementation, since the trees are "young", and thus still well organized.
-The COSB-tree implementation is also compared to Berkeley DB **cite berkeley db**, and achieves performance comparable to Berkeley DB with tuned parameters.
-The results seem to show that there is a window in which large and small-block size B-trees perform poorly, and that COSB-trees may be a reasonable alternative to improve performance.
+
+In order for B-trees to remain versatile, they must be able to accommodate
+various forms of data. However, traditional B-trees and string B-trees have
+suboptimal performance when variable length strings are used as keys. In
+addition, string B-trees do not support compression, and commonly used front
+compression is handled sub-optimally. Regular B-trees and string B-trees are
+not cache oblivious, and thus incur poorer performance as described
+previously. Range queries have poor performance in B-trees. In
+\cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
+Oblivious String B-tree (COSB-tree), which they argue works efficiently in
+these problem areas for variable length string keys. They also provide a
+modified front-compression scheme, which gives their approach the advantages
+of using less space and having improved locality within the compressed data.
+
+Two experiments are performed, in which a regular B-tree implementation tuned
+to various block sizes is compared to a static COSB-tree implementation. The
+experiment involved measuring the amount of time required to perform: 440,000
+and 450,000 inserts of random values, a range query of all values, and 1000
+random searches. The results of the experiment show that COSB-trees
+outperformed B-trees most of the time for random searches, range queries and
+random inserts. However, some point between 440,000 and 450,000 inserts the
+COSB-tree had to rebalance the entire tree, causing it to fall behind in this
+measurement. However, Bender et al. argue that the experiment was biased
+towards the B-tree implementation, since the trees are "young", and thus still
+well organized. The COSB-tree implementation is also compared to Berkeley DB
+**cite berkeley db**, and achieves performance comparable to Berkeley DB with
+tuned parameters. The results seem to show that there is a window in which
+large and small-block size B-trees perform poorly, and that COSB-trees may be
+a reasonable alternative to improve performance.
+
 \subsection{Description}
@@
 \subsubsection{General COSB-tree Structure}
-The static COSB-tree consists of 2 layers: a centroid tree at the top, and an array of the keys stored in lexicographic order on the bottom.
-The leaves of the centroid tree contain pointers to keys in the array.
-The centroid tree is structured such that a search through the centroid tree identifies the local area of the desired key in the array.
-Then a sequential scan is performed across the array to find the key.
-
-The Dynamic COSB-tree consists of 3 major components: a centroid tree at the top; a packed memory array (PMA) structure of hash data for the keys known as the \emph{hashdata} layer; and on the bottom a PMA structure for the original key values, compressed using an augmented version of forward compression known as the \emph{keydata} layer.
-
-The centroid tree on a high level behaves similarly for static and dynamic COSB-trees.
-The leaves of the centroid tree point into the hashdata layer, where a sequential scan is performed to find the correct key by hash value.
-The hashdata layer points directly into the keydata layer, where the appropriate key is decoded, compared and returned.
-
-In the following subsections, centroid trees are explained, and a summary of how to modify them to preserve balance and structure in the dynamic case is provided.
-Two iterations on forward compression are also summarized to show how to incrementally add support for variable length strings as keys for dynamic COSB-trees.
+
+The static COSB-tree consists of 2 layers: a centroid tree at the top, and an
+array of the keys stored in lexicographic order on the bottom. The leaves of
+the centroid tree contain pointers to keys in the array. The centroid tree is
+structured such that a search through the centroid tree identifies the local
+area of the desired key in the array. Then a sequential scan is performed
+across the array to find the key.
+
+The Dynamic COSB-tree consists of 3 major components: a centroid tree at the
+top; a packed memory array (PMA) structure of hash data for the keys known as
+the \emph{hashdata} layer; and on the bottom a PMA structure for the original
+key values, compressed using an augmented version of forward compression known
+as the \emph{keydata} layer.
+
+The centroid tree on a high level behaves similarly for static and dynamic
+COSB-trees. The leaves of the centroid tree point into the hashdata layer,
+where a sequential scan is performed to find the correct key by hash value.
+The hashdata layer points directly into the keydata layer, where the
+appropriate key is decoded, compared and returned.
+
+In the following subsections, centroid trees are explained, and a summary of
+how to modify them to preserve balance and structure in the dynamic case is
+provided. Two iterations on forward compression are also summarized to show
+how to incrementally add support for variable length strings as keys for
+dynamic COSB-trees.
 
@@
 \end{figure}
-In this approach to solving the dictionary matching problem *cite dictionary matching problem paper*, a dictionary $D$ of $N$ keys is preprocessed using hashing and divide-and-conquer via the construction of a centroid tree over the keys.
-This technique forms the basis for the Cache Oblivious String B-tree (COSB-tree) design.
-Consider the compacted trie $T$ of $D$.
-Bender et al.
-observe that there exists a centroid vertex $\rho$ in $T$ such that $\rho$ has between N/3 and 2N/3 descendants, inclusive.
-When searching for a key $k$ in the compacted trie, you traverse the tree as follows: for each node $y$ that you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on into the trie rooted at $y$ or \emph{down trie} of $y$.
-Otherwise traverse the trie resulting from excluding $y$ and its subtree.
-The \emph{centroid tree} is obtained by rooting the tree at \emph{centroid vertex} $\rho$, and making its children recursively defined centroid trees of its up and down tries.
-
-At each step through the centroid tree, a consistent fraction of the nodes are eliminated from the search, resulting in an $O(logN)$ time bound when ignoring the time required to operate on keys.
-To make key comparisons faster, the keys are hashed and hash values are compared in place of the original key values.
-To avoid mismatches due to overlap in hash values, the original key values are compared after finding a match in hash values.
-The resulting time bound for \Search{} (interchangeable with \Member{}) is $O(||k||+\log{}N)$.
-If each node in the centroid tree knows its successor and predecessor descendants, \Pred{} and \Succ{} can be achieved in $O(||k||+||k'||+logN)$, where $k'$ is the successor or predecessor of $k$.
-Note that the additive $logN$ incurred by the use of centroid trees can be avoided when using main memory.
-Centroid trees are used to capitalize on locality for the cache oblivious string B-tree design, described later in this section.
-
-The centroid tree is built using $O(N/\log{}N)$ keys, and the centroid tree is layed out in memory using a modified van Emde Boas (vEB) layout for weight-balanced binary trees.
-The \emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into equal-size subtrees for the vEB layout.
-The \emph{hyperhyperfloor} of x is denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to the nearest power of a power of 2.
-This ensures trees of equal size on all \emph{levels of detail}, as the centroid tree is traversed.
-
-The following approach is used to rebalance the centroid tree: Put all the elements in DFS or Euler tour order for the trie.
-Given this order, scan the trie to identify the centroid vertex, partition the trie into its respective up and down tries, and recurse on the up and down tries.
-The Euler tour order can be achieved via $O(\log{}N)$ scans, since the trie is only $O(\log{}N)$ deep.
-Thus updates have amortized cost of $O(1+\log^{2}N/B)$ memory transfers.
-Note that this doesn't include updating successor or predecessor pointers, which is described in more detail below.
-
-The following approach is used for maintaining the vEB layout for centroid trees in the dynamic COSB-tree: first, weight balance is relaxed from a maximum weight difference of 2 between left and right subtrees, to a larger constant $c$.
-Thus a node $v$ only becomes out of balance every $\Omega(\textsc{weight(v)})$ insertions or deletions of descendants of v, reducing the frequency of rebuilds.
-The nodes are stored in a PMA structure, preserving order of the nodes using $\Theta(N)$space, apart from inserts and deletes.
-On insert or delete, the structure is scanned in both directions starting from the update point for sections of the array to rearrange based on a "density" threshold of the section.
-Rearranging here refers to evenly distributing the nodes across the section to make room for more nodes.
-To ensure that updating pointers isn't too expensive, any node included in the section must have its subtrees included.
-Specifically, the section is built up from the the lowest subtree level that contains or contained the update point, and neighboring subtrees are continuously added until either the density threshold is met or the entire current subtree layout at this "level of detail" is included.
-For example, suppose a node is inserted at the leaf of some lower recursive subtree $L_{i}$.
-To continue building up this section, $L_{i-1}$ or $L_{i+1}$ could be added.
-If all the lower recursive subtrees $L_{1}$,$L_{2}$,...,$L_{x}$ have been added, the top level tree $U$ is added.
-This corresponds to a subtree $L'_{j}$ one level above.
-Thus repeat this process until the density threshold is met, and then rebalance the tree.
-This approach depends on the fact that all the recursive subtrees at the same level of detail asymptotically have the same size, which Bender et al.
-prove about the hyperhyperfloor-based vEB layout described previously.
-This approach results in an amortized update cost of $O(1+(\log{}N)/B)$ memory transfers per update.
-In conjunction with the cost of rebuilds described above, this is only an additive $O(\log{}N/B)$ factor, so the overall bound is amortized $O(1+\log^{2}N/B)$ memory transfers.
-
-Note that making the COSB-trees dynamic causes updates to successor and predecessor pointers in the centroid tree to be expensive.
-Thus the notion of ancestors is relaxed to the following (Lemma 10 in \cite{BenderFaKu06}): Let $\alpha$ be the parent node of $\gamma$ in the compressed trie.
-Then either $\alpha$ is a descendant of $\gamma$ or $\gamma$ is a descendant of $\alpha$.
-The proof is as follows: As you recursively traverse the trie, both $\alpha$ and $\gamma$ are in the up trie or down trie, until at some point one of them must be the centroid vertex.
-This is used to redefine what the predecessor successor pointers refer to.
-The centroid may no loner point to the correct predecessor or successor with respect to the relevant subtrie, but one of the matched ancestors by construction must point to the correct predecessor or successor.
-Thus for the dynamic case, when dealing with successors and predecessors, $O(logN)$ ancestors' successor and predecessor pointers in addition to the successor and predecessor pointers of the node in question must be checked.
-However, this is an additive factor, so the same time bound for update described above is still achieved.
+
+In this approach to solving the dictionary matching problem *cite dictionary
+matching problem paper*, a dictionary $D$ of $N$ keys is preprocessed using
+hashing and divide-and-conquer via the construction of a centroid tree over
+the keys. This technique forms the basis for the Cache Oblivious String B-tree
+(COSB-tree) design. Consider the compacted trie $T$ of $D$. Bender et al.
+observe that there exists a centroid vertex $\rho$ in $T$ such that $\rho$ has
+between N/3 and 2N/3 descendants, inclusive. When searching for a key $k$ in
+the compacted trie, you traverse the tree as follows: for each node $y$ that
+you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on into the
+trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the trie
+resulting from excluding $y$ and its subtree. The \emph{centroid tree} is
+obtained by rooting the tree at \emph{centroid vertex} $\rho$, and making its
+children recursively defined centroid trees of its up and down tries.
+
+At each step through the centroid tree, a consistent fraction of the nodes are
+eliminated from the search, resulting in an $O(logN)$ time bound when ignoring
+the time required to operate on keys. To make key comparisons faster, the keys
+are hashed and hash values are compared in place of the original key values.
+To avoid mismatches due to overlap in hash values, the original key values are
+compared after finding a match in hash values. The resulting time bound for
+\Search{} (interchangeable with \Member{}) is $O(||k||+\log{}N)$. If each node
+in the centroid tree knows its successor and predecessor descendants, \Pred{}
+and \Succ{} can be achieved in $O(||k||+||k'||+logN)$, where $k'$ is the
+successor or predecessor of $k$. Note that the additive $logN$ incurred by the
+use of centroid trees can be avoided when using main memory. Centroid trees
+are used to capitalize on locality for the cache oblivious string B-tree
+design, described later in this section.
+
+The centroid tree is built using $O(N/\log{}N)$ keys, and the centroid tree is
+layed out in memory using a modified van Emde Boas (vEB) layout for
+weight-balanced binary trees. The \emph{hyperhyperfloors} of nodes are used as
+a guide for cutting the tree into equal-size subtrees for the vEB layout. The
+\emph{hyperhyperfloor} of x is denoted as $\lhyperfloor{}x\rhyperfloor{}$, and
+refers to rounding x down to the nearest power of a power of 2. This ensures
+trees of equal size on all \emph{levels of detail}, as the centroid tree is
+traversed.
+
+The following approach is used to rebalance the centroid tree: Put all the
+elements in DFS or Euler tour order for the trie. Given this order, scan the
+trie to identify the centroid vertex, partition the trie into its respective
+up and down tries, and recurse on the up and down tries. The Euler tour order
+can be achieved via $O(\log{}N)$ scans, since the trie is only $O(\log{}N)$
+deep. Thus updates have amortized cost of $O(1+\log^{2}N/B)$ memory transfers.
+Note that this doesn't include updating successor or predecessor pointers,
+which is described in more detail below.
+
+The following approach is used for maintaining the vEB layout for centroid
+trees in the dynamic COSB-tree: first, weight balance is relaxed from a
+maximum weight difference of 2 between left and right subtrees, to a larger
+constant $c$. Thus a node $v$ only becomes out of balance every
+$\Omega(\textsc{weight(v)})$ insertions or deletions of descendants of v,
+reducing the frequency of rebuilds. The nodes are stored in a PMA structure,
+preserving order of the nodes using $\Theta(N)$space, apart from inserts and
+deletes. On insert or delete, the structure is scanned in both directions
+starting from the update point for sections of the array to rearrange based on
+a "density" threshold of the section. Rearranging here refers to evenly
+distributing the nodes across the section to make room for more nodes. To
+ensure that updating pointers isn't too expensive, any node included in the
+section must have its subtrees included. Specifically, the section is built up
+from the the lowest subtree level that contains or contained the update point,
+and neighboring subtrees are continuously added until either the density
+threshold is met or the entire current subtree layout at this "level of
+detail" is included. For example, suppose a node is inserted at the leaf of
+some lower recursive subtree $L_{i}$. To continue building up this section,
+$L_{i-1}$ or $L_{i+1}$ could be added. If all the lower recursive subtrees
+$L_{1}$,$L_{2}$,...,$L_{x}$ have been added, the top level tree $U$ is added.
+This corresponds to a subtree $L'_{j}$ one level above. Thus repeat this
+process until the density threshold is met, and then rebalance the tree. This
+approach depends on the fact that all the recursive subtrees at the same level
+of detail asymptotically have the same size, which Bender et al. prove about
+the hyperhyperfloor-based vEB layout described previously. This approach
+results in an amortized update cost of $O(1+(\log{}N)/B)$ memory transfers per
+update. In conjunction with the cost of rebuilds described above, this is only
+an additive $O(\log{}N/B)$ factor, so the overall bound is amortized
+$O(1+\log^{2}N/B)$ memory transfers.
+
+Note that making the COSB-trees dynamic causes updates to successor and
+predecessor pointers in the centroid tree to be expensive. Thus the notion of
+ancestors is relaxed to the following (Lemma 10 in \cite{BenderFaKu06}): Let
+$\alpha$ be the parent node of $\gamma$ in the compressed trie. Then either
+$\alpha$ is a descendant of $\gamma$ or $\gamma$ is a descendant of $\alpha$.
+The proof is as follows: As you recursively traverse the trie, both $\alpha$
+and $\gamma$ are in the up trie or down trie, until at some point one of them
+must be the centroid vertex. This is used to redefine what the predecessor
+successor pointers refer to. The centroid may no loner point to the correct
+predecessor or successor with respect to the relevant subtrie, but one of the
+matched ancestors by construction must point to the correct predecessor or
+successor. Thus for the dynamic case, when dealing with successors and
+predecessors, $O(logN)$ ancestors' successor and predecessor pointers in
+addition to the successor and predecessor pointers of the node in question
+must be checked. However, this is an additive factor, so the same time bound
+for update described above is still achieved.
 
 \subsubsection{Modifying Front Compression to Preserve Locality}
-This section provides a brief summary of how front compression is modified to optimize compression and decoding for static and dynamic cache-oblivious string B-trees.
-Consider a sequence of $j$ keys to store $k_{1}$,  $k_{2}$, ...,  $k_{j}$.
-Without compression, this requires $\sum_{j}||k_{j}||$ space in memory.
-To compress these keys using front compression, let $\pi_{j}$ be the longest common prefix between $k_{j}$ and $k_{j+1}$.
-Now instead we store the keys in the following format:
+
+This section provides a brief summary of how front compression is modified to
+optimize compression and decoding for static and dynamic cache-oblivious
+string B-trees. Consider a sequence of $j$ keys to store $k_{1}$, $k_{2}$,
+..., $k_{j}$. Without compression, this requires $\sum_{j}||k_{j}||$ space in
+memory. To compress these keys using front compression, let $\pi_{j}$ be the
+longest common prefix between $k_{j}$ and $k_{j+1}$. Now instead we store the
+keys in the following format:
+
 \begin{center}
@@
 \end{center}
-where $\sigma_{j}$ is the suffix of $k_{j}$ after removing the first $\pi_{j}$ bits.
-This saves almost $\sum_{j}||\pi_{j}||$ memory.
-However, since decoding $k_{j}$ involves concatenating the first $\pi_{j}$ bits of $k_{j-1}$ to $\sigma_{j}$, $k_{j-1}$ must be decoded.
-$k_{j-1}$ may depend on previous keys as well, making $k_{j}$ expensive to decode.
-
-The goal of modifying front compression is to maintain some of the space benefits to compression while reducing the high decoding cost and keeping the size of D small.
-Bender et al.
-provide a modification for static COSB-trees that reduces the high decoding cost of front compression of D to $O(1+||k||B)$, with size $O(\llangle{D}\rrangle{})$ after compression.
-The modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be the first key added to the compression stream.
-For each subsequent key $k_{i}$, if $k_{i}$ can be decoded with just the last $c||k_{i}||$ characters, add $\pi_{i}$,$\sigma_{i}$ like before.
-Otherwise, add 0,$k_{i}$, which means add $k_{i}$ to the stream in its uncompressed form.
-By construction, this scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.%todo: add something about why this isO(<<D>>)?
-
-However, the above modification falls short for anticipating insertions and deletions.
-The previous modification relies on each key being decoded within $c||k^{*}||$ elements of the compressed array, but this is no longer guaranteed when new keys can be inserted to the left of $k$.
-To make the locality-preserving front compression technique suitable for dynamic COSB-trees, Bender et al.
-modify the algorithm to include an extra copied prefix for each key.
-Consider the $3c||k||$ characters that would be to the left of a new key$k^{*}$ after insertion.
-Let $k'$ be the key at the end of this range.
-The idea is to prevent the insertion of $k^{*}$ from affecting any non-copied keys just to the right of $k'$, since this could potentially increase their decode cost.
-Compute the length $l$ of the largest common prefix (lcp) between $k^{*}$ and $k'$, and compare $l$ to all the other lcp's in this range.
-$l$ represents the minimum length for any lcp between $k^{*}$ and $k'$, since the keys are stored in lexicographic order.
-If there is a copied prefix of length at least $l$ or a copied key within this range, insert $k^{*}$ as usual.
-Inserting $k^{*}$ will have no negative effect in this respect, since the decode length is effectively reset by the copied key or prefix like before.
-However, if inserting $k^{*}$ causes the minimum length of the lcp's in this range to increase, we need to fix the compression.
-To do this, we update the lcp of the key directly to the right of $k^{*}$ and store an extra copy of this lcp with $k^{*}$ itself.
-In this way, we can charge the work of copying extra prefixes for a new key to the appropriate number of elements before it.
-This number is determined by the length of the extra copied prefix stored with the key.
+
+where $\sigma_{j}$ is the suffix of $k_{j}$ after removing the first $\pi_{j}$
+bits. This saves almost $\sum_{j}||\pi_{j}||$ memory. However, since decoding
+$k_{j}$ involves concatenating the first $\pi_{j}$ bits of $k_{j-1}$ to
+$\sigma_{j}$, $k_{j-1}$ must be decoded. $k_{j-1}$ may depend on previous keys
+as well, making $k_{j}$ expensive to decode.
+
+The goal of modifying front compression is to maintain some of the space
+benefits to compression while reducing the high decoding cost and keeping the
+size of D small. Bender et al. provide a modification for static COSB-trees
+that reduces the high decoding cost of front compression of D to
+$O(1+||k||B)$, with size $O(\llangle{D}\rrangle{})$ after compression. The
+modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be
+the first key added to the compression stream. For each subsequent key
+$k_{i}$, if $k_{i}$ can be decoded with just the last $c||k_{i}||$ characters,
+add $\pi_{i}$,$\sigma_{i}$ like before. Otherwise, add 0,$k_{i}$, which means
+add $k_{i}$ to the stream in its uncompressed form. By construction, this
+scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory
+transfers.%todo: add something about why this isO(<<D>>)?
+
+However, the above modification falls short for anticipating insertions and
+deletions. The previous modification relies on each key being decoded within
+$c||k^{*}||$ elements of the compressed array, but this is no longer
+guaranteed when new keys can be inserted to the left of $k$. To make the
+locality-preserving front compression technique suitable for dynamic
+COSB-trees, Bender et al. modify the algorithm to include an extra copied
+prefix for each key. Consider the $3c||k||$ characters that would be to the
+left of a new key$k^{*}$ after insertion. Let $k'$ be the key at the end of
+this range. The idea is to prevent the insertion of $k^{*}$ from affecting any
+non-copied keys just to the right of $k'$, since this could potentially
+increase their decode cost. Compute the length $l$ of the largest common
+prefix (lcp) between $k^{*}$ and $k'$, and compare $l$ to all the other lcp's
+in this range. $l$ represents the minimum length for any lcp between $k^{*}$
+and $k'$, since the keys are stored in lexicographic order. If there is a
+copied prefix of length at least $l$ or a copied key within this range, insert
+$k^{*}$ as usual. Inserting $k^{*}$ will have no negative effect in this
+respect, since the decode length is effectively reset by the copied key or
+prefix like before. However, if inserting $k^{*}$ causes the minimum length of
+the lcp's in this range to increase, we need to fix the compression. To do
+this, we update the lcp of the key directly to the right of $k^{*}$ and store
+an extra copy of this lcp with $k^{*}$ itself. In this way, we can charge the
+work of copying extra prefixes for a new key to the appropriate number of
+elements before it. This number is determined by the length of the extra
+copied prefix stored with the key.
 
 \subsection{Operators}
+
 %todo: figure out how to do range queries from this paper somehow...
+
 The operations supported are dictated by the Dictionary Matching problem.
-These operations are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{}, and \Range{}.
-The behavior of each operation for dynamic COSB-trees is described below.
-\Insertkonly{} and \Delete{} are also supported but not described here.
-
-\Search{}: works as follows for a given key $k$.
-Compute the hash value for $k$ and call it $\rho$, and traverse the centroid tree of the COSB-tree using $\rho$ to obtain the successor $\rho_{s}$ and predecessor $\rho_{p}$ of $k$.
-Index into the hashdata layer and perform a sequential scan across the PMA structure from $\rho_{p}$  to $\rho_{s}$ to find $\rho$.
-On a match with value $\rho'$, index into the keydata layer and decode $\rho'$ to obtain the resulting key $k'$.
-If $k' = k$, return true.
-If $k' \ne k$ or $k$ was not found during the sequential scan return false.
-
-\Pred{},\Succ{}: works in the same way as \Search{} until the sequential scan in the hashdata layer.
-When scanning the hashdata layer, look for $\rho_{p}$ or $\rho_{s}$, respectively, in place of $\rho$, and compare the appropriate keys in the keydata layer.
-
-\Range{}: The paper doesn't say, but presumably this is very similar to \Search{}.
-Let $k_{b}$ be the beginning of the range and let $k_{e}$ be the end of the range.
-Hash the range keys and identify the predecessor of $\rho_{b}$ and the successor of $\rho_{e}$ using the centroid tree.
-Scan the hashdata layer across from  $\rho_{b,p}$ to  $\rho_{e,s}$.
-Only start noting the beginning of range when the first key with hash value after $\rho_{b}$, and stop once we reach the first key with hash value after $\rho_{e}$.
-Decode and return the set of keys within the given range in the keydata layer.
+These operations are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{},
+and \Range{}. The behavior of each operation for dynamic COSB-trees is
+described below. \Insertkonly{} and \Delete{} are also supported but not
+described here.
+
+\Search{}: works as follows for a given key $k$. Compute the hash value for
+$k$ and call it $\rho$, and traverse the centroid tree of the COSB-tree using
+$\rho$ to obtain the successor $\rho_{s}$ and predecessor $\rho_{p}$ of $k$.
+Index into the hashdata layer and perform a sequential scan across the PMA
+structure from $\rho_{p}$ to $\rho_{s}$ to find $\rho$. On a match with value
+$\rho'$, index into the keydata layer and decode $\rho'$ to obtain the
+resulting key $k'$. If $k' = k$, return true. If $k' \ne k$ or $k$ was not
+found during the sequential scan return false.
+
+\Pred{},\Succ{}: works in the same way as \Search{} until the sequential scan
+in the hashdata layer. When scanning the hashdata layer, look for $\rho_{p}$
+or $\rho_{s}$, respectively, in place of $\rho$, and compare the appropriate
+keys in the keydata layer.
+
+\Range{}: The paper doesn't say, but presumably this is very similar to
+\Search{}. Let $k_{b}$ be the beginning of the range and let $k_{e}$ be the
+end of the range. Hash the range keys and identify the predecessor of
+$\rho_{b}$ and the successor of $\rho_{e}$ using the centroid tree. Scan the
+hashdata layer across from $\rho_{b,p}$ to $\rho_{e,s}$. Only start noting the
+beginning of range when the first key with hash value after $\rho_{b}$, and
+stop once we reach the first key with hash value after $\rho_{e}$. Decode and
+return the set of keys within the given range in the keydata layer.
 
 \subsection{Complexity}
-\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high probability (w.h.p.).
-Traversing the centroid tree requires  $O(1+\log_{B}N)$ memory transfers, since the tree is built using$O(N/logN)$ keys and is balanced.
-Once in the hashdata layer, we are already within $O(logN)$ keys of $k$.
-Thus the local scan requires $O(1+(||k||+\log{}N)/B)$ memory transfers.
-Obtaining $k$ from the keydata layer requires no more memory transfers than the scan in the hashdata layer.
-The \Pred{} and \Succ{} operations are almost identical in behavior to searches, and thus only have the added cost of having to operate on $k'$, where $k'$ is the predecessor or successor, respectively.
-These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers w.h.p..
-Similarly, range queries take $O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory transfers, where in this case $k'$ is the other end of the range and $Q$ is the set of keys within the given range.
-
-\Insertkonly{} and \Delete{} w.h.p.
-require $O(1+\log_{B}N+\log^{2}N||k||/B)$ memory transfers.
-As mentioned previously, due to rebuilds updates have amortized cost of $O(1+(\log^{2}N)/B)$ memory transfers.
-The rest of the cost for insertions and deletions is due to having to find the correct key to delete or location for insert, which can be seen from the cost of \Search{}.
+
+\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high
+probability (w.h.p.). Traversing the centroid tree requires $O(1+\log_{B}N)$
+memory transfers, since the tree is built using$O(N/logN)$ keys and is
+balanced. Once in the hashdata layer, we are already within $O(logN)$ keys of
+$k$. Thus the local scan requires $O(1+(||k||+\log{}N)/B)$ memory transfers.
+Obtaining $k$ from the keydata layer requires no more memory transfers than
+the scan in the hashdata layer. The \Pred{} and \Succ{} operations are almost
+identical in behavior to searches, and thus only have the added cost of having
+to operate on $k'$, where $k'$ is the predecessor or successor, respectively.
+These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory
+transfers w.h.p.. Similarly, range queries take
+$O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory transfers,
+where in this case $k'$ is the other end of the range and $Q$ is the set of
+keys within the given range.
+
+\Insertkonly{} and \Delete{} w.h.p. require $O(1+\log_{B}N+\log^{2}N||k||/B)$
+memory transfers. As mentioned previously, due to rebuilds updates have
+amortized cost of $O(1+(\log^{2}N)/B)$ memory transfers. The rest of the cost
+for insertions and deletions is due to having to find the correct key to
+delete or location for insert, which can be seen from the cost of \Search{}.
 
@@
 than eight member transfers. As the COLA follows lookahead pointers to the
-higher levels, it will eventually get to a level where $k \geq 3$. Note that the
-COLA does not ``enter'' this level at the start of the array; rather, its
+higher levels, it will eventually get to a level where $k \geq 3$. Note that
+the COLA does not ``enter'' this level at the start of the array; rather, its
 arrived here by following the real lookahead pointers from the levels below
@@
 each level in the COLA every eighth element in a level is present in the
-previous level. Thus, it follows that in the next level only 8 elements need to be
-searched. Furthermore, since the current level was reach from a lower level,
-it follows that only 8 elements need to be searched on this level to find the
-appropriate $i$. Thus, overall, the COLA uses $O(8log(N)) = O(log(N))$ memory
-transfers.
+previous level. Thus, it follows that in the next level only 8 elements need
+to be searched. Furthermore, since the current level was reach from a lower
+level, it follows that only 8 elements need to be searched on this level to
+find the appropriate $i$. Thus, overall, the COLA uses $O(8log(N)) =
+O(log(N))$ memory transfers.
 
@@
 \subsection{Motivation}
-B-trees are often used as an indexing data structure for various
-database and file systems. Thus, for a cache-oblivious B-tree data structure
-to be of practical value, it must be able to handle concurrent access and modification
-of its internal state. The straw-man solution is to simply ensure mutual exclusion
-on the entire data structure; this approach, however, yields poor performance.
+
+B-trees are often used as an indexing data structure for various database and
+file systems. Thus, for a cache-oblivious B-tree data structure to be of
+practical value, it must be able to handle concurrent access and modification
+of its internal state. The straw-man solution is to simply ensure mutual
+exclusion on the entire data structure; this approach, however, yields poor
+performance.
 
@@
 data structures to solve this problem: the first is an exponential CO B-tree,
-the second is a CO B-tree based on a packed memory array (PMA). Bender
-also presents a lock-free variant of the latter data structure. For
-the remainder of the discussion, assume that the keys are fixed-size integer
-keys.
+the second is a CO B-tree based on a packed memory array (PMA). Bender also
+presents a lock-free variant of the latter data structure. For the remainder
+of the discussion, assume that the keys are fixed-size integer keys.
 
 \subsection{Description}
+
 Before describing the data structures in more detail, we first consider
@@
 \subsubsection{Concurrency Model}
-The concurrency model starts with a system that has a total cache
-size of $M$ and block size of $B$. There are $P$ processors in the
-system, each with a cache of size $\frac{M}{P}$. A particular block
-can reside in multiple caches in \textit{shared} mode. However, if a
-processor wants to perform a write to a block, it must acquire it in 
-\textit{exclusive} mode, in which case that block can only reside in
-a single cache. In this model, a request for a block costs a single
-memory transfer.
-
-For concurrency control, two primitives are used. The first is
-locks. The second is \textit{load-linked/store-conditional} (LL/SC).
-The authors argue that LL/SC provides better performance than 
-using a \textit{compare-and-swap} (CAS) primitive, but their 
-techniques can be extended to use CAS.
+
+The concurrency model starts with a system that has a total cache size of $M$
+and block size of $B$. There are $P$ processors in the system, each with a
+cache of size $\frac{M}{P}$. A particular block can reside in multiple caches
+in \textit{shared} mode. However, if a processor wants to perform a write to a
+block, it must acquire it in \textit{exclusive} mode, in which case that block
+can only reside in a single cache. In this model, a request for a block costs
+a single memory transfer.
+
+For concurrency control, two primitives are used. The first is locks. The
+second is \textit{load-linked/store-conditional} (LL/SC). The authors argue
+that LL/SC provides better performance than using a \textit{compare-and-swap}
+(CAS) primitive, but their techniques can be extended to use CAS.
 
 \subsection{Exponential CO B-tree}
-This data structure is based on a strongly weight-balanced 
-exponential tree found in \cite{Bender2002}. A tuning parameter
-$1 < \alpha < 2$ is chosen for the data structure. Each node
-in the tree maintains both keys which divide up its children,
-in addition to a \textit{right-link} pointer, which contains
-both a pointer to a node's right sibling, plus the key
-value of the minimum key in the node's right sibling. If no such
-sibling exists, a sentinel $\infty$ pointer is used.
-
-A node's layout in memory is as follows. Suppose a node has $k$
-elements in it. The first part of the node consists of a static
-CO search B-tree, containing the $k$ elements. The tree is sized
-to be a complete $\lhyperceil k \rhyperceil$-leaf tree. The second
-part of the node consists of a $\lhyperceil k \rhyperceil$-size array
-contained the $k$ elements in sorted order. The leaf nodes of the
-CO search tree point to the associated entries in the array.
+
+This data structure is based on a strongly weight-balanced exponential tree
+found in \cite{Bender2002}. A tuning parameter $1 < \alpha < 2$ is chosen for
+the data structure. Each node in the tree maintains both keys which divide up
+its children, in addition to a \textit{right-link} pointer, which contains
+both a pointer to a node's right sibling, plus the key value of the minimum
+key in the node's right sibling. If no such sibling exists, a sentinel
+$\infty$ pointer is used.
+
+A node's layout in memory is as follows. Suppose a node has $k$ elements in
+it. The first part of the node consists of a static CO search B-tree,
+containing the $k$ elements. The tree is sized to be a complete $\lhyperceil k
+\rhyperceil$-leaf tree. The second part of the node consists of a $\lhyperceil
+k \rhyperceil$-size array contained the $k$ elements in sorted order. The leaf
+nodes of the CO search tree point to the associated entries in the array.
 
 \subsubsection{Operations}
-The operations supported by the exponential CO B-tree are \Search{}
-and \Insert{}. The authors do not discuss range-queries, and this
-is presumably due to the concurrency issues in implementing a fully-consistent
-range scan (weakly-consistent range scans could be implemented in the data
-structure with very little modification), although interestingly enough
-the PMA variant of B-tree (discussed in Section \ref{sec:pma}) does
-cover range scans.
+
+The operations supported by the exponential CO B-tree are \Search{} and
+\Insert{}. The authors do not discuss range-queries, and this is presumably
+due to the concurrency issues in implementing a fully-consistent range scan
+(weakly-consistent range scans could be implemented in the data structure with
+very little modification), although interestingly enough the PMA variant of
+B-tree (discussed in Section \ref{sec:pma}) does cover range scans.
 
 \Search{} works as follows. For a non-leaf node, first check the
-\textit{right-link} key. If the \textit{right-link} key is less than
-$k$, then traverse the right link, and recurse. Otherwise, follow
-the appropriate child pointer. When a leaf node is reached, only
-follow \textit{right-link} pointers, until either $k$ has been located
-or some element greater than $k$ is reached. No locks are acquired
-during the entire search operation.
+\textit{right-link} key. If the \textit{right-link} key is less than $k$, then
+traverse the right link, and recurse. Otherwise, follow the appropriate child
+pointer. When a leaf node is reached, only follow \textit{right-link}
+pointers, until either $k$ has been located or some element greater than $k$
+is reached. No locks are acquired during the entire search operation.
 
@@
 inserted, and the lock is released. Unlike a standard B-tree, there is no
-maximum number of elements in a node; instead we split nodes probabilistically.
-With probability ${1}/{2^{\alpha^h}}$, a node at height $h$ is split (leaf
-nodes have $h = 0$). A split works as follows.  We reacquire a lock on the node
-to be split. We split the node into $u$ and $u'$, where $u$ contains all the
-keys $< k$, and $u'$ contains the keys $\geq k$. We release the lock, and
-recursively insert $k$ into the parent of the split node. It is important
-to note that, since $u$ contains a prefix of the original node to be split,
-that we reuse the original node in constructing $u$ (that way, concurrent
-searches can still locate the keys in $u$).
+maximum number of elements in a node; instead we split nodes
+probabilistically. With probability ${1}/{2^{\alpha^h}}$, a node at height $h$
+is split (leaf nodes have $h = 0$). A split works as follows. We reacquire a
+lock on the node to be split. We split the node into $u$ and $u'$, where $u$
+contains all the keys $< k$, and $u'$ contains the keys $\geq k$. We release
+the lock, and recursively insert $k$ into the parent of the split node. It is
+important to note that, since $u$ contains a prefix of the original node to be
+split, that we reuse the original node in constructing $u$ (that way,
+concurrent searches can still locate the keys in $u$).
 
@@
 \subsubsection{Complexity}
+
 Sequential search takes $O(\log_B{N} + \log_{\alpha}{\log_2{B}})$ block
@@
 bound comes from the argument that when the tree has height $h =
-\Omega(\log_{\alpha}{\log_2{\log_2{N}}})$, then the total number of keys in the
-tree is $O(2^{\alpha^h})$. The proof is then completed by summing over the
+\Omega(\log_{\alpha}{\log_2{\log_2{N}}})$, then the total number of keys in
+the tree is $O(2^{\alpha^h})$. The proof is then completed by summing over the
 range of $h$ which is probabilistic $O(\log_{\alpha}{\log_2{N}})$, the access
@@
 Sequential insert takes $O(\log_B{N} + \log{\alpha}{\log_2{B}})$ block
-transfers (in expectation). This proof is constructed in a similar 
-fashion as the search proof: consider inserting a tree at some height $h$.
-By noting that the probability a particular node reaches height $h$
-is $2^{-(\alpha^h-1)/(\alpha-1)}$, a summation over all levels
-of the cost of rebuilding a node yields the time-bound.
+transfers (in expectation). This proof is constructed in a similar fashion as
+the search proof: consider inserting a tree at some height $h$. By noting that
+the probability a particular node reaches height $h$ is
+$2^{-(\alpha^h-1)/(\alpha-1)}$, a summation over all levels of the cost of
+rebuilding a node yields the time-bound.
 
@@
 \label{sec:pma}
-The PMA CO B-tree is a two level data structure, containing a static
-CO B-tree which indexes into a packed-memory array. Bender introduces
-a new packed-memory array, called a \textit{one-way packed-memory array},
-which they argue is easier to manage concurrently.
+
+The PMA CO B-tree is a two level data structure, containing a static CO B-tree
+which indexes into a packed-memory array. Bender introduces a new
+packed-memory array, called a \textit{one-way packed-memory array}, which they
+argue is easier to manage concurrently.
 
@@
 
-When the \textit{active region} becomes too large/small due to inserts/deletes,
-the PMA is rebalanced. A new array is allocated, and the elements are
-spread out evenly over the new array (too minimize element density).
+When the \textit{active region} becomes too large/small due to
+inserts/deletes, the PMA is rebalanced. A new array is allocated, and the
+elements are spread out evenly over the new array (too minimize element
+density).
 
@@
 
-\Search{} works as follows. Search the static tree for $k$ until
-a leaf node is reached. Follow the leaf node pointer into the PMA.
-Scan right in the PMA until either $k$ is located, or some element
-$> k$ is reached. Do not acquire any locks during the search.
-
-\Insert{} works as follows. Use \Search{} to locate the $\Theta(\log N)$-region
-of the PMA which (would) contain $k$. Lock the region, and scan right to find
-the appropriate slot $s$ to place $k$ into. $s$ is defined such that slot $s -
-1$ contains the largest key smaller than $k$. If scanning forward requires
-going into the next $\Theta(\log N)$-region, lock the next region and release
-the current region (\textit{hand-over-hand} locking technique). Once $s$
-is located, either $s$ is free, in which case just insert $k$, or $s$ is not free,
-in which case a \textit{rebalance} operation is required.
+\Search{} works as follows. Search the static tree for $k$ until a leaf node
+is reached. Follow the leaf node pointer into the PMA. Scan right in the PMA
+until either $k$ is located, or some element $> k$ is reached. Do not acquire
+any locks during the search.
+
+\Insert{} works as follows. Use \Search{} to locate the $\Theta(\log
+N)$-region of the PMA which (would) contain $k$. Lock the region, and scan
+right to find the appropriate slot $s$ to place $k$ into. $s$ is defined such
+that slot $s - 1$ contains the largest key smaller than $k$. If scanning
+forward requires going into the next $\Theta(\log N)$-region, lock the next
+region and release the current region (\textit{hand-over-hand} locking
+technique). Once $s$ is located, either $s$ is free, in which case just insert
+$k$, or $s$ is not free, in which case a \textit{rebalance} operation is
+required.
 
@@
 region from $s$ to the current position $s'$ is not \textit{too dense}. The
-notion of density is defined in detail in \cite{BenderFiGi05}. The region
-$[s, s']$ is then rebalanced as follows. Start from the rightmost element,
-and move elements to the right (maintaining order) such that no suffix of
-$[s, s']$ is too dense. This rebalancing creates room for $k$; we insert $k$
-and then release the locks.
+notion of density is defined in detail in \cite{BenderFiGi05}. The region $[s,
+s']$ is then rebalanced as follows. Start from the rightmost element, and move
+elements to the right (maintaining order) such that no suffix of $[s, s']$ is
+too dense. This rebalancing creates room for $k$; we insert $k$ and then
+release the locks.
 
@@
 % TODO: can somebody check this time-bound? seems kind of bad
-Sequential search time-bounds are not presented in the paper,
-but presumably since search is a static CO tree lookup followed
-by a scan, the time bound is $O(\log_{B}{N} + {\log_2{N}}/{B} + 1)$.
-
-Sequential insert is $O(\log_{B}{N} + \log_2^2{N}/{B} + 1)$. The proof
-is based off an accounting argument which places $\Theta(\log_2^2{m}/B)$
-dollars in each array slot. The proof is not presented in any detail in the
-paper, but rather it is a slightly-modified version of the proof presented
-in \cite{Katriel02}.
+
+Sequential search time-bounds are not presented in the paper, but presumably
+since search is a static CO tree lookup followed by a scan, the time bound is
+$O(\log_{B}{N} + {\log_2{N}}/{B} + 1)$.
+
+Sequential insert is $O(\log_{B}{N} + \log_2^2{N}/{B} + 1)$. The proof is
+based off an accounting argument which places $\Theta(\log_2^2{m}/B)$ dollars
+in each array slot. The proof is not presented in any detail in the paper, but
+rather it is a slightly-modified version of the proof presented in
+\cite{Katriel02}.
 
 \subsubsection{Lock-free Variant}
+
 The lock-free variant is based on the PMA CO tree, except the locks acquired
@@
 are initiated by denoting the cell in question with the marker. If a process
-ever encounters a cell with a marker, it drops what it is doing and first helps
-to complete the operation. When the originating process gets context-switched
-back in, any SCs will fail, and then the process can check to see if the
-operation was already completed. The implementation of the non-blocking
-primitives is spelled out in more detail in \cite{BenderFiGi05}, but the basic
-idea is using LL/SC to detect when concurrent operations are on-going.
+ever encounters a cell with a marker, it drops what it is doing and first
+helps to complete the operation. When the originating process gets
+context-switched back in, any SCs will fail, and then the process can check to
+see if the operation was already completed. The implementation of the
+non-blocking primitives is spelled out in more detail in \cite{BenderFiGi05},
+but the basic idea is using LL/SC to detect when concurrent operations are
+on-going.
 
@@
 \textit{cell-insertion} operation is performed on $s$. If it succeeds, we are
-done. If it fails, then a rebalance is necessary. Rebalance proceeds as before,
-scanning right to find the region $[s, s']$. However, whenever a non-empty cell
-is encountered, a load-link (LL) is performed on the cell.  Once the region is
-found, the rebalancing-to-the-right is performed as before.  However, a
-store-conditional (SC) is performed on the destination of each cell.  If any SC
-fails, then this means concurrent modification had occurred, in which case the
-algorithm restarts from the beginning.
+done. If it fails, then a rebalance is necessary. Rebalance proceeds as
+before, scanning right to find the region $[s, s']$. However, whenever a
+non-empty cell is encountered, a load-link (LL) is performed on the cell. Once
+the region is found, the rebalancing-to-the-right is performed as before.
+However, a store-conditional (SC) is performed on the destination of each
+cell. If any SC fails, then this means concurrent modification had occurred,
+in which case the algorithm restarts from the beginning.
 

### 2011-12-12 paper.tex
finished restructuring co string b-trees section for now

@@
 \usepackage{graphicx}
+\usepackage{url}
 
@@
 
-Two experiments are performed, in which a regular B-tree implementation tuned
-to various block sizes is compared to a static COSB-tree implementation. The
-experiment involved measuring the amount of time required to perform: 440,000
-and 450,000 inserts of random values, a range query of all values, and 1000
-random searches. The results of the experiment show that COSB-trees
+\subsection{Description}
+Bender et al.
+perform 2 experiments in \cite{BenderFaKu06}, in which a regular B-tree implementation tuned to various block sizes is compared to a static COSB-tree implementation.
+The experiment involved measuring the amount of time required to perform: 440,000 and 450,000 inserts of random values, a range query of all values, and 1000 random searches.
+The results of the experiment show that COSB-trees
 outperformed B-trees most of the time for random searches, range queries and
@@
 well organized. The COSB-tree implementation is also compared to Berkeley DB
-**cite berkeley db**, and achieves performance comparable to Berkeley DB with
+\cite{BerkeleyDB}, and achieves performance comparable to Berkeley DB with
 tuned parameters. The results seem to show that there is a window in which
@@
 
-\subsection{Description}
-
 \subsubsection{General COSB-tree Structure}
 
-The static COSB-tree consists of 2 layers: a centroid tree at the top, and an
-array of the keys stored in lexicographic order on the bottom. The leaves of
-the centroid tree contain pointers to keys in the array. The centroid tree is
-structured such that a search through the centroid tree identifies the local
-area of the desired key in the array. Then a sequential scan is performed
-across the array to find the key.
-
 The Dynamic COSB-tree consists of 3 major components: a centroid tree at the
-top; a packed memory array (PMA) structure of hash data for the keys known as
+top; a PMA structure of hash data for the keys known as
 the \emph{hashdata} layer; and on the bottom a PMA structure for the original
-key values, compressed using an augmented version of forward compression known
-as the \emph{keydata} layer.
-
-The centroid tree on a high level behaves similarly for static and dynamic
-COSB-trees. The leaves of the centroid tree point into the hashdata layer,
-where a sequential scan is performed to find the correct key by hash value.
-The hashdata layer points directly into the keydata layer, where the
-appropriate key is decoded, compared and returned.
+key values, compressed using an augmented version of forward compression known as the \emph{keydata} layer.
+Each leaf of the centroid tree contains a pointer into the hashdata array.
+The centroid tree only contains a subset of the keys, so each leaf points to the respective \emph{neighborhood} of the key being searched for.
+After traversing the centroid tree and following the pointer into the hashdata array layer, we are within $O(\log{}N)$ keys of $k$.
+Then a sequential scan is performed in the hashdata array layer to find $k$.
+Each entry in the hashdata array contains the hash value of the corresponding key, and a pointer to the compressed key value in the keydata layer.
 
@@
 
-\caption{A centroid tree example}
+\caption{A centroid tree example.
+Dotted lines represent centroid-rooted up tries, where the parent is not a prefix of the child.
+Solid lines represent centroid-rooted down tries, where the parent is a prefix of the child.}
 \label{fig:centroid}
@@
 
-In this approach to solving the dictionary matching problem *cite dictionary
-matching problem paper*, a dictionary $D$ of $N$ keys is preprocessed using
-hashing and divide-and-conquer via the construction of a centroid tree over
-the keys. This technique forms the basis for the Cache Oblivious String B-tree
-(COSB-tree) design. Consider the compacted trie $T$ of $D$. Bender et al.
-observe that there exists a centroid vertex $\rho$ in $T$ such that $\rho$ has
-between N/3 and 2N/3 descendants, inclusive. When searching for a key $k$ in
-the compacted trie, you traverse the tree as follows: for each node $y$ that
+COSB-trees are a solution to the dictionary matching problem.
+Dictionary matching, or string matching, refers to matching strings using prefix search.
+Let $D$ be a dictionary of $N$ keys.
+To perform faster searches, the keys in $D$ are preprocessed using hashing and divide-and-conquer via centroid trees.
+Figure \ref{fig:centroid} depicts an example centroid tree.
+
+Consider the compacted\footnote{where compacted means all non-branching paths are reduced to edges.} trie $T$ constructed over $D$ for string matching.
+Consider the vertex $\rho$ in $T$ such that $\rho$ has
+between N/3 and 2N/3 descendants, inclusive. Call $\rho$ the \emph{centroid vertex} of $T$. When searching for a key $k$ in
+the compacted trie, you traverse the trie as follows: for each node $y$ that
 you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on into the
@@
 resulting from excluding $y$ and its subtree. The \emph{centroid tree} is
-obtained by rooting the tree at \emph{centroid vertex} $\rho$, and making its
+obtained by rooting the tree at the centroid vertex $\rho$, and making its
 children recursively defined centroid trees of its up and down tries.
 
-At each step through the centroid tree, a consistent fraction of the nodes are
-eliminated from the search, resulting in an $O(logN)$ time bound when ignoring
-the time required to operate on keys. To make key comparisons faster, the keys
-are hashed and hash values are compared in place of the original key values.
-To avoid mismatches due to overlap in hash values, the original key values are
-compared after finding a match in hash values. The resulting time bound for
-\Search{} (interchangeable with \Member{}) is $O(||k||+\log{}N)$. If each node
-in the centroid tree knows its successor and predecessor descendants, \Pred{}
-and \Succ{} can be achieved in $O(||k||+||k'||+logN)$, where $k'$ is the
-successor or predecessor of $k$. Note that the additive $logN$ incurred by the
-use of centroid trees can be avoided when using main memory. Centroid trees
-are used to capitalize on locality for the cache oblivious string B-tree
-design, described later in this section.
-
-The centroid tree is built using $O(N/\log{}N)$ keys, and the centroid tree is
-layed out in memory using a modified van Emde Boas (vEB) layout for
+The centroid tree in the COSB-tree is layed out in memory using a modified van Emde Boas (vEB) layout for
 weight-balanced binary trees. The \emph{hyperhyperfloors} of nodes are used as
@@
 The following approach is used to rebalance the centroid tree: Put all the
-elements in DFS or Euler tour order for the trie. Given this order, scan the
-trie to identify the centroid vertex, partition the trie into its respective
+elements in DFS or Euler tour order for the compacted trie. Given this order, scan the trie to identify the centroid vertex. Then partition the trie into its respective
 up and down tries, and recurse on the up and down tries. The Euler tour order
-can be achieved via $O(\log{}N)$ scans, since the trie is only $O(\log{}N)$
-deep. Thus updates have amortized cost of $O(1+\log^{2}N/B)$ memory transfers.
-Note that this doesn't include updating successor or predecessor pointers,
-which is described in more detail below.
-
-The following approach is used for maintaining the vEB layout for centroid
-trees in the dynamic COSB-tree: first, weight balance is relaxed from a
+can be found in $O(\log{}N)$ scans, since the trie is only $O(\log{}N)$
+deep. Thus these updates have amortized cost of $O(1+\log^{2}N/B)$ memory transfers.
+
+The following approach is used for maintaining the vEB layout of the centroid
+tree in the COSB-tree: first, weight balance is relaxed from a
 maximum weight difference of 2 between left and right subtrees, to a larger
 constant $c$. Thus a node $v$ only becomes out of balance every
-$\Omega(\textsc{weight(v)})$ insertions or deletions of descendants of v,
-reducing the frequency of rebuilds. The nodes are stored in a PMA structure,
-preserving order of the nodes using $\Theta(N)$space, apart from inserts and
-deletes. On insert or delete, the structure is scanned in both directions
-starting from the update point for sections of the array to rearrange based on
-a "density" threshold of the section. Rearranging here refers to evenly
-distributing the nodes across the section to make room for more nodes. To
+$\Omega(\textsc{weight(v)})$ insertions or deletions of descendants of v. The nodes are stored in a PMA, preserving order of the nodes using $\Theta(N)$space.
+
+On insert or delete, the area around the update point is scanned for sections of the array to rearrange.
+If a section around the update point reaches a certain "density" threshold, the nodes are evenly
+distributed across the section to make room for more nodes. To
 ensure that updating pointers isn't too expensive, any node included in the
-section must have its subtrees included. Specifically, the section is built up
-from the the lowest subtree level that contains or contained the update point,
-and neighboring subtrees are continuously added until either the density
-threshold is met or the entire current subtree layout at this "level of
-detail" is included. For example, suppose a node is inserted at the leaf of
-some lower recursive subtree $L_{i}$. To continue building up this section,
-$L_{i-1}$ or $L_{i+1}$ could be added. If all the lower recursive subtrees
-$L_{1}$,$L_{2}$,...,$L_{x}$ have been added, the top level tree $U$ is added.
-This corresponds to a subtree $L'_{j}$ one level above. Thus repeat this
-process until the density threshold is met, and then rebalance the tree. This
-approach depends on the fact that all the recursive subtrees at the same level
-of detail asymptotically have the same size, which Bender et al. prove about
-the hyperhyperfloor-based vEB layout described previously. This approach
+section must have all of its subtrees included.
+The subtrees are dictated by the vEB layout, and subtrees are chosen from the bottom up.
+The smallest neighboring trees (single nodes) in the vEB layout are added first, and larger trees are included when all of their subtrees have been added to the section.
+Root trees are only added after all bottom trees are added (in terms of the vEB layout).
+Figure \ref{fig:centroid_relayout} gives an example of how subtrees are chosen for rearranging.
+
+This approach
 results in an amortized update cost of $O(1+(\log{}N)/B)$ memory transfers per
@@
 $O(1+\log^{2}N/B)$ memory transfers.
-
-Note that making the COSB-trees dynamic causes updates to successor and
-predecessor pointers in the centroid tree to be expensive. Thus the notion of
-ancestors is relaxed to the following (Lemma 10 in \cite{BenderFaKu06}): Let
-$\alpha$ be the parent node of $\gamma$ in the compressed trie. Then either
-$\alpha$ is a descendant of $\gamma$ or $\gamma$ is a descendant of $\alpha$.
-The proof is as follows: As you recursively traverse the trie, both $\alpha$
-and $\gamma$ are in the up trie or down trie, until at some point one of them
-must be the centroid vertex. This is used to redefine what the predecessor
-successor pointers refer to. The centroid may no loner point to the correct
-predecessor or successor with respect to the relevant subtrie, but one of the
-matched ancestors by construction must point to the correct predecessor or
-successor. Thus for the dynamic case, when dealing with successors and
-predecessors, $O(logN)$ ancestors' successor and predecessor pointers in
-addition to the successor and predecessor pointers of the node in question
-must be checked. However, this is an additive factor, so the same time bound
-for update described above is still achieved.
-
+\begin{figure}
+
+\begin{center}
+	\includegraphics[width=0.8\columnwidth]{figures/centroidrelayoutexample.pdf}
+\end{center}
+
+\caption{A centroid layout example.
+Assume the smallest subtrees in this example only contain a single node.
+Suppose a node was inserted into a leaf position in the bottom left of the tree.
+The numbers represent the order in which each subtree is added to the section to be rearranged.}
+\label{fig:centroid_relayout}
+\end{figure}
 \subsubsection{Modifying Front Compression to Preserve Locality}
@@
 This section provides a brief summary of how front compression is modified to
-optimize compression and decoding for static and dynamic cache-oblivious
-string B-trees. Consider a sequence of $j$ keys to store $k_{1}$, $k_{2}$,
+optimize compression and decoding for dynamic COSB-trees. Consider a sequence of $j$ keys to store $k_{1}$, $k_{2}$,
 ..., $k_{j}$. Without compression, this requires $\sum_{j}||k_{j}||$ space in
-memory. To compress these keys using front compression, let $\pi_{j}$ be the
+memory. To compress these keys using standard front compression, let $\pi_{j}$ be the
 longest common prefix between $k_{j}$ and $k_{j+1}$. Now instead we store the
@@
 
-The goal of modifying front compression is to maintain some of the space
-benefits to compression while reducing the high decoding cost and keeping the
-size of D small. Bender et al. provide a modification for static COSB-trees
-that reduces the high decoding cost of front compression of D to
-$O(1+||k||B)$, with size $O(\llangle{D}\rrangle{})$ after compression. The
-modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be
+Bender et al.
+provide a modification that reduces the high decoding cost of front compressing D to $O(1+||k||B)$, while maintaining a compressed size of $O(\llangle{D}\rrangle{})$\footnote{$\llangle{D}\rrangle{}$ is shorthand for the size of $D$ after front compression} after compression.
+The modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be
 the first key added to the compression stream. For each subsequent key
 $k_{i}$, if $k_{i}$ can be decoded with just the last $c||k_{i}||$ characters,
-add $\pi_{i}$,$\sigma_{i}$ like before. Otherwise, add 0,$k_{i}$, which means
-add $k_{i}$ to the stream in its uncompressed form. By construction, this
-scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory
-transfers.%todo: add something about why this isO(<<D>>)?
+add $\pi_{i}$,$\sigma_{i}$ as described above.
+Otherwise, add $k_{i}$ to the stream without compressing it.
+By construction, this scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.
 
@@
 deletions. The previous modification relies on each key being decoded within
-$c||k^{*}||$ elements of the compressed array, but this is no longer
+$c||k||$ elements of the compressed array, but this is no longer
 guaranteed when new keys can be inserted to the left of $k$. To make the
 locality-preserving front compression technique suitable for dynamic
-COSB-trees, Bender et al. modify the algorithm to include an extra copied
-prefix for each key. Consider the $3c||k||$ characters that would be to the
-left of a new key$k^{*}$ after insertion. Let $k'$ be the key at the end of
-this range. The idea is to prevent the insertion of $k^{*}$ from affecting any
-non-copied keys just to the right of $k'$, since this could potentially
-increase their decode cost. Compute the length $l$ of the largest common
-prefix (lcp) between $k^{*}$ and $k'$, and compare $l$ to all the other lcp's
-in this range. $l$ represents the minimum length for any lcp between $k^{*}$
-and $k'$, since the keys are stored in lexicographic order. If there is a
-copied prefix of length at least $l$ or a copied key within this range, insert
-$k^{*}$ as usual. Inserting $k^{*}$ will have no negative effect in this
-respect, since the decode length is effectively reset by the copied key or
-prefix like before. However, if inserting $k^{*}$ causes the minimum length of
-the lcp's in this range to increase, we need to fix the compression. To do
-this, we update the lcp of the key directly to the right of $k^{*}$ and store
-an extra copy of this lcp with $k^{*}$ itself. In this way, we can charge the
-work of copying extra prefixes for a new key to the appropriate number of
-elements before it. This number is determined by the length of the extra
-copied prefix stored with the key.
+COSB-trees, Bender et al. modify the algorithm to include extra copied
+prefixes.
+Suppose a key $k$ is being inserted.
+If keys further down in the array will be negatively impacted, the longest common prefix between $k$ and the key right after $k$ is copied to $k$ along with $k$'s original compression data.
+The idea behind adding extra copied prefixes is to maintain $O(||k||)$ decode distances around the area where the new key is being inserted.
+The characters preceding $k$ are charged for the copied prefix, accounting for the work to add the copied prefix to $k$.
 
@@
 
-The operations supported are dictated by the Dictionary Matching problem.
-These operations are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{},
+The operations supported by the COSB-tree are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{},
 and \Range{}. The behavior of each operation for dynamic COSB-trees is
@@
 
-\Range{}: The paper doesn't say, but presumably this is very similar to
+\Range{}: This operation is not explicitly described in \cite{BenderFaKu06}, but presumably this is very similar to
 \Search{}. Let $k_{b}$ be the beginning of the range and let $k_{e}$ be the
@@
 \subsection{Complexity}
+The keys in are hashed and hash values are compared in place of the original key values to speed up comparisons in the centroid tree and hashdata layer.
+To avoid mismatches due to overlapping hash values, the original key values are compared only once at the bottom layer, resulting an additive $O(||k||/B)$ memory transfers in each of the operators for each key $k$ that needs to be compared.
 
 \Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high
-probability (w.h.p.). Traversing the centroid tree requires $O(1+\log_{B}N)$
-memory transfers, since the tree is built using$O(N/logN)$ keys and is
-balanced. Once in the hashdata layer, we are already within $O(logN)$ keys of
-$k$. Thus the local scan requires $O(1+(||k||+\log{}N)/B)$ memory transfers.
-Obtaining $k$ from the keydata layer requires no more memory transfers than
-the scan in the hashdata layer. The \Pred{} and \Succ{} operations are almost
-identical in behavior to searches, and thus only have the added cost of having
-to operate on $k'$, where $k'$ is the predecessor or successor, respectively.
-These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory
-transfers w.h.p.. Similarly, range queries take
+probability (w.h.p.).
+The centroid tree is stored in a PMA with a dynamic CO layout.
+This layout reduces the number of memory transfers required when traversing the tree to $O(log_{B}N)$.
+Once in the hashdata layer, we are already within $O(logN)$ keys of $k$, which require $O(\log{}N)/B)$ memory transfers to scan.
+Decoding and comparing the original key values requires $O(1+||k||/B)$ memory transfers, due to the modified front compression scheme.
+
+The \Pred{} and \Succ{} operations are similar in behavior to searches, and thus only have the added cost of having to operate on $k'$, where $k'$ is the predecessor or successor, respectively.
+These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers w.h.p..
+Similarly, range queries take
 $O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory transfers,

### 2011-12-12 paper.tex
edits + latex fixes

--- 2011-12-11 paper.tex
@@
 
-\documentclass{style}
+% preprint + pagenumbering for page numbers
+\documentclass[preprint]{style}
+\pagenumbering{arabic}
 
@@
 \usepackage{graphicx}
+
+% This gets rid of the block of whitespace on the first page
+\makeatletter
+\let\@copyrightspace\relax
+\makeatother
 
@@
 This paper presents a survey of cache-oblivious B-trees (CO B-trees). The CO
-B-tree was first introduced by Bender et al. in 2000. It is an external memory
-data structure - meaning that it optimizes the number of memory transfers it
-performs rather than the number of CPU instructions that it executes. Unlike
-other external memory data structures, like the widely used B-tree, the CO
-B-tree does not require any information about the number of level in the
-machine's memory hierarchy, the block size and capacity at each level, or the
-relative speeds at each level. Hence, the CO B-tree is ``cache-oblivious''.
-Despite this, the CO B-tree is able to achieve the same optimal search bound
-as the B-tree and near optimal insertion and deletion bounds as well. In this
-paper we first describe the original CO B-tree and explain how it is able to
-achieve its memory performance. We then present and analyze three followup CO
-B-tree data structures that improve upon weaknesses of the original: the
-cache-oblivious string B-tree, which introduces support for variable length
-keys, the cache-oblivious streaming B-tree, which incorporates improvements to
-the insertion bound at the expense of the search bound, and lastly, the
-cache-oblivious concurrent B-tree, which introduces support for concurrency.
+B-tree was first introduced by Bender et al \cite{BenderDemainColton}. It is
+an external memory data structure; it optimizes the number of memory transfers
+performed rather than the number of CPU instructions that executed. Unlike
+other external memory data structures such as the widely used B-tree, the CO
+B-tree does not have access to the parameter's of the machine's memory
+hierarchy, including the block size/capacity at each level, or the relative
+speeds. Hence, the CO B-tree is \textit{cache-oblivious}.  Despite this, the CO
+B-tree is able to achieve the same provably optimal search bound as the B-tree
+and near optimal insertion and deletion bounds. In this paper we first describe
+the original CO B-tree and explain how it is able to achieve its performance.
+We then present and analyze three followup CO B-tree data structures that
+improve upon weaknesses of the original: the \textit{cache-oblivious string
+B-tree}, which introduces support for variable length keys, the
+\textit{cache-oblivious streaming B-tree}, which improves the insertion bound
+at the expense of the search bound, and lastly, the \textit{cache-oblivious
+concurrent B-tree}, which introduces support for concurrency.
 
@@
 
-Most algorithms assume an idealized model of computation that consists of a
-single CPU and memory as a flat array
-that can be read and written. However, in reality memory is not a flat array
-but rather a hierarchy from super fast on chip caches, to slower main memory,
-to even slower disks. When a program requests data, the data is found in the
-memory hierarchy and typically carried up the hierarchy to the fastest memory
-layer possible, usually a cache. The CPU is then able to access it to perform
-any necessary operations. Unfortunately, this cache often has a small capacity
-and once it is full any new data requests cause some existing cached data to
-be evicted. Transferring data from lower to higher levels in the memory
-hierarchy often takes a long time. In fact, access times between the different
-layers of the hierarchy can differ by many orders of magnitude and thus the
-time to transfer data often dominates the overall runtime performance of an
-algorithm. External memory algorithms are wary of this fact and aim to
-optimizing the number of memory transfers it performs rather than the number
-of instructions it executes. They often try to exploit the fact the memory
-hierarchies transfer data in blocks rather than individual atomic units. Thus,
-external memory algorithms are designed to not only limit the number of memory
-transfers it performs but also preserve the locality of related data so that
-when a memory access is performed, its cost is amortized.
+Most algorithms assume an idealized model of computation that consists of
+memory as a flat byte-addressable array which can be read and modified in some
+fixed time. However, in practice memory is not a flat array, but rather is
+organized as a hierarchy consisting of fast on-chip caches (L1, L2, etc.), to
+slower DRAM, to even slower disks. When a program requests data, the data is
+located in the memory hierarchy and typically carried up the chain to the
+fastest memory layer (closest to the CPU), usually a cache. Unfortunately, this
+cache can only have a limited capacity, and once it is full any new data
+requests necessitate some existing cached data to be evicted. Transferring data
+from lower to higher levels in the memory hierarchy often takes an
+exponentially long time; access times between the different layers can differ
+by many orders of magnitude. Therefore, the time to transfer data often
+dominates the overall runtime performance of an algorithm.
+
+External memory algorithms are wary of this fact and seek to optimize the
+number of memory transfers performed rather than the number of CPU instructions
+executed. They often try to exploit the fact the memory hierarchies transfer
+data in blocks rather than individual machine word units. Thus, external memory
+algorithms are designed to not only limit the number of memory transfers, but
+also preserve the locality of related data so that when a costly memory access
+is performed, its cost is amortized.
 
@@
 it is no surprise that many external memory problems can be reduced to finding
-a suitable data structure that efficiently implement certain operations. A
-well known external memory data structure is the B-tree. At a high level, the
-B-tree works by making each one of its nodes the same size as a memory block
-in a particular level of the hierarchy. Furthermore, the fanout in the tree
-(the number of children of a node) is proportional to the block size as well.
-The B-tree incurs $O(log(N))$ memory transfers for \Search{}, \Insert{}, and
-\Delete{} and is widely used in databases and filesystems.
+a suitable data structure that efficiently implement certain operations. A well
+known external memory data structure is the B-tree \cite{BayerM72}. The B-tree
+works by making each one of its nodes the same size as a memory block in a
+particular level of the hierarchy. Furthermore, the fanout in the tree (the
+number of children of a node) is proportional to the block size as well.  The
+B-tree incurs $O(\log_{B}{N})$\footnote{Throughout this paper, $\log$ with the
+base omitted is base-$2$} memory transfers for \Search{}, \Insert{}, and
+\Delete{} and is widely used in databases and file-systems.
 
@@
 
-The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in
-a two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}.
-However, in reality, computing architectures typically have memory hierarchies
-that are much larger than two. The argument for the DAM model is that in
-practice only one level in the memory hierarchy tends to dominate the
-performance and thus a two-level memory model is sufficient for all intents
-and purposes. However, this does mean that, theoretically, B-trees can begin
-to incur an increasing number of memory transfers if the dominating level in
-the hierarchy changes.
-
-Additionally, the B-tree expects to know the block size of its memory layer in
-advance. B-Trees are implemented with parameterized code that needs to be
-changed for every level in the memory hierarchy. Needless to say, this tuning
+The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in a
+two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}.
+However, in practice, architectures typically have memory hierarchies that are
+larger than two. The argument for the DAM model is that in practice, only one
+level in the memory hierarchy tends to dominate the performance, and thus a
+two-level memory model is sufficient.  However, this means that B-trees can
+begin to incur an increasing number of memory transfers if the dominating level
+in the hierarchy changes.
+
+Additionally, the organization of a B-tree requires knowing the block size of
+the memory layer in advance. B-trees are implemented with parameterized code
+that needs to be changed for every level in the memory hierarchy. This tuning
 can be very difficult and error prone. Additionally, tuning this parameter may
@@
 The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
-of computation. In the CO model an algorithm has no information about the
-memory hierarchy. It does not know anything about the speed or the block size
-of any level. Conceptually, the CO model allows algorithm designers to reason
-about the simple two-level model like DAM but also prove results for any
-arbitrary multi-level memory hierarchy. This is because algorithms that
+of computation \cite{Frigo99}. In the CO model an algorithm has no information
+about the memory hierarchy. It does not know anything about the speed or the
+block size of any level. Conceptually, the CO model allows algorithm designers
+to reason about the simple two-level model like DAM but also prove results for
+any arbitrary multi-level memory hierarchy. This is because algorithms that
 perform well in a CO model should, in theory, perform well regardless of
@@
 
-In \cite{BenderDemainColton} Bender et al. introduce the cache-oblivious
+In \cite{BenderDemainColton}, Bender et al. introduce the cache-oblivious
 B-tree (CO B-tree). The CO B-tree is similar to the normal B-tree except that
-it does not require explicit parameters about the memory hierarch.
-
-The CO B-tree is able to achieve the optimal search bound of $\Theta(log(N))$
-memory transfers, the optimal scan bound of $\Theta(N/B)$ memory transfers and
-near optimal insertion and deletion bound of $\Theta(log_B N +
-\frac{log^2{N}}{B})$. The CO B-tree was a seminal data structure that gave
-rise to many others.
-
-This paper provides a survey of the different types of CO B-trees and
-discusses the various trade offs that they make. We start by analyzing the
-original CO B-tree and explain how it works. This includes a detailed
-description of the packed memory array (PMA) and strongly weight-balance
-binary search trees (SWSBT) that are integral to the CO B-tree's memory
-performance. Following this discussion we provide an analysis of the CO String
-B-tree,a CO B-tree with variable length keys, the CO Streaming B-tree, a CO
-B-tree that improves the cost of insertion at the expense of search, and the
-CO Concurrent B-tree, which allows multiple processes to access the data at
-the same time.
+it does not require explicit parameters about the memory hierarchy.
+
+The CO B-tree is able to achieve the optimal search bound of
+$\Theta(\log_{B}{N})$ memory transfers, the optimal scan bound of $\Theta(N/B)$
+memory transfers and near optimal insertion and deletion bound of 
+$\Theta(\log_{B}{N} + \frac{\log^2{N}}{B})$. 
+The CO B-tree was a seminal data structure that gave rise to many others.
+
+This paper provides a survey of the different types of CO B-trees and discusses
+the various trade offs that they make. We start by analyzing the original CO
+B-tree and explain how it works. This includes a detailed description of the
+packed memory array (PMA) and strongly weight-balance binary search trees
+(SWBST) that are integral to the CO B-tree's memory performance. Following this
+discussion we provide an analysis of the CO string B-tree (a CO B-tree with
+variable length keys), the CO streaming B-tree (a CO B-tree that improves the
+cost of insertion at the expense of search), and the CO concurrent B-tree (a CO
+B-tree that allows multiple processes to efficiently access data concurrently).
 
@@
 
-Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
-\cite{BenderDemainColton}. The CO B-tree achieves its number of memory
-transfers by structuring its elements as a strongly weight-balanced search
-tree (SWBST). The SWBST is then laid out into a packed memory array (PMA)
-using a van Emde Boas (vEB) layout. The SWBST and PMA play crucial roles in
-the other cache-oblivious data structures that were developed after the CO
-B-tree and thus are described at length in the next two sections.
+The CO B-tree optimizes the number of memory transfers by structuring its
+elements as a strongly weight-balanced search tree (SWBST). The SWBST is then
+laid out into a packed memory array (PMA) using a van Emde Boas (vEB) layout.
+The SWBST and PMA play crucial roles in the other cache-oblivious data
+structures that were developed after the CO B-tree and thus are described at
+length in the next two sections.
 
@@
 The ultimate goal of a CO B-tree is to reduce the number of memory transfers
-while supporting \Search{}, \Insert{}, and \Delete{} operations. Thus, at a
-fundamental level the CO B-tree still needs to maintain its data in some
-structure. That structure is a balanced search tree. The search tree is laid
-out according to a van Emde Boas (vEB) layout \cite{veb1}, \cite{veb2}. The
-idea of a vEB layout is fairly straight forward. Let $h$ denote the height of
-a tree. the vEB layout splits the tree approximately at height $h/2$. This
-splits the original tree into a ``top'' subtree and many ``bottom'' subtrees.
-Let $A$ denote the top subtree and $B_1... B_k$ denote the bottom subtrees.
-The vEB layout is defined recursively using the order: $A, B1, ... B_k$. This
-layout is also shown in Figure~\ref{fig:veb}.
-
-A nice consequence of using a vEB layout is that it is known that it will use
-$O(log_B N)$ memory transfers for searches \cite{veb1}, \cite{veb2}. Thus, the
-CO B-tree has already achieved the desired number of memory transfers for
-search queries. It still remains to be shown that the CO B-tree achieves the
-desired number of memory transfers for dynamic inserts and deletes.
-
-It is known that the CO B-tree is going to use a balance search tree, but what
-kind? This is where the strongly weight-balanced search tree becomes useful.
-The weight of a node is $u$ is equal to the total number of $u$'s descendants
-plus 1. It can also be thought of as the sum of the weights of $u$'s children
-plus 1. Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.
-
-Most weight-balanced search trees require that the heights of a node's left
-and right subtrees are simply off by a constant factor. The SWBST enforces the
-stronger property that a node at height $h$ in the tree (height is the node's
-distance from the leaves; leaves have a height of 1) has $\Theta c^h$
-descendants where $c$ is some constant. This property becomes important later
-on when laying the tree out in memory (see Section~\ref{sec:layout}).
-
-A known data structure that achieves this property is the ``weight-balanced
-B-tree'' \cite{swbst}. The CO B-tree leverages this data structure to maintain
-its elements. Before SWBST can be used, however, it remains to be proven that
-it maintains the the $c^h$ descendants property even after inserts and
-deletes.
-
-The definition of a SWBST given in \cite{swbst} provides us with this bound
-for the weight of a node:
-
+while supporting \Search{}, \Insert{}, and \Delete{} operations. Thus, it is
+not surprising that the CO B-tree maintains its data in a balanced search tree.
+The search tree is laid out in memory according to a van Emde Boas (vEB) layout
+\cite{veb1,veb2}. The idea of a vEB layout is fairly straight forward. Let $h$
+denote the height of a tree. The vEB layout splits the tree approximately at
+height $h/2$. This splits the original tree into a \textit{top} subtree and
+many \textit{bottom} subtrees.  Let $A$ denote the top subtree and
+$B_1,...,B_k$ denote the bottom subtrees.  The vEB layout is defined
+recursively using the order: $A,B_1,...,B_k$. This layout is also shown in
+Figure~\ref{fig:veb}.
+
+A nice consequence of using a vEB layout is that it can be shown to use
+$O(\log_{B}{N})$ memory transfers for searches \cite{veb1, veb2}. Thus, the CO
+B-tree has already achieved the desired bound for search queries. It still
+remains to be shown that the CO B-tree achieves the desired number of memory
+transfers for dynamic inserts and deletes.
+
+To support dynamic inserts/deletes, a \textit{strongly weight-balanced search
+tree} (SWBST) is used \cite{swbst}.  The weight of a node $u$ is equal to the
+total number of $u$'s descendants plus 1. It can also be thought of as the sum
+of the weights of $u$'s children plus 1. Formally, $u$'s weight is $w(u) = 1 +
+\sum_{v \in children} w(v)$.
+
+Most weight-balanced search trees require that the heights of a node's left and
+right subtrees differ by at most a constant factor. The SWBST enforces the
+stronger property that a node at height $h$ in the tree (leaves have a height
+of 1) has $\Theta(c^h)$ descendants where $c$ is some constant ($c > 4$). This
+property becomes important later on when laying the tree out in memory (see
+Section~\ref{sec:layout}).
+
+The definition of a SWBST given in \cite{swbst} provides this bound for the
+weight of a node $u$:
 \begin{equation}
@@
 
-Let us now consider insertions. To insert an element into a WSBST we search
+Let us now consider insertions. To insert an element, search
 down the tree and find the natural place to insert the element - just like in
-any tree structure. After inserting the element into some node $w$, we have to
+any tree structure. After inserting the element into some node $w$, 
 check to see if any of $w$'s ancestors have become unbalanced by having a
-weight greater than $2^{d-1}$. If such an ancestor exists, it follows that
-every node along the path between that ancestor and $w$ will be un balanced as
-well. Thus, we start with $w$ and move up the tree.
-
-When we split a node, we naturally need to determine which children belong to
-which of the two nodes. We cannot just divide the children evenly because it
-may cause another violation. Thus, instead, we find the longest sequence of
-elements such that the weight of that subsequence is $\lceil w(u)/2 \rceil$.
-By doing this, we maintain the bound that $\frac{c^{h-1}}{2} \leq w(u) \leq
-2c^{h-1}$. We cascade these changes up the tree until all of the ancestors are
-balanced.
-
-Deletions are supported in the exact same way, except, we check if a node has
-a weight $w(u) < \frac{c^{h-1}}{2}$. In this case, we will merge two neighbors
-together and proceed up the tree. There is one edge case that needs to be
-handled. After merging, you may produce a new node that has a weight larger
-than the bounds. If this happens we immediately split the node again. This can
+weight greater than $2c^{h-1}$. If such an ancestor exists, it follows that
+every node along the path between that ancestor and $w$ will be unbalanced as
+well. This requires splitting nodes to restore balance, starting from $w$
+and moving up the tree.
+
+When a node is split, we need to determine which children belong to which of
+the two nodes. Dividing the children evenly is insufficient because it may
+cause another violation. Instead, find the longest sequence of elements such
+that the weight of the subsequence is $\lceil w(u)/2 \rceil$.  By doing this,
+we maintain the bound that $\frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}$. Cascade
+these changes up the tree until all of the ancestors are balanced.
+
+Deletions are supported similarly, except we check if a node has a weight $w(u)
+< \frac{c^{h-1}}{2}$. In this case, we merge two neighbors together and proceed
+up the tree. There is one edge case that needs to be handled. The merging
+procedure may produce a new node that has a weight larger than the bounds
+allow. If this happens, immediately split the node again, recursively. This can
 be thought of as stealing nodes from a neighbor.
 
-We have now shown that the SWBST achieves the desired property that the number
-of descendants of a node is $c^h$. We can now show how to lay this tree out in
-memory.
+We have now sketched that the SWBST has the desired property that the number of
+descendants of a node is $\Theta(c^h)$. We now describe the tree's memory layout.
 
@@
 
-Now that we have our elements structured in a SWBST, we are still left with
-mapping this tree to memory and proving that it achieves our desired number of
-memory transfers. The packed memory array (PMA) was developed to solve this
-problem \cite{BenderDemainColton}. There are three main costs that the PMA is
-concerned with.
+Now that the elements structured in a SWBST, we are still left with the problem
+of mapping this tree to memory and proving that it achieves the desired number
+of memory transfers per operation. The packed memory array (PMA) was developed
+to solve this problem \cite{BenderDemainColton}. There are three main costs
+that the PMA is concerned with: \textit{inserts}, \textit{splits/merges}, and
+\textit{pointer updates}.
 
 First, the PMA needs to maintain open space in its memory for newly inserted
-items. The PMA is able to accomplish this using $O(1 + (log^2 N) / B)$ memory
-transfers. Doing so may seem like attempting to achieve contradictory goals:
-maintaining the locality of elements while keeping space available for new
-inserts. The PMA uses a technique described in \cite{packedmemoryarray}. At a
-high level, the PMA maintains windows into the array. When a window becomes
-too full or too empty, the elements are spread out within the context of a
-larger window. Two parameters are required, the window size and the threshold
-for trigger the spreading out. However, these parameters are not related to
-the block size of the memory level and thus do not violate the CO model of
-computation. Overall, the PMA is able to keep extra space in $O(1 + (log^2 N)
-/ B)$ memory transfers.
-
-Second, the PMA needs to handle the splitting and merging caused by the SWBST
-discussed in the previous section. The PMA is able to accomplish this using
-$O(1 + (log N) /B)$ memory transfers. The proof is based on the fact that each
-node has $c^h$ descendants (recall, this is the key property that SWBST
-provided). Thus, merging or splitting a node will cause $c^h$ nodes to move,
-requiring $O(1 + c^h/B)$ memory transfers. It follows from the definition of
-SWBST in \cite{swbst} that when a node is rebalanced, it will require $c^h$
-insertions or deletions until it is rebalanced again. Thus, the amortized cost
-of each insertion and/or deletion is $O(1/B)$. When you split a node, it may
-go all the way up the tree and thus overall it will incur $O((log N) / B)$
-memory transfers.
-
-Third, the PMA needs to handle updating the parent and child pointers and may
-become invalidated because of how the nodes moved around. The PMA is able to
-update the parent and child pointers. If the pointers are all close together
-in memory (e.g. within $B$ distance of each other) it is not big deal because
-only a constant number of memory transfers are required. However, if they are
-further apart the PMA introduces buffer nodes. The buffer nodes don't actually
-do anything. All they do is serve to ``stretch'' the vEB layout so that nodes
-do not move far away from one another. By using these buffers the PMA is able
-to achieve a the pointer update cost using $O(1+ logB / \sqrt{B}) log^2 N)$
-memory transfers.
+items. The PMA is able to handle an insertion into the SWBST using only $O(1 +
+(\log^2{N}) / B)$ memory transfers (this is the cost of updating the memory
+layout of the SWBST in the PMA, not the cost of updating the data structure
+itself). Doing so may seem contradictory: maintaining the locality of elements
+while keeping space available for new inserts. The PMA uses a technique
+described in \cite{packedmemoryarray}. At a high level, the PMA maintains
+windows into the array. When a window becomes too full or too empty, the
+elements are spread out within the context of a larger window. Two parameters
+are required: the window size and the threshold for triggering a spread out.
+However, these parameters are not related to the block size of the memory level
+and thus do not violate the CO model.
+
+Second, the PMA needs to handle splits and merges caused by inserting into the
+SWBST discussed in the previous section. The PMA is able to accomplish this
+using $O(1 + (\log{N}) /B)$ memory transfers. The proof is based on the fact
+that each node has $\Theta(c^h)$ descendants (recall, this is the key property
+that SWBST provided). Thus, merging or splitting a node will cause
+$\Theta(c^h)$ nodes to move, requiring $O(1 + c^h/B)$ memory transfers. It
+follows from the definition of SWBST in \cite{swbst} that when a node is
+rebalanced, it will require $\Theta(c^h)$ insertions or deletions until it is
+rebalanced again. Thus, the amortized cost of each insertion and/or deletion is
+$O(1/B)$. When a node is split, it may traverse all the way up the tree, and
+thus the split/merge bound is $O((\log N) / B)$ memory transfers.
+
+Third, the PMA needs to handle updating the parent and child pointers that may
+become invalidated because of how the nodes moved around. If the pointers are
+all packed together in memory (e.g. within $B$ distance of each other) then
+only $O(1)$ memory transfers are required. However, if they are further apart
+then a problem arises. To deal with this, buffer nodes are introduced.  Buffer
+nodes ``stretch'' the vEB layout so that nodes do not move far away from one
+another. By using buffers, the PMA is able to achieve a pointer update cost
+using $O(1+ \frac{\log{B}}{\sqrt{B}}\log^2{N})$ memory transfers.
 
 Of these three terms, the dominating factor is maintaining room for newly
-inserted elements - $O((log^2N)/B)$. We know that the SWBST achieves
-insertions and deletions in $O(log_B N)$ if it is cache-aware. Thus, adding
-PMA makes the SWBST cache-oblivious at the cost of incurring an additional
-$O((log^2N)/B)$ memory transfers.
-
-Therefore, overall, the CO B-Tree achieves insertions and deletions in
-$O(log_B N + (log^2N)/B)$ memory transfers. However, it still achieves the
-optimal $O(log_B N)$ memory transfers for searches, matching the performance
-of ordinary, cache-aware, B-trees.
+inserted elements - $O((\log^2{N})/B)$. Therefore, overall the CO B-tree
+achieves insertions and deletions in $O(\log_B{N} + (\log^2{N})/B)$ memory
+transfers. It still achieves the optimal $O(\log_B{N})$ memory transfers for
+searches, matching the performance of static cache-aware B-trees.
 
@@
 While B-trees achieve an optimal bound for the number of memory transfers for
-search, they do not perform as well for updates and insertions. The
-buffered-repository tree (BRT) is a common alternative to the B-tree that
-achieves an improved amortized $O(log(N)/B)$ memory transfers for inserts
-compared to the $O(log_{B+1}N)$ memory access of B-trees. Of course, the
-B-tree and BRT are analyzed with the DMA model and are not cache-oblivious.
-
-This search vs insertion trade-off also exists in the cache-oblivious model.
-The original BO B-tree described in Section~\ref{sec:original} is designed to
-optimize for the search in this trade off. However, Bender et al. propose two
-data structures that and achieve the opposite end: the shuttle tree and the
+search, they do not perform as well for bulk updates and insertions. The
+buffered-repository tree (BRT) is an alternative to the B-tree that achieves an
+improved amortized $O(\log{\frac{N}{B}})$ memory transfers for inserts compared
+to the $O(\log_{B}{N})$ memory access of B-trees. In practice, log-structured
+merge (LSM) trees \cite{ONeil96} are used for high-insert workloads.
+
+This search vs. insertion trade-off also exists in the cache-oblivious model.
+The original CO B-tree described in Section~\ref{sec:original} is designed to
+optimize for search in this trade off. However, Bender et al. propose two data
+structures that and achieve the opposite end: the shuttle tree and the
 cache-oblivious lookahead array (COLA) \cite{BenderFaFi07}. These two data
-structures are refereed to as ``streaming B-trees'', indicating how they are
-optimized for insertions at the cost of search.
+structures are refereed to as \textit{streaming B-trees}, indicating how they
+are optimized for bulk insertions at the cost of search.
 
@@
 
-The shuttle tree is similar in layout and structure as the original CO B-Tree.
+The shuttle tree is similar in layout and structure as the original CO B-tree.
 It uses a strongly weight-balanced search tree described previously (see
 Section~\ref{sec:structure}) that is laid out in a vEB (see
-Section~\ref{sec:veb}) layout into a PMA (see Section~\ref{sec:layout}). A
+Figure~\ref{fig:veb}) layout into a PMA (see Section~\ref{sec:layout}). A
 notable difference between the shuttle tree and the CO B-tree is that the CO
-B-tree splits the tree at height $h/2$ when creating the vEB layout while the
-shuttle tree splits at height $F_k$, where $F_k$ is the $kth$ Fibonacci number
-that is stickily less than the height of the tree. Thus if a tree's height
-happens to be a Fibonacci number, $F_k$, then the shuttle tree will choose to
+B-tree splits the tree at height $h/2$ when creating the vEB layout, while the
+shuttle tree splits at height $F_k$, where $F_k$ is the $k$-th Fibonacci number
+that is strictly less than the height of the tree. Thus if a tree's height
+happens to be a Fibonacci number $F_k$, then the shuttle tree will choose to
 split the tree at height $F_{k-1}$. The top and bottom ``halves'' produced by
@@
 At a high level, the shuttle tree improves the efficiency of insertion by
-adding buffers to each node in the tree. In fact, shuttle tree nodes have a
+adding buffers to each node in the tree. Shuttle tree nodes have a
 linked list of buffers rather than just a single buffer like the BRT. Each of
-these buffers is another shuttle tree. The a shuttle tree node enforces two
+these buffers is another shuttle tree. A shuttle tree node enforces two
 constraints: the length of its linked list as well as the overall height (and
 thus capacity) of each of the buffered shuttle trees contained within its
-linked list.
-
-The length of a node's linked list is related to the node's height in the
-tree. A node has a buffer for every time it the leaf of a sub trees that is
-split when creating the vEB layout. This idea is illustrated in Figure
+linked list are bounded, described below.
+
+The length of a node's linked list is related to the node's height in the tree.
+A node has a buffer for every time it is the leaf of a sub-tree that is split
+when creating the vEB layout. This idea is illustrated in Figure
 \ref{fig:buffers}.
 
-The shuttle tree's node's buffers have a maximum capacity that is enforced by
+The shuttle tree's node buffers have a maximum capacity that is enforced by
 bounding its height. The size of the buffers increases going across the linked
-list. Thus, a node's first buffer is smaller than its last buffer. The maximum
-height of a buffer is given by $F_{H(k)} = F_{k-2 \lceil log_{\phi} k \rceil}
-$ where $\phi$ is the golden ratio and $k$ is chosen such that $F_k$ =
-$\xi(h)$ and $\xi(x)$ is defined recursively. If $h$ is a Fibonacci number
-then $\xi(h) = h$ and otherwise $\xi(h) = \xi(h-f)$ where f is the largest
-Fibonacci number less than $h$. The maximum height of an individual buffer is
-related to the size of the recursive subtree for which the node is a leaf.
-Since the recursive subtrees are split based on the largest Fibonacci number
-less than $h$ (as described previously) it follows that the buffer sizes are
-also based on Fibonacci numbers.
+list from left-to-right. Thus, a node's first buffer is smaller than its last
+buffer. The maximum height of a buffer is given by $F_{H(k)} = F_{k-2 \lceil
+\log_{\phi} k \rceil}$, where $\phi$ is the golden ratio and $k$ is chosen such
+that $F_k$ = $\xi(h)$ and $\xi(x)$ is defined recursively: If $h$ is a
+Fibonacci number then $\xi(h) = h$, otherwise $\xi(h) = \xi(h-f)$ where $f$ is
+the largest Fibonacci number less than $h$. The maximum height of an individual
+buffer is related to the size of the recursive subtree for which the node is a
+leaf.  Since the recursive subtrees are split based on the largest Fibonacci
+number less than $h$ (as described previously) it follows that the buffer sizes
+are also a function of the Fibonacci numbers.
 
@@
 
-The operations that are described by Bender et al. for the shuttle tree are
-\Search{}, \Insert{}, and \Scan{}. An explicit discussion on \Delete{} is
-missing; however, it is assumed that it is symmetric to \Insert{}.
-
-\Search{} and \Scan{} work virtually the same way as the CO B-tree. The only
-difference is that you have to search a child node's buffer before following
-the pointer down the tree.
-
-\Insert{} is also similar to the CO B-tree except you buffer elements as you
-add them. When you insert an element you move down the tree from the root to
-leaves and stop when you find a node that has a buffer (note that because of
-the nature of Fibonacci numbers that it is possible that a node does not have
-any buffers). You then attempt to insert the element into the smallest buffer
-in the node's linked list. If this insertion causes the buffer to overflow
-(e.g. the height of the buffer exceeds its maximum), you move all of the
-elements from this buffer to the next largest buffer. When the largest buffer
-overflows you take all of its elements and ``shuttle'' them down the tree.
-Eventually you will hit a leaf node and inserting here may cause the balance
+The shuttle tree supports \Search{}, \Insert{}, and \Scan{}. An explicit
+discussion on \Delete{} is missing; it is assumed that it is similar 
+to \Insert{}.
+
+\Search{} and \Scan{} work mostly the same way as the CO B-tree. The only
+difference is that a node's buffer must be searched before following a
+pointer down the tree.
+
+\Insert{} is also similar to the CO B-tree except elements are buffered as they
+are inserted. \Insert{} starts by move down the tree from the root to leaves
+and stopping when a node that has a buffer is located (note that because of the
+nature of Fibonacci numbers that it is possible that a node does not have any
+buffers). The element is inserted into the smallest buffer in the node's linked
+list. If this insertion causes the buffer to overflow (e.g. the height of the
+buffer exceeds its maximum), all of the elements from the buffer are moved into
+the next largest buffer (and so on, recursively).  When the largest buffer
+overflows all of its elements are taken and \textit{shuttled} down the tree.
+Eventually a leaf node will be reached and inserting may cause the balance
 constraints on the SWBST to become invalidated. The shuttle tree rebalances
@@
 split cascades up the tree until balance is restored.
-
-The shuttle tree also needs to take care of the buffers belonging to a split
-node. When a node $u$ is split into two nodes $u_1$ and $u_2$ the shuttle tree
-immediately allocates enough space for all of the $u_2$'s buffers ($u_1$
-simply reuses the space allocated for $u$'s buffers). We then scan over $u$'s
-buffers and find the ones rooted in $u_1$ and $u_2$ and move them
-respectively. This can be achieved by a single scan.
-
-The core problem of maintaining room for inserts, bounding the costs of
+%% TODO: explain what it means to be "shuttled" down the tree
+
+The shuttle tree also needs to take care of the buffers when a node is split.
+When a node $u$ is split into two nodes $u_1$ and $u_2$, the shuttle tree
+immediately allocates enough space for all of the $u_2$'s buffers ($u_1$ simply
+reuses the space allocated for $u$'s buffers). $u$'s buffers are then scanned
+over (by a single scan) and the elements are placed in the new buffers
+accordingly.
+
+The problem of maintaining room for inserts, bounding the costs of
 merging and splitting, and bounding the costs of updating parent and child
 pointers in the shuttle tree is achieved using the same PMA used with the CO
-B-tree. This is decried in detail in Section~\ref{sec:layout}.
+B-tree. This is described in detail in Section~\ref{sec:layout}.
 
@@
 
-\Search{} uses $O(log_B N)$ memory transfers where $N = \Theta(c^{F_k})$ and
-$c$ is your fanout. Note that $c$ is not treated as a constant since this is a
-cache-oblivious data structure. This bound is achieved by searching over the
+\Search{} uses $O(\log_{B}{N})$ memory transfers where $N = \Theta(c^{F_k})$ and
+$c$ is the fanout. This bound is achieved by searching over the
 top and bottom recursive subtrees at each level when traversing the tree. A
 key trick is that when you visit a node you only search its largest buffer
-(that last buffer in the linked list for that level). The smaller buffers will
+(the last buffer in the linked list for that level). The smaller buffers will
 be visited eventually as you recurse down the tree. Eventually you will get
-down to a single subtree that can fit into $O(1)$ blocks and thus can be
-searched with $O(1)$ memory accesses. Since the fanout is proportional to our
-block size it will take us $O(log_B N)$ block accesses until we get down to a
-subtree and thus, by induction, the worst case runtime is $O(log_B N)$.
+down to a single subtree that can fit into a single blocks and thus can be
+searched with one memory accesses. Since the fanout is proportional to the
+block size it will take us $O(\log_B N)$ block accesses until we get down to a
+subtree and thus, by induction, the worst case runtime is $O(\log_B N)$.
+
+% Why is fanout propto block-size? doesn't this violate cache-obliviousness
 
@@
 
-\Scan{} uses an amortized $O(L/B)$ memory transfers to scan in $L$ consecutive
-elements. This cost obviously does not include the single \Search{} operation
-that is required to find the first starting element of the range. Once that
-first element's block is transferred, the shuttle tree proceeds by traversing
-the tree and scanning in elements. Since the vEB layout enforces the spatial
-locality of nearby nodes in the tree, the shuttle tree only has to read in a
-new block once every $B$ memory accesses. Thus, each individual memory access
-is is amortized to $1/B$ and the overall number of memory transfers is
+\Scan{} uses an amortized $O(L/B)$ memory transfers to scan $L$ consecutive
+elements. This cost does not include the single \Search{} operation that is
+required to find the first starting element of the range. Once the first
+element's block is transferred, the shuttle tree proceeds by traversing the
+tree and scanning in elements in order. Since the vEB layout enforces the
+spatial locality of nearby nodes in the tree, the shuttle tree only has to read
+in a new block about once every $B$ elements. Thus, each individual memory
+access is is amortized to $1/B$ and the overall number of memory transfers is
 $O(L/B)$.
@@
 The last thing that is worth mentioning is that the overall space requirements
-of the shuttle tree is $O(n)$. The high level intuition for this proof is that
-the majority of the buffers are really small and thus do not impact the
-overall size of the tree.
+of the shuttle tree is $O(N)$, since the majority of the buffers are really
+small and thus do not impact the overall size of the tree.
 
@@
 
-The cache-oblivious lookahead array (COLA) is actually quite different from
-the other cache-oblivious data structures seen so far. Instead of an explicit
-tree structure with child and parent pointers, the COLA is organized as a list
-of arrays or levels. Each array has size $2^k$. Furthermore, each array is
-either completely full or completely empty. As such, the COLA is very similar
-to the binomial list structure described in \cite{BentleySaxe}.
+The cache-oblivious lookahead array (COLA) is quite different from the other
+cache-oblivious data structures discussed so far. Instead of an explicit tree
+structure, the COLA is organized as a list of arrays or levels. Each array has
+size $2^k$, where $k$ is the level of the array ($k = 0$ is the bottom level).
+Furthermore, each array is either completely full or completely empty. As such,
+the COLA is very similar to the binomial list structure described in
+\cite{BentleySaxe}.
 
@@
 The operations supported by COLA are \Search{}, \Insert{}, and \Scan{}. Once
-again, \Delete{} can be inferred symmetrically from \Insert{}.
+again, \Delete{} can be inferred from \Insert{}.
 
@@
 This array has size $2^0 = 1$, thus the element is in an array on it own.
-Whenever you insert an element you check to see if there is another array of
+Whenever an element is inserted, check to see if there is another array of
 the same size (this can be thought of as checking to see if there is another
@@
 array that is twice the size and thus goes into the next level. The COLA
-continues to merge until there are no two arrays of the same size.
-
-\Search{} can naively be performed by simply scanning each of the levels in
-the COLA. Since everything is sorted the COLA can use a binary search to
-quickly find the element it is looking for. However, this is not very
-efficient. As an improvement COLA, as the name suggests, uses look ahead
-pointers to reduce the number of memory accesses required. Every 8th element
-from the $(k+1)$st level exists in the $k$ level array in its proper sorted
-location. This element also contains a pointer to its real location in the
-$(k+1)$st level. Thus this pointer is referred to as ``real lookahead
-pointer''. Additionally, every 4th cell in $k$th level array contains two
-pointers pointing to the next and previous ``real lookahead pointers'' on this
-level. The next and previous pointers are referred to as ``duplicate lookahead
-pointers''. To search a level the COLA performs a binary search within an
-appropriate range, determine the range in the next level using the ``duplicate
-lookahead pointers'' and then recursively cascade up the tree.
+recursively merges until there are no two arrays of the same size.
+
+\Search{} can naively be implemented by scanning each of the levels in the
+COLA. Since everything is sorted the COLA can use a binary search to find the
+element it is looking for. However, this is not very efficient. As an
+improvement, COLA, as the name suggests, uses look ahead pointers to reduce the
+number of memory accesses required. Every 8th element from the $(k+1)$-st level
+exists in the $k$-th level array in its proper sorted location. This element
+also contains a pointer to its real location in the $(k+1)$-st level. Thus this
+pointer is referred to as \textit{real lookahead pointer}. Additionally, every
+4th cell in $k$-th level array contains two pointers pointing to the next and
+previous real lookahead pointers on this level. The next and previous pointers
+are referred to as \textit{duplicate lookahead pointers}. To search a level the
+COLA performs a binary search within an appropriate range, determine the range
+in the next level using the duplicate lookahead pointers and then recursively
+cascades up the tree.
 
@@
 
-\Insert{} achieves an amortized $O(log N / B)$ memory transfers. The proof is
-based on the fact that an element will never be merged more than $O(log(N))$
+\Insert{} achieves an amortized $O(\log N / B)$ memory transfers. The proof is
+based on the fact that an element will never be merged more than $O(\log{N})$
 times. This is because at that point it will be in the last array.
-Additionally, the first $log(B)$ merges that a element participates in uses a
-constant number of memory transfers since we can assume that the relevant
-blocks are already cached. Thus, we only need to worry about memory transfers
-when we merge levels that have size greater than $B$. If the size of the array
-is $k$ then the cost of merging is $O(k/B)$ and thus the amortized cost of a
-merging a single element is simply $1/B$. Since we just proved that an element
-will only be merged $O(log(N))$ time, it follows that the overall amortized
-memory transfers to insert an element is $O((log N)/B)$.
-
-\Search{} achieves $O(log N)$ memory transfers. The proof uses induction to
+Additionally, the first $\log{B}$ merges that a element participates in uses
+$O(1)$ memory transfers since we assume that the relevant blocks are already in
+memory. Thus, we only need to worry about memory transfers when we merge levels
+that have size greater than $B$. If the size of the array is $k$, then the cost
+of merging is $O(k/B)$, and thus the amortized cost of a merging a single
+element is $1/B$. Since an element will only be merged $O(\log{N})$ time, it
+follows that the overall amortized memory transfers to insert an element is
+$O((\log N)/B)$.
+
+\Search{} runs in $O(\log N)$ memory transfers. The proof uses induction to
 show that the COLA never needs to search more than eight elements on each
-level. Since there are $O(log(N))$ levels and since eight is a constant, it
-follows that overall, search uses $O(log N)$ memory transfers. The fact that
+level. Since there are $O(\log N))$ levels and since eight is a constant, it
+follows that overall search uses $O(\log N)$ memory transfers. The fact that
 only eight elements need to be searched per level is shown through induction.
 The base case is trivial. The first three levels of the COLA each have at most
-eight elements (note that these elements are either actual real elements or
-lookahead pointers). The COLA can obviously search these levels with no more
+eight elements (note that these elements are either real elements or
+lookahead pointers). The COLA can search these levels with no more
 than eight member transfers. As the COLA follows lookahead pointers to the
@@
 find a new $i$ such that $e_i \leq e < e_{i+1}$. Once this $i$ is found the
-duplicate lookahead pointers can be used to locate the ``next'' and
-``previous'' real lookahead pointers in the next level. The two real lookahead
-pointers are used to and restrict the search in the next level. Recall that at
-each level in the COLA every eighth element in a level is present in the
+duplicate lookahead pointers can be used to locate the \textit{next} and
+\textit{previous} real lookahead pointers in the next level. The two real lookahead
+pointers are used to restrict the search in the next level. Recall that at
+each level in the COLA, every eighth element in a level is present in the
 previous level. Thus, it follows that in the next level only 8 elements need
@@
 level, it follows that only 8 elements need to be searched on this level to
-find the appropriate $i$. Thus, overall, the COLA uses $O(8log(N)) =
-O(log(N))$ memory transfers.
+find the appropriate $i$. Thus, overall, the COLA uses $O(8\log{N}) =
+O(\log{N})$ memory transfers.
 
@@
 In experimental results, the COLA achieved significant improvements over
-normal (cache-aware) B-trees, performing 790 times faster. However this
+normal (cache-aware) B-trees, performing $790$ times faster. However this
 improvement comes at the cost of sorted inserts and normal searches. Sorted
-inserts are 3.1 times slower and searches are 3.5 times slower. Nevertheless,
-these experimental results indicate that the the COLA is a suitable data
-structure when random insertions are common.
+inserts are $3.1$ times slower and searches are $3.5$ times slower. Nevertheless,
+these experimental results indicate that the COLA is a suitable data
+structure when lots of random insertions are common.
 
@@
 of its internal state. The straw-man solution is to simply ensure mutual
-exclusion on the entire data structure; this approach, however, yields poor
-performance.
+exclusion on the entire data structure and use the pre-existing data structures
+without modification; this approach, however, yields poor performance.
 
@@
 the second is a CO B-tree based on a packed memory array (PMA). Bender also
-presents a lock-free variant of the latter data structure. For the remainder
-of the discussion, assume that the keys are fixed-size integer keys.
+presents a lock-free variant of the latter data structure.
 
@@
 
-This data structure is based on a strongly weight-balanced exponential tree
-found in \cite{Bender2002}. A tuning parameter $1 < \alpha < 2$ is chosen for
-the data structure. Each node in the tree maintains both keys which divide up
-its children, in addition to a \textit{right-link} pointer, which contains
-both a pointer to a node's right sibling, plus the key value of the minimum
-key in the node's right sibling. If no such sibling exists, a sentinel
-$\infty$ pointer is used.
+This data structure is based on the strongly weight-balanced exponential tree
+described in Section~\ref{sec:structure}. A tuning parameter $1 < \alpha < 2$
+is chosen for the data structure. Each node in the tree maintains both keys
+which divide up its children, in addition to a \textit{right-link} pointer,
+which contains both a pointer to a node's right sibling, plus the key value of
+the minimum key in the node's right sibling. If no such sibling exists, a
+sentinel $\infty$ pointer is used.
 
@@
 \rhyperceil$-leaf tree. The second part of the node consists of a $\lhyperceil
-k \rhyperceil$-size array contained the $k$ elements in sorted order. The leaf
+k \rhyperceil$-size array containing the $k$ elements in sorted order. The leaf
 nodes of the CO search tree point to the associated entries in the array.
@@
 
-Sequential search takes $O(\log_B{N} + \log_{\alpha}{\log_2{B}})$ block
-transfers (with high probability). The probabilistic bound comes from the
-probabilistic key promotion property. The intuition behind the proof of this
-bound comes from the argument that when the tree has height $h =
-\Omega(\log_{\alpha}{\log_2{\log_2{N}}})$, then the total number of keys in
-the tree is $O(2^{\alpha^h})$. The proof is then completed by summing over the
-range of $h$ which is probabilistic $O(\log_{\alpha}{\log_2{N}})$, the access
-time from some level $h$ to find the correct child.
-
-Sequential insert takes $O(\log_B{N} + \log{\alpha}{\log_2{B}})$ block
+Sequential search takes $O(\log_B{N} + \log_{\alpha}{\log{B}})$ block transfers
+(with high probability). The probabilistic bound comes from the probabilistic
+key promotion property. The intuition behind the proof of this bound comes from
+the argument that when the tree has height $h =
+\Omega(\log_{\alpha}{\log{\log{N}}})$, then the total number of keys in the
+tree is $O(2^{\alpha^h})$. The proof is then completed by summing over the
+range of $h$ which is probabilistic $[0, O(\log_{\alpha}{\log{N}})]$, the
+access time from some level $h$ to find the correct child.
+
+Sequential insert takes $O(\log_B{N} + \log{\alpha}{\log{B}})$ block
 transfers (in expectation). This proof is constructed in a similar fashion as
@@
 which indexes into a packed-memory array. Bender introduces a new
-packed-memory array, called a \textit{one-way packed-memory array}, which they
-argue is easier to manage concurrently.
+packed-memory array, called a \textit{one-way packed-memory array}, which he
+argues is easier to manage concurrently.
 
@@
 inserts/deletes, the PMA is rebalanced. A new array is allocated, and the
-elements are spread out evenly over the new array (too minimize element
+elements are spread out evenly over the new array (to minimize element
 density).
 
-Each leaf in the static CO B-tree used is a pointer into a $\Theta(\log
-N)$-size region in the packed-memory array.
+Each leaf in the static CO B-tree is a pointer into a $\Theta(\log N)$-size
+region in the packed-memory array.
 
@@
 since search is a static CO tree lookup followed by a scan, the time bound is
-$O(\log_{B}{N} + {\log_2{N}}/{B} + 1)$.
-
-Sequential insert is $O(\log_{B}{N} + \log_2^2{N}/{B} + 1)$. The proof is
-based off an accounting argument which places $\Theta(\log_2^2{m}/B)$ dollars
-in each array slot. The proof is not presented in any detail in the paper, but
-rather it is a slightly-modified version of the proof presented in
-\cite{Katriel02}.
+$O(\log_{B}{N} + ({\log{N}})/{B} + 1)$.
+
+Sequential insert is $O(\log_{B}{N} + (\log^2{N})/{B} + 1)$. The proof is based
+off an accounting argument which places $\Theta((\log^2{m})/B)$ dollars in each
+array slot. The proof is not presented in any detail in the paper, but rather
+the authors argue that it is simply a slightly-modified version of the proof
+presented in \cite{Katriel02}.
 
@@
 during the PMA modification are replaced by augmenting each element in the PMA
-with a marker. A marker is used to denote the ongoing (concurrent) operation
-of a particular cell.
+with a marker. A marker is used to denote the ongoing (concurrent) operation of
+a particular cell. The authors assume that a marker is able to fit within a
+single machine-word (so that the atomic operations are available and
+performant).
 
@@
 \end{inparaenum}. These primitives are self-explanatory, except they
-all work in a non-blocking manner (they can fail at any point).
+all work in a non-blocking manner (they can fail at any point, and it is
+up to the current process to restart the operation on failure).
 

### 2011-12-12 paper.tex
edits of streaming section + fix a lost citation
I lost a citation from merging- oops

@@
 compression is handled sub-optimally. Regular B-trees and string B-trees are
-not cache oblivious, and thus incur poorer performance as described
-previously. Range queries have poor performance in B-trees. In
-\cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
-Oblivious String B-tree (COSB-tree), which they argue works efficiently in
-these problem areas for variable length string keys. They also provide a
-modified front-compression scheme, which gives their approach the advantages
-of using less space and having improved locality within the compressed data.
-
-\subsection{Description}
-Bender et al.
-perform 2 experiments in \cite{BenderFaKu06}, in which a regular B-tree implementation tuned to various block sizes is compared to a static COSB-tree implementation.
-The experiment involved measuring the amount of time required to perform: 440,000 and 450,000 inserts of random values, a range query of all values, and 1000 random searches.
-The results of the experiment show that COSB-trees
-outperformed B-trees most of the time for random searches, range queries and
-random inserts. However, some point between 440,000 and 450,000 inserts the
-COSB-tree had to rebalance the entire tree, causing it to fall behind in this
-measurement. However, Bender et al. argue that the experiment was biased
-towards the B-tree implementation, since the trees are "young", and thus still
-well organized. The COSB-tree implementation is also compared to Berkeley DB
-\cite{BerkeleyDB}, and achieves performance comparable to Berkeley DB with
-tuned parameters. The results seem to show that there is a window in which
-large and small-block size B-trees perform poorly, and that COSB-trees may be
-a reasonable alternative to improve performance.
+not cache oblivious, and thus incur poorer performance as described previously.
+In \cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
+Oblivious String B-tree (COSB-tree), which they argue works efficiently for
+variable length string keys. They also provide a modified front-compression
+scheme, which gives their approach the advantages of using less space and
+having improved locality within the compressed data.
 
@@
 
-The Dynamic COSB-tree consists of 3 major components: a centroid tree at the
-top; a PMA structure of hash data for the keys known as
-the \emph{hashdata} layer; and on the bottom a PMA structure for the original
-key values, compressed using an augmented version of forward compression known as the \emph{keydata} layer.
-Each leaf of the centroid tree contains a pointer into the hashdata array.
-The centroid tree only contains a subset of the keys, so each leaf points to the respective \emph{neighborhood} of the key being searched for.
-After traversing the centroid tree and following the pointer into the hashdata array layer, we are within $O(\log{}N)$ keys of $k$.
-Then a sequential scan is performed in the hashdata array layer to find $k$.
-Each entry in the hashdata array contains the hash value of the corresponding key, and a pointer to the compressed key value in the keydata layer.
-
-In the following subsections, centroid trees are explained, and a summary of
+The dynamic COSB-tree consists of three major components: a centroid tree at
+the top, a PMA structure of hash data for the keys known as the \emph{hashdata}
+layer; and on the bottom a PMA structure for the original key values,
+compressed using an augmented version of forward compression known as the
+\emph{keydata} layer. Each leaf of the centroid tree contains a pointer into
+the hashdata array. The centroid tree only contains a subset of the keys, so
+each leaf points to the respective \emph{neighborhood} of the key being
+searched for. After traversing the centroid tree and following the pointer into
+the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
+sequential scan is performed in the hashdata array layer to find $k$. Each
+entry in the hashdata array contains the hash value of the corresponding key,
+and a pointer to the compressed key value in the keydata layer.
+
+In the following sections, centroid trees are described, and a summary of
 how to modify them to preserve balance and structure in the dynamic case is
@@
 
-COSB-trees are a solution to the dictionary matching problem.
-Dictionary matching, or string matching, refers to matching strings using prefix search.
-Let $D$ be a dictionary of $N$ keys.
-To perform faster searches, the keys in $D$ are preprocessed using hashing and divide-and-conquer via centroid trees.
-Figure \ref{fig:centroid} depicts an example centroid tree.
-
-Consider the compacted\footnote{where compacted means all non-branching paths are reduced to edges.} trie $T$ constructed over $D$ for string matching.
-Consider the vertex $\rho$ in $T$ such that $\rho$ has
-between N/3 and 2N/3 descendants, inclusive. Call $\rho$ the \emph{centroid vertex} of $T$. When searching for a key $k$ in
-the compacted trie, you traverse the trie as follows: for each node $y$ that
-you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on into the
-trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the trie
-resulting from excluding $y$ and its subtree. The \emph{centroid tree} is
+COSB-trees are a solution to the \textit{dictionary matching problem}.
+Dictionary matching, or string matching, refers to matching strings using
+prefix search. Let $D$ be a dictionary of $N$ keys. To perform faster searches,
+the keys in $D$ are preprocessed using hashing and divide-and-conquer via
+centroid trees. Figure \ref{fig:centroid} depicts an example centroid tree.
+
+Consider the compacted\footnote{Compacted means all non-branching paths are
+reduced to edges.} trie $T$ constructed over $D$ for string matching.  Consider
+the vertex $\rho$ in $T$ such that $\rho$ has between N/3 and 2N/3 descendants,
+inclusive. Call $\rho$ the \emph{centroid vertex} of $T$. When searching for a
+key $k$ in the compacted trie, you traverse the trie as follows: for each node
+$y$ that you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on
+into the trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the
+trie resulting from excluding $y$ and its subtree. The \emph{centroid tree} is
 obtained by rooting the tree at the centroid vertex $\rho$, and making its
@@
 
-The centroid tree in the COSB-tree is layed out in memory using a modified van Emde Boas (vEB) layout for
-weight-balanced binary trees. The \emph{hyperhyperfloors} of nodes are used as
-a guide for cutting the tree into equal-size subtrees for the vEB layout. The
-\emph{hyperhyperfloor} of x is denoted as $\lhyperfloor{}x\rhyperfloor{}$, and
-refers to rounding x down to the nearest power of a power of 2. This ensures
-trees of equal size on all \emph{levels of detail}, as the centroid tree is
-traversed.
-
-The following approach is used to rebalance the centroid tree: Put all the
-elements in DFS or Euler tour order for the compacted trie. Given this order, scan the trie to identify the centroid vertex. Then partition the trie into its respective
-up and down tries, and recurse on the up and down tries. The Euler tour order
-can be found in $O(\log{}N)$ scans, since the trie is only $O(\log{}N)$
-deep. Thus these updates have amortized cost of $O(1+\log^{2}N/B)$ memory transfers.
+The centroid tree in the COSB-tree is layed out in memory using a modified van
+Emde Boas (vEB) layout for weight-balanced binary trees. The
+\emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into
+equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x is
+denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to
+the nearest power of a power of 2. This ensures trees of equal size on all
+\emph{levels of detail}, as the centroid tree is traversed.
+% Should explain what "level of detail" means here
+
+The following approach is used to rebalance the centroid tree: put all the
+elements in DFS or Euler tour order for the compacted trie. Given this order,
+scan the trie to identify the centroid vertex. Then partition the trie into its
+respective up and down tries, and recurse on the up and down tries. The Euler
+tour order can be found in $O(\log{N})$ scans, since the trie is only
+$O(\log{N})$ deep. Thus these updates have amortized cost of
+$O(1+(\log^{2}{N})/B)$ memory transfers.
 
 The following approach is used for maintaining the vEB layout of the centroid
-tree in the COSB-tree: first, weight balance is relaxed from a
-maximum weight difference of 2 between left and right subtrees, to a larger
-constant $c$. Thus a node $v$ only becomes out of balance every
-$\Omega(\textsc{weight(v)})$ insertions or deletions of descendants of v. The nodes are stored in a PMA, preserving order of the nodes using $\Theta(N)$space.
-
-On insert or delete, the area around the update point is scanned for sections of the array to rearrange.
-If a section around the update point reaches a certain "density" threshold, the nodes are evenly
-distributed across the section to make room for more nodes. To
-ensure that updating pointers isn't too expensive, any node included in the
-section must have all of its subtrees included.
-The subtrees are dictated by the vEB layout, and subtrees are chosen from the bottom up.
-The smallest neighboring trees (single nodes) in the vEB layout are added first, and larger trees are included when all of their subtrees have been added to the section.
-Root trees are only added after all bottom trees are added (in terms of the vEB layout).
-Figure \ref{fig:centroid_relayout} gives an example of how subtrees are chosen for rearranging.
-
-This approach
-results in an amortized update cost of $O(1+(\log{}N)/B)$ memory transfers per
-update. In conjunction with the cost of rebuilds described above, this is only
-an additive $O(\log{}N/B)$ factor, so the overall bound is amortized
-$O(1+\log^{2}N/B)$ memory transfers.
+tree in the COSB-tree: first, weight balance is relaxed from a maximum weight
+difference of 2 between left and right subtrees, to a larger constant $c$. Thus
+a node $v$ only becomes out of balance every $\Omega(\textsc{weight(v)})$
+insertions or deletions of descendants of $v$. The nodes are stored in a PMA,
+preserving order of the nodes using $\Theta(N)$ space.
+
+On insert or delete, the area around the update point is scanned for sections
+of the array to rearrange. If a section around the update point reaches a
+certain \textit{density} threshold, the nodes are evenly distributed across the
+section to make room for more nodes. To ensure that updating pointers is not too
+expensive, any node included in the section must have all of its subtrees
+included. The subtrees are dictated by the vEB layout, and subtrees are chosen
+from the bottom up. The smallest neighboring trees (single nodes) in the vEB
+layout are added first, and larger trees are included when all of their
+subtrees have been added to the section. Root trees are only added after all
+bottom trees are added (in terms of the vEB layout). Figure
+\ref{fig:centroid_relayout} gives an example of how subtrees are chosen for
+rearranging.
+
+This approach results in an amortized update cost of $O(1+(\log{N})/B)$ memory
+transfers per update. In conjunction with the cost of rebuilds described above,
+this is only an additive $O((\log{N})/B)$ factor, so the overall bound is
+amortized $O(1+(\log^{2}N)/B)$ memory transfers.
 \begin{figure}
@@
 
-\caption{A centroid layout example.
-Assume the smallest subtrees in this example only contain a single node.
-Suppose a node was inserted into a leaf position in the bottom left of the tree.
-The numbers represent the order in which each subtree is added to the section to be rearranged.}
+\caption{A centroid layout example. Assume the smallest subtrees in this
+example only contain a single node. Suppose a node was inserted into a leaf
+position in the bottom left of the tree. The numbers represent the order in
+which each subtree is added to the section to be rearranged.}
 \label{fig:centroid_relayout}
 \end{figure}
+
 \subsubsection{Modifying Front Compression to Preserve Locality}
@@
 This section provides a brief summary of how front compression is modified to
-optimize compression and decoding for dynamic COSB-trees. Consider a sequence of $j$ keys to store $k_{1}$, $k_{2}$,
-..., $k_{j}$. Without compression, this requires $\sum_{j}||k_{j}||$ space in
-memory. To compress these keys using standard front compression, let $\pi_{j}$ be the
-longest common prefix between $k_{j}$ and $k_{j+1}$. Now instead we store the
-keys in the following format:
-
-\begin{center}
-$k_{1}$,$||\pi_{2}||$,$\sigma_{2}$,$||\pi_{3}||$$\sigma_{3}$,...,$||\pi_{j}||$,$\sigma_{j}$
-\end{center}
-
+optimize compression and decoding for dynamic COSB-trees. Consider a sequence
+of $j$ keys to store $k_{1},k_{2},...,k_{j}$. Without compression, this
+requires $\sum_{j}||k_{j}||$ space in memory. To compress these keys using
+standard front compression, let $\pi_{j}$ be the longest common prefix between
+$k_{j}$ and $k_{j+1}$. Now instead we store the keys in the following format:
+\begin{equation}
+  k_{1}||\pi_{2}||\sigma_{2}||\pi_{3}||\sigma_{3}||...||\pi_{j}||\sigma_{j}
+\end{equation} 
 where $\sigma_{j}$ is the suffix of $k_{j}$ after removing the first $\pi_{j}$
@@
 
-Bender et al.
-provide a modification that reduces the high decoding cost of front compressing D to $O(1+||k||B)$, while maintaining a compressed size of $O(\llangle{D}\rrangle{})$\footnote{$\llangle{D}\rrangle{}$ is shorthand for the size of $D$ after front compression} after compression.
-The modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be
-the first key added to the compression stream. For each subsequent key
-$k_{i}$, if $k_{i}$ can be decoded with just the last $c||k_{i}||$ characters,
-add $\pi_{i}$,$\sigma_{i}$ as described above.
-Otherwise, add $k_{i}$ to the stream without compressing it.
-By construction, this scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.
+Bender et al. provide a modification that reduces the high decoding cost of
+front compressing $D$ to $O(1+||k||B)$, while maintaining a compressed size of
+$O(\llangle{D}\rrangle{})$\footnote{$\llangle{D}\rrangle{}$ is shorthand for
+the size of $D$ after front compression} after compression. The modification
+works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be the first key
+added to the compression stream. For each subsequent key $k_{i}$, if $k_{i}$
+can be decoded with just the last $c||k_{i}||$ characters, add
+$\pi_{i}$, $\sigma_{i}$ as described above. Otherwise, add $k_{i}$ to the stream
+without compressing it. By construction, this scheme guarantees that any key
+$k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.
 
@@
 The operations supported by the COSB-tree are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{},
-and \Range{}. The behavior of each operation for dynamic COSB-trees is
+and \Scan{}. The behavior of each operation for dynamic COSB-trees is
 described below. \Insertkonly{} and \Delete{} are also supported but not
@@
 
-\Search{}: works as follows for a given key $k$. Compute the hash value for
+\Search{} works as follows for a given key $k$. Compute the hash value for
 $k$ and call it $\rho$, and traverse the centroid tree of the COSB-tree using
@@
 
-\Pred{},\Succ{}: works in the same way as \Search{} until the sequential scan
-in the hashdata layer. When scanning the hashdata layer, look for $\rho_{p}$
-or $\rho_{s}$, respectively, in place of $\rho$, and compare the appropriate
-keys in the keydata layer.
-
-\Range{}: This operation is not explicitly described in \cite{BenderFaKu06}, but presumably this is very similar to
-\Search{}. Let $k_{b}$ be the beginning of the range and let $k_{e}$ be the
-end of the range. Hash the range keys and identify the predecessor of
-$\rho_{b}$ and the successor of $\rho_{e}$ using the centroid tree. Scan the
-hashdata layer across from $\rho_{b,p}$ to $\rho_{e,s}$. Only start noting the
-beginning of range when the first key with hash value after $\rho_{b}$, and
-stop once we reach the first key with hash value after $\rho_{e}$. Decode and
-return the set of keys within the given range in the keydata layer.
+\Pred{}, \Succ{} work in the same way as \Search{} until the sequential scan in
+the hashdata layer. When scanning the hashdata layer, look for $\rho_{p}$ or
+$\rho_{s}$, respectively, in place of $\rho$, and compare the appropriate keys
+in the keydata layer.
+
+\Scan{}: This operation is not explicitly described in \cite{BenderFaKu06}, but
+presumably this is very similar to \Search{}. Let $k_{b}$ be the beginning of
+the range and let $k_{e}$ be the end of the range. Hash the range keys and
+identify the predecessor of $\rho_{b}$ and the successor of $\rho_{e}$ using
+the centroid tree. Scan the hashdata layer across from $\rho_{b,p}$ to
+$\rho_{e,s}$. Only start noting the beginning of range when the first key with
+hash value after $\rho_{b}$, and stop once we reach the first key with hash
+value after $\rho_{e}$. Decode and return the set of keys within the given
+range in the keydata layer.
 
 \subsection{Complexity}
-The keys in are hashed and hash values are compared in place of the original key values to speed up comparisons in the centroid tree and hashdata layer.
-To avoid mismatches due to overlapping hash values, the original key values are compared only once at the bottom layer, resulting an additive $O(||k||/B)$ memory transfers in each of the operators for each key $k$ that needs to be compared.
-
-\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high
-probability (w.h.p.).
-The centroid tree is stored in a PMA with a dynamic CO layout.
-This layout reduces the number of memory transfers required when traversing the tree to $O(log_{B}N)$.
-Once in the hashdata layer, we are already within $O(logN)$ keys of $k$, which require $O(\log{}N)/B)$ memory transfers to scan.
-Decoding and comparing the original key values requires $O(1+||k||/B)$ memory transfers, due to the modified front compression scheme.
-
-The \Pred{} and \Succ{} operations are similar in behavior to searches, and thus only have the added cost of having to operate on $k'$, where $k'$ is the predecessor or successor, respectively.
-These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers w.h.p..
-Similarly, range queries take
-$O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory transfers,
-where in this case $k'$ is the other end of the range and $Q$ is the set of
-keys within the given range.
+The keys are hashed and hash values are compared in place of the original key
+values to speed up comparisons in the centroid tree and hashdata layer. To
+avoid mismatches due to hash collisions, the original key values are compared
+only once at the bottom layer, resulting an additive $O(||k||/B)$ memory
+transfers in each of the operators for each key $k$ that needs to be compared.
+
+\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high probability
+(w.h.p.).  The centroid tree is stored in a PMA with a dynamic CO layout. This
+layout reduces the number of memory transfers required when traversing the tree
+to $O(\log_{B}{N})$.  Once in the hashdata layer, we are already within $O(\log{N})$
+keys of $k$, which require $O((\log{N})/B)$ memory transfers to scan. Decoding
+and comparing the original key values requires $O(1+||k||/B)$ memory transfers,
+due to the modified front compression scheme.
+
+The \Pred{} and \Succ{} operations are similar in behavior to searches, and
+thus only have the added cost of having to operate on $k'$, where $k'$ is the
+predecessor or successor, respectively. These operations as a result take
+$O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers (w.h.p.). Similarly, range
+queries take $O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory
+transfers, where in this case $k'$ is the other end of the range and $Q$ is the
+set of keys within the given range.
 
@@
 delete or location for insert, which can be seen from the cost of \Search{}.
+
+\subsection{Experimental Results}
+Bender et al. perform two experiments in \cite{BenderFaKu06}, in which a
+regular B-tree implementation tuned to various block sizes is compared to a
+static COSB-tree implementation. The experiment involved measuring the amount
+of time required to perform:
+\begin{inparaenum}[(a)]
+  \item 440,000 and 450,000 inserts of random values, 
+  \item a range query of all values, and 
+  \item 1000 random searches.
+\end{inparaenum} The results of the
+experiment show that COSB-trees outperformed B-trees most of the time for
+random searches, range queries and random inserts. However, some point between
+440,000 and 450,000 inserts the COSB-tree had to rebalance the entire tree,
+causing it to fall behind in this measurement. However, Bender et al. argue
+that the experiment was biased towards the B-tree implementation, since the
+trees are ``young'', and thus still well organized. The COSB-tree
+implementation is also compared to Berkeley DB \cite{BerkeleyDB}, and achieves
+performance comparable to Berkeley DB with tuned parameters. The results seem
+to show that there is a window in which large and small-block size B-trees
+perform poorly, and that COSB-trees may be a reasonable alternative to improve
+performance.
 
@@
   \item \textit{cell-deletion}, and
-  \item \textit{read}
-\end{inparaenum}. These primitives are self-explanatory, except they
+  \item \textit{read}.
+\end{inparaenum} These primitives are self-explanatory, except they
 all work in a non-blocking manner (they can fail at any point, and it is

### 2011-12-12 paper.tex
Adding conclusion

@@
 
+\begin{figure*}[h!t]
+\begin{center}
+    \begin{tabular}{ |   l   |  l     |     l     |     }
+    \hline
+	Data Structure & Search & Insert / Delete  \\ \hline \hline
+	 \textit{Original CO B-tree}  & $O(\log_{B}{N})$ & $O(\log_{B}{N} + \frac{\log^2{N}}{B})$  \\ \hline
+	\textit{String CO B-tree}  & $O(1+\log_{B}N+||k||/B)$ &   $O(1+\log_{B}N+\log^{2}N||k||/B)$  \\ \hline
+	\textit{Streaming Shuttle Tree}  & $O(\log_{B}{N})$ &  $O(log_{B+1} N / B^{\Theta(1/(loglogB)^2)} + (log^2 N) / B)$  \\ \hline
+	\textit{Streaming COLA}  & $O(\log N)$ & $O((\log N) / B)$  \\ \hline
+	\textit{Concurrent Exponential CO B-tree} &   $O(\log_B{N} + \log_{\alpha}{\log{B}})$  &   $O(\log_B{N} + \log{\alpha}{\log{B}})$  \\ \hline
+	\textit{Concurrent PMA CO B-tree}  & $O(\log_{B}{N} + ({\log{N}})/{B} + 1)$ &   $O(\log_{B}{N} + (\log^2{N})/{B} + 1)$  \\ \hline
+    \end{tabular}
+    \caption{A summary of the surveyed CO B-trees. The bounds are in terms of the number of memory transfers. The bounds for \textit{Concurrent Exponential CO B-tree} are in expectation only.}
+    \label{fig:table}
+\end{center}
+\end{figure*}
+
 \subsubsection{Lock-free Variant Operations}
@@
 \section{Conclusion}
-Fill this in.
-%\end{document}  % This is where a 'short' article might terminate
-
-%ACKNOWLEDGMENTS are optional
+
+This paper presented a survey on different types of cache-oblivious B-trees: the 
+original \textit{cache-oblivious B-tree}, the \textit{cache-oblivious string
+B-tree}, the \textit{cache-oblivious streaming B-tree}, and the \textit{cache-oblivious
+concurrent B-tree}. A summary of these data structures is provided
+in Figure~\ref{fig:table}.
+
 \section{Acknowledgments}
-Fill this in.
+We sincerely thank Dr. David Karger and all of the TAs for 6.854 
+for their guidance, help, and support. May the Flow (Max-Flow, that is) be with you all.
 

### 2011-12-12 paper.tex
added citations to motivation for co string b-trees, and fixed figure 3

--- 2011-12-12 paper.tex
@@
 
-% preprint + pagenumbering for page numbers
-\documentclass[preprint]{style}
-\pagenumbering{arabic}
+\documentclass{style}
 
@@
 \usepackage{url}
-
-% This gets rid of the block of whitespace on the first page
-\makeatletter
-\let\@copyrightspace\relax
-\makeatother
 
@@
 This paper presents a survey of cache-oblivious B-trees (CO B-trees). The CO
-B-tree was first introduced by Bender et al \cite{BenderDemainColton}. It is
-an external memory data structure; it optimizes the number of memory transfers
-performed rather than the number of CPU instructions that executed. Unlike
-other external memory data structures such as the widely used B-tree, the CO
-B-tree does not have access to the parameter's of the machine's memory
-hierarchy, including the block size/capacity at each level, or the relative
-speeds. Hence, the CO B-tree is \textit{cache-oblivious}.  Despite this, the CO
-B-tree is able to achieve the same provably optimal search bound as the B-tree
-and near optimal insertion and deletion bounds. In this paper we first describe
-the original CO B-tree and explain how it is able to achieve its performance.
-We then present and analyze three followup CO B-tree data structures that
-improve upon weaknesses of the original: the \textit{cache-oblivious string
-B-tree}, which introduces support for variable length keys, the
-\textit{cache-oblivious streaming B-tree}, which improves the insertion bound
-at the expense of the search bound, and lastly, the \textit{cache-oblivious
-concurrent B-tree}, which introduces support for concurrency.
+B-tree was first introduced by Bender et al. in 2000. It is an external memory
+data structure - meaning that it optimizes the number of memory transfers it
+performs rather than the number of CPU instructions that it executes. Unlike
+other external memory data structures, like the widely used B-tree, the CO
+B-tree does not require any information about the number of level in the
+machine's memory hierarchy, the block size and capacity at each level, or the
+relative speeds at each level. Hence, the CO B-tree is ``cache-oblivious''.
+Despite this, the CO B-tree is able to achieve the same optimal search bound
+as the B-tree and near optimal insertion and deletion bounds as well. In this
+paper we first describe the original CO B-tree and explain how it is able to
+achieve its memory performance. We then present and analyze three followup CO
+B-tree data structures that improve upon weaknesses of the original: the
+cache-oblivious string B-tree, which introduces support for variable length
+keys, the cache-oblivious streaming B-tree, which incorporates improvements to
+the insertion bound at the expense of the search bound, and lastly, the
+cache-oblivious concurrent B-tree, which introduces support for concurrency.
 
@@
 
-Most algorithms assume an idealized model of computation that consists of
-memory as a flat byte-addressable array which can be read and modified in some
-fixed time. However, in practice memory is not a flat array, but rather is
-organized as a hierarchy consisting of fast on-chip caches (L1, L2, etc.), to
-slower DRAM, to even slower disks. When a program requests data, the data is
-located in the memory hierarchy and typically carried up the chain to the
-fastest memory layer (closest to the CPU), usually a cache. Unfortunately, this
-cache can only have a limited capacity, and once it is full any new data
-requests necessitate some existing cached data to be evicted. Transferring data
-from lower to higher levels in the memory hierarchy often takes an
-exponentially long time; access times between the different layers can differ
-by many orders of magnitude. Therefore, the time to transfer data often
-dominates the overall runtime performance of an algorithm.
-
-External memory algorithms are wary of this fact and seek to optimize the
-number of memory transfers performed rather than the number of CPU instructions
-executed. They often try to exploit the fact the memory hierarchies transfer
-data in blocks rather than individual machine word units. Thus, external memory
-algorithms are designed to not only limit the number of memory transfers, but
-also preserve the locality of related data so that when a costly memory access
-is performed, its cost is amortized.
+Most algorithms assume an idealized model of computation that consists of a
+single CPU and memory as a flat array
+that can be read and written. However, in reality memory is not a flat array
+but rather a hierarchy from super fast on chip caches, to slower main memory,
+to even slower disks. When a program requests data, the data is found in the
+memory hierarchy and typically carried up the hierarchy to the fastest memory
+layer possible, usually a cache. The CPU is then able to access it to perform
+any necessary operations. Unfortunately, this cache often has a small capacity
+and once it is full any new data requests cause some existing cached data to
+be evicted. Transferring data from lower to higher levels in the memory
+hierarchy often takes a long time. In fact, access times between the different
+layers of the hierarchy can differ by many orders of magnitude and thus the
+time to transfer data often dominates the overall runtime performance of an
+algorithm. External memory algorithms are wary of this fact and aim to
+optimizing the number of memory transfers it performs rather than the number
+of instructions it executes. They often try to exploit the fact the memory
+hierarchies transfer data in blocks rather than individual atomic units. Thus,
+external memory algorithms are designed to not only limit the number of memory
+transfers it performs but also preserve the locality of related data so that
+when a memory access is performed, its cost is amortized.
 
@@
 it is no surprise that many external memory problems can be reduced to finding
-a suitable data structure that efficiently implement certain operations. A well
-known external memory data structure is the B-tree \cite{BayerM72}. The B-tree
-works by making each one of its nodes the same size as a memory block in a
-particular level of the hierarchy. Furthermore, the fanout in the tree (the
-number of children of a node) is proportional to the block size as well.  The
-B-tree incurs $O(\log_{B}{N})$\footnote{Throughout this paper, $\log$ with the
-base omitted is base-$2$} memory transfers for \Search{}, \Insert{}, and
-\Delete{} and is widely used in databases and file-systems.
+a suitable data structure that efficiently implement certain operations. A
+well known external memory data structure is the B-tree. At a high level, the
+B-tree works by making each one of its nodes the same size as a memory block
+in a particular level of the hierarchy. Furthermore, the fanout in the tree
+(the number of children of a node) is proportional to the block size as well.
+The B-tree incurs $O(log(N))$ memory transfers for \Search{}, \Insert{}, and
+\Delete{} and is widely used in databases and filesystems.
 
@@
 
-The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in a
-two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}.
-However, in practice, architectures typically have memory hierarchies that are
-larger than two. The argument for the DAM model is that in practice, only one
-level in the memory hierarchy tends to dominate the performance, and thus a
-two-level memory model is sufficient.  However, this means that B-trees can
-begin to incur an increasing number of memory transfers if the dominating level
-in the hierarchy changes.
-
-Additionally, the organization of a B-tree requires knowing the block size of
-the memory layer in advance. B-trees are implemented with parameterized code
-that needs to be changed for every level in the memory hierarchy. This tuning
+The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in
+a two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}.
+However, in reality, computing architectures typically have memory hierarchies
+that are much larger than two. The argument for the DAM model is that in
+practice only one level in the memory hierarchy tends to dominate the
+performance and thus a two-level memory model is sufficient for all intents
+and purposes. However, this does mean that, theoretically, B-trees can begin
+to incur an increasing number of memory transfers if the dominating level in
+the hierarchy changes.
+
+Additionally, the B-tree expects to know the block size of its memory layer in
+advance. B-Trees are implemented with parameterized code that needs to be
+changed for every level in the memory hierarchy. Needless to say, this tuning
 can be very difficult and error prone. Additionally, tuning this parameter may
@@
 The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
-of computation \cite{Frigo99}. In the CO model an algorithm has no information
-about the memory hierarchy. It does not know anything about the speed or the
-block size of any level. Conceptually, the CO model allows algorithm designers
-to reason about the simple two-level model like DAM but also prove results for
-any arbitrary multi-level memory hierarchy. This is because algorithms that
+of computation. In the CO model an algorithm has no information about the
+memory hierarchy. It does not know anything about the speed or the block size
+of any level. Conceptually, the CO model allows algorithm designers to reason
+about the simple two-level model like DAM but also prove results for any
+arbitrary multi-level memory hierarchy. This is because algorithms that
 perform well in a CO model should, in theory, perform well regardless of
@@
 
-In \cite{BenderDemainColton}, Bender et al. introduce the cache-oblivious
+In \cite{BenderDemainColton} Bender et al. introduce the cache-oblivious
 B-tree (CO B-tree). The CO B-tree is similar to the normal B-tree except that
-it does not require explicit parameters about the memory hierarchy.
-
-The CO B-tree is able to achieve the optimal search bound of
-$\Theta(\log_{B}{N})$ memory transfers, the optimal scan bound of $\Theta(N/B)$
-memory transfers and near optimal insertion and deletion bound of 
-$\Theta(\log_{B}{N} + \frac{\log^2{N}}{B})$. 
-The CO B-tree was a seminal data structure that gave rise to many others.
-
-This paper provides a survey of the different types of CO B-trees and discusses
-the various trade offs that they make. We start by analyzing the original CO
-B-tree and explain how it works. This includes a detailed description of the
-packed memory array (PMA) and strongly weight-balance binary search trees
-(SWBST) that are integral to the CO B-tree's memory performance. Following this
-discussion we provide an analysis of the CO string B-tree (a CO B-tree with
-variable length keys), the CO streaming B-tree (a CO B-tree that improves the
-cost of insertion at the expense of search), and the CO concurrent B-tree (a CO
-B-tree that allows multiple processes to efficiently access data concurrently).
+it does not require explicit parameters about the memory hierarch.
+
+The CO B-tree is able to achieve the optimal search bound of $\Theta(log(N))$
+memory transfers, the optimal scan bound of $\Theta(N/B)$ memory transfers and
+near optimal insertion and deletion bound of $\Theta(log_B N +
+\frac{log^2{N}}{B})$. The CO B-tree was a seminal data structure that gave
+rise to many others.
+
+This paper provides a survey of the different types of CO B-trees and
+discusses the various trade offs that they make. We start by analyzing the
+original CO B-tree and explain how it works. This includes a detailed
+description of the packed memory array (PMA) and strongly weight-balance
+binary search trees (SWSBT) that are integral to the CO B-tree's memory
+performance. Following this discussion we provide an analysis of the CO String
+B-tree,a CO B-tree with variable length keys, the CO Streaming B-tree, a CO
+B-tree that improves the cost of insertion at the expense of search, and the
+CO Concurrent B-tree, which allows multiple processes to access the data at
+the same time.
 
@@
 
-The CO B-tree optimizes the number of memory transfers by structuring its
-elements as a strongly weight-balanced search tree (SWBST). The SWBST is then
-laid out into a packed memory array (PMA) using a van Emde Boas (vEB) layout.
-The SWBST and PMA play crucial roles in the other cache-oblivious data
-structures that were developed after the CO B-tree and thus are described at
-length in the next two sections.
+Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
+\cite{BenderDemainColton}. The CO B-tree achieves its number of memory
+transfers by structuring its elements as a strongly weight-balanced search
+tree (SWBST). The SWBST is then laid out into a packed memory array (PMA)
+using a van Emde Boas (vEB) layout. The SWBST and PMA play crucial roles in
+the other cache-oblivious data structures that were developed after the CO
+B-tree and thus are described at length in the next two sections.
 
@@
 The ultimate goal of a CO B-tree is to reduce the number of memory transfers
-while supporting \Search{}, \Insert{}, and \Delete{} operations. Thus, it is
-not surprising that the CO B-tree maintains its data in a balanced search tree.
-The search tree is laid out in memory according to a van Emde Boas (vEB) layout
-\cite{veb1,veb2}. The idea of a vEB layout is fairly straight forward. Let $h$
-denote the height of a tree. The vEB layout splits the tree approximately at
-height $h/2$. This splits the original tree into a \textit{top} subtree and
-many \textit{bottom} subtrees.  Let $A$ denote the top subtree and
-$B_1,...,B_k$ denote the bottom subtrees.  The vEB layout is defined
-recursively using the order: $A,B_1,...,B_k$. This layout is also shown in
-Figure~\ref{fig:veb}.
-
-A nice consequence of using a vEB layout is that it can be shown to use
-$O(\log_{B}{N})$ memory transfers for searches \cite{veb1, veb2}. Thus, the CO
-B-tree has already achieved the desired bound for search queries. It still
-remains to be shown that the CO B-tree achieves the desired number of memory
-transfers for dynamic inserts and deletes.
-
-To support dynamic inserts/deletes, a \textit{strongly weight-balanced search
-tree} (SWBST) is used \cite{swbst}.  The weight of a node $u$ is equal to the
-total number of $u$'s descendants plus 1. It can also be thought of as the sum
-of the weights of $u$'s children plus 1. Formally, $u$'s weight is $w(u) = 1 +
-\sum_{v \in children} w(v)$.
-
-Most weight-balanced search trees require that the heights of a node's left and
-right subtrees differ by at most a constant factor. The SWBST enforces the
-stronger property that a node at height $h$ in the tree (leaves have a height
-of 1) has $\Theta(c^h)$ descendants where $c$ is some constant ($c > 4$). This
-property becomes important later on when laying the tree out in memory (see
-Section~\ref{sec:layout}).
-
-The definition of a SWBST given in \cite{swbst} provides this bound for the
-weight of a node $u$:
+while supporting \Search{}, \Insert{}, and \Delete{} operations. Thus, at a
+fundamental level the CO B-tree still needs to maintain its data in some
+structure. That structure is a balanced search tree. The search tree is laid
+out according to a van Emde Boas (vEB) layout \cite{veb1}, \cite{veb2}. The
+idea of a vEB layout is fairly straight forward. Let $h$ denote the height of
+a tree. the vEB layout splits the tree approximately at height $h/2$. This
+splits the original tree into a ``top'' subtree and many ``bottom'' subtrees.
+Let $A$ denote the top subtree and $B_1... B_k$ denote the bottom subtrees.
+The vEB layout is defined recursively using the order: $A, B1, ... B_k$. This
+layout is also shown in Figure~\ref{fig:veb}.
+
+A nice consequence of using a vEB layout is that it is known that it will use
+$O(log_B N)$ memory transfers for searches \cite{veb1}, \cite{veb2}. Thus, the
+CO B-tree has already achieved the desired number of memory transfers for
+search queries. It still remains to be shown that the CO B-tree achieves the
+desired number of memory transfers for dynamic inserts and deletes.
+
+It is known that the CO B-tree is going to use a balance search tree, but what
+kind? This is where the strongly weight-balanced search tree becomes useful.
+The weight of a node is $u$ is equal to the total number of $u$'s descendants
+plus 1. It can also be thought of as the sum of the weights of $u$'s children
+plus 1. Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.
+
+Most weight-balanced search trees require that the heights of a node's left
+and right subtrees are simply off by a constant factor. The SWBST enforces the
+stronger property that a node at height $h$ in the tree (height is the node's
+distance from the leaves; leaves have a height of 1) has $\Theta c^h$
+descendants where $c$ is some constant. This property becomes important later
+on when laying the tree out in memory (see Section~\ref{sec:layout}).
+
+A known data structure that achieves this property is the ``weight-balanced
+B-tree'' \cite{swbst}. The CO B-tree leverages this data structure to maintain
+its elements. Before SWBST can be used, however, it remains to be proven that
+it maintains the the $c^h$ descendants property even after inserts and
+deletes.
+
+The definition of a SWBST given in \cite{swbst} provides us with this bound
+for the weight of a node:
+
 \begin{equation}
@@
 
-Let us now consider insertions. To insert an element, search
+Let us now consider insertions. To insert an element into a WSBST we search
 down the tree and find the natural place to insert the element - just like in
-any tree structure. After inserting the element into some node $w$, 
+any tree structure. After inserting the element into some node $w$, we have to
 check to see if any of $w$'s ancestors have become unbalanced by having a
-weight greater than $2c^{h-1}$. If such an ancestor exists, it follows that
-every node along the path between that ancestor and $w$ will be unbalanced as
-well. This requires splitting nodes to restore balance, starting from $w$
-and moving up the tree.
-
-When a node is split, we need to determine which children belong to which of
-the two nodes. Dividing the children evenly is insufficient because it may
-cause another violation. Instead, find the longest sequence of elements such
-that the weight of the subsequence is $\lceil w(u)/2 \rceil$.  By doing this,
-we maintain the bound that $\frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}$. Cascade
-these changes up the tree until all of the ancestors are balanced.
-
-Deletions are supported similarly, except we check if a node has a weight $w(u)
-< \frac{c^{h-1}}{2}$. In this case, we merge two neighbors together and proceed
-up the tree. There is one edge case that needs to be handled. The merging
-procedure may produce a new node that has a weight larger than the bounds
-allow. If this happens, immediately split the node again, recursively. This can
+weight greater than $2^{d-1}$. If such an ancestor exists, it follows that
+every node along the path between that ancestor and $w$ will be un balanced as
+well. Thus, we start with $w$ and move up the tree.
+
+When we split a node, we naturally need to determine which children belong to
+which of the two nodes. We cannot just divide the children evenly because it
+may cause another violation. Thus, instead, we find the longest sequence of
+elements such that the weight of that subsequence is $\lceil w(u)/2 \rceil$.
+By doing this, we maintain the bound that $\frac{c^{h-1}}{2} \leq w(u) \leq
+2c^{h-1}$. We cascade these changes up the tree until all of the ancestors are
+balanced.
+
+Deletions are supported in the exact same way, except, we check if a node has
+a weight $w(u) < \frac{c^{h-1}}{2}$. In this case, we will merge two neighbors
+together and proceed up the tree. There is one edge case that needs to be
+handled. After merging, you may produce a new node that has a weight larger
+than the bounds. If this happens we immediately split the node again. This can
 be thought of as stealing nodes from a neighbor.
 
-We have now sketched that the SWBST has the desired property that the number of
-descendants of a node is $\Theta(c^h)$. We now describe the tree's memory layout.
+We have now shown that the SWBST achieves the desired property that the number
+of descendants of a node is $c^h$. We can now show how to lay this tree out in
+memory.
 
@@
 
-Now that the elements structured in a SWBST, we are still left with the problem
-of mapping this tree to memory and proving that it achieves the desired number
-of memory transfers per operation. The packed memory array (PMA) was developed
-to solve this problem \cite{BenderDemainColton}. There are three main costs
-that the PMA is concerned with: \textit{inserts}, \textit{splits/merges}, and
-\textit{pointer updates}.
+Now that we have our elements structured in a SWBST, we are still left with
+mapping this tree to memory and proving that it achieves our desired number of
+memory transfers. The packed memory array (PMA) was developed to solve this
+problem \cite{BenderDemainColton}. There are three main costs that the PMA is
+concerned with.
 
 First, the PMA needs to maintain open space in its memory for newly inserted
-items. The PMA is able to handle an insertion into the SWBST using only $O(1 +
-(\log^2{N}) / B)$ memory transfers (this is the cost of updating the memory
-layout of the SWBST in the PMA, not the cost of updating the data structure
-itself). Doing so may seem contradictory: maintaining the locality of elements
-while keeping space available for new inserts. The PMA uses a technique
-described in \cite{packedmemoryarray}. At a high level, the PMA maintains
-windows into the array. When a window becomes too full or too empty, the
-elements are spread out within the context of a larger window. Two parameters
-are required: the window size and the threshold for triggering a spread out.
-However, these parameters are not related to the block size of the memory level
-and thus do not violate the CO model.
-
-Second, the PMA needs to handle splits and merges caused by inserting into the
-SWBST discussed in the previous section. The PMA is able to accomplish this
-using $O(1 + (\log{N}) /B)$ memory transfers. The proof is based on the fact
-that each node has $\Theta(c^h)$ descendants (recall, this is the key property
-that SWBST provided). Thus, merging or splitting a node will cause
-$\Theta(c^h)$ nodes to move, requiring $O(1 + c^h/B)$ memory transfers. It
-follows from the definition of SWBST in \cite{swbst} that when a node is
-rebalanced, it will require $\Theta(c^h)$ insertions or deletions until it is
-rebalanced again. Thus, the amortized cost of each insertion and/or deletion is
-$O(1/B)$. When a node is split, it may traverse all the way up the tree, and
-thus the split/merge bound is $O((\log N) / B)$ memory transfers.
-
-Third, the PMA needs to handle updating the parent and child pointers that may
-become invalidated because of how the nodes moved around. If the pointers are
-all packed together in memory (e.g. within $B$ distance of each other) then
-only $O(1)$ memory transfers are required. However, if they are further apart
-then a problem arises. To deal with this, buffer nodes are introduced.  Buffer
-nodes ``stretch'' the vEB layout so that nodes do not move far away from one
-another. By using buffers, the PMA is able to achieve a pointer update cost
-using $O(1+ \frac{\log{B}}{\sqrt{B}}\log^2{N})$ memory transfers.
+items. The PMA is able to accomplish this using $O(1 + (log^2 N) / B)$ memory
+transfers. Doing so may seem like attempting to achieve contradictory goals:
+maintaining the locality of elements while keeping space available for new
+inserts. The PMA uses a technique described in \cite{packedmemoryarray}. At a
+high level, the PMA maintains windows into the array. When a window becomes
+too full or too empty, the elements are spread out within the context of a
+larger window. Two parameters are required, the window size and the threshold
+for trigger the spreading out. However, these parameters are not related to
+the block size of the memory level and thus do not violate the CO model of
+computation. Overall, the PMA is able to keep extra space in $O(1 + (log^2 N)
+/ B)$ memory transfers.
+
+Second, the PMA needs to handle the splitting and merging caused by the SWBST
+discussed in the previous section. The PMA is able to accomplish this using
+$O(1 + (log N) /B)$ memory transfers. The proof is based on the fact that each
+node has $c^h$ descendants (recall, this is the key property that SWBST
+provided). Thus, merging or splitting a node will cause $c^h$ nodes to move,
+requiring $O(1 + c^h/B)$ memory transfers. It follows from the definition of
+SWBST in \cite{swbst} that when a node is rebalanced, it will require $c^h$
+insertions or deletions until it is rebalanced again. Thus, the amortized cost
+of each insertion and/or deletion is $O(1/B)$. When you split a node, it may
+go all the way up the tree and thus overall it will incur $O((log N) / B)$
+memory transfers.
+
+Third, the PMA needs to handle updating the parent and child pointers and may
+become invalidated because of how the nodes moved around. The PMA is able to
+update the parent and child pointers. If the pointers are all close together
+in memory (e.g. within $B$ distance of each other) it is not big deal because
+only a constant number of memory transfers are required. However, if they are
+further apart the PMA introduces buffer nodes. The buffer nodes don't actually
+do anything. All they do is serve to ``stretch'' the vEB layout so that nodes
+do not move far away from one another. By using these buffers the PMA is able
+to achieve a the pointer update cost using $O(1+ logB / \sqrt{B}) log^2 N)$
+memory transfers.
 
 Of these three terms, the dominating factor is maintaining room for newly
-inserted elements - $O((\log^2{N})/B)$. Therefore, overall the CO B-tree
-achieves insertions and deletions in $O(\log_B{N} + (\log^2{N})/B)$ memory
-transfers. It still achieves the optimal $O(\log_B{N})$ memory transfers for
-searches, matching the performance of static cache-aware B-trees.
+inserted elements - $O((log^2N)/B)$. We know that the SWBST achieves
+insertions and deletions in $O(log_B N)$ if it is cache-aware. Thus, adding
+PMA makes the SWBST cache-oblivious at the cost of incurring an additional
+$O((log^2N)/B)$ memory transfers.
+
+Therefore, overall, the CO B-Tree achieves insertions and deletions in
+$O(log_B N + (log^2N)/B)$ memory transfers. However, it still achieves the
+optimal $O(log_B N)$ memory transfers for searches, matching the performance
+of ordinary, cache-aware, B-trees.
 
@@
 In order for B-trees to remain versatile, they must be able to accommodate
-various forms of data. However, traditional B-trees and string B-trees have
-suboptimal performance when variable length strings are used as keys. In
-addition, string B-trees do not support compression, and commonly used front
-compression is handled sub-optimally. Regular B-trees and string B-trees are
+various forms of data. Bender et al. assert the following weaknesses in existing string B-tree designs: traditional B-trees have
+suboptimal performance when variable length strings are used as keys\cite{BenderDemainColton}.
+String B-trees\cite{Ferragina98} handle variable-length strings well but do not use compression.
+Front compression\cite{BayerUn77,ClarkSaSt69,Comer79,Wagner73} is used in many B-trees, but is handled sub-optimally\cite{BenderFaKu06}.
+Regular B-trees and string B-trees are
 not cache oblivious, and thus incur poorer performance as described
-previously. Range queries have poor performance in B-trees. In
+previously by Bender et al.\cite{BenderDemainColton}. Range queries have poor performance in B-trees\cite{BenderFaKu06}.
+
+In
 \cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
@@
 
-\caption{A centroid tree example.
+\caption{A centroid tree example from \cite{BenderFaKu06}.
 Dotted lines represent centroid-rooted up tries, where the parent is not a prefix of the child.
@@
 While B-trees achieve an optimal bound for the number of memory transfers for
-search, they do not perform as well for bulk updates and insertions. The
-buffered-repository tree (BRT) is an alternative to the B-tree that achieves an
-improved amortized $O(\log{\frac{N}{B}})$ memory transfers for inserts compared
-to the $O(\log_{B}{N})$ memory access of B-trees. In practice, log-structured
-merge (LSM) trees \cite{ONeil96} are used for high-insert workloads.
-
-This search vs. insertion trade-off also exists in the cache-oblivious model.
-The original CO B-tree described in Section~\ref{sec:original} is designed to
-optimize for search in this trade off. However, Bender et al. propose two data
-structures that and achieve the opposite end: the shuttle tree and the
+search, they do not perform as well for updates and insertions. The
+buffered-repository tree (BRT) is a common alternative to the B-tree that
+achieves an improved amortized $O(log(N)/B)$ memory transfers for inserts
+compared to the $O(log_{B+1}N)$ memory access of B-trees. Of course, the
+B-tree and BRT are analyzed with the DMA model and are not cache-oblivious.
+
+This search vs insertion trade-off also exists in the cache-oblivious model.
+The original BO B-tree described in Section~\ref{sec:original} is designed to
+optimize for the search in this trade off. However, Bender et al. propose two
+data structures that and achieve the opposite end: the shuttle tree and the
 cache-oblivious lookahead array (COLA) \cite{BenderFaFi07}. These two data
-structures are refereed to as \textit{streaming B-trees}, indicating how they
-are optimized for bulk insertions at the cost of search.
+structures are refereed to as ``streaming B-trees'', indicating how they are
+optimized for insertions at the cost of search.
 
@@
 
-The shuttle tree is similar in layout and structure as the original CO B-tree.
+The shuttle tree is similar in layout and structure as the original CO B-Tree.
 It uses a strongly weight-balanced search tree described previously (see
 Section~\ref{sec:structure}) that is laid out in a vEB (see
-Figure~\ref{fig:veb}) layout into a PMA (see Section~\ref{sec:layout}). A
+Section~\ref{sec:veb}) layout into a PMA (see Section~\ref{sec:layout}). A
 notable difference between the shuttle tree and the CO B-tree is that the CO
-B-tree splits the tree at height $h/2$ when creating the vEB layout, while the
-shuttle tree splits at height $F_k$, where $F_k$ is the $k$-th Fibonacci number
-that is strictly less than the height of the tree. Thus if a tree's height
-happens to be a Fibonacci number $F_k$, then the shuttle tree will choose to
+B-tree splits the tree at height $h/2$ when creating the vEB layout while the
+shuttle tree splits at height $F_k$, where $F_k$ is the $kth$ Fibonacci number
+that is stickily less than the height of the tree. Thus if a tree's height
+happens to be a Fibonacci number, $F_k$, then the shuttle tree will choose to
 split the tree at height $F_{k-1}$. The top and bottom ``halves'' produced by
@@
 At a high level, the shuttle tree improves the efficiency of insertion by
-adding buffers to each node in the tree. Shuttle tree nodes have a
+adding buffers to each node in the tree. In fact, shuttle tree nodes have a
 linked list of buffers rather than just a single buffer like the BRT. Each of
-these buffers is another shuttle tree. A shuttle tree node enforces two
+these buffers is another shuttle tree. The a shuttle tree node enforces two
 constraints: the length of its linked list as well as the overall height (and
 thus capacity) of each of the buffered shuttle trees contained within its
-linked list are bounded, described below.
-
-The length of a node's linked list is related to the node's height in the tree.
-A node has a buffer for every time it is the leaf of a sub-tree that is split
-when creating the vEB layout. This idea is illustrated in Figure
+linked list.
+
+The length of a node's linked list is related to the node's height in the
+tree. A node has a buffer for every time it the leaf of a sub trees that is
+split when creating the vEB layout. This idea is illustrated in Figure
 \ref{fig:buffers}.
 
-The shuttle tree's node buffers have a maximum capacity that is enforced by
+The shuttle tree's node's buffers have a maximum capacity that is enforced by
 bounding its height. The size of the buffers increases going across the linked
-list from left-to-right. Thus, a node's first buffer is smaller than its last
-buffer. The maximum height of a buffer is given by $F_{H(k)} = F_{k-2 \lceil
-\log_{\phi} k \rceil}$, where $\phi$ is the golden ratio and $k$ is chosen such
-that $F_k$ = $\xi(h)$ and $\xi(x)$ is defined recursively: If $h$ is a
-Fibonacci number then $\xi(h) = h$, otherwise $\xi(h) = \xi(h-f)$ where $f$ is
-the largest Fibonacci number less than $h$. The maximum height of an individual
-buffer is related to the size of the recursive subtree for which the node is a
-leaf.  Since the recursive subtrees are split based on the largest Fibonacci
-number less than $h$ (as described previously) it follows that the buffer sizes
-are also a function of the Fibonacci numbers.
+list. Thus, a node's first buffer is smaller than its last buffer. The maximum
+height of a buffer is given by $F_{H(k)} = F_{k-2 \lceil log_{\phi} k \rceil}
+$ where $\phi$ is the golden ratio and $k$ is chosen such that $F_k$ =
+$\xi(h)$ and $\xi(x)$ is defined recursively. If $h$ is a Fibonacci number
+then $\xi(h) = h$ and otherwise $\xi(h) = \xi(h-f)$ where f is the largest
+Fibonacci number less than $h$. The maximum height of an individual buffer is
+related to the size of the recursive subtree for which the node is a leaf.
+Since the recursive subtrees are split based on the largest Fibonacci number
+less than $h$ (as described previously) it follows that the buffer sizes are
+also based on Fibonacci numbers.
 
@@
 
-The shuttle tree supports \Search{}, \Insert{}, and \Scan{}. An explicit
-discussion on \Delete{} is missing; it is assumed that it is similar 
-to \Insert{}.
-
-\Search{} and \Scan{} work mostly the same way as the CO B-tree. The only
-difference is that a node's buffer must be searched before following a
-pointer down the tree.
-
-\Insert{} is also similar to the CO B-tree except elements are buffered as they
-are inserted. \Insert{} starts by move down the tree from the root to leaves
-and stopping when a node that has a buffer is located (note that because of the
-nature of Fibonacci numbers that it is possible that a node does not have any
-buffers). The element is inserted into the smallest buffer in the node's linked
-list. If this insertion causes the buffer to overflow (e.g. the height of the
-buffer exceeds its maximum), all of the elements from the buffer are moved into
-the next largest buffer (and so on, recursively).  When the largest buffer
-overflows all of its elements are taken and \textit{shuttled} down the tree.
-Eventually a leaf node will be reached and inserting may cause the balance
+The operations that are described by Bender et al. for the shuttle tree are
+\Search{}, \Insert{}, and \Scan{}. An explicit discussion on \Delete{} is
+missing; however, it is assumed that it is symmetric to \Insert{}.
+
+\Search{} and \Scan{} work virtually the same way as the CO B-tree. The only
+difference is that you have to search a child node's buffer before following
+the pointer down the tree.
+
+\Insert{} is also similar to the CO B-tree except you buffer elements as you
+add them. When you insert an element you move down the tree from the root to
+leaves and stop when you find a node that has a buffer (note that because of
+the nature of Fibonacci numbers that it is possible that a node does not have
+any buffers). You then attempt to insert the element into the smallest buffer
+in the node's linked list. If this insertion causes the buffer to overflow
+(e.g. the height of the buffer exceeds its maximum), you move all of the
+elements from this buffer to the next largest buffer. When the largest buffer
+overflows you take all of its elements and ``shuttle'' them down the tree.
+Eventually you will hit a leaf node and inserting here may cause the balance
 constraints on the SWBST to become invalidated. The shuttle tree rebalances
@@
 split cascades up the tree until balance is restored.
-%% TODO: explain what it means to be "shuttled" down the tree
-
-The shuttle tree also needs to take care of the buffers when a node is split.
-When a node $u$ is split into two nodes $u_1$ and $u_2$, the shuttle tree
-immediately allocates enough space for all of the $u_2$'s buffers ($u_1$ simply
-reuses the space allocated for $u$'s buffers). $u$'s buffers are then scanned
-over (by a single scan) and the elements are placed in the new buffers
-accordingly.
-
-The problem of maintaining room for inserts, bounding the costs of
+
+The shuttle tree also needs to take care of the buffers belonging to a split
+node. When a node $u$ is split into two nodes $u_1$ and $u_2$ the shuttle tree
+immediately allocates enough space for all of the $u_2$'s buffers ($u_1$
+simply reuses the space allocated for $u$'s buffers). We then scan over $u$'s
+buffers and find the ones rooted in $u_1$ and $u_2$ and move them
+respectively. This can be achieved by a single scan.
+
+The core problem of maintaining room for inserts, bounding the costs of
 merging and splitting, and bounding the costs of updating parent and child
 pointers in the shuttle tree is achieved using the same PMA used with the CO
-B-tree. This is described in detail in Section~\ref{sec:layout}.
+B-tree. This is decried in detail in Section~\ref{sec:layout}.
 
@@
 
-\Search{} uses $O(\log_{B}{N})$ memory transfers where $N = \Theta(c^{F_k})$ and
-$c$ is the fanout. This bound is achieved by searching over the
+\Search{} uses $O(log_B N)$ memory transfers where $N = \Theta(c^{F_k})$ and
+$c$ is your fanout. Note that $c$ is not treated as a constant since this is a
+cache-oblivious data structure. This bound is achieved by searching over the
 top and bottom recursive subtrees at each level when traversing the tree. A
 key trick is that when you visit a node you only search its largest buffer
-(the last buffer in the linked list for that level). The smaller buffers will
+(that last buffer in the linked list for that level). The smaller buffers will
 be visited eventually as you recurse down the tree. Eventually you will get
-down to a single subtree that can fit into a single blocks and thus can be
-searched with one memory accesses. Since the fanout is proportional to the
-block size it will take us $O(\log_B N)$ block accesses until we get down to a
-subtree and thus, by induction, the worst case runtime is $O(\log_B N)$.
-
-% Why is fanout propto block-size? doesn't this violate cache-obliviousness
+down to a single subtree that can fit into $O(1)$ blocks and thus can be
+searched with $O(1)$ memory accesses. Since the fanout is proportional to our
+block size it will take us $O(log_B N)$ block accesses until we get down to a
+subtree and thus, by induction, the worst case runtime is $O(log_B N)$.
 
@@
 
-\Scan{} uses an amortized $O(L/B)$ memory transfers to scan $L$ consecutive
-elements. This cost does not include the single \Search{} operation that is
-required to find the first starting element of the range. Once the first
-element's block is transferred, the shuttle tree proceeds by traversing the
-tree and scanning in elements in order. Since the vEB layout enforces the
-spatial locality of nearby nodes in the tree, the shuttle tree only has to read
-in a new block about once every $B$ elements. Thus, each individual memory
-access is is amortized to $1/B$ and the overall number of memory transfers is
+\Scan{} uses an amortized $O(L/B)$ memory transfers to scan in $L$ consecutive
+elements. This cost obviously does not include the single \Search{} operation
+that is required to find the first starting element of the range. Once that
+first element's block is transferred, the shuttle tree proceeds by traversing
+the tree and scanning in elements. Since the vEB layout enforces the spatial
+locality of nearby nodes in the tree, the shuttle tree only has to read in a
+new block once every $B$ memory accesses. Thus, each individual memory access
+is is amortized to $1/B$ and the overall number of memory transfers is
 $O(L/B)$.
@@
 The last thing that is worth mentioning is that the overall space requirements
-of the shuttle tree is $O(N)$, since the majority of the buffers are really
-small and thus do not impact the overall size of the tree.
+of the shuttle tree is $O(n)$. The high level intuition for this proof is that
+the majority of the buffers are really small and thus do not impact the
+overall size of the tree.
 
@@
 
-The cache-oblivious lookahead array (COLA) is quite different from the other
-cache-oblivious data structures discussed so far. Instead of an explicit tree
-structure, the COLA is organized as a list of arrays or levels. Each array has
-size $2^k$, where $k$ is the level of the array ($k = 0$ is the bottom level).
-Furthermore, each array is either completely full or completely empty. As such,
-the COLA is very similar to the binomial list structure described in
-\cite{BentleySaxe}.
+The cache-oblivious lookahead array (COLA) is actually quite different from
+the other cache-oblivious data structures seen so far. Instead of an explicit
+tree structure with child and parent pointers, the COLA is organized as a list
+of arrays or levels. Each array has size $2^k$. Furthermore, each array is
+either completely full or completely empty. As such, the COLA is very similar
+to the binomial list structure described in \cite{BentleySaxe}.
 
@@
 The operations supported by COLA are \Search{}, \Insert{}, and \Scan{}. Once
-again, \Delete{} can be inferred from \Insert{}.
+again, \Delete{} can be inferred symmetrically from \Insert{}.
 
@@
 This array has size $2^0 = 1$, thus the element is in an array on it own.
-Whenever an element is inserted, check to see if there is another array of
+Whenever you insert an element you check to see if there is another array of
 the same size (this can be thought of as checking to see if there is another
@@
 array that is twice the size and thus goes into the next level. The COLA
-recursively merges until there are no two arrays of the same size.
-
-\Search{} can naively be implemented by scanning each of the levels in the
-COLA. Since everything is sorted the COLA can use a binary search to find the
-element it is looking for. However, this is not very efficient. As an
-improvement, COLA, as the name suggests, uses look ahead pointers to reduce the
-number of memory accesses required. Every 8th element from the $(k+1)$-st level
-exists in the $k$-th level array in its proper sorted location. This element
-also contains a pointer to its real location in the $(k+1)$-st level. Thus this
-pointer is referred to as \textit{real lookahead pointer}. Additionally, every
-4th cell in $k$-th level array contains two pointers pointing to the next and
-previous real lookahead pointers on this level. The next and previous pointers
-are referred to as \textit{duplicate lookahead pointers}. To search a level the
-COLA performs a binary search within an appropriate range, determine the range
-in the next level using the duplicate lookahead pointers and then recursively
-cascades up the tree.
+continues to merge until there are no two arrays of the same size.
+
+\Search{} can naively be performed by simply scanning each of the levels in
+the COLA. Since everything is sorted the COLA can use a binary search to
+quickly find the element it is looking for. However, this is not very
+efficient. As an improvement COLA, as the name suggests, uses look ahead
+pointers to reduce the number of memory accesses required. Every 8th element
+from the $(k+1)$st level exists in the $k$ level array in its proper sorted
+location. This element also contains a pointer to its real location in the
+$(k+1)$st level. Thus this pointer is referred to as ``real lookahead
+pointer''. Additionally, every 4th cell in $k$th level array contains two
+pointers pointing to the next and previous ``real lookahead pointers'' on this
+level. The next and previous pointers are referred to as ``duplicate lookahead
+pointers''. To search a level the COLA performs a binary search within an
+appropriate range, determine the range in the next level using the ``duplicate
+lookahead pointers'' and then recursively cascade up the tree.
 
@@
 
-\Insert{} achieves an amortized $O(\log N / B)$ memory transfers. The proof is
-based on the fact that an element will never be merged more than $O(\log{N})$
+\Insert{} achieves an amortized $O(log N / B)$ memory transfers. The proof is
+based on the fact that an element will never be merged more than $O(log(N))$
 times. This is because at that point it will be in the last array.
-Additionally, the first $\log{B}$ merges that a element participates in uses
-$O(1)$ memory transfers since we assume that the relevant blocks are already in
-memory. Thus, we only need to worry about memory transfers when we merge levels
-that have size greater than $B$. If the size of the array is $k$, then the cost
-of merging is $O(k/B)$, and thus the amortized cost of a merging a single
-element is $1/B$. Since an element will only be merged $O(\log{N})$ time, it
-follows that the overall amortized memory transfers to insert an element is
-$O((\log N)/B)$.
-
-\Search{} runs in $O(\log N)$ memory transfers. The proof uses induction to
+Additionally, the first $log(B)$ merges that a element participates in uses a
+constant number of memory transfers since we can assume that the relevant
+blocks are already cached. Thus, we only need to worry about memory transfers
+when we merge levels that have size greater than $B$. If the size of the array
+is $k$ then the cost of merging is $O(k/B)$ and thus the amortized cost of a
+merging a single element is simply $1/B$. Since we just proved that an element
+will only be merged $O(log(N))$ time, it follows that the overall amortized
+memory transfers to insert an element is $O((log N)/B)$.
+
+\Search{} achieves $O(log N)$ memory transfers. The proof uses induction to
 show that the COLA never needs to search more than eight elements on each
-level. Since there are $O(\log N))$ levels and since eight is a constant, it
-follows that overall search uses $O(\log N)$ memory transfers. The fact that
+level. Since there are $O(log(N))$ levels and since eight is a constant, it
+follows that overall, search uses $O(log N)$ memory transfers. The fact that
 only eight elements need to be searched per level is shown through induction.
 The base case is trivial. The first three levels of the COLA each have at most
-eight elements (note that these elements are either real elements or
-lookahead pointers). The COLA can search these levels with no more
+eight elements (note that these elements are either actual real elements or
+lookahead pointers). The COLA can obviously search these levels with no more
 than eight member transfers. As the COLA follows lookahead pointers to the
@@
 find a new $i$ such that $e_i \leq e < e_{i+1}$. Once this $i$ is found the
-duplicate lookahead pointers can be used to locate the \textit{next} and
-\textit{previous} real lookahead pointers in the next level. The two real lookahead
-pointers are used to restrict the search in the next level. Recall that at
-each level in the COLA, every eighth element in a level is present in the
+duplicate lookahead pointers can be used to locate the ``next'' and
+``previous'' real lookahead pointers in the next level. The two real lookahead
+pointers are used to and restrict the search in the next level. Recall that at
+each level in the COLA every eighth element in a level is present in the
 previous level. Thus, it follows that in the next level only 8 elements need
@@
 level, it follows that only 8 elements need to be searched on this level to
-find the appropriate $i$. Thus, overall, the COLA uses $O(8\log{N}) =
-O(\log{N})$ memory transfers.
+find the appropriate $i$. Thus, overall, the COLA uses $O(8log(N)) =
+O(log(N))$ memory transfers.
 
@@
 In experimental results, the COLA achieved significant improvements over
-normal (cache-aware) B-trees, performing $790$ times faster. However this
+normal (cache-aware) B-trees, performing 790 times faster. However this
 improvement comes at the cost of sorted inserts and normal searches. Sorted
-inserts are $3.1$ times slower and searches are $3.5$ times slower. Nevertheless,
-these experimental results indicate that the COLA is a suitable data
-structure when lots of random insertions are common.
+inserts are 3.1 times slower and searches are 3.5 times slower. Nevertheless,
+these experimental results indicate that the the COLA is a suitable data
+structure when random insertions are common.
 
@@
 of its internal state. The straw-man solution is to simply ensure mutual
-exclusion on the entire data structure and use the pre-existing data structures
-without modification; this approach, however, yields poor performance.
+exclusion on the entire data structure; this approach, however, yields poor
+performance.
 
@@
 the second is a CO B-tree based on a packed memory array (PMA). Bender also
-presents a lock-free variant of the latter data structure.
+presents a lock-free variant of the latter data structure. For the remainder
+of the discussion, assume that the keys are fixed-size integer keys.
 
@@
 
-This data structure is based on the strongly weight-balanced exponential tree
-described in Section~\ref{sec:structure}. A tuning parameter $1 < \alpha < 2$
-is chosen for the data structure. Each node in the tree maintains both keys
-which divide up its children, in addition to a \textit{right-link} pointer,
-which contains both a pointer to a node's right sibling, plus the key value of
-the minimum key in the node's right sibling. If no such sibling exists, a
-sentinel $\infty$ pointer is used.
+This data structure is based on a strongly weight-balanced exponential tree
+found in \cite{Bender2002}. A tuning parameter $1 < \alpha < 2$ is chosen for
+the data structure. Each node in the tree maintains both keys which divide up
+its children, in addition to a \textit{right-link} pointer, which contains
+both a pointer to a node's right sibling, plus the key value of the minimum
+key in the node's right sibling. If no such sibling exists, a sentinel
+$\infty$ pointer is used.
 
@@
 \rhyperceil$-leaf tree. The second part of the node consists of a $\lhyperceil
-k \rhyperceil$-size array containing the $k$ elements in sorted order. The leaf
+k \rhyperceil$-size array contained the $k$ elements in sorted order. The leaf
 nodes of the CO search tree point to the associated entries in the array.
@@
 
-Sequential search takes $O(\log_B{N} + \log_{\alpha}{\log{B}})$ block transfers
-(with high probability). The probabilistic bound comes from the probabilistic
-key promotion property. The intuition behind the proof of this bound comes from
-the argument that when the tree has height $h =
-\Omega(\log_{\alpha}{\log{\log{N}}})$, then the total number of keys in the
-tree is $O(2^{\alpha^h})$. The proof is then completed by summing over the
-range of $h$ which is probabilistic $[0, O(\log_{\alpha}{\log{N}})]$, the
-access time from some level $h$ to find the correct child.
-
-Sequential insert takes $O(\log_B{N} + \log{\alpha}{\log{B}})$ block
+Sequential search takes $O(\log_B{N} + \log_{\alpha}{\log_2{B}})$ block
+transfers (with high probability). The probabilistic bound comes from the
+probabilistic key promotion property. The intuition behind the proof of this
+bound comes from the argument that when the tree has height $h =
+\Omega(\log_{\alpha}{\log_2{\log_2{N}}})$, then the total number of keys in
+the tree is $O(2^{\alpha^h})$. The proof is then completed by summing over the
+range of $h$ which is probabilistic $O(\log_{\alpha}{\log_2{N}})$, the access
+time from some level $h$ to find the correct child.
+
+Sequential insert takes $O(\log_B{N} + \log{\alpha}{\log_2{B}})$ block
 transfers (in expectation). This proof is constructed in a similar fashion as
@@
 which indexes into a packed-memory array. Bender introduces a new
-packed-memory array, called a \textit{one-way packed-memory array}, which he
-argues is easier to manage concurrently.
+packed-memory array, called a \textit{one-way packed-memory array}, which they
+argue is easier to manage concurrently.
 
@@
 inserts/deletes, the PMA is rebalanced. A new array is allocated, and the
-elements are spread out evenly over the new array (to minimize element
+elements are spread out evenly over the new array (too minimize element
 density).
 
-Each leaf in the static CO B-tree is a pointer into a $\Theta(\log N)$-size
-region in the packed-memory array.
+Each leaf in the static CO B-tree used is a pointer into a $\Theta(\log
+N)$-size region in the packed-memory array.
 
@@
 since search is a static CO tree lookup followed by a scan, the time bound is
-$O(\log_{B}{N} + ({\log{N}})/{B} + 1)$.
-
-Sequential insert is $O(\log_{B}{N} + (\log^2{N})/{B} + 1)$. The proof is based
-off an accounting argument which places $\Theta((\log^2{m})/B)$ dollars in each
-array slot. The proof is not presented in any detail in the paper, but rather
-the authors argue that it is simply a slightly-modified version of the proof
-presented in \cite{Katriel02}.
+$O(\log_{B}{N} + {\log_2{N}}/{B} + 1)$.
+
+Sequential insert is $O(\log_{B}{N} + \log_2^2{N}/{B} + 1)$. The proof is
+based off an accounting argument which places $\Theta(\log_2^2{m}/B)$ dollars
+in each array slot. The proof is not presented in any detail in the paper, but
+rather it is a slightly-modified version of the proof presented in
+\cite{Katriel02}.
 
@@
 during the PMA modification are replaced by augmenting each element in the PMA
-with a marker. A marker is used to denote the ongoing (concurrent) operation of
-a particular cell. The authors assume that a marker is able to fit within a
-single machine-word (so that the atomic operations are available and
-performant).
+with a marker. A marker is used to denote the ongoing (concurrent) operation
+of a particular cell.
 
@@
 \end{inparaenum}. These primitives are self-explanatory, except they
-all work in a non-blocking manner (they can fail at any point, and it is
-up to the current process to restart the operation on failure).
+all work in a non-blocking manner (they can fail at any point).
 

### 2011-12-12 paper.tex

--- 2011-12-12 paper.tex
@@
 In order for B-trees to remain versatile, they must be able to accommodate
-various forms of data. However, traditional B-trees and string B-trees have
-suboptimal performance when variable length strings are used as keys. In
-addition, string B-trees do not support compression, and commonly used front
-compression is handled sub-optimally. Regular B-trees and string B-trees are
-not cache oblivious, and thus incur poorer performance as described previously.
+various forms of data. Bender et al. assert the following weaknesses in existing string B-tree designs: traditional B-trees have
+suboptimal performance when variable length strings are used as keys\cite{BenderDemainColton}.
+String B-trees\cite{Ferragina98} handle variable-length strings well but do not use compression.
+Front compression\cite{BayerUn77,ClarkSaSt69,Comer79,Wagner73} is used in many B-trees, but is handled sub-optimally\cite{BenderFaKu06}.
+Regular B-trees and string B-trees are
+not cache oblivious, and thus incur poorer performance as described
+previously by Bender et al.\cite{BenderDemainColton}. Range queries have poor performance in B-trees\cite{BenderFaKu06}.
+
 In \cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
-Oblivious String B-tree (COSB-tree), which they argue works efficiently for
-variable length string keys. They also provide a modified front-compression
-scheme, which gives their approach the advantages of using less space and
-having improved locality within the compressed data.
+Oblivious String B-tree (COSB-tree), which they argue works efficiently in
+these problem areas for variable length string keys. They also provide a
+modified front-compression scheme, which gives their approach the advantages
+of using less space and having improved locality within the compressed data.
+
+\subsection{Description}
+Bender et al.
+perform 2 experiments in \cite{BenderFaKu06}, in which a regular B-tree implementation tuned to various block sizes is compared to a static COSB-tree implementation.
+The experiment involved measuring the amount of time required to perform: 440,000 and 450,000 inserts of random values, a range query of all values, and 1000 random searches.
+The results of the experiment show that COSB-trees
+outperformed B-trees most of the time for random searches, range queries and
+random inserts. However, some point between 440,000 and 450,000 inserts the
+COSB-tree had to rebalance the entire tree, causing it to fall behind in this
+measurement. However, Bender et al. argue that the experiment was biased
+towards the B-tree implementation, since the trees are "young", and thus still
+well organized. The COSB-tree implementation is also compared to Berkeley DB
+\cite{BerkeleyDB}, and achieves performance comparable to Berkeley DB with
+tuned parameters. The results seem to show that there is a window in which
+large and small-block size B-trees perform poorly, and that COSB-trees may be
+a reasonable alternative to improve performance.
 

### 2011-12-12 paper.tex
fix the modifications lost during merge

--- 2011-12-12 paper.tex
@@
 In order for B-trees to remain versatile, they must be able to accommodate
-various forms of data. However, traditional B-trees and string B-trees have
-suboptimal performance when variable length strings are used as keys. In
-addition, string B-trees do not support compression, and commonly used front
-compression is handled sub-optimally. Regular B-trees and string B-trees are
-not cache oblivious, and thus incur poorer performance as described previously.
+various forms of data. Bender et al. argue the following weaknesses in
+existing B-tree designs: 
+\begin{inparaenum}[(a)]
+  \item traditional B-trees have suboptimal performance when variable length
+  strings are used as keys \cite{BenderDemainColton},
+  \item string B-trees \cite{Ferragina98} handle variable-length strings well
+  but do not use compression, 
+  \item front compression \cite{BayerUn77,ClarkSaSt69,Comer79,Wagner73} is used
+  in many B-trees, but is handled sub-optimally \cite{BenderFaKu06}. 
+\end{inparaenum}
+
 In \cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
-Oblivious String B-tree (COSB-tree), which they argue works efficiently for
-variable length string keys. They also provide a modified front-compression
-scheme, which gives their approach the advantages of using less space and
-having improved locality within the compressed data.
+Oblivious String B-tree (COSB-tree), which they argue works more efficiently
+for variable length string keys. They also provide a modified front-compression
+scheme, which lets their data structure use less space and have improved
+compressed data locality.
 
@@
 
-\caption{A centroid tree example.
-Dotted lines represent centroid-rooted up tries, where the parent is not a prefix of the child.
-Solid lines represent centroid-rooted down tries, where the parent is a prefix of the child.}
+\caption{A centroid tree example from \cite{BenderFaKu06}. Dotted lines
+represent centroid-rooted up tries, where the parent is not a prefix of the
+child. Solid lines represent centroid-rooted down tries, where the parent is a
+prefix of the child.}
 \label{fig:centroid}
@@
 deletions. The previous modification relies on each key being decoded within
-$c||k||$ elements of the compressed array, but this is no longer
-guaranteed when new keys can be inserted to the left of $k$. To make the
+$c||k||$ elements of the compressed array, but this is no longer guaranteed
+when new keys can be inserted to the left of $k$. To make the
 locality-preserving front compression technique suitable for dynamic
 COSB-trees, Bender et al. modify the algorithm to include extra copied
-prefixes.
-Suppose a key $k$ is being inserted.
-If keys further down in the array will be negatively impacted, the longest common prefix between $k$ and the key right after $k$ is copied to $k$ along with $k$'s original compression data.
-The idea behind adding extra copied prefixes is to maintain $O(||k||)$ decode distances around the area where the new key is being inserted.
-The characters preceding $k$ are charged for the copied prefix, accounting for the work to add the copied prefix to $k$.
+prefixes. Suppose a key $k$ is being inserted. If keys further down in the
+array will be negatively impacted, the longest common prefix between $k$ and
+the key right after $k$ is copied to $k$ along with $k$'s original compression
+data. The idea behind adding extra copied prefixes is to maintain $O(||k||)$
+decode distances around the area where the new key is being inserted. The
+characters preceding $k$ are charged for the copied prefix, accounting for the
+work to add the copied prefix to $k$.
 

### 2011-12-13 paper.tex
Edits

@@
 \usepackage{url}
+\usepackage{amsmath}
 
@@
 
-\caption{An illustration of the van Emde Boas (vEB) layout}
+\caption{An illustration of the recursive van Emde Boas (vEB) layout}
 \label{fig:veb}
@@
 \begin{equation}
+  \label{eq:bounds}
   \frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}
@@
 First, the PMA needs to maintain open space in its memory for newly inserted
-items. The PMA is able to handle an insertion into the SWBST using only $O(1 +
+items. Doing so may seem contradictory: maintaining the locality of elements
+while keeping space available for new inserts. The achieve a balance between
+these extremes the PMA uses a technique
+described in \cite{packedmemoryarray}.
+The PMA is able to handle an insertion into the SWBST using only $O(1 +
 (\log^2{N}) / B)$ memory transfers (this is the cost of updating the memory
 layout of the SWBST in the PMA, not the cost of updating the data structure
-itself). Doing so may seem contradictory: maintaining the locality of elements
-while keeping space available for new inserts. The PMA uses a technique
-described in \cite{packedmemoryarray}. At a high level, the PMA maintains
-windows into the array. When a window becomes too full or too empty, the
+itself). At a high level, the PMA maintains
+``windows'' into the array. When a window becomes too full or too empty, the
 elements are spread out within the context of a larger window. Two parameters
@@
 that each node has $\Theta(c^h)$ descendants (recall, this is the key property
-that SWBST provided). Thus, merging or splitting a node will cause
+that SWBST provided; shown in Equation~\ref{eq:bounds}). Thus, merging or splitting a node will cause
 $\Theta(c^h)$ nodes to move, requiring $O(1 + c^h/B)$ memory transfers. It
@@
 In \cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
-Oblivious String B-tree (COSB-tree), which they argue works more efficiently
-for variable length string keys. They also provide a modified front-compression
-scheme, which lets their data structure use less space and have improved
+Oblivious String B-tree (COSB-tree), which works more efficiently
+for variable length string keys. The COSB-tree uses a modified front-compression
+scheme, which allows it to use less space and have improved
 compressed data locality.
@@
 
-The dynamic COSB-tree consists of three major components: a centroid tree at
-the top, a PMA structure of hash data for the keys known as the \emph{hashdata}
-layer; and on the bottom a PMA structure for the original key values,
-compressed using an augmented version of forward compression known as the
-\emph{keydata} layer. Each leaf of the centroid tree contains a pointer into
+The dynamic COSB-tree consists of three layers: a centroid tree, a PMA 
+of hash data for the keys (known as the 
+\emph{hashdata} layer), and a PMA of compressed values for the original key
+ (known as the 
+\emph{keydata} layer). Each leaf of the centroid tree contains a pointer into
 the hashdata array. The centroid tree only contains a subset of the keys, so
@@
 entry in the hashdata array contains the hash value of the corresponding key,
-and a pointer to the compressed key value in the keydata layer.
+and a pointer to the compressed key value in the keydata layer. The keydata layer
+is consulted to resolve hash collisions in the hashdata layer.
+
+%a centroid tree at
+%the top, a PMA structure of hash data for the keys known as the \emph{hashdata}
+%layer; and on the bottom a PMA structure for the original key values,
+%compressed using an augmented version of forward compression known as the
+%\emph{keydata} layer. Each leaf of the centroid tree contains a pointer into
+%the hashdata array. The centroid tree only contains a subset of the keys, so
+%each leaf points to the respective \emph{neighborhood} of the key being
+%searched for. After traversing the centroid tree and following the pointer into
+%the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
+%sequential scan is performed in the hashdata array layer to find $k$. Each
+%entry in the hashdata array contains the hash value of the corresponding key,
+%and a pointer to the compressed key value in the keydata layer.
 
@@
 
-\subsubsection{The Dictionary Matching Problem and Centroid Trees} %todo:cite dictionary matching problem--sources in cosb-tree paper
+\subsubsection{Centroid Trees} %todo:cite dictionary matching problem--sources in cosb-tree paper
 \begin{figure}
@@
 
-The centroid tree in the COSB-tree is layed out in memory using a modified van
-Emde Boas (vEB) layout for weight-balanced binary trees. The
+The centroid tree in the COSB-tree is layed out in memory using a modified vEB
+(see Figure~\ref{fig:veb}) layout for weight-balanced binary trees. The
 \emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into
-equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x is
-denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to
-the nearest power of a power of 2. This ensures trees of equal size on all
-\emph{levels of detail}, as the centroid tree is traversed.
+equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x refers
+to rounding x down to the nearest power of a power of 2. This ensures trees
+of equal size on all \emph{levels of detail}, as the centroid tree is traversed.
+
+********* LEILANI: Can you provide a 5 word description of levels of details ***
+
+%The \emph{hyperhyperfloor} of x is
+%denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to
+%the nearest power of a power of 2. This ensures trees of equal size on all
+%\emph{levels of detail}, as the centroid tree is traversed.
 % Should explain what "level of detail" means here
@@
 $O(1+(\log^{2}{N})/B)$ memory transfers.
+
+********* LEILANI: What is a centroid vertex? ***
 
@@
 
-\subsubsection{Modifying Front Compression to Preserve Locality}
+\subsubsection{Modifying Front Compression}
 
@@
 
-\subsection{Operators}
+\subsection{Operations}
 
@@
 improved amortized $O(\log{\frac{N}{B}})$ memory transfers for inserts compared
-to the $O(\log_{B}{N})$ memory access of B-trees. In practice, log-structured
+to the $O(\log_{B}{N})$ memory access of B-trees \cite{Buchsbaum}. In practice, log-structured
 merge (LSM) trees \cite{ONeil96} are used for high-insert workloads.
@@
 
-\subsubsection{Operators}
+\subsubsection{Operations}
 
@@
 down to a single subtree that can fit into a single blocks and thus can be
-searched with one memory accesses. Since the fanout is proportional to the
+searched with one memory transfer. Since the fanout is proportional to the
 block size it will take us $O(\log_B N)$ block accesses until we get down to a

### 2011-12-13 paper.tex
Edits
addressed Salman's comments on the co string b-tree section. moved co streaming b-tree section to be first.

@@
 
-\section{Variable Length Strings}
-\begin{figure}
-
-\begin{center}
-	\includegraphics[width=0.8\columnwidth]{figures/COSB-treegeneralstructurediagram.pdf}
-\end{center}
-
-\caption{An illustration of the dynamic COSB-tree structure}
-\label{fig:string_structure}
-\end{figure}
-\subsection{Motivation}
-
-In order for B-trees to remain versatile, they must be able to accommodate
-various forms of data. Bender et al. argue the following weaknesses in
-existing B-tree designs: 
-\begin{inparaenum}[(a)]
-  \item traditional B-trees have suboptimal performance when variable length
-  strings are used as keys \cite{BenderDemainColton},
-  \item string B-trees \cite{Ferragina98} handle variable-length strings well
-  but do not use compression, 
-  \item front compression \cite{BayerUn77,ClarkSaSt69,Comer79,Wagner73} is used
-  in many B-trees, but is handled sub-optimally \cite{BenderFaKu06}. 
-\end{inparaenum}
-
-In \cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
-Oblivious String B-tree (COSB-tree), which works more efficiently
-for variable length string keys. The COSB-tree uses a modified front-compression
-scheme, which allows it to use less space and have improved
-compressed data locality.
-
-\subsubsection{General COSB-tree Structure}
-
-The dynamic COSB-tree consists of three layers: a centroid tree, a PMA 
-of hash data for the keys (known as the 
-\emph{hashdata} layer), and a PMA of compressed values for the original key
- (known as the 
-\emph{keydata} layer). Each leaf of the centroid tree contains a pointer into
-the hashdata array. The centroid tree only contains a subset of the keys, so
-each leaf points to the respective \emph{neighborhood} of the key being
-searched for. After traversing the centroid tree and following the pointer into
-the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
-sequential scan is performed in the hashdata array layer to find $k$. Each
-entry in the hashdata array contains the hash value of the corresponding key,
-and a pointer to the compressed key value in the keydata layer. The keydata layer
-is consulted to resolve hash collisions in the hashdata layer.
-
-%a centroid tree at
-%the top, a PMA structure of hash data for the keys known as the \emph{hashdata}
-%layer; and on the bottom a PMA structure for the original key values,
-%compressed using an augmented version of forward compression known as the
-%\emph{keydata} layer. Each leaf of the centroid tree contains a pointer into
-%the hashdata array. The centroid tree only contains a subset of the keys, so
-%each leaf points to the respective \emph{neighborhood} of the key being
-%searched for. After traversing the centroid tree and following the pointer into
-%the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
-%sequential scan is performed in the hashdata array layer to find $k$. Each
-%entry in the hashdata array contains the hash value of the corresponding key,
-%and a pointer to the compressed key value in the keydata layer.
-
-In the following sections, centroid trees are described, and a summary of
-how to modify them to preserve balance and structure in the dynamic case is
-provided. Two iterations on forward compression are also summarized to show
-how to incrementally add support for variable length strings as keys for
-dynamic COSB-trees.
-
-\subsubsection{Centroid Trees} %todo:cite dictionary matching problem--sources in cosb-tree paper
-\begin{figure}
-
-\begin{center}
-	\includegraphics[width=0.8\columnwidth]{figures/centroidtreeexample.pdf}
-\end{center}
-
-\caption{A centroid tree example from \cite{BenderFaKu06}. Dotted lines
-represent centroid-rooted up tries, where the parent is not a prefix of the
-child. Solid lines represent centroid-rooted down tries, where the parent is a
-prefix of the child.}
-\label{fig:centroid}
-\end{figure}
-
-COSB-trees are a solution to the \textit{dictionary matching problem}.
-Dictionary matching, or string matching, refers to matching strings using
-prefix search. Let $D$ be a dictionary of $N$ keys. To perform faster searches,
-the keys in $D$ are preprocessed using hashing and divide-and-conquer via
-centroid trees. Figure \ref{fig:centroid} depicts an example centroid tree.
-
-Consider the compacted\footnote{Compacted means all non-branching paths are
-reduced to edges.} trie $T$ constructed over $D$ for string matching.  Consider
-the vertex $\rho$ in $T$ such that $\rho$ has between N/3 and 2N/3 descendants,
-inclusive. Call $\rho$ the \emph{centroid vertex} of $T$. When searching for a
-key $k$ in the compacted trie, you traverse the trie as follows: for each node
-$y$ that you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on
-into the trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the
-trie resulting from excluding $y$ and its subtree. The \emph{centroid tree} is
-obtained by rooting the tree at the centroid vertex $\rho$, and making its
-children recursively defined centroid trees of its up and down tries.
-
-The centroid tree in the COSB-tree is layed out in memory using a modified vEB
-(see Figure~\ref{fig:veb}) layout for weight-balanced binary trees. The
-\emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into
-equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x refers
-to rounding x down to the nearest power of a power of 2. This ensures trees
-of equal size on all \emph{levels of detail}, as the centroid tree is traversed.
-
-********* LEILANI: Can you provide a 5 word description of levels of details ***
-
-%The \emph{hyperhyperfloor} of x is
-%denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to
-%the nearest power of a power of 2. This ensures trees of equal size on all
-%\emph{levels of detail}, as the centroid tree is traversed.
-% Should explain what "level of detail" means here
-
-The following approach is used to rebalance the centroid tree: put all the
-elements in DFS or Euler tour order for the compacted trie. Given this order,
-scan the trie to identify the centroid vertex. Then partition the trie into its
-respective up and down tries, and recurse on the up and down tries. The Euler
-tour order can be found in $O(\log{N})$ scans, since the trie is only
-$O(\log{N})$ deep. Thus these updates have amortized cost of
-$O(1+(\log^{2}{N})/B)$ memory transfers.
-
-********* LEILANI: What is a centroid vertex? ***
-
-The following approach is used for maintaining the vEB layout of the centroid
-tree in the COSB-tree: first, weight balance is relaxed from a maximum weight
-difference of 2 between left and right subtrees, to a larger constant $c$. Thus
-a node $v$ only becomes out of balance every $\Omega(\textsc{weight(v)})$
-insertions or deletions of descendants of $v$. The nodes are stored in a PMA,
-preserving order of the nodes using $\Theta(N)$ space.
-
-On insert or delete, the area around the update point is scanned for sections
-of the array to rearrange. If a section around the update point reaches a
-certain \textit{density} threshold, the nodes are evenly distributed across the
-section to make room for more nodes. To ensure that updating pointers is not too
-expensive, any node included in the section must have all of its subtrees
-included. The subtrees are dictated by the vEB layout, and subtrees are chosen
-from the bottom up. The smallest neighboring trees (single nodes) in the vEB
-layout are added first, and larger trees are included when all of their
-subtrees have been added to the section. Root trees are only added after all
-bottom trees are added (in terms of the vEB layout). Figure
-\ref{fig:centroid_relayout} gives an example of how subtrees are chosen for
-rearranging.
-
-This approach results in an amortized update cost of $O(1+(\log{N})/B)$ memory
-transfers per update. In conjunction with the cost of rebuilds described above,
-this is only an additive $O((\log{N})/B)$ factor, so the overall bound is
-amortized $O(1+(\log^{2}N)/B)$ memory transfers.
-\begin{figure}
-
-\begin{center}
-	\includegraphics[width=0.8\columnwidth]{figures/centroidrelayoutexample.pdf}
-\end{center}
-
-\caption{A centroid layout example. Assume the smallest subtrees in this
-example only contain a single node. Suppose a node was inserted into a leaf
-position in the bottom left of the tree. The numbers represent the order in
-which each subtree is added to the section to be rearranged.}
-\label{fig:centroid_relayout}
-\end{figure}
-
-\subsubsection{Modifying Front Compression}
-
-This section provides a brief summary of how front compression is modified to
-optimize compression and decoding for dynamic COSB-trees. Consider a sequence
-of $j$ keys to store $k_{1},k_{2},...,k_{j}$. Without compression, this
-requires $\sum_{j}||k_{j}||$ space in memory. To compress these keys using
-standard front compression, let $\pi_{j}$ be the longest common prefix between
-$k_{j}$ and $k_{j+1}$. Now instead we store the keys in the following format:
-\begin{equation}
-  k_{1}||\pi_{2}||\sigma_{2}||\pi_{3}||\sigma_{3}||...||\pi_{j}||\sigma_{j}
-\end{equation} 
-where $\sigma_{j}$ is the suffix of $k_{j}$ after removing the first $\pi_{j}$
-bits. This saves almost $\sum_{j}||\pi_{j}||$ memory. However, since decoding
-$k_{j}$ involves concatenating the first $\pi_{j}$ bits of $k_{j-1}$ to
-$\sigma_{j}$, $k_{j-1}$ must be decoded. $k_{j-1}$ may depend on previous keys
-as well, making $k_{j}$ expensive to decode.
-
-Bender et al. provide a modification that reduces the high decoding cost of
-front compressing $D$ to $O(1+||k||B)$, while maintaining a compressed size of
-$O(\llangle{D}\rrangle{})$\footnote{$\llangle{D}\rrangle{}$ is shorthand for
-the size of $D$ after front compression} after compression. The modification
-works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be the first key
-added to the compression stream. For each subsequent key $k_{i}$, if $k_{i}$
-can be decoded with just the last $c||k_{i}||$ characters, add
-$\pi_{i}$, $\sigma_{i}$ as described above. Otherwise, add $k_{i}$ to the stream
-without compressing it. By construction, this scheme guarantees that any key
-$k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.
-
-However, the above modification falls short for anticipating insertions and
-deletions. The previous modification relies on each key being decoded within
-$c||k||$ elements of the compressed array, but this is no longer guaranteed
-when new keys can be inserted to the left of $k$. To make the
-locality-preserving front compression technique suitable for dynamic
-COSB-trees, Bender et al. modify the algorithm to include extra copied
-prefixes. Suppose a key $k$ is being inserted. If keys further down in the
-array will be negatively impacted, the longest common prefix between $k$ and
-the key right after $k$ is copied to $k$ along with $k$'s original compression
-data. The idea behind adding extra copied prefixes is to maintain $O(||k||)$
-decode distances around the area where the new key is being inserted. The
-characters preceding $k$ are charged for the copied prefix, accounting for the
-work to add the copied prefix to $k$.
-
-\subsection{Operations}
-
-%todo: figure out how to do range queries from this paper somehow...
-
-The operations supported by the COSB-tree are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{},
-and \Scan{}. The behavior of each operation for dynamic COSB-trees is
-described below. \Insertkonly{} and \Delete{} are also supported but not
-described here.
-
-\Search{} works as follows for a given key $k$. Compute the hash value for
-$k$ and call it $\rho$, and traverse the centroid tree of the COSB-tree using
-$\rho$ to obtain the successor $\rho_{s}$ and predecessor $\rho_{p}$ of $k$.
-Index into the hashdata layer and perform a sequential scan across the PMA
-structure from $\rho_{p}$ to $\rho_{s}$ to find $\rho$. On a match with value
-$\rho'$, index into the keydata layer and decode $\rho'$ to obtain the
-resulting key $k'$. If $k' = k$, return true. If $k' \ne k$ or $k$ was not
-found during the sequential scan return false.
-
-\Pred{}, \Succ{} work in the same way as \Search{} until the sequential scan in
-the hashdata layer. When scanning the hashdata layer, look for $\rho_{p}$ or
-$\rho_{s}$, respectively, in place of $\rho$, and compare the appropriate keys
-in the keydata layer.
-
-\Scan{}: This operation is not explicitly described in \cite{BenderFaKu06}, but
-presumably this is very similar to \Search{}. Let $k_{b}$ be the beginning of
-the range and let $k_{e}$ be the end of the range. Hash the range keys and
-identify the predecessor of $\rho_{b}$ and the successor of $\rho_{e}$ using
-the centroid tree. Scan the hashdata layer across from $\rho_{b,p}$ to
-$\rho_{e,s}$. Only start noting the beginning of range when the first key with
-hash value after $\rho_{b}$, and stop once we reach the first key with hash
-value after $\rho_{e}$. Decode and return the set of keys within the given
-range in the keydata layer.
-
-\subsection{Complexity}
-The keys are hashed and hash values are compared in place of the original key
-values to speed up comparisons in the centroid tree and hashdata layer. To
-avoid mismatches due to hash collisions, the original key values are compared
-only once at the bottom layer, resulting an additive $O(||k||/B)$ memory
-transfers in each of the operators for each key $k$ that needs to be compared.
-
-\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high probability
-(w.h.p.).  The centroid tree is stored in a PMA with a dynamic CO layout. This
-layout reduces the number of memory transfers required when traversing the tree
-to $O(\log_{B}{N})$.  Once in the hashdata layer, we are already within $O(\log{N})$
-keys of $k$, which require $O((\log{N})/B)$ memory transfers to scan. Decoding
-and comparing the original key values requires $O(1+||k||/B)$ memory transfers,
-due to the modified front compression scheme.
-
-The \Pred{} and \Succ{} operations are similar in behavior to searches, and
-thus only have the added cost of having to operate on $k'$, where $k'$ is the
-predecessor or successor, respectively. These operations as a result take
-$O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers (w.h.p.). Similarly, range
-queries take $O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory
-transfers, where in this case $k'$ is the other end of the range and $Q$ is the
-set of keys within the given range.
-
-\Insertkonly{} and \Delete{} w.h.p. require $O(1+\log_{B}N+\log^{2}N||k||/B)$
-memory transfers. As mentioned previously, due to rebuilds updates have
-amortized cost of $O(1+(\log^{2}N)/B)$ memory transfers. The rest of the cost
-for insertions and deletions is due to having to find the correct key to
-delete or location for insert, which can be seen from the cost of \Search{}.
-
-\subsection{Experimental Results}
-Bender et al. perform two experiments in \cite{BenderFaKu06}, in which a
-regular B-tree implementation tuned to various block sizes is compared to a
-static COSB-tree implementation. The experiment involved measuring the amount
-of time required to perform:
-\begin{inparaenum}[(a)]
-  \item 440,000 and 450,000 inserts of random values, 
-  \item a range query of all values, and 
-  \item 1000 random searches.
-\end{inparaenum} The results of the
-experiment show that COSB-trees outperformed B-trees most of the time for
-random searches, range queries and random inserts. However, some point between
-440,000 and 450,000 inserts the COSB-tree had to rebalance the entire tree,
-causing it to fall behind in this measurement. However, Bender et al. argue
-that the experiment was biased towards the B-tree implementation, since the
-trees are ``young'', and thus still well organized. The COSB-tree
-implementation is also compared to Berkeley DB \cite{BerkeleyDB}, and achieves
-performance comparable to Berkeley DB with tuned parameters. The results seem
-to show that there is a window in which large and small-block size B-trees
-perform poorly, and that COSB-trees may be a reasonable alternative to improve
-performance.
-
 \section{Streaming B-trees}
@@
 structure when lots of random insertions are common.
+
+\section{Variable Length Strings}
+\begin{figure}
+
+\begin{center}
+	\includegraphics[width=0.8\columnwidth]{figures/COSB-treegeneralstructurediagram.pdf}
+\end{center}
+
+\caption{An illustration of the dynamic COSB-tree structure}
+\label{fig:string_structure}
+\end{figure}
+\subsection{Motivation}
+
+In order for B-trees to remain versatile, they must be able to accommodate
+various forms of data. Bender et al. argue the following weaknesses in
+existing B-tree designs: 
+\begin{inparaenum}[(a)]
+  \item traditional B-trees have suboptimal performance when variable length
+  strings are used as keys \cite{BenderDemainColton},
+  \item string B-trees \cite{Ferragina98} handle variable-length strings well
+  but do not use compression, 
+  \item front compression \cite{BayerUn77,ClarkSaSt69,Comer79,Wagner73} is used
+  in many B-trees, but is handled sub-optimally \cite{BenderFaKu06}. 
+\end{inparaenum}
+
+In \cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
+Oblivious String B-tree (COSB-tree), which works more efficiently
+for variable length string keys. The COSB-tree uses a modified front-compression
+scheme, which allows it to use less space and have improved
+compressed data locality.
+
+\subsubsection{General COSB-tree Structure}
+
+The dynamic COSB-tree consists of three layers: a centroid tree, a PMA 
+of hash data for the keys (known as the 
+\emph{hashdata} layer), and a PMA of compressed values for the original key
+ (known as the 
+\emph{keydata} layer). Each leaf of the centroid tree contains a pointer into
+the hashdata array. The centroid tree contains only hash values, and is designed such that
+each leaf points to the respective \emph{neighborhood} of the key being
+searched for. After traversing the centroid tree and following the pointer into
+the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
+sequential scan is performed in the hashdata array layer to find $k$. Each
+entry in the hashdata array contains only the hash value of the corresponding key,
+and a pointer to the compressed key value in the keydata layer. The keydata layer
+is consulted to resolve hash collisions in the hashdata layer, and to obtain the original value of the key.
+
+%a centroid tree at
+%the top, a PMA structure of hash data for the keys known as the \emph{hashdata}
+%layer; and on the bottom a PMA structure for the original key values,
+%compressed using an augmented version of forward compression known as the
+%\emph{keydata} layer. Each leaf of the centroid tree contains a pointer into
+%the hashdata array. The centroid tree only contains a subset of the keys, so
+%each leaf points to the respective \emph{neighborhood} of the key being
+%searched for. After traversing the centroid tree and following the pointer into
+%the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
+%sequential scan is performed in the hashdata array layer to find $k$. Each
+%entry in the hashdata array contains the hash value of the corresponding key,
+%and a pointer to the compressed key value in the keydata layer.
+
+In the following sections, centroid trees are described, and a summary of
+how to modify them to preserve balance and structure in the dynamic case is
+provided. Two iterations on forward compression are also summarized to show
+how to incrementally add support for variable length strings as keys for
+dynamic COSB-trees.
+
+\subsubsection{Centroid Trees} %todo:cite dictionary matching problem--sources in cosb-tree paper
+\begin{figure}
+
+\begin{center}
+	\includegraphics[width=0.8\columnwidth]{figures/centroidtreeexample.pdf}
+\end{center}
+
+\caption{A centroid tree example from \cite{BenderFaKu06}. Dotted lines
+represent centroid-rooted up tries, where the parent is not a prefix of the
+child. Solid lines represent centroid-rooted down tries, where the parent is a
+prefix of the child.}
+\label{fig:centroid}
+\end{figure}
+
+COSB-trees are a solution to the \textit{dictionary matching problem}.
+Dictionary matching, or string matching, refers to matching strings using
+prefix search. Let $D$ be a dictionary of $N$ keys. To perform faster searches,
+the keys in $D$ are preprocessed using hashing and divide-and-conquer via
+centroid trees. Figure \ref{fig:centroid} depicts an example centroid tree.
+
+Consider the compacted\footnote{Compacted means all non-branching paths are
+reduced to edges.} trie $T$ constructed over $D$ for string matching. There exists a
+vertex $\rho$ in $T$ such that $\rho$ has between N/3 and 2N/3 descendants,
+inclusive. Call $\rho$ the \emph{centroid vertex} of $T$. When searching for a
+key $k$ in the compacted trie, you traverse the trie as follows: for each node
+$y$ that you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on
+into the trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the
+trie resulting from excluding $y$ and its subtree, and call this the \emph{up trie} of $y$. The \emph{centroid tree} is
+obtained by rooting the tree at the centroid vertex $\rho$, and making its
+children recursively defined centroid trees of its up and down tries.
+
+The centroid tree in the COSB-tree is layed out in memory using a modified vEB
+(see Figure~\ref{fig:veb}) layout for weight-balanced binary trees. Instead of using height $h/2$, the
+\emph{hyperhyperfloors} of the weights of nodes are used as a guide for cutting the tree into
+equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x refers
+to rounding x down to the nearest power of a power of 2.
+
+%The \emph{hyperhyperfloor} of x is
+%denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to
+%the nearest power of a power of 2. This ensures trees of equal size on all
+%\emph{levels of detail}, as the centroid tree is traversed.
+% Should explain what "level of detail" means here
+
+The following approach is used to rebalance the centroid tree: put all the
+elements in DFS or Euler tour order for the compacted trie. Given this order,
+scan the trie to identify the centroid vertex. Then partition the trie into its
+respective up and down tries, and recurse on the up and down tries. The Euler
+tour order can be found in $O(\log{N})$ scans, since the trie is only
+$O(\log{N})$ deep. Thus these updates have amortized cost of
+$O(1+(\log^{2}{N})/B)$ memory transfers.
+
+The following approach is used for maintaining the vEB layout of the centroid
+tree in the COSB-tree: first, weight balance is relaxed from a maximum weight
+difference of 2 between left and right subtrees, to a larger constant $c$. Thus
+a node $v$ only becomes out of balance every $\Omega(\textsc{weight(v)})$
+insertions or deletions of descendants of $v$. The nodes are stored in a PMA,
+preserving order of the nodes using $\Theta(N)$ space.
+
+On insert or delete, the area around the update point is scanned for sections
+of the array to rearrange. If a section around the update point reaches a
+certain \textit{density} threshold, the nodes are evenly distributed across the
+section to make room for more nodes. To ensure that updating pointers is not too
+expensive, any node included in the section must have all of its subtrees
+included. The subtrees are dictated by the vEB layout, and subtrees are chosen
+from the bottom up. The smallest neighboring trees (single nodes) in the vEB
+layout are added first, and larger trees are included when all of their
+subtrees have been added to the section. Root trees are only added after all
+bottom trees are added (in terms of the vEB layout). Figure
+\ref{fig:centroid_relayout} gives an example of how subtrees are chosen for
+rearranging.
+
+This approach results in an amortized update cost of $O(1+(\log{N})/B)$ memory
+transfers per update. In conjunction with the cost of rebuilds described above,
+this is only an additive $O((\log{N})/B)$ factor, so the overall bound is
+amortized $O(1+(\log^{2}N)/B)$ memory transfers.
+\begin{figure}
+
+\begin{center}
+	\includegraphics[width=0.8\columnwidth]{figures/centroidrelayoutexample.pdf}
+\end{center}
+
+\caption{A centroid layout example. Assume the smallest subtrees in this
+example only contain a single node. Suppose a node was inserted into a leaf
+position in the bottom left of the tree. The numbers represent the order in
+which each subtree is added to the section to be rearranged.}
+\label{fig:centroid_relayout}
+\end{figure}
+
+\subsubsection{Modifying Front Compression}
+
+This section provides a brief summary of how front compression is modified to
+optimize compression and decoding for dynamic COSB-trees. This approach 
+increases the locality of the compressed data. Consider a sequence
+of $j$ keys to store $k_{1},k_{2},...,k_{j}$. Without compression, this
+requires $\sum_{j}||k_{j}||$ space in memory. To compress these keys using
+standard front compression, let $\pi_{j}$ be the longest common prefix between
+$k_{j}$ and $k_{j+1}$. Now instead we store the keys in the following format:
+\begin{equation}
+  k_{1}||\pi_{2}||\sigma_{2}||\pi_{3}||\sigma_{3}||...||\pi_{j}||\sigma_{j}
+\end{equation} 
+where $\sigma_{j}$ is the suffix of $k_{j}$ after removing the first $\pi_{j}$
+bits. This saves almost $\sum_{j}||\pi_{j}||$ memory. However, since decoding
+$k_{j}$ involves concatenating the first $\pi_{j}$ bits of $k_{j-1}$ to
+$\sigma_{j}$, $k_{j-1}$ must be decoded. $k_{j-1}$ may depend on previous keys
+as well, making $k_{j}$ expensive to decode.
+
+Bender et al. provide a modification that reduces the high decoding cost of
+front compressing $D$ to $O(1+||k||B)$, while maintaining a compressed size of
+$O(\llangle{D}\rrangle{})$\footnote{$\llangle{D}\rrangle{}$ is shorthand for
+the size of $D$ after front compression} after compression. The modification
+works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be the first key
+added to the compression stream. For each subsequent key $k_{i}$, if $k_{i}$
+can be decoded with just the last $c||k_{i}||$ characters, add
+$\pi_{i}$, $\sigma_{i}$ as described above. Otherwise, add $k_{i}$ to the stream
+without compressing it. By construction, this scheme guarantees that any key
+$k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.
+
+However, the above modification falls short for anticipating insertions and
+deletions. The previous modification relies on each key being decoded within
+$c||k||$ elements of the compressed array, but this is no longer guaranteed
+when new keys can be inserted to the left of $k$. To make the
+locality-preserving front compression technique suitable for dynamic
+COSB-trees, Bender et al. modify the algorithm to include extra copied
+prefixes. Suppose a key $k$ is being inserted. If keys further down in the
+array will be negatively impacted, the longest common prefix between $k$ and
+the key right after $k$ is copied to $k$ along with $k$'s original compression
+data. The idea behind adding extra copied prefixes is to maintain $O(||k||)$
+decode distances around the area where the new key is being inserted. The
+characters preceding $k$ are charged for the copied prefix, accounting for the
+work to add the copied prefix to $k$.
+
+\subsection{Operations}
+
+%todo: figure out how to do range queries from this paper somehow...
+
+The operations supported by the COSB-tree are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{},
+and \Scan{}. The behavior of each operation for dynamic COSB-trees is
+described below. \Insertkonly{} and \Delete{} are also supported but not
+described here.
+
+\Search{} works as follows for a given key $k$. Compute the hash value for
+$k$ and call it $\rho$, and traverse the centroid tree of the COSB-tree using
+$\rho$ to obtain the successor $\rho_{s}$ and predecessor $\rho_{p}$ of $k$.
+Index into the hashdata layer and perform a sequential scan across the PMA
+structure from $\rho_{p}$ to $\rho_{s}$ to find $\rho$. On a match with value
+$\rho'$, index into the keydata layer and decode $\rho'$ to obtain the
+resulting key $k'$. If $k' = k$, return true. If $k' \ne k$ or $k$ was not
+found during the sequential scan return false.
+
+\Pred{}, \Succ{} work in the same way as \Search{} until the sequential scan in
+the hashdata layer. When scanning the hashdata layer, look for $\rho_{p}$ or
+$\rho_{s}$, respectively, in place of $\rho$, and compare the appropriate keys
+in the keydata layer.
+
+\Scan{}: This operation is not explicitly described in \cite{BenderFaKu06}, but
+presumably this is very similar to \Search{}. Let $k_{b}$ be the beginning of
+the range and let $k_{e}$ be the end of the range. Hash the range keys and
+identify the predecessor of $\rho_{b}$ and the successor of $\rho_{e}$ using
+the centroid tree. Scan the hashdata layer across from $\rho_{b,p}$ to
+$\rho_{e,s}$. Only start noting the beginning of range when the first key with
+hash value after $\rho_{b}$, and stop once we reach the first key with hash
+value after $\rho_{e}$. Decode and return the set of keys within the given
+range in the keydata layer.
+
+\subsection{Complexity}
+The keys are hashed and hash values are compared in place of the original key
+values to speed up comparisons in the centroid tree and hashdata layer. To
+avoid mismatches due to hash collisions, the original key values are compared
+only once at the bottom layer, resulting an additive $O(||k||/B)$ memory
+transfers in each of the operators for each key $k$ that needs to be compared.
+
+\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high probability
+(w.h.p.).  The centroid tree is stored in a PMA with a dynamic CO layout. This
+layout reduces the number of memory transfers required when traversing the tree
+to $O(\log_{B}{N})$.  Once in the hashdata layer, we are already within $O(\log{N})$
+keys of $k$, which require $O((\log{N})/B)$ memory transfers to scan. Decoding
+and comparing the original key values requires $O(1+||k||/B)$ memory transfers,
+due to the modified front compression scheme.
+
+The \Pred{} and \Succ{} operations are similar in behavior to searches, and
+thus only have the added cost of having to operate on $k'$, where $k'$ is the
+predecessor or successor, respectively. These operations as a result take
+$O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers (w.h.p.). Similarly, range
+queries take $O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory
+transfers, where in this case $k'$ is the other end of the range and $Q$ is the
+set of keys within the given range.
+
+\Insertkonly{} and \Delete{} w.h.p. require $O(1+\log_{B}N+\log^{2}N||k||/B)$
+memory transfers. As mentioned previously, due to rebuilds updates have
+amortized cost of $O(1+(\log^{2}N)/B)$ memory transfers. The rest of the cost
+for insertions and deletions is due to having to find the correct key to
+delete or location for insert, which can be seen from the cost of \Search{}.
+
+\subsection{Experimental Results}
+Bender et al. perform two experiments in \cite{BenderFaKu06}, in which a
+regular B-tree implementation tuned to various block sizes is compared to a
+static COSB-tree implementation. The experiment involved measuring the amount
+of time required to perform:
+\begin{inparaenum}[(a)]
+  \item 440,000 and 450,000 inserts of random values, 
+  \item a range query of all values, and 
+  \item 1000 random searches.
+\end{inparaenum} The results of the
+experiment show that COSB-trees outperformed B-trees most of the time for
+random searches, range queries and random inserts. However, some point between
+440,000 and 450,000 inserts the COSB-tree had to rebalance the entire tree,
+causing it to fall behind in this measurement. However, Bender et al. argue
+that the experiment was biased towards the B-tree implementation, since the
+trees are ``young'', and thus still well organized. The COSB-tree
+implementation is also compared to Berkeley DB \cite{BerkeleyDB}, and achieves
+performance comparable to Berkeley DB with tuned parameters. The results seem
+to show that there is a window in which large and small-block size B-trees
+perform poorly, and that COSB-trees may be a reasonable alternative to improve
+performance.
 

### 2011-12-14 paper.tex
Adding new figures
did my leg of the edits. Salman please skim for typos if you can. I found a few and may have missed/created some

@@
 This paper presents a survey of cache-oblivious B-trees (CO B-trees). The CO
-B-tree was first introduced by Bender et al \cite{BenderDemainColton}. It is
-an external memory data structure; it optimizes the number of memory transfers
-performed rather than the number of CPU instructions that executed. Unlike
-other external memory data structures such as the widely used B-tree, the CO
-B-tree does not have access to the parameter's of the machine's memory
-hierarchy, including the block size/capacity at each level, or the relative
+B-tree was first introduced by Bender et al. \cite{BenderDemainColton}. It is
+an external memory data structure, and thus it optimizes the number of memory transfers
+performed rather than the number of CPU instructions executed. Unlike
+other external memory data structures, like the widely used B-tree, the CO
+B-tree does not have access to the parameters of the machine's memory
+hierarchy, such as the block size/capacity at each level, or the relative
 speeds. Hence, the CO B-tree is \textit{cache-oblivious}.  Despite this, the CO
-B-tree is able to achieve the same provably optimal search bound as the B-tree
+B-tree achieves the same provably optimal search bound as the B-tree
 and near optimal insertion and deletion bounds. In this paper we first describe
-the original CO B-tree and explain how it is able to achieve its performance.
+the original CO B-tree and explain how it achieves its performance bounds.
 We then present and analyze three followup CO B-tree data structures that
 improve upon weaknesses of the original: the \textit{cache-oblivious string
-B-tree}, which introduces support for variable length keys, the
+B-tree}, which introduces support for variable length keys; the
 \textit{cache-oblivious streaming B-tree}, which improves the insertion bound
-at the expense of the search bound, and lastly, the \textit{cache-oblivious
+at the expense of the search bound; and the \textit{cache-oblivious
 concurrent B-tree}, which introduces support for concurrency.
@@
 
-Most algorithms assume an idealized model of computation that consists of
-memory as a flat byte-addressable array which can be read and modified in some
-fixed time. However, in practice memory is not a flat array, but rather is
-organized as a hierarchy consisting of fast on-chip caches (L1, L2, etc.), to
-slower DRAM, to even slower disks. When a program requests data, the data is
+Most external memory algorithms assume an idealized model of computation in which
+memory is a flat byte-addressable array that can be read and modified in some
+fixed time. However, in practice memory is typically
+organized as a hierarchy consisting of fast on-chip caches (L1, L2, etc.),
+slower DRAM, and even slower disks. When a program requests data, the data is
 located in the memory hierarchy and typically carried up the chain to the
-fastest memory layer (closest to the CPU), usually a cache. Unfortunately, this
-cache can only have a limited capacity, and once it is full any new data
+fastest memory layer (closest to the CPU), which is usually a cache. Unfortunately, this
+cache has limited capacity, and once it is full any new data
 requests necessitate some existing cached data to be evicted. Transferring data
 from lower to higher levels in the memory hierarchy often takes an
-exponentially long time; access times between the different layers can differ
+exponentially long time, and access times between the different layers can differ
 by many orders of magnitude. Therefore, the time to transfer data often
@@
 
-External memory algorithms are wary of this fact and seek to optimize the
-number of memory transfers performed rather than the number of CPU instructions
-executed. They often try to exploit the fact the memory hierarchies transfer
-data in blocks rather than individual machine word units. Thus, external memory
-algorithms are designed to not only limit the number of memory transfers, but
-also preserve the locality of related data so that when a costly memory access
-is performed, its cost is amortized.
+External memory algorithms are wary of this fact, and focus on optimizing the
+number of memory transfers performed instead of the number of CPU instructions
+executed. They often try to exploit the fact that memory hierarchies transfer
+data in blocks rather than individual machine word units. These
+algorithms are designed to limit the number of memory transfers, as well as
+preserve the locality of related data to amortize costly memory accesses.
 
@@
 
-Since external memory algorithms are primarily concerned with accessing data,
-it is no surprise that many external memory problems can be reduced to finding
-a suitable data structure that efficiently implement certain operations. A well
-known external memory data structure is the B-tree \cite{BayerM72}. The B-tree
-works by making each one of its nodes the same size as a memory block in a
-particular level of the hierarchy. Furthermore, the fanout in the tree (the
-number of children of a node) is proportional to the block size as well.  The
+Because external memory algorithms are primarily concerned with data transfers,
+many external memory problems can be reduced to finding
+a suitable data structure that efficiently implements certain operations. A well
+known external memory data structure is the B-tree \cite{BayerM72}, which is 
+widely used in databases and file-systems. The B-tree
+works by making every node in the tree the same size as a memory block in a
+particular level of the hierarchy. The fanout of the tree
+is also proportional to the block size.  The
 B-tree incurs $O(\log_{B}{N})$\footnote{Throughout this paper, $\log$ with the
 base omitted is base-$2$} memory transfers for \Search{}, \Insert{}, and
-\Delete{} and is widely used in databases and file-systems.
+\Delete{}.
 
@@
 two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}.
-However, in practice, architectures typically have memory hierarchies that are
-larger than two. The argument for the DAM model is that in practice, only one
+However, in practice, architectures typically have more than two levels in their memory hierarchies.
+The argument for the DAM model is that in practice, only one
 level in the memory hierarchy tends to dominate the performance, and thus a
-two-level memory model is sufficient.  However, this means that B-trees can
-begin to incur an increasing number of memory transfers if the dominating level
+two-level memory model is sufficient.  However, this means that B-trees may
+incur an increasing number of memory transfers if the dominating level
 in the hierarchy changes.
@@
 the memory layer in advance. B-trees are implemented with parameterized code
-that needs to be changed for every level in the memory hierarchy. This tuning
-can be very difficult and error prone. Additionally, tuning this parameter may
+that needs to be modified for every level in the memory hierarchy. This tuning
+can be very difficult and error prone, and may
 even be impossible in certain heterogeneous computing environments.
@@
 of computation \cite{Frigo99}. In the CO model an algorithm has no information
-about the memory hierarchy. It does not know anything about the speed or the
+about the memory hierarchy, and knows nothing about the speed or the
 block size of any level. Conceptually, the CO model allows algorithm designers
-to reason about the simple two-level model like DAM but also prove results for
+to reason about the simple two-level model like DAM, but also prove results for
 any arbitrary multi-level memory hierarchy. This is because algorithms that
@@
 changes in block size, cache capacity, and cache speed. Consequently, CO
-algorithm should gracefully scale to multi-level hierarchies as well.
+algorithms should gracefully scale to multi-level hierarchies as well.
 
@@
 B-tree (CO B-tree). The CO B-tree is similar to the normal B-tree except that
-it does not require explicit parameters about the memory hierarchy.
+it does not require explicit information about the memory hierarchy.
 
@@
 $\Theta(\log_{B}{N} + \frac{\log^2{N}}{B})$. 
-The CO B-tree was a seminal data structure that gave rise to many others.
+%The CO B-tree was a seminal data structure that gave rise to many others.
 
@@
 the various trade offs that they make. We start by analyzing the original CO
-B-tree and explain how it works. This includes a detailed description of the
-packed memory array (PMA) and strongly weight-balance binary search trees
-(SWBST) that are integral to the CO B-tree's memory performance. Following this
-discussion we provide an analysis of the CO string B-tree (a CO B-tree with
+B-tree and explain how it works. This includes a detailed description of two data
+structures integral to the CO B-trees performance: the
+packed memory array (PMA) and strongly weight-balanced binary search tree
+(SWBST). Following this
+discussion we analyze three variants of the CO B-tree: the CO string B-tree (a CO B-tree with
 variable length keys), the CO streaming B-tree (a CO B-tree that improves the
@@
 
-The CO B-tree optimizes the number of memory transfers by structuring its
+The CO B-tree optimizes the number of memory transfers by organizing its
 elements as a strongly weight-balanced search tree (SWBST). The SWBST is then
@@
 The SWBST and PMA play crucial roles in the other cache-oblivious data
-structures that were developed after the CO B-tree and thus are described at
+structures that were developed after the CO B-tree, and thus are described at
 length in the next two sections.
@@
 The search tree is laid out in memory according to a van Emde Boas (vEB) layout
-\cite{veb1,veb2}. The idea of a vEB layout is fairly straight forward. Let $h$
+\cite{veb1,veb2}: Let $h$
 denote the height of a tree. The vEB layout splits the tree approximately at
@@
 
-A nice consequence of using a vEB layout is that it can be shown to use
-$O(\log_{B}{N})$ memory transfers for searches \cite{veb1, veb2}. Thus, the CO
-B-tree has already achieved the desired bound for search queries. It still
+The vEB layout can be shown to use $O(\log_{B}{N})$ memory transfers for 
+searches \cite{veb1, veb2}. Thus, using just the vEB layout, the CO
+B-tree has already achieved the desired bound for search queries. However, it still
 remains to be shown that the CO B-tree achieves the desired number of memory
@@
 To support dynamic inserts/deletes, a \textit{strongly weight-balanced search
-tree} (SWBST) is used \cite{swbst}.  The weight of a node $u$ is equal to the
-total number of $u$'s descendants plus 1. It can also be thought of as the sum
-of the weights of $u$'s children plus 1. Formally, $u$'s weight is $w(u) = 1 +
+tree} (SWBST) is used \cite{swbst}.  Let the weight of a node $u$ be equal to the
+total number of $u$'s descendants plus 1. A recursive definition for the weight of $u$ is the sum
+of the weights of $u$'s children plus 1. Formally, $w(u) = 1 +
 \sum_{v \in children} w(v)$.
@@
 stronger property that a node at height $h$ in the tree (leaves have a height
-of 1) has $\Theta(c^h)$ descendants where $c$ is some constant ($c > 4$). This
+of 1) has $\Theta(c^h)$ descendants, where $c$ is some constant ($c > 4$). This
 property becomes important later on when laying the tree out in memory (see
@@
 
-Let us now consider insertions. To insert an element, search
-down the tree and find the natural place to insert the element - just like in
-any tree structure. After inserting the element into some node $w$, 
+Let us now consider insertions. To insert an element, simply search through
+the tree to find where the element would go. After inserting the element into some node $w$, 
 check to see if any of $w$'s ancestors have become unbalanced by having a
 weight greater than $2c^{h-1}$. If such an ancestor exists, it follows that
-every node along the path between that ancestor and $w$ will be unbalanced as
-well. This requires splitting nodes to restore balance, starting from $w$
+every node along the path between that ancestor and $w$ will also be unbalanced.
+This requires splitting nodes to restore balance, starting from $w$
 and moving up the tree.
@@
 
-Deletions are supported similarly, except we check if a node has a weight $w(u)
+Deletions are supported similarly, except we check if a node $u$ has a weight $w(u)
 < \frac{c^{h-1}}{2}$. In this case, we merge two neighbors together and proceed
-up the tree. There is one edge case that needs to be handled. The merging
-procedure may produce a new node that has a weight larger than the bounds
+up the tree. However, the merging procedure may produce a new node that has a 
+weight larger than the bounds
 allow. If this happens, immediately split the node again, recursively. This can
@@
 
-Now that the elements structured in a SWBST, we are still left with the problem
-of mapping this tree to memory and proving that it achieves the desired number
-of memory transfers per operation. The packed memory array (PMA) was developed
-to solve this problem \cite{BenderDemainColton}. There are three main costs
-that the PMA is concerned with: \textit{inserts}, \textit{splits/merges}, and
+Assuming the elements are now organized in a SWBST, we will show how to
+of map this tree to memory and explain how it achieves the desired number
+of memory transfers per operation. Do do this, we will use a packed memory 
+array (PMA)\cite{BenderDemainColton}. There are three main costs
+that the PMA addresses: \textit{inserts}, \textit{splits/merges}, and
 \textit{pointer updates}.
 
-First, the PMA needs to maintain open space in its memory for newly inserted
-items. Doing so may seem contradictory: maintaining the locality of elements
-while keeping space available for new inserts. The achieve a balance between
-these extremes the PMA uses a technique
+First, the PMA needs to maintain free space for future inserts.
+%Doing so may seem contradictory: maintaining the locality of elements while keeping space available for new inserts. 
+To achieve a balance between maintaining locality 
+and ensuring free space for new items, the PMA uses a technique
 described in \cite{packedmemoryarray}.
-The PMA is able to handle an insertion into the SWBST using only $O(1 +
-(\log^2{N}) / B)$ memory transfers (this is the cost of updating the memory
-layout of the SWBST in the PMA, not the cost of updating the data structure
-itself). At a high level, the PMA maintains
+The PMA performs an insertion into the SWBST using only $O(1 +
+(\log^2{N}) / B)$ memory transfers. Note that this is the cost of updating the memory
+layout of the SWBST in the PMA, not the cost of updating the SWBST%data structure
+itself. At a high level, the PMA maintains
 ``windows'' into the array. When a window becomes too full or too empty, the
@@
 using $O(1 + (\log{N}) /B)$ memory transfers. The proof is based on the fact
-that each node has $\Theta(c^h)$ descendants (recall, this is the key property
-that SWBST provided; shown in Equation~\ref{eq:bounds}). Thus, merging or splitting a node will cause
-$\Theta(c^h)$ nodes to move, requiring $O(1 + c^h/B)$ memory transfers. It
-follows from the definition of SWBST in \cite{swbst} that when a node is
-rebalanced, it will require $\Theta(c^h)$ insertions or deletions until it is
-rebalanced again. Thus, the amortized cost of each insertion and/or deletion is
-$O(1/B)$. When a node is split, it may traverse all the way up the tree, and
-thus the split/merge bound is $O((\log N) / B)$ memory transfers.
-
-Third, the PMA needs to handle updating the parent and child pointers that may
-become invalidated because of how the nodes moved around. If the pointers are
+that each node has $\Theta(c^h)$ descendants (recall that this is the key property
+that SWBST provides, and is shown in Equation~\ref{eq:bounds}). Thus, merging 
+or splitting a node will cause $\Theta(c^h)$ nodes to move, requiring 
+$O(1 + c^h/B)$ memory transfers. It follows from the definition of SWBST in 
+\cite{swbst} that when a node is rebalanced, it will require $\Theta(c^h)$ insertions 
+or deletions until it is rebalanced again. Thus, the amortized cost of each insertion 
+and/or deletion is $O(1/B)$. When a node is split, it may traverse all the way up the 
+tree, and thus the split/merge bound is $O((\log N) / B)$ memory transfers.
+
+Third, the PMA needs to update the parent and child pointers that
+become invalidated as a result of the movement of the nodes. If the pointers are
 all packed together in memory (e.g. within $B$ distance of each other) then
-only $O(1)$ memory transfers are required. However, if they are further apart
-then a problem arises. To deal with this, buffer nodes are introduced.  Buffer
+only $O(1)$ memory transfers are required. However, a problem arises if they 
+are further apart. To address this issue, buffer nodes are introduced.  Buffer
 nodes ``stretch'' the vEB layout so that nodes do not move far away from one
-another. By using buffers, the PMA is able to achieve a pointer update cost
+another. Using buffers, the PMA is able to achieve a pointer update cost
 using $O(1+ \frac{\log{B}}{\sqrt{B}}\log^2{N})$ memory transfers.
 
-Of these three terms, the dominating factor is maintaining room for newly
-inserted elements - $O((\log^2{N})/B)$. Therefore, overall the CO B-tree
+Of these three terms, the cost of maintaining space for newly
+inserted elements dominates - $O((\log^2{N})/B)$. Therefore, overall the CO B-tree
 achieves insertions and deletions in $O(\log_B{N} + (\log^2{N})/B)$ memory
@@
 
-While B-trees achieve an optimal bound for the number of memory transfers for
+While B-trees achieve an optimal bound in terms of memory transfers for
 search, they do not perform as well for bulk updates and insertions. The
 buffered-repository tree (BRT) is an alternative to the B-tree that achieves an
-improved amortized $O(\log{\frac{N}{B}})$ memory transfers for inserts compared
+improved amortized $O(\log{\frac{N}{B}})$ memory transfers for inserts, compared
 to the $O(\log_{B}{N})$ memory access of B-trees \cite{Buchsbaum}. In practice, log-structured
@@
 adding buffers to each node in the tree. Shuttle tree nodes have a
-linked list of buffers rather than just a single buffer like the BRT. Each of
-these buffers is another shuttle tree. A shuttle tree node enforces two
-constraints: the length of its linked list as well as the overall height (and
-thus capacity) of each of the buffered shuttle trees contained within its
-linked list are bounded, described below.
+linked list of buffers, compared to use of just a single buffer in the BRT. Each of
+these buffers is another shuttle tree. A shuttle tree node bounds both the length 
+of its linked list, and the overall height (and thus capacity) of each of the 
+buffered shuttle trees contained within its linked list.
 
@@
 The shuttle tree's node buffers have a maximum capacity that is enforced by
-bounding its height. The size of the buffers increases going across the linked
-list from left-to-right. Thus, a node's first buffer is smaller than its last
+bounding its height. The buffers are arranged in increasing order of size from
+left-to-right across the linked list. Thus, a node's first buffer is smaller than its last
 buffer. The maximum height of a buffer is given by $F_{H(k)} = F_{k-2 \lceil
@@
 leaf.  Since the recursive subtrees are split based on the largest Fibonacci
-number less than $h$ (as described previously) it follows that the buffer sizes
+number less than $h$ (as described previously), it follows that the buffer sizes
 are also a function of the Fibonacci numbers.
@@
 
-\Search{} and \Scan{} work mostly the same way as the CO B-tree. The only
-difference is that a node's buffer must be searched before following a
-pointer down the tree.
-
-\Insert{} is also similar to the CO B-tree except elements are buffered as they
-are inserted. \Insert{} starts by move down the tree from the root to leaves
-and stopping when a node that has a buffer is located (note that because of the
-nature of Fibonacci numbers that it is possible that a node does not have any
-buffers). The element is inserted into the smallest buffer in the node's linked
-list. If this insertion causes the buffer to overflow (e.g. the height of the
+\Search{} and \Scan{} work mostly the same as their counterparts in the original CO 
+B-tree. The only difference is that a node's buffer must be searched before continuing
+down the tree.
+
+\Insert{} also behaves like insertions into the original CO B-tree, except that 
+elements are buffered as they are inserted. \Insert{} starts by moving 
+down the tree from the root to the leaves, and stopping when a node
+that has a buffer is located. Note that because of the
+nature of Fibonacci numbers, a node may not have any buffers. 
+The element is inserted into the smallest buffer in the node's linked
+list. If this insertion causes the chosen buffer to overflow (e.g. the height of the
 buffer exceeds its maximum), all of the elements from the buffer are moved into
-the next largest buffer (and so on, recursively).  When the largest buffer
-overflows all of its elements are taken and \textit{shuttled} down the tree.
-Eventually a leaf node will be reached and inserting may cause the balance
-constraints on the SWBST to become invalidated. The shuttle tree rebalances
-itself just like any other SWBST: the leaf node is split into two and this
-split cascades up the tree until balance is restored.
+the next largest buffer, and so on recursively.  When the largest buffer
+overflows, all of its elements are \textit{shuttled} down the tree.
+Eventually, a leaf node is reached and the insertion may invalidate the balance
+constraints on the SWBST. The shuttle tree rebalances
+itself as described previously in Section~\ref{sec:structure}: the leaf node is 
+split into two and this split cascades up the tree until balance is restored.
 %% TODO: explain what it means to be "shuttled" down the tree
 
-The shuttle tree also needs to take care of the buffers when a node is split.
+The shuttle tree also needs to manage buffers when a node is split.
 When a node $u$ is split into two nodes $u_1$ and $u_2$, the shuttle tree
-immediately allocates enough space for all of the $u_2$'s buffers ($u_1$ simply
-reuses the space allocated for $u$'s buffers). $u$'s buffers are then scanned
-over (by a single scan) and the elements are placed in the new buffers
+immediately allocates enough space for all of $u_2$'s buffers. $u_1$
+reuses the space allocated for $u$'s buffers. Then a single scan is performed
+over $u$'s buffers, and the elements are placed in the new buffers
 accordingly.
 
-The problem of maintaining room for inserts, bounding the costs of
-merging and splitting, and bounding the costs of updating parent and child
-pointers in the shuttle tree is achieved using the same PMA used with the CO
-B-tree. This is described in detail in Section~\ref{sec:layout}.
+The PMA structure described in Section~\ref{sec:layout} is used to: maintain
+room for inserts, bound the costs of merging and splitting, and bound the 
+costs of updating parent and child pointers in the shuttle tree
 
@@
 
-\Search{} uses $O(\log_{B}{N})$ memory transfers where $N = \Theta(c^{F_k})$ and
+\Search{} uses $O(\log_{B}{N})$ memory transfers, where $N = \Theta(c^{F_k})$ and
 $c$ is the fanout. This bound is achieved by searching over the
 top and bottom recursive subtrees at each level when traversing the tree. A
-key trick is that when you visit a node you only search its largest buffer
-(the last buffer in the linked list for that level). The smaller buffers will
-be visited eventually as you recurse down the tree. Eventually you will get
-down to a single subtree that can fit into a single blocks and thus can be
+key observation is that when you visit a node you only search its largest buffer
+(the last buffer in the linked list for that level). The smaller buffers will eventually
+be visited as you recurse down the tree. Eventually, you will traverse
+down to a single subtree that fits in a single block, and thus can be
 searched with one memory transfer. Since the fanout is proportional to the
-block size it will take us $O(\log_B N)$ block accesses until we get down to a
-subtree and thus, by induction, the worst case runtime is $O(\log_B N)$.
+block size, it will only take $O(\log_B N)$ block accesses to reach this
+subtree. Thus, by induction, the worst case runtime is $O(\log_B N)$.
 
@@
 \Insert{} uses $O(log_{B+1} N / B^{\Theta(1/(loglogB)^2)} + (log^2 N) / B)$
-memory transfers. This improvement over the CO B-tree is achieved because the
-shuttle tree's buffers are used to maintain locality of the elements long
+memory transfers. This improvement over the CO B-tree is achieved using
+the shuttle tree's buffers to maintain locality of the elements long
 enough to amortize the cost of moving them down the tree and possibly into
 another memory block. The proof is based on the fact that when an element is
-moved from one block to another it is due to either a leaf node splitting or a
+moved from one block to another, it is due to either a leaf node splitting or a
 node's buffer overflowing. Recall that when inserting a new element causes a
 buffer to overflow, all of the the elements from that buffer move down the
-tree, not just the single one that caused the overflow. Thus, the cost of
-inserting an element is amortize by the number of elements in a node's largest
+tree, not just the single element that caused the overflow. Thus, the cost of
+insertion is amortized by the number of elements in a node's largest
 buffer. The number of elements in the largest buffer is
 $B^{\Theta(1/(loglogB)^2)}$ leading to the $B^{\Theta(1/(loglogB)^2)}$ factor
-in the big-Oh notation.
+in the big-O notation.
 
@@
 elements. This cost does not include the single \Search{} operation that is
-required to find the first starting element of the range. Once the first
+required to find the starting element of the range. Once the first
 element's block is transferred, the shuttle tree proceeds by traversing the
-tree and scanning in elements in order. Since the vEB layout enforces the
+tree and scanning the elements in order. Since the vEB layout enforces
 spatial locality of nearby nodes in the tree, the shuttle tree only has to read
-in a new block about once every $B$ elements. Thus, each individual memory
-access is is amortized to $1/B$ and the overall number of memory transfers is
-$O(L/B)$.
-
-The last thing that is worth mentioning is that the overall space requirements
-of the shuttle tree is $O(N)$, since the majority of the buffers are really
-small and thus do not impact the overall size of the tree.
+in a new block roughly once every $B$ elements. Thus, each individual memory
+access is is amortized to $1/B$ memory transfers, and the total number of 
+memory transfers is $O(L/B)$.
+
+Note that the shuttle tree uses $O(N)$ space, since 
+the majority of the buffers are very small and thus do not impact the 
+overall size of the tree.
 
@@
 
-The cache-oblivious lookahead array (COLA) is quite different from the other
-cache-oblivious data structures discussed so far. Instead of an explicit tree
+The cache-oblivious lookahead array (COLA) is notably different from the other
+CO data structures discussed thus far. Instead of an explicit tree
 structure, the COLA is organized as a list of arrays or levels. Each array has
@@
 \Insert{} works by initially inserting an element into the first-level array.
-This array has size $2^0 = 1$, thus the element is in an array on it own.
-Whenever an element is inserted, check to see if there is another array of
+This array has size $2^0 = 1$, thus the element is in its own array.
+Each time an element is inserted, check to see if there is another array of
 the same size (this can be thought of as checking to see if there is another
@@
 COLA merges the two together in sorted order. This merging will produce a new
-array that is twice the size and thus goes into the next level. The COLA
+array that is twice the size and thus is moved into the next level. The COLA
 recursively merges until there are no two arrays of the same size.
@@
 \Search{} can naively be implemented by scanning each of the levels in the
-COLA. Since everything is sorted the COLA can use a binary search to find the
+COLA. Since all the arrays are sorted the COLA can use a binary search to find the
 element it is looking for. However, this is not very efficient. As an
-improvement, COLA, as the name suggests, uses look ahead pointers to reduce the
-number of memory accesses required. Every 8th element from the $(k+1)$-st level
+improvement COLA, as the name suggests, uses look ahead pointers to reduce the
+number of required memory accesses. Every 8th element from the $(k+1)$-st level
 exists in the $k$-th level array in its proper sorted location. This element
-also contains a pointer to its real location in the $(k+1)$-st level. Thus this
-pointer is referred to as \textit{real lookahead pointer}. Additionally, every
-4th cell in $k$-th level array contains two pointers pointing to the next and
+also contains a pointer to its real location in the $(k+1)$-st level. Call this
+pointer a \textit{real lookahead pointer}. Additionally, every
+4th cell in a $k$-th level array contains two pointers pointing to the next and
 previous real lookahead pointers on this level. The next and previous pointers
 are referred to as \textit{duplicate lookahead pointers}. To search a level the
-COLA performs a binary search within an appropriate range, determine the range
-in the next level using the duplicate lookahead pointers and then recursively
+COLA performs a binary search within an appropriate range, determines the range
+in the next level using the duplicate lookahead pointers, and then recursively
 cascades up the tree.
@@
 based on the fact that an element will never be merged more than $O(\log{N})$
-times. This is because at that point it will be in the last array.
+times, since at that point it will be in the last array.
 Additionally, the first $\log{B}$ merges that a element participates in uses
@@
 of merging is $O(k/B)$, and thus the amortized cost of a merging a single
-element is $1/B$. Since an element will only be merged $O(\log{N})$ time, it
-follows that the overall amortized memory transfers to insert an element is
+element is $1/B$. Thus it follows from our earlier assumption that the 
+overall amortized number of memory transfers needed to insert an element is 
 $O((\log N)/B)$.
 
-\Search{} runs in $O(\log N)$ memory transfers. The proof uses induction to
-show that the COLA never needs to search more than eight elements on each
-level. Since there are $O(\log N))$ levels and since eight is a constant, it
+\Search{} runs in $O(\log N)$ memory transfers. The proof shows that
+the COLA never needs to search more than 8 elements on each
+level. Since there are $O(\log N))$ levels and 8 is a constant, it
 follows that overall search uses $O(\log N)$ memory transfers. The fact that
-only eight elements need to be searched per level is shown through induction.
+only 8 elements need to be searched per level is shown through induction.
 The base case is trivial. The first three levels of the COLA each have at most
-eight elements (note that these elements are either real elements or
+8 elements (note that these elements are either real elements or
 lookahead pointers). The COLA can search these levels with no more
-than eight member transfers. As the COLA follows lookahead pointers to the
-higher levels, it will eventually get to a level where $k \geq 3$. Note that
-the COLA does not ``enter'' this level at the start of the array; rather, its
-arrived here by following the real lookahead pointers from the levels below
-and thus is at some index $i$ that contains an element $e_i$ such that $e_i <
-e$ and $e$ is the element being searched for. The goal at this point is to
-find a new $i$ such that $e_i \leq e < e_{i+1}$. Once this $i$ is found the
-duplicate lookahead pointers can be used to locate the \textit{next} and
+than 8 member transfers. As the COLA follows lookahead pointers to the
+higher levels, it will eventually reach a level where $k \geq 3$. Note that
+the COLA does not ``enter'' this level at the start of the array. It
+arrives there by following the real lookahead pointers from the levels below,
+and thus is located at some index $i$ that contains an element $e_i$ such that $e_i <
+e$ and $e$ is the element being searched for. Now we want to
+find a new $i$ such that $e_i \leq e < e_{i+1}$. Once this $i$ is found, the
+duplicate lookahead pointers are used to locate the \textit{next} and
 \textit{previous} real lookahead pointers in the next level. The two real lookahead
 pointers are used to restrict the search in the next level. Recall that at
-each level in the COLA, every eighth element in a level is present in the
+each level in the COLA, every 8 element in a level is present in the
 previous level. Thus, it follows that in the next level only 8 elements need
-to be searched. Furthermore, since the current level was reach from a lower
+to be searched. Furthermore, since the current level was reached from a lower
 level, it follows that only 8 elements need to be searched on this level to
-find the appropriate $i$. Thus, overall, the COLA uses $O(8\log{N}) =
+find the appropriate $i$. Therefore, overall, the COLA uses $O(8\log{N}) =
 O(\log{N})$ memory transfers.
@@
 \begin{center}
-	\includegraphics[width=0.8\columnwidth]{figures/COSB-treegeneralstructurediagram.pdf}
+	\includegraphics[width=\columnwidth]{figures/COSB-treegeneralstructurediagram.pdf}
 \end{center}
@@
 for variable length string keys. The COSB-tree uses a modified front-compression
-scheme, which allows it to use less space and have improved
-compressed data locality.
+scheme, which uses less space and improves compressed data locality.
 
@@
 
-The dynamic COSB-tree consists of three layers: a centroid tree, a PMA 
-of hash data for the keys (known as the 
-\emph{hashdata} layer), and a PMA of compressed values for the original key
- (known as the 
-\emph{keydata} layer). Each leaf of the centroid tree contains a pointer into
+The dynamic COSB-tree consists of three layers: a weight-balanced centroid tree stored in
+a packed-memory array (PMA, see Section~\ref{sec:layout}); a PMA of hash data for the keys, known as the 
+\emph{hashdata} layer; and a PMA of compressed values for the original key,
+known as the 
+\emph{keydata} layer (see Figure~\ref{fig:string_structure}). Each leaf of the centroid tree contains a pointer into
 the hashdata array. The centroid tree contains only hash values, and is designed such that
-each leaf points to the respective \emph{neighborhood} of the key being
+each leaf points to the respective \emph{neighborhood} of keys being
 searched for. After traversing the centroid tree and following the pointer into
@@
 
-%a centroid tree at
-%the top, a PMA structure of hash data for the keys known as the \emph{hashdata}
-%layer; and on the bottom a PMA structure for the original key values,
-%compressed using an augmented version of forward compression known as the
-%\emph{keydata} layer. Each leaf of the centroid tree contains a pointer into
-%the hashdata array. The centroid tree only contains a subset of the keys, so
-%each leaf points to the respective \emph{neighborhood} of the key being
-%searched for. After traversing the centroid tree and following the pointer into
-%the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
-%sequential scan is performed in the hashdata array layer to find $k$. Each
-%entry in the hashdata array contains the hash value of the corresponding key,
-%and a pointer to the compressed key value in the keydata layer.
-
 In the following sections, centroid trees are described, and a summary of
@@
 
-\subsubsection{Centroid Trees} %todo:cite dictionary matching problem--sources in cosb-tree paper
+\subsubsection{Centroid Trees}
+\label{sec:centroid_trees}
 \begin{figure}
@@
 the keys in $D$ are preprocessed using hashing and divide-and-conquer via
-centroid trees. Figure \ref{fig:centroid} depicts an example centroid tree.
+centroid trees. Figure~\ref{fig:centroid} depicts an example centroid tree from
+\cite{BenderFaKu06}.
 
@@
 into the trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the
-trie resulting from excluding $y$ and its subtree, and call this the \emph{up trie} of $y$. The \emph{centroid tree} is
-obtained by rooting the tree at the centroid vertex $\rho$, and making its
-children recursively defined centroid trees of its up and down tries.
-
-The centroid tree in the COSB-tree is layed out in memory using a modified vEB
-(see Figure~\ref{fig:veb}) layout for weight-balanced binary trees. Instead of using height $h/2$, the
-\emph{hyperhyperfloors} of the weights of nodes are used as a guide for cutting the tree into
-equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x refers
-to rounding x down to the nearest power of a power of 2.
-
-%The \emph{hyperhyperfloor} of x is
-%denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to
-%the nearest power of a power of 2. This ensures trees of equal size on all
-%\emph{levels of detail}, as the centroid tree is traversed.
-% Should explain what "level of detail" means here
-
-The following approach is used to rebalance the centroid tree: put all the
+trie resulting from excluding $y$ and its subtree, and call this the \emph{up trie} of
+$y$. The \emph{centroid tree} is obtained by rooting the tree at the centroid vertex
+$\rho$, and making its children recursively defined centroid trees of its up and
+down tries.
+
+The centroid tree in terms of structure is essentially a weight-balanced binary tree (see
+Section~\ref{sec:structure}), and is laid out in memory using a modified vEB
+layout (see Figure~\ref{fig:veb}). However, instead of using height $h/2$ as
+described previously, the \emph{hyperhyperfloors} of the weights of nodes are
+used as a guide for cutting the tree into equal-size subtrees for the vEB layout. The
+\emph{hyperhyperfloor} of x refers to rounding x down to the nearest power of a 
+power of 2.
+
+In addition, a different approach is used to rebalance the centroid tree: put all the
 elements in DFS or Euler tour order for the compacted trie. Given this order,
@@
 
-The following approach is used for maintaining the vEB layout of the centroid
-tree in the COSB-tree: first, weight balance is relaxed from a maximum weight
-difference of 2 between left and right subtrees, to a larger constant $c$. Thus
-a node $v$ only becomes out of balance every $\Omega(\textsc{weight(v)})$
-insertions or deletions of descendants of $v$. The nodes are stored in a PMA,
-preserving order of the nodes using $\Theta(N)$ space.
-
-On insert or delete, the area around the update point is scanned for sections
-of the array to rearrange. If a section around the update point reaches a
-certain \textit{density} threshold, the nodes are evenly distributed across the
-section to make room for more nodes. To ensure that updating pointers is not too
-expensive, any node included in the section must have all of its subtrees
+Similar to the original CO B-tree, the centroid tree is stored in a PMA
+(see Section~\ref{sec:layout}). However, the centroid tree PMA is more 
+flexible with a modified updating scheme. The changes made to updating 
+"windows" into the array are as follows (see Section~\ref{sec:layout} 
+for more on "window" updates): on insert or 
+delete, the area around the update point is scanned for sections
+or windows of the array to rearrange. If a window of a
+certain \textit{density} threshold is found (as described previously in~\ref{sec:layout}), the 
+nodes are evenly distributed across the
+window to make room for more nodes. To ensure that updating pointers is not too
+expensive, any node included in the window must have all of its subtrees
 included. The subtrees are dictated by the vEB layout, and subtrees are chosen
-from the bottom up. The smallest neighboring trees (single nodes) in the vEB
-layout are added first, and larger trees are included when all of their
-subtrees have been added to the section. Root trees are only added after all
-bottom trees are added (in terms of the vEB layout). Figure
+from the bottom up. The smallest neighboring subtrees (single nodes) in the vEB
+layout are added first, and larger subtrees are included when all of their own
+subtrees have already been added to the section. top subtrees are only added after all
+bottom subtrees are added (in terms of the vEB layout). Figure
 \ref{fig:centroid_relayout} gives an example of how subtrees are chosen for
-rearranging.
+redistribution of nodes.
 
 This approach results in an amortized update cost of $O(1+(\log{N})/B)$ memory
-transfers per update. In conjunction with the cost of rebuilds described above,
+transfers per update. In conjunction with the cost of rebalancing described above,
 this is only an additive $O((\log{N})/B)$ factor, so the overall bound is
@@
 position in the bottom left of the tree. The numbers represent the order in
-which each subtree is added to the section to be rearranged.}
+which each subtree is added to the window to be rearranged.}
 \label{fig:centroid_relayout}
@@
 
-This section provides a brief summary of how front compression is modified to
+This section provides a brief summary of front compression, and how it is modified to
 optimize compression and decoding for dynamic COSB-trees. This approach 
-increases the locality of the compressed data. Consider a sequence
+increases the locality of the compressed data.
+
+First, front compression is described: consider a sequence
 of $j$ keys to store $k_{1},k_{2},...,k_{j}$. Without compression, this
@@
 Bender et al. provide a modification that reduces the high decoding cost of
-front compressing $D$ to $O(1+||k||B)$, while maintaining a compressed size of
+front compressing $D$ to $O(1+||k||B)$ memory transfers, while 
+maintaining a compressed size of
 $O(\llangle{D}\rrangle{})$\footnote{$\llangle{D}\rrangle{}$ is shorthand for
@@
 added to the compression stream. For each subsequent key $k_{i}$, if $k_{i}$
-can be decoded with just the last $c||k_{i}||$ characters, add
-$\pi_{i}$, $\sigma_{i}$ as described above. Otherwise, add $k_{i}$ to the stream
+can be decoded with just the last $c||k_{i}||$ characters in the compression stream, add
+$\pi_{i}$, $\sigma_{i}$ as described above for standard front compression.
+Otherwise, add $k_{i}$ to the stream
 without compressing it. By construction, this scheme guarantees that any key
@@
 
-%todo: figure out how to do range queries from this paper somehow...
-
 The operations supported by the COSB-tree are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{},
@@
 \Search{} works as follows for a given key $k$. Compute the hash value for
-$k$ and call it $\rho$, and traverse the centroid tree of the COSB-tree using
+$k$ and call it $\rho$. Traverse the centroid tree of the COSB-tree using
 $\rho$ to obtain the successor $\rho_{s}$ and predecessor $\rho_{p}$ of $k$.
-Index into the hashdata layer and perform a sequential scan across the PMA
-structure from $\rho_{p}$ to $\rho_{s}$ to find $\rho$. On a match with value
+Index into the hashdata layer with the resulting pointer from the centroid tree traversal,
+and perform a sequential scan across the PMA
+from $\rho_{p}$ to $\rho_{s}$ to find $\rho$. On a match with hash value
 $\rho'$, index into the keydata layer and decode $\rho'$ to obtain the
@@
 the centroid tree. Scan the hashdata layer across from $\rho_{b,p}$ to
-$\rho_{e,s}$. Only start noting the beginning of range when the first key with
+$\rho_{e,s}$. Only start noting the beginning of the range when the first key with
 hash value after $\rho_{b}$, and stop once we reach the first key with hash
@@
 transfers, where in this case $k'$ is the other end of the range and $Q$ is the
-set of keys within the given range.
-
-\Insertkonly{} and \Delete{} w.h.p. require $O(1+\log_{B}N+\log^{2}N||k||/B)$
-memory transfers. As mentioned previously, due to rebuilds updates have
-amortized cost of $O(1+(\log^{2}N)/B)$ memory transfers. The rest of the cost
-for insertions and deletions is due to having to find the correct key to
+set of keys in $D$ that are within the given range.
+
+\Insertkonly{} and \Delete{} require $O(1+\log_{B}N+\log^{2}N||k||/B)$
+memory transfers (w.h.p.). As mentioned previously in Section~\ref{sec:centroid_trees}, 
+due to rebalancing updates have amortized cost of $O(1+(\log^{2}N)/B)$ memory transfers.
+The rest of the cost for insertions and deletions is due to having to find the correct key to
 delete or location for insert, which can be seen from the cost of \Search{}.
@@
 exclusion on the entire data structure and use the pre-existing data structures
-without modification; this approach, however, yields poor performance.
+without modification. However, this approach yields poor performance.
 
@@
 proposed \cite{BayerS77, LehmanY81}. We now survey a concurrency scheme for CO
-B-trees, first presented in \cite{BenderFiGi05}. Bender develops two different
-data structures to solve this problem: the first is an exponential CO B-tree,
-the second is a CO B-tree based on a packed memory array (PMA). Bender also
-presents a lock-free variant of the latter data structure.
+B-trees, first presented in \cite{BenderFiGi05}. Bender et al. develops two different
+data structures to solve this problem: first an exponential CO B-tree,
+and second a CO B-tree based on a PMA (see Section~\ref{sec:layout} for more on PMA's). Bender et al. also
+present a lock-free variant of the latter data structure.
 
@@
 Before describing the data structures in more detail, we first consider
-the concurrency model which is used throughout \cite{BenderFiGi05}.
+the concurrency model used throughout \cite{BenderFiGi05}.
 
@@
 described in Section~\ref{sec:structure}. A tuning parameter $1 < \alpha < 2$
-is chosen for the data structure. Each node in the tree maintains both keys
-which divide up its children, in addition to a \textit{right-link} pointer,
-which contains both a pointer to a node's right sibling, plus the key value of
+is chosen for the data structure. Each node in the tree maintains the keys
+which divide up its children, in addition to a \textit{right-link} pointer. The right-link
+pointer contains a pointer to the node's right sibling, and the key value of
 the minimum key in the node's right sibling. If no such sibling exists, a
@@
 (weakly-consistent range scans could be implemented in the data structure with
-very little modification), although interestingly enough the PMA variant of
+very little modification). However, the PMA variant of CO
 B-tree (discussed in Section \ref{sec:pma}) does cover range scans.
@@
 \Insert{} works as follows. First, \Search{} is used to locate the leaf node
-where $k$ will be inserted into. A lock is acquired on that node, $k$ is
+that $k$ will be inserted into. A lock is acquired on that node, $k$ is
 inserted, and the lock is released. Unlike a standard B-tree, there is no
-maximum number of elements in a node; instead we split nodes
-probabilistically. With probability ${1}/{2^{\alpha^h}}$, a node at height $h$
-is split (leaf nodes have $h = 0$). A split works as follows. We reacquire a
+maximum number of elements in a node. Instead, we split nodes
+probabilistically: with probability ${1}/{2^{\alpha^h}}$, a node at height $h$
+is split (leaf nodes have $h = 0$). To perform a split, we first reacquire a
 lock on the node to be split. We split the node into $u$ and $u'$, where $u$
-contains all the keys $< k$, and $u'$ contains the keys $\geq k$. We release
+contains all the keys $< k$, and $u'$ contains the keys $\geq k$. We then release
 the lock, and recursively insert $k$ into the parent of the split node. It is
 important to note that, since $u$ contains a prefix of the original node to be
-split, that we reuse the original node in constructing $u$ (that way,
-concurrent searches can still locate the keys in $u$).
+split, we reuse the original node in constructing $u$. Doing so ensures that
+concurrent searches can still locate the keys in $u$.
 
@@
 Sequential search takes $O(\log_B{N} + \log_{\alpha}{\log{B}})$ block transfers
-(with high probability). The probabilistic bound comes from the probabilistic
+(w. h. p.). The probabilistic bound comes from the probabilistic
 key promotion property. The intuition behind the proof of this bound comes from
 the argument that when the tree has height $h =
-\Omega(\log_{\alpha}{\log{\log{N}}})$, then the total number of keys in the
+\Omega(\log_{\alpha}{\log{\log{N}}})$, the total number of keys in the
 tree is $O(2^{\alpha^h})$. The proof is then completed by summing over the
-range of $h$ which is probabilistic $[0, O(\log_{\alpha}{\log{N}})]$, the
-access time from some level $h$ to find the correct child.
+range of $h$ which is probabilistic $[0, O(\log_{\alpha}{\log{N}})]$, which is the
+the number of memory transfers made, starting from some level $h$,
+to find the correct child.
 
 Sequential insert takes $O(\log_B{N} + \log{\alpha}{\log{B}})$ block
-transfers (in expectation). This proof is constructed in a similar fashion as
-the search proof: consider inserting a tree at some height $h$. By noting that
-the probability a particular node reaches height $h$ is
+transfers (in expectation). This proof is similar to the search proof: 
+consider inserting a tree at some height $h$. By noting that
+the probability that a particular node reaches height $h$ is
 $2^{-(\alpha^h-1)/(\alpha-1)}$, a summation over all levels of the cost of
-rebuilding a node yields the time-bound.
+rebuilding a node yields the desired bound.
 
@@
 
-The PMA CO B-tree is a two level data structure, containing a static CO B-tree
-which indexes into a packed-memory array. Bender introduces a new
-packed-memory array, called a \textit{one-way packed-memory array}, which he
+The packed-memory array (PMA) CO B-tree is a two level data structure, containing a static CO B-tree
+which indexes into a PMA (see Section~\ref{sec:layout} for more on PMA's). 
+Bender et al. introduce a new PMA, 
+called a \textit{one-way packed-memory array}, which he
 argues is easier to manage concurrently.
@@
 
-A one-way packed-memory array is a contiguous memory structure which has three
+A one-way PMA is a contiguous memory structure which has three
 distinct regions: a leftmost region of empty space, an \textit{active region}
@@
 space. The array only grows into the rightmost region, and shrinks from the
-leftmost region (hence one-way). The PMA is of size $m = \Theta(N)$, subject
+leftmost region, hence one-way. The PMA is of size $m = \Theta(N)$, subject
 to $m$ being a power of $2$, where $N$ is the number of elements.
@@
 \Insert{} works as follows. Use \Search{} to locate the $\Theta(\log
-N)$-region of the PMA which (would) contain $k$. Lock the region, and scan
-right to find the appropriate slot $s$ to place $k$ into. $s$ is defined such
-that slot $s - 1$ contains the largest key smaller than $k$. If scanning
+N)$-region of the PMA that would contain $k$. Let $s$ be the slot
+such that slot $s - 1$ contains the largest key smaller than $k$.
+Lock the current region and scan right to find $s$. If scanning
 forward requires going into the next $\Theta(\log N)$-region, lock the next
 region and release the current region (\textit{hand-over-hand} locking
-technique). Once $s$ is located, either $s$ is free, in which case just insert
-$k$, or $s$ is not free, in which case a \textit{rebalance} operation is
-required.
-
-To do a rebalance operation, we scan right (locking as necessary) until the
-region from $s$ to the current position $s'$ is not \textit{too dense}. The
+technique). Once $s$ is located, if $s$ is free, just insert
+$k$. Otherwise, a \textit{rebalance} operation is required.
+
+To perform a rebalance operation, we scan right (locking as necessary) until the
+region from $s$ to the current position $s'$ is not "too dense". The
 notion of density is defined in detail in \cite{BenderFiGi05}. The region $[s,
@@
 
-Sequential search time-bounds are not presented in the paper, but presumably
+Sequential search memory transfer bounds are not presented in the paper, but presumably
 since search is a static CO tree lookup followed by a scan, the time bound is
@@
 Sequential insert is $O(\log_{B}{N} + (\log^2{N})/{B} + 1)$. The proof is based
-off an accounting argument which places $\Theta((\log^2{m})/B)$ dollars in each
-array slot. The proof is not presented in any detail in the paper, but rather
+on an accounting argument that places $\Theta((\log^2{m})/B)$ dollars in each
+array slot. The proof is not presented in any detail in \cite{BenderFiGi05}, but
 the authors argue that it is simply a slightly-modified version of the proof
@@
 
-The lock-free variant is based on the PMA CO tree, except the locks acquired
+The lock-free variant is based on the PMA CO B-tree, except the locks acquired
 during the PMA modification are replaced by augmenting each element in the PMA
-with a marker. A marker is used to denote the ongoing (concurrent) operation of
+with a marker. The marker is used to denote the ongoing (concurrent) operation of
 a particular cell. The authors assume that a marker is able to fit within a
-single machine-word (so that the atomic operations are available and
-performant).
+single machine-word, so that the atomic operations are available and
+performant.
 
@@
 \end{inparaenum} These primitives are self-explanatory, except they
-all work in a non-blocking manner (they can fail at any point, and it is
-up to the current process to restart the operation on failure).
+all work in a non-blocking manner. They can fail at any point, and it is
+up to the current process to restart the operation on failure.
 
@@
 are initiated by denoting the cell in question with the marker. If a process
-ever encounters a cell with a marker, it drops what it is doing and first
-helps to complete the operation. When the originating process gets
-context-switched back in, any SCs will fail, and then the process can check to
-see if the operation was already completed. The implementation of the
-non-blocking primitives is spelled out in more detail in \cite{BenderFiGi05},
+ever encounters a cell with a marker, it temporarily suspends its current task and
+helps to complete the pending operation. When the originating process is
+context-switched back in, any store-conditionals (SCs) will fail, and then the process can check to
+see if the operation has already been completed. The implementation of the
+non-blocking primitives is described in more detail in \cite{BenderFiGi05},
 but the basic idea is using LL/SC to detect when concurrent operations are
@@
 \textit{cell-insertion} operation is performed on $s$. If it succeeds, we are
-done. If it fails, then a rebalance is necessary. Rebalance proceeds as
+done. Otherwise, a rebalance is necessary. The rebalance proceeds as
 before, scanning right to find the region $[s, s']$. However, whenever a
@@
 
-\section{Acknowledgments}
-We sincerely thank Dr. David Karger and all of the TAs for 6.854 
-for their guidance, help, and support. May the Flow (Max-Flow, that is) be with you all.
+% are we putting this in in the final copy?
+%\section{Acknowledgments}
+%We sincerely thank Dr. David Karger and all of the TAs for 6.854 
+f%or their guidance, help, and support. May the Flow (Max-Flow, that is) be with you all.
 

### 2011-12-14 paper.tex
Tweaking figure placement
Making the paper more 'professional' (*rollseyes*) at Leilani's request. Also fixing the header on one section and fixing a couple of the citations.
added a description section that is not a one-liner to variable length strings for consistency

--- 2011-12-13 paper.tex
@@
 
+\addtolength{\topmargin}{-.07in}
+\addtolength{\textheight}{.07in}
+
+%\addtolength{\oddsidemargin}{-0.08in}
+%\addtolength{\evensidemargin}{-0.08in}
+%\addtolength{\textwidth}{0.08in}
+
 \begin{document}
@@
 Salman Ahmad\\
-\affaddr{MIT CSAIL}
+\affaddr{MIT Media Lab}
 \email{saahmad@mit.edu}
@@
 \begin{center}
-	\includegraphics[width=0.8\columnwidth]{figures/veb.pdf}
+	\includegraphics[width=0.75\columnwidth]{figures/veb.pdf}
 \end{center}
@@
 descendants of a node is $\Theta(c^h)$. We now describe the tree's memory layout.
+
+\vspace{-0.1in}
 
@@
 \begin{figure}
-
 \begin{center}
@@
 \end{center}
-
 \caption{An illustration of how the shuttle tree determines the size of a node's linked list. The 
@@
 \section{Variable Length Strings}
-\begin{figure}
-
-\begin{center}
-	\includegraphics[width=0.8\columnwidth]{figures/COSB-treegeneralstructurediagram.pdf}
-\end{center}
-
-\caption{An illustration of the dynamic COSB-tree structure}
-\label{fig:string_structure}
-\end{figure}
+
 \subsection{Motivation}
@@
 
+\subsection{Description}
+The following subsections describe the structure of the COSB-tree in more detail,
+as well as summarize the modifications Bender et al. make to front compression
+for COSB-trees.
 \subsubsection{General COSB-tree Structure}
+
+\begin{figure}
+\begin{center}
+	\includegraphics[width=\columnwidth]{figures/COSB-treegeneralstructurediagram.pdf}
+\end{center}
+\caption{An illustration of the dynamic COSB-tree structure}
+\label{fig:string_structure}
+\end{figure}
 
@@
 is consulted to resolve hash collisions in the hashdata layer, and to obtain the original value of the key.
+The structure is described in Figure~\ref{fig:string_structure}.
 
@@
 
-\subsubsection{Centroid Trees} %todo:cite dictionary matching problem--sources in cosb-tree paper
-\begin{figure}
-
-\begin{center}
-	\includegraphics[width=0.8\columnwidth]{figures/centroidtreeexample.pdf}
-\end{center}
-
-\caption{A centroid tree example from \cite{BenderFaKu06}. Dotted lines
-represent centroid-rooted up tries, where the parent is not a prefix of the
-child. Solid lines represent centroid-rooted down tries, where the parent is a
-prefix of the child.}
-\label{fig:centroid}
-\end{figure}
-
 COSB-trees are a solution to the \textit{dictionary matching problem}.
@@
 children recursively defined centroid trees of its up and down tries.
+
+\subsubsection{Centroid Trees} %todo:cite dictionary matching problem--sources in cosb-tree paper
+\begin{figure}
+\begin{center}
+	\includegraphics[width=0.8\columnwidth]{figures/centroidtreeexample.pdf}
+\end{center}
+\caption{A centroid tree example from \cite{BenderFaKu06}. Dotted lines
+represent centroid-rooted up tries, where the parent is not a prefix of the
+child. Solid lines represent centroid-rooted down tries, where the parent is a
+prefix of the child.}
+\label{fig:centroid}
+\end{figure}
 
@@
 \begin{center}
-	\includegraphics[width=0.8\columnwidth]{figures/centroidrelayoutexample.pdf}
+	\includegraphics[width=0.7\columnwidth]{figures/centroidrelayoutexample.pdf}
 \end{center}
@@
 
-Sequential insert takes $O(\log_B{N} + \log{\alpha}{\log{B}})$ block
-transfers (in expectation). This proof is constructed in a similar fashion as
-the search proof: consider inserting a tree at some height $h$. By noting that
-the probability a particular node reaches height $h$ is
-$2^{-(\alpha^h-1)/(\alpha-1)}$, a summation over all levels of the cost of
-rebuilding a node yields the time-bound.
-
-\subsection{Packed Memory Array CO B-tree}
-\label{sec:pma}
-
-The PMA CO B-tree is a two level data structure, containing a static CO B-tree
-which indexes into a packed-memory array. Bender introduces a new
-packed-memory array, called a \textit{one-way packed-memory array}, which he
-argues is easier to manage concurrently.
-
-\subsubsection{One-Way PMA Primer}
-
-A one-way packed-memory array is a contiguous memory structure which has three
-distinct regions: a leftmost region of empty space, an \textit{active region}
-which contains the elements of the array, and a rightmost region of empty
-space. The array only grows into the rightmost region, and shrinks from the
-leftmost region (hence one-way). The PMA is of size $m = \Theta(N)$, subject
-to $m$ being a power of $2$, where $N$ is the number of elements.
-
-When the \textit{active region} becomes too large/small due to
-inserts/deletes, the PMA is rebalanced. A new array is allocated, and the
-elements are spread out evenly over the new array (to minimize element
-density).
-
-Each leaf in the static CO B-tree is a pointer into a $\Theta(\log N)$-size
-region in the packed-memory array.
-
-\subsubsection{Operations}
-
-\Search{} works as follows. Search the static tree for $k$ until a leaf node
-is reached. Follow the leaf node pointer into the PMA. Scan right in the PMA
-until either $k$ is located, or some element $> k$ is reached. Do not acquire
-any locks during the search.
-
-\Insert{} works as follows. Use \Search{} to locate the $\Theta(\log
-N)$-region of the PMA which (would) contain $k$. Lock the region, and scan
-right to find the appropriate slot $s$ to place $k$ into. $s$ is defined such
-that slot $s - 1$ contains the largest key smaller than $k$. If scanning
-forward requires going into the next $\Theta(\log N)$-region, lock the next
-region and release the current region (\textit{hand-over-hand} locking
-technique). Once $s$ is located, either $s$ is free, in which case just insert
-$k$, or $s$ is not free, in which case a \textit{rebalance} operation is
-required.
-
-To do a rebalance operation, we scan right (locking as necessary) until the
-region from $s$ to the current position $s'$ is not \textit{too dense}. The
-notion of density is defined in detail in \cite{BenderFiGi05}. The region $[s,
-s']$ is then rebalanced as follows. Start from the rightmost element, and move
-elements to the right (maintaining order) such that no suffix of $[s, s']$ is
-too dense. This rebalancing creates room for $k$; we insert $k$ and then
-release the locks.
-
-\subsubsection{Complexity}
-
-% TODO: can somebody check this time-bound? seems kind of bad
-
-Sequential search time-bounds are not presented in the paper, but presumably
-since search is a static CO tree lookup followed by a scan, the time bound is
-$O(\log_{B}{N} + ({\log{N}})/{B} + 1)$.
-
-Sequential insert is $O(\log_{B}{N} + (\log^2{N})/{B} + 1)$. The proof is based
-off an accounting argument which places $\Theta((\log^2{m})/B)$ dollars in each
-array slot. The proof is not presented in any detail in the paper, but rather
-the authors argue that it is simply a slightly-modified version of the proof
-presented in \cite{Katriel02}.
-
-\subsubsection{Lock-free Variant}
-
-The lock-free variant is based on the PMA CO tree, except the locks acquired
-during the PMA modification are replaced by augmenting each element in the PMA
-with a marker. A marker is used to denote the ongoing (concurrent) operation of
-a particular cell. The authors assume that a marker is able to fit within a
-single machine-word (so that the atomic operations are available and
-performant).
-
-In addition, four non-blocking primitives are introduced:
-\begin{inparaenum}[(a)]
-  \item \textit{move},
-  \item \textit{cell-insertion},
-  \item \textit{cell-deletion}, and
-  \item \textit{read}.
-\end{inparaenum} These primitives are self-explanatory, except they
-all work in a non-blocking manner (they can fail at any point, and it is
-up to the current process to restart the operation on failure).
-
-The concurrency protocol works as follows. Any of the non-blocking primitives
-are initiated by denoting the cell in question with the marker. If a process
-ever encounters a cell with a marker, it drops what it is doing and first
-helps to complete the operation. When the originating process gets
-context-switched back in, any SCs will fail, and then the process can check to
-see if the operation was already completed. The implementation of the
-non-blocking primitives is spelled out in more detail in \cite{BenderFiGi05},
-but the basic idea is using LL/SC to detect when concurrent operations are
-on-going.
-
-\begin{figure*}[h!t]
+\begin{figure*}[t]
 \begin{center}
@@
 
+Sequential insert takes $O(\log_B{N} + \log{\alpha}{\log{B}})$ block
+transfers (in expectation). This proof is constructed in a similar fashion as
+the search proof: consider inserting a tree at some height $h$. By noting that
+the probability a particular node reaches height $h$ is
+$2^{-(\alpha^h-1)/(\alpha-1)}$, a summation over all levels of the cost of
+rebuilding a node yields the time-bound.
+
+\subsection{Packed Memory Array CO B-tree}
+\label{sec:pma}
+
+The PMA CO B-tree is a two level data structure, containing a static CO B-tree
+which indexes into a packed-memory array. Bender introduces a new
+packed-memory array, called a \textit{one-way packed-memory array}, which he
+argues is easier to manage concurrently.
+
+\subsubsection{One-Way PMA Primer}
+
+A one-way packed-memory array is a contiguous memory structure which has three
+distinct regions: a leftmost region of empty space, an \textit{active region}
+which contains the elements of the array, and a rightmost region of empty
+space. The array only grows into the rightmost region, and shrinks from the
+leftmost region (hence one-way). The PMA is of size $m = \Theta(N)$, subject
+to $m$ being a power of $2$, where $N$ is the number of elements.
+
+When the \textit{active region} becomes too large/small due to
+inserts/deletes, the PMA is rebalanced. A new array is allocated, and the
+elements are spread out evenly over the new array (to minimize element
+density).
+
+Each leaf in the static CO B-tree is a pointer into a $\Theta(\log N)$-size
+region in the packed-memory array.
+
+\subsubsection{Operations}
+
+\Search{} works as follows. Search the static tree for $k$ until a leaf node
+is reached. Follow the leaf node pointer into the PMA. Scan right in the PMA
+until either $k$ is located, or some element $> k$ is reached. Do not acquire
+any locks during the search.
+
+\Insert{} works as follows. Use \Search{} to locate the $\Theta(\log
+N)$-region of the PMA which (would) contain $k$. Lock the region, and scan
+right to find the appropriate slot $s$ to place $k$ into. $s$ is defined such
+that slot $s - 1$ contains the largest key smaller than $k$. If scanning
+forward requires going into the next $\Theta(\log N)$-region, lock the next
+region and release the current region (\textit{hand-over-hand} locking
+technique). Once $s$ is located, either $s$ is free, in which case just insert
+$k$, or $s$ is not free, in which case a \textit{rebalance} operation is
+required.
+
+To do a rebalance operation, we scan right (locking as necessary) until the
+region from $s$ to the current position $s'$ is not \textit{too dense}. The
+notion of density is defined in detail in \cite{BenderFiGi05}. The region $[s,
+s']$ is then rebalanced as follows. Start from the rightmost element, and move
+elements to the right (maintaining order) such that no suffix of $[s, s']$ is
+too dense. This rebalancing creates room for $k$; we insert $k$ and then
+release the locks.
+
+\vfill
+
+\subsubsection{Complexity}
+
+% TODO: can somebody check this time-bound? seems kind of bad
+
+Sequential search time-bounds are not presented in the paper, but presumably
+since search is a static CO tree lookup followed by a scan, the time bound is
+$O(\log_{B}{N} + ({\log{N}})/{B} + 1)$.
+
+Sequential insert is $O(\log_{B}{N} + (\log^2{N})/{B} + 1)$. The proof is based
+off an accounting argument which places $\Theta((\log^2{m})/B)$ dollars in each
+array slot. The proof is not presented in any detail in the paper, but rather
+the authors argue that it is simply a slightly-modified version of the proof
+presented in \cite{Katriel02}.
+
+\vfill
+
+\subsubsection{Lock-free Variant}
+
+The lock-free variant is based on the PMA CO tree, except the locks acquired
+during the PMA modification are replaced by augmenting each element in the PMA
+with a marker. A marker is used to denote the ongoing (concurrent) operation of
+a particular cell. The authors assume that a marker is able to fit within a
+single machine-word (so that the atomic operations are available and
+performant).
+
+In addition, four non-blocking primitives are introduced:
+\begin{inparaenum}[(a)]
+  \item \textit{move},
+  \item \textit{cell-insertion},
+  \item \textit{cell-deletion}, and
+  \item \textit{read}.
+\end{inparaenum} These primitives are self-explanatory, except they
+all work in a non-blocking manner (they can fail at any point, and it is
+up to the current process to restart the operation on failure).
+
+The concurrency protocol works as follows. Any of the non-blocking primitives
+are initiated by denoting the cell in question with the marker. If a process
+ever encounters a cell with a marker, it drops what it is doing and first
+helps to complete the operation. When the originating process gets
+context-switched back in, any SCs will fail, and then the process can check to
+see if the operation was already completed. The implementation of the
+non-blocking primitives is spelled out in more detail in \cite{BenderFiGi05},
+but the basic idea is using LL/SC to detect when concurrent operations are
+on-going.
+
+\vfill
+
 \subsubsection{Lock-free Variant Operations}
@@
 
+\newpage
+
 \section{Conclusion}
@@
 
+\vspace{0.1in}
+
 \section{Acknowledgments}
-We sincerely thank Dr. David Karger and all of the TAs for 6.854 
-for their guidance, help, and support. May the Flow (Max-Flow, that is) be with you all.
+
+We sincerely thank Dr. David Karger and all of the 6.854 TAs
+for their guidance, help, and support.
+
+\vspace{0.1in}
+
+%%%% May the Flow (Max-Flow, that is) be with you all.
 

### 2011-12-14 paper.tex
Fixing the citations and tweaking layouts

@@
 
-\subsection{Description}
-The following subsections describe the structure of the COSB-tree in more detail,
-as well as summarize the modifications Bender et al. make to front compression
-for COSB-trees.
 \subsubsection{General COSB-tree Structure}
@@
 \begin{center}
-	\includegraphics[width=0.7\columnwidth]{figures/centroidrelayoutexample.pdf}
+	\includegraphics[width=0.8\columnwidth]{figures/centroidrelayoutexample.pdf}
 \end{center}
@@
 Before describing the data structures in more detail, we first consider
-the concurrency model which is used throughout \cite{BenderFiGi05}.
+the concurrency model which is used in \cite{BenderFiGi05}.
+
+\newpage
 
@@
 \end{center}
+\vspace{-0.2in}
 \end{figure*}
@@
 
-\vfill
-
 \subsubsection{Complexity}
@@
 presented in \cite{Katriel02}.
-
-\vfill
 
@@
 
-\vfill
-
 \subsubsection{Lock-free Variant Operations}
@@
 
-\newpage
+\vspace{0.1in}
 
@@
 
-\vspace{0.1in}
+\vspace{0.15in}
 
